### 2015-10-21 paper3.Rnw
still need bib to compile

\documentclass{article}
\usepackage{natbib}
\bibliographystyle{chicago}
\title{ HMM}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

\setkeys{Gin}{width=0.9\textwidth}
\providecommand{\keywords}[1]{\textbf{\textit{Keywords:}} #1}

\begin{document}
%\SweaveOpts{concordance=TRUE}
\maketitle

<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2")
@

\begin{abstract}
Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal heterogeneity and
autocorrelation (persistence) in group occupancy can obscure such
signals. We use latent-state and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers. Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely explored model extension, resolves previous HMM modeling issues
and clarifies the behavioural patterns of panthers.
...\end{abstract}

\keywords{abc}

\section{Introduction}

Animal ecologists are interested in factors that determine behavioural sequences of movement patterns (i.e. when animals move in a certain pattern, and under what circumstances they switch to alternative movement pattern). It is extremely difficult to observe, distinguish and classify the underlying behavioural modes/states exhibited by animals. Thus, the only observable information are the behavioural results from their location records. Over limited time scales, the path of a  moving individual can often be characterized by biased/correlated random walks \citep{okubo_diffusion_1980,turchin_quantitative_1998,okubo_diffusion_2001}, however; they fail to capture the movement patterns over longer periods due to change in movement behaviour \citep{firle_influence_1998,morales_scaling_2002}.

Recent years, animal ecologists proposed the multiphasic movement framework, examining the complex functions of internal states and external ecological and environmental factors, attempting to capture the tendency of an individual to stay in one behaivoural state for an extended time period and switching to other behavioural states \citep{firle_influence_1998,nathan_movement_2008}. Behavioural states of a study animal in movement studies via biotelemetry are unobserved directly, however, the multiphasic movement framework provides a mean to interpret, characterize and distinguish movement patterns with specific functions. Models such as finite mixture model (FMM) \citep{tracey_mapping_2012} and hidden Markov model (HMM) \citep{van_de_kerk_hidden_2014} are able to both group observed movements into behavioural states and characterize movement hehaviour within the states. In this paper, hidden Markov models and extensions are the model of choice but they are one choice out of a universe of possibilities.

Finite mixture models (can be model as a hidden Markov model with constant probabilities between all states) do not account for serial dependence in the behavioural results thus, unable to make inference on underlying behavioural sequence. The hidden Markov model, a discrete-time time-series model that allows serial dependence and focuses on observations that are regularly spaced in time, has been used to study animal movement and space use patterns \citep{franke_analysis_2004,franke_prediction_2006,patterson_classifying_2009}. The Hidden Markov model provides a better fit than latent-state models, and a transition component \citep{morales_extracting_2004} explaining the changes in behavioural sequence \citep{fryxell_multiple_2008}. The hidden Markov framework captures the observational aspect of describing the underlying behavioural sequence and the behavioural result, however, the transitions estimated by hidden Markov models are homogeneous, which unnatural realistically and biologically. Thus, to date, there is no existing framework that fits these two things together.

The number of movement states directly affects the complexity of the movement model, thus, it is restricted by the number of observations and biologically distinctive properties. Most hidden Markov models limit to two underlying behavioural states based on binary assumption mixtures, three states are not unheard of, and beyond three states are difficult to interpret biologically given the behavioural states are unobserved directly. Higher quality and quantity of data are available over the years, thus the restriction of limited data can be overlooked. Model sections by goodness of fit are identifying numerous clusters suggesting potential behavioural states, beyond our understanding and interpretations biologically \citep{van_de_kerk_hidden_2014}. Thus, the question arises whether (1) it is reasonable to have numerous states beyond biological distinguishable interpretation and their meaning, (2) over predicting number of clusters under the absents of casual factors, and (3) a framework that can handle these assumptions.

Endangered spicies such as Florida panthers (Puma Concolor Coryi) revealed multiphasic movement when fitting hidden semi-Markov models (extension of hidden Markov models) \citep{van_de_kerk_hidden_2014}. Three state models were used but greater than three states suggest better fits but were difficult to interpret biologically. In this paper, we will investigate the movement patterns of the Florida panther and use time covariates to model strong diurnal cycle observed in Florida panthers that was not previously accounted for and see if (1) temporal heterogeneity can reduce the number of states to an appropriate and biologically interpretable level, (2) readjusting the complexity levels of the hidden Markov framework can capture the behavioural inhomogeneous transition patterns, (3) capture the movement patterns over long period of time and able to produce biologically reasonable movements. The ultimate goal is to see if increasing the complexity in the hidden Markov framework can decrease the overall model complexity by reducing number of states.

\section{Materials and Methods}

GPS collars were fitted to 18 Florida Panther (Puma Concolor Coryi) in 2005-2012 by Florida Fish and Wildlife and Conservation Commission (FWC) staff using trained hounds and houndsmen. Locations used in this study were the hourly GPS coordinates then decomposed into step lengths (in meters) and turning angles (in radians). For the movement trajectory of each individual, the step lengths and turning angles between successive locations were calculated using the package adehabitatLT \citep{calenge2012package}.

In the previous study, hidden semi-Markov models (HSMM), an extension of HMM that permits explicit modelling of dwell times (ie. Time spent in a specific state) using alternative distributions \citep{langrock_flexible_2012} were used to reveal multiphasic movement of Florida panthers. Of 18 Florida panthers GPS-tracked in hourly intervals, 13 panthers were modelled independently with 2 to 6 states time homogeneous HSMM model with alternative distributions for step lengths (log-Normal, gamma, and Weibull), turning angle (von Mises and wrapped Cauchy) and dwell times (Poisson and negative binomial). Weibull and wrapped Cauchy were the choice of distributions chosen via Akaike information criterion (AIC) among the alternative distributions for step length and turning angle respectively. Furthermore, Bayesian information criterion (BIC) were used to evaluate goodness of fit between models with different number of states. BIC values of these models decrease as the number of states increased from three to six states suggesting the six states model was favoured statistically, on the other hand, it was hard to interpret the six states biologically. Despite the flexibily to model dwell time, it did not affect results, thus, in this paper, we will use HMM instead of HSMM. Furthermore, the state occupancies based on the Viterbi algorithm, which predicts the most likely sequence of movement states based on observed movement results \citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict future movement patterns and cannot capture which movement modes panthers are likely to occupy depending on time of day from the time homogeneous transitions (ie, the transition between all states are constant through time).

\subsection{Models}

In a HMM, the joint likelihood of emissions $\textbf{Y} =$ (\boldmath{$ y_{1},...,y_{T}$} \unboldmath) and hidden states $\textbf{Z}, z_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters \boldmath{$\theta$} \unboldmath and covariates $\textbf{X}_{1:T} =$ (\boldmath{$x_{1},...x_{T}$} \unboldmath), can be written as: \\

$P(\textbf{Y}_{1:T},\textbf{Z}_{1:T}\mid$\boldmath{$\theta$}\unboldmath$,\textbf{X}_{1:T})=P(z_{1}\mid $\boldmath{$x_{1}$}\unboldmath$)P($\boldmath{$y_{1}$}\unboldmath$\mid z_{1},$\boldmath{$x_{1}$}\unboldmath$) \\ \prod\limits_{k=2}^{T}P(z_{k} \mid z_{k-1},$\boldmath{$x_{k}$}\unboldmath$)P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$ \\

where we have the following three distinct components:

\begin{enumerate}
\item Initial Probability: $P(z_{1} = i \mid $\boldmath{$x_{1}$}\unboldmath$)P($\boldmath{$y_{1}$}\unboldmath $\mid z_{1},$ \boldmath{$x_{1}$}\unboldmath $)$, the probability of state $i$ at time $t=1$ where covariate \boldmath{$x_{1}$} \unboldmath and the vector of observation density \boldmath{$y_{1}$} \unboldmath condition on covariates \boldmath{$x_{1}$} \unboldmath at state $z_{1}$.

\item Transition Probability: $P(z_{k}=j \mid z_{k-1}=i,$\boldmath{$x_{k}$}\unboldmath$)$, the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $z_{k}$ at time $t=k$.

\item Emission Probability: $P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$, a vector of observation density \boldmath{$y_{k}$} \unboldmath condition on covariates \boldmath{$x_{k}$} \unboldmath at state $z_{k}$ at time $t=k$.  
\end{enumerate}

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{van_de_kerk_hidden_2014} for HMM fitting. BIC was used for model selection because we were interested in explaning the data generation process by identifying the 'true' number of underlying movement states.

We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by Morale et al. (2004). For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:

$\pi_{ij} = \frac{\mathrm{exp}(\eta_{ij}(t))}{1+\sum_{j=2}^{n}\mathrm{exp}(\eta_{ij}(t))},$ for $j={2,...,n}$

$\pi_{i1} = 1 - \sum_{j=2}^{n}\pi_{ij}$

where $\eta_{ij}(t)$ can be modelled as the following:

\begin{enumerate}
\item Multiple block transition:  The transitioning probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:

$\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm}b_{m,t}$

where $a_{ijm}$ are parameters, and $b_{m,t}$ is a binary indicator (1 to the time block at the corresponding time $t$, and 0 otherwise).

\item Quadratic functional transition with a time covariate:

$\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}$

\item Sinusoidal functional transition with a time covariate:

$\eta_{ij}(t)= b_{ij1}+b_{ij2}\mathrm{cos}(\frac{2\pi t}{24})+b_{ij3}\mathrm{sin}(\frac{2\pi t}{24})$

\end{enumerate}

We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches. First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\bigtriangleup$B = BIC - min(BIC)). Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity. Lastly, autocorrelation comparison between the observed and simulated data.

The package depmixS4 \citep{visser2010depmixs4} for R
(R Development Core Team 2011) was used for fitting covariate dependent transition HMM
models, simulating states and step lengths using the
estimated parameters, and for estimating states with the Viterbi
algorithm.

\section{Results}
<<BICred_plot_setup,echo=FALSE,warning=FALSE,message=FALSE,cache=TRUE>>=
load('tables.Rdata')
@

<<BICred_plot, echo=FALSE,warning=FALSE,message=FALSE,fig.cap="this is the first figure",dev="tikz">>=

biccomp <- ggplot(ReducedDF, aes(x=statesred,y=V1,colour=typered))+
  scale_y_continuous()+
  scale_size_continuous( name="Num of Para") + 
  geom_point(aes(size=V1.1))+ scale_colour_brewer(palette="Set1") + 
  labs(y=expression("$\\Delta$ BIC")) + 
  geom_line(aes(colour=typered),size=0.5)+  ggtitle("Reduction BIC Comparison") +theme_bw()
biccomp
@

Figure~\ref{fig:BICred_plot}

Adding temporal heterogeneity complexity in the transitions for 3 to 6 state HSMM was a challenge and the number of parameters increase dramatically, creating overfitting problems. Model reduction was used to balance out the increase in parameters and see if there are evidence suggesting adding temporal heterogeneity complexity can reduce the number of states to a statistically and biologically interpretable level. Without loss of generality, we assessed time homogeneous transition HMM models instead of HSMM and made sure each model reduction does not change the BIC selection of number of states, which HMM heavily depends on. Goodness of fit will decrease, but the information on the number of states remains the same.

We simulated simple covariate dependent transitions and fitted with covariate independent transitions to confirm if the covariate independent transition HMM over fit the true number of states from the covariate dependent transition HMM. From simulation, it is likely to over estimate the number of states by fitting time homogenous transition HMMs if the true transition is time inhomogeneous. See Appendix.

<<BIC plot, echo=FALSE,warning=FALSE,message=FALSE,dev="tikz">>=
biccomp1 <- ggplot(FullDF, aes(x=state,y=V1.1,colour=type))+scale_y_continuous()+ 
  labs(y=expression("$\\Delta$ BIC"))+
  scale_size_continuous(breaks=c(25,50,100,200,300), name="Num of Para") + 
  geom_point(aes(size=V1.2))+ scale_colour_brewer(palette="Set1") + geom_line(aes(colour=type),size=0.5)+  
  theme_bw()+ggtitle("Cat130 Simulation BIC Comparison")
biccomp1

@

Looking at figure ... time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HSMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1,2,3,4,5,6,21,22,23,24}$, $b_{2}$ for $ t={7,8,9,10,11,12,13,14,15,16}$, and $b_{3}$ for $t={17,18,19,20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.

<<Average step by time of day plot, echo=FALSE,warning=FALSE,message=FALSE>>=
simavgdistplot <- ggplot(AvgSLDF,aes(x=time2,y=avgdist,colour=type2)) + 
  geom_point() + ggtitle('Cat130 Average Distance Obs vs Sim')+geom_line()+
  scale_x_continuous(breaks=0:23) +scale_colour_brewer(palette="Set1")+ theme_bw()

simavgdistplot
@

Next, we used the fitted models and simulate the same number of observations as the observed movement data and compare the average distance by time with the observed data. The average distance from the time homogeneous simulation was consistent with minimal variation and flat through time, which does not agree with the marginal average distance distribution of the observed data. This suggest strong evidence in distinishing the goodness of fit between time homogeneous transition and time inhomogeneous transition models. All the time inhomogeneous transition models' simulation can capture the strong diurnal cycle exhabit by Florida panthers from the observed movement data (ie. longer step lengths during night time and shorter steplength during the day), where as the time homogeneous transition simulated data cannot.

<<acf plot, echo=FALSE,warning=FALSE,message=FALSE>>=
allmacf <- ggplot(MacfDF,aes(lag,ACF,colour=.id))+scale_colour_brewer(palette="Set1")+
  geom_point()+geom_line() + ggtitle('Cat130 Observation vs Simulation ACF')+theme_bw()
allmacf
@

Lastly, the observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity. Again, time homogeneous transition simulations fail to capture the autocorrelation, where as the time inhomogeneous transition simulations somewhat captures it. The autocorrelation cannot determine goodness of fit, but with supporting results from $\bigtriangleup$BIC and average distance by time of day, time inhomogeneous transition HMM models provides both statistical and biological evidence of handling temporal heterogenitiy that was not considered before.

<<msd plot, echo=FALSE,warning=FALSE,message=FALSE>>=
msdplot<- ggplot(MsdDF,aes(x=Mean,y=SD,colour=Type))+geom_point()+geom_line()+ggtitle("Cat130 Step length Mean vs SD") + theme_bw()

msdplot
@

The biggest issue with multiple states is identifying them biologically. The states itself have no fixed meaning, though it suggest distinct characteristics present in the data. We really do not know if there Florida panthers really moves according to hidden states, and even if they do, we don't know the true number nor able to observe them. Three distinct movement states are ideal and biologically interpretable for Florida panthers: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state. Long step lengths are easy to capture in all models, where as short step lengths are hard to capture given the GPS errors and unstable movements patterns suggesting multiple states associate with short step lengths. Thus, lower number of short step length states with higher variance makes more biologically than multiple distinguishable short step length states with low variance.

\section{Discussion}

Identifying behavioural states based on some set of observations is a common methodological problem in behavioural ecology, and finding a mutual agreement biologically and statistically is challenging. Models which suggest a higher number of states fit the data well but are hard to interpret biologically; models that are biologically interpretable are not statistically supported. There are numerous literatures of HMM framework but very limited on modeling more than three states. HMM is a simple straight forward framework, but HMM extensions of adding a small amount of complexity can complicate the model greatly and reveal deeper understanding. For example, \citet{van_de_kerk_hidden_2014} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitant distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.

We have presented a relatively simple concept yet unheard of extension by which covariate independent HMM can overestimate the number of true states in covariate dependent HMM (transition in our case). HMM with more than three states are not biologically interpretable and often discarded, but we cannot ignore that increasing complexity (number of states) states can increase goodness of fit statistically. Our analyses revealed modeling temporal heterogeneity in HMM transition can reduce the number of states estimated in time homogeneous transitions HMM movement of Florida panthers and reduce the overall model complexity.

Depending on what kind of movement predictions we want to make, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of estimating the states by the Viterbi algorithm and can easily overlook more complicated models. The Viterbi algorithm predicts the most likely sequence of movement states based on the observations which allows the time homogenous transition model's prediction to capture the temporal heterogeneity from the observations. In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.

Although the time inhomogeneous transition offer an
objective approach to (1) reduce the number of states to a biologically interpretable level, (2) captures the observational autocorrelation and temporal pattern, (3) reduce the overall complexity as illustrated by our study, several challenges
must be overcome in order to maximize its potential. The current model is based on the reduced univariate step length response emission without fitting turning angles.Although turning angles are strongly correlated with step length does not effect the overall prediction on number of states (see Appendix), movement trajectories predictions still require both step length and turning angles. Furthermore, modelling temporal heterogeneity only reduce the number of predicted states by 1 resulting in 4/5 states might suggest (1) other factors that might also contribute to over predicting number of states, or (2) there really are 1 or 2 more different movement characteristics that are biologically unknown. Despite the challenges, our study provides an important extension in HMM modeling that is not in the seen before and a real life application that it exist and applicable in animal movement ecology.

\bibliography{paper}

\end{document}

### 2015-10-22 paper3.Rnw
uploaded Rnw, should work now

@@
 
-\setkeys{Gin}{width=0.9\textwidth}
 \providecommand{\keywords}[1]{\textbf{\textit{Keywords:}} #1}
@@
 \begin{document}
+%\SweaveOpts{concordance=TRUE}
 %\SweaveOpts{concordance=TRUE}
@@
 
-Recent years, animal ecologists proposed the multiphasic movement framework, examining the complex functions of internal states and external ecological and environmental factors, attempting to capture the tendency of an individual to stay in one behaivoural state for an extended time period and switching to other behavioural states \citep{firle_influence_1998,nathan_movement_2008}.
-Behavioural states of a study animal in movement studies via biotelemetry are unobserved directly, however, the multiphasic movement framework provides a mean to interpret, characterize and distinguish movement patterns with specific functions.
-Models such as finite mixture model (FMM) \citep{tracey_mapping_2012} and hidden Markov model (HMM) \citep{van_de_kerk_hidden_2014} are able to both group observed movements into behavioural states and characterize movement hehaviour within the states.
+In recent years, animal ecologists have proposed the multiphasic movement framework, which considers combinations of complex unobserved functions of internal states and external (ecological and environmental) factors in an attempt to capture the tendency of an individual to stay in one behaivoural state for an extended time period and switching to other behavioural states \citep{firle_influence_1998,nathan_movement_2008}.
+Behavioural states of an animal in movement studies via biotelemetry are often unobserved directly, however, the multiphasic movement framework provides a mean to interpret, characterize and distinguish movement patterns with specific functions.
+Models such as finite mixture model (FMM) \citep{tracey_mapping_2012} and hidden Markov model (HMM) \citep{van_de_kerk_hidden_2014, patterson2009classifying, langrock_flexible_2012} are able both to group observed movements into behavioural states and characterize movement hehaviour within the states.
 In this paper, hidden Markov models and extensions are the model of choice but they are one choice out of a universe of possibilities.
 
-Finite mixture models (can be model as a hidden Markov model with constant probabilities between all states) do not account for serial dependence in the behavioural results thus, unable to make inference on underlying behavioural sequence.
+Finite mixture models (a special case of hidden Markov model with constant probabilities between all states) are designed to accommodate a particular form of heterogeneity in the movement pattern; that is, the movement pattern may consist of clusters/ states/ components each having a distinct distribution.
+Although FMMs are a useful tool, they do not handle one of the key features of movement date, sequential dependence.
+FMMs do not account for serial dependency in the behavioural results thus, unable to make inference on underlying behavioural sequence.
+Markov models are useful for this scenario, allowing sequential dependence relationships in the movements.
 The hidden Markov model, a discrete-time time-series model that allows serial dependence and focuses on observations that are regularly spaced in time, has been used to study animal movement and space use patterns \citep{franke_analysis_2004,franke_prediction_2006,patterson_classifying_2009}.
-The Hidden Markov model provides a better fit than latent-state models, and a transition component \citep{morales_extracting_2004} explaining the changes in behavioural sequence \citep{fryxell_multiple_2008}.
+Hidden Markov models provide a better fit than FMM (latent-state models for unobserved states), and a transition component \citep{morales_extracting_2004} explaining the changes in behavioural sequence \citep{fryxell_multiple_2008}.
 The hidden Markov framework captures the observational aspect of describing the underlying behavioural sequence and the behavioural result, however, the transitions estimated by hidden Markov models are homogeneous, which unnatural realistically and biologically.
@@
 
-The number of movement states directly affects the complexity of the movement model, thus, it is restricted by the number of observations and biologically distinctive properties.
-Most hidden Markov models limit to two underlying behavioural states based on binary assumption mixtures, three states are not unheard of, and beyond three states are difficult to interpret biologically given the behavioural states are unobserved directly.
+In many applications, the hidden states often have some significant characteristics so that the observed data can be reasonably grouped into distinguishable clusters.
+For example, two-state models are often used with dichotomous features \citep{schliehe-diecks_application_2012, mckellar_using_2014,langrock_flexible_2012, fryxell_multiple_2008}, and sometimes ( very few examples) three states  \citep{dean2012behavioural, morales_extracting_2004,franke_prediction_2006, kerk2015hidden}.
+Beyond three states are difficult to interpret biologically given the behavioural states are unobserved directly, and will require a lot more parameters and data to model.
 Higher quality and quantity of data are available over the years, thus the restriction of limited data can be overlooked.
@@
 
-Endangered spicies such as Florida panthers (Puma Concolor Coryi) revealed multiphasic movement when fitting hidden semi-Markov models (extension of hidden Markov models) \citep{van_de_kerk_hidden_2014}.
+Endangered spices such as Florida panthers (Puma Concolor Coryi) revealed multiphasic movement when fitting hidden semi-Markov models (extension of hidden Markov models) \citep{van_de_kerk_hidden_2014}.
 Three state models were used but greater than three states suggest better fits but were difficult to interpret biologically.
@@
 
-\section{Materials and Methods}
+\section{Motivation of Hidden Markov Extensions}
 
@@
 Locations used in this study were the hourly GPS coordinates then decomposed into step lengths (in meters) and turning angles (in radians).
-For the movement trajectory of each individual, the step lengths and turning angles between successive locations were calculated using the package adehabitatLT \citep{calenge2012package}.
 
-In the previous study, hidden semi-Markov models (HSMM), an extension of HMM that permits explicit modelling of dwell times (ie.
-Time spent in a specific state) using alternative distributions \citep{langrock_flexible_2012} were used to reveal multiphasic movement of Florida panthers.
+Hidden semi-Markov models (HSMM), an extension of HMM that permits explicit modelling of dwell times (ie.
+Time spent in a specific state) using alternative distributions \citep{langrock_flexible_2012} were used to reveal multiphasic movement of Florida panthers \citep{van_de_kerk_hidden_2014}.
 Of 18 Florida panthers GPS-tracked in hourly intervals, 13 panthers were modelled independently with 2 to 6 states time homogeneous HSMM model with alternative distributions for step lengths (log-Normal, gamma, and Weibull), turning angle (von Mises and wrapped Cauchy) and dwell times (Poisson and negative binomial).
@@
 
-\subsection{Models}
-
-In a HMM, the joint likelihood of emissions $\textbf{Y} =$ (\boldmath{$ y_{1},...,y_{T}$} \unboldmath) and hidden states $\textbf{Z}, z_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters \boldmath{$\theta$} \unboldmath and covariates $\textbf{X}_{1:T} =$ (\boldmath{$x_{1},...x_{T}$} \unboldmath), can be written as: \\
+In a HMM, the joint likelihood of emissions $\textbf{Y} =$ (\boldmath{$ y_{1},...,y_{T}$} \unboldmath) and an hidden states sequence $\textbf{Z}, z_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters \boldmath{$\theta$} \unboldmath and covariates $\textbf{X}_{1:T} =$ (\boldmath{$x_{1},...x_{T}$} \unboldmath), can be written as: \\
 
@@
 \end{enumerate}
+The likelihood above is the likelihood of the observed sequence given a particular hidden sequence.
+In order to calculate the likelihood of the observed sequence given the HMM, we need to sum over all possible hidden sequences.
+This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and estimate parameters numerically \citep{zucchini_hidden_2009}.
 
@@
 where $\eta_{ij}(t)$ can be modelled as the following: 
-
 \begin{enumerate}
@@
 
-The package depmixS4 \citep{visser2010depmixs4} for R
+The package depmixS4 \citep{visser2012package} for R
 (R Development Core Team 2011) was used for fitting covariate dependent transition HMM
@@
 \section{Results}
-<<BICred_plot_setup,echo=FALSE,warning=FALSE,message=FALSE,cache=TRUE>>=
-load('tables.Rdata')
-@
 
-<<BICred_plot, echo=FALSE,warning=FALSE,message=FALSE,fig.cap="this is the first figure",dev="tikz">>=
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figures/BICred_plot-1} \caption[this is the first figure]{this is the first figure}\label{fig:BICred_plot}
+\end{figure}
 
-biccomp <- ggplot(ReducedDF, aes(x=statesred,y=V1,colour=typered))+
-  scale_y_continuous()+
-  scale_size_continuous( name="Num of Para") + 
-  geom_point(aes(size=V1.1))+ scale_colour_brewer(palette="Set1") + 
-  labs(y=expression("$\\Delta$ BIC")) + 
-  geom_line(aes(colour=typered),size=0.5)+  ggtitle("Reduction BIC Comparison") +theme_bw()
-biccomp
-@
-
-Figure~\ref{fig:BICred_plot}
+\end{knitrout}
 
@@
 
-<<BIC plot, echo=FALSE,warning=FALSE,message=FALSE,dev="tikz">>=
-biccomp1 <- ggplot(FullDF, aes(x=state,y=V1.1,colour=type))+scale_y_continuous()+ 
-  labs(y=expression("$\\Delta$ BIC"))+
-  scale_size_continuous(breaks=c(25,50,100,200,300), name="Num of Para") + 
-  geom_point(aes(size=V1.2))+ scale_colour_brewer(palette="Set1") + geom_line(aes(colour=type),size=0.5)+  
-  theme_bw()+ggtitle("Cat130 Simulation BIC Comparison")
-biccomp1
+\begin{knitrout}
+\begin{figure}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figures/130all} \caption{this is second figure}
+\end{figure}
 
-@
+\end{knitrout}
 
-Looking at figure ...
-time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC.
-The time inhomogeneous transition HMMs reduced the number of states from the HSMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC.
+As a compliment, we also added FMM and FMM with temporal heterogeneities to compare the temporal effects in goodness of fit.
+Looking at figure 2 time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC.
+The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC.
 Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level.
@@
 
-<<Average step by time of day plot, echo=FALSE,warning=FALSE,message=FALSE>>=
-simavgdistplot <- ggplot(AvgSLDF,aes(x=time2,y=avgdist,colour=type2)) + 
-  geom_point() + ggtitle('Cat130 Average Distance Obs vs Sim')+geom_line()+
-  scale_x_continuous(breaks=0:23) +scale_colour_brewer(palette="Set1")+ theme_bw()
+Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC.
+The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1,2,3,4,5,6,21,22,23,24}$, $b_{2}$ for $ t={7,8,9,10,11,12,13,14,15,16}$, and $b_{3}$ for $t={17,18,19,20}$) and it reduced the number of states by 1 but had no effect on the females.
+The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.
+We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.
 
-simavgdistplot
-@
+\begin{knitrout}
+\begin{figure}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figures/130avg} \caption{Average step length by time of day}
+\end{figure}
+
+\end{knitrout}
 
 Next, we used the fitted models and simulate the same number of observations as the observed movement data and compare the average distance by time with the observed data.
-The average distance from the time homogeneous simulation was consistent with minimal variation and flat through time, which does not agree with the marginal average distance distribution of the observed data.
-This suggest strong evidence in distinishing the goodness of fit between time homogeneous transition and time inhomogeneous transition models.
-All the time inhomogeneous transition models' simulation can capture the strong diurnal cycle exhabit by Florida panthers from the observed movement data (ie.
-longer step lengths during night time and shorter steplength during the day), where as the time homogeneous transition simulated data cannot.
+Models that assume temporal homogeneity (FMM and HMM) fail to capture the diurnal activity cycle.
+Models incorporating temporal heterogeneity can capture the observed patterns in the hourly average step.
 
-<<acf plot, echo=FALSE,warning=FALSE,message=FALSE>>=
-allmacf <- ggplot(MacfDF,aes(lag,ACF,colour=.id))+scale_colour_brewer(palette="Set1")+
-  geom_point()+geom_line() + ggtitle('Cat130 Observation vs Simulation ACF')+theme_bw()
-allmacf
-@
+\begin{knitrout}
+\begin{figure}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figures/130acf} \caption{ACF }
+\end{figure}
+\end{knitrout}
 
-Lastly, the observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity.
-Again, time homogeneous transition simulations fail to capture the autocorrelation, where as the time inhomogeneous transition simulations somewhat captures it.
-The autocorrelation cannot determine goodness of fit, but with supporting results from $\bigtriangleup$BIC and average distance by time of day, time inhomogeneous transition HMM models provides both statistical and biological evidence of handling temporal heterogenitiy that was not considered before.
+Lastly,the observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity.
+FMM has no autocorrelation at all.
+HMM without temporal heterogeneity can capture the first few lags of the ACF but is unable to capture lags beyond 7.
+FMM with temporal heterogeneity can capture lags beyond 12, but not short-term correlation.
+HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data.
 
-<<msd plot, echo=FALSE,warning=FALSE,message=FALSE>>=
-msdplot<- ggplot(MsdDF,aes(x=Mean,y=SD,colour=Type))+geom_point()+geom_line()+ggtitle("Cat130 Step length Mean vs SD") + theme_bw()
-
-msdplot
-@
+\begin{knitrout}
+\begin{figure}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figures/msd_plot-1} 
+\end{figure}
+\end{knitrout}
 
@@
 
-Identifying behavioural states based on some set of observations is a common methodological problem in behavioural ecology, and finding a mutual agreement biologically and statistically is challenging.
-Models which suggest a higher number of states fit the data well but are hard to interpret biologically; models that are biologically interpretable are not statistically supported.
-There are numerous literatures of HMM framework but very limited on modeling more than three states.
+Identifying behavioural states based on some set of observations is a common methodological problem in behavioural ecology; finding solutions that are both biologically and statistically interpretable and defensible is challenging.
+Models that suggest a higher number of states fit the data well but are hard to interpret biologically; models that are biologically interpretable are not statistically supported.
+While the HMM framework has been applied in many research areas, very few studies have considered modeling with more than 3 states.
 HMM is a simple straight forward framework, but HMM extensions of adding a small amount of complexity can complicate the model greatly and reveal deeper understanding.
@@
 
-We have presented a relatively simple concept yet unheard of extension by which covariate independent HMM can overestimate the number of true states in covariate dependent HMM (transition in our case).
+We have presented a relatively simple concept yet litted-studied extension by covariate dependent HMM can overestimate the number of true states in covariate dependent HMM (in our case, time-dependent transitions).
 HMM with more than three states are not biologically interpretable and often discarded, but we cannot ignore that increasing complexity (number of states) states can increase goodness of fit statistically.
@@
 
-Depending on what kind of movement predictions we want to make, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of estimating the states by the Viterbi algorithm and can easily overlook more complicated models.
+For some kinds of model predictions, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of estimating the states by the Viterbi algorithm (the tradition state prediction method) and can easily overlook more complicated models.
 The Viterbi algorithm predicts the most likely sequence of movement states based on the observations which allows the time homogenous transition model's prediction to capture the temporal heterogeneity from the observations.
 In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.
+See Appendix.
 
@@
 
-\bibliography{paper}
+\bibliography{paper.bib}
 

### 2015-10-25 Fitting_and_Simulation_code.R
to new computer

@@ -184,26 +184,4 @@ simmixsin <- function(state,cat,fit){
   return(df)
 }
 
-
-hmmfits36 <- function(dat){
-  for(numstates in 3:6){
-    assign(paste("cat",dat$cat[1],"fithomo",numstates,sep=''),fithomo(numstates,dat))
-    assign(paste("cat",dat$cat[1],"fitsin",numstates,sep=''),fitsin(numstates,dat))
-    assign(paste("cat",dat$cat[1],"fitquad",numstates,sep=''),fitquad(numstates,dat))
-    assign(paste("cat",dat$cat[1],"fitblock",numstates,sep=''),fitblock(numstates,dat))
-    assign(paste("cat",dat$cat[1],"fitmix",numstates,sep=''),fitmix(numstates,dat))
-    assign(paste("cat",dat$cat[1],"fitmixsin",numstates,sep=''),fitblock(numstates,dat))
-  }
-}
-  
-hourly <- function(dat){
-  for(numstates in 3:4){
-    assign(paste("cat",dat$cat[1],"fithourly",numstates,sep=''),fithourly(numstates,dat))
-  }
-}
- 
-## simulate function
-
-    
-    
     

### 2015-10-25 pantherDataFrame.R

@@ -33,9 +33,9 @@ dat <- (dat %>%
 
 cats <- unique(dat$cat)
 
-for(cat in cats){
-    temp <- subset(dat,dat$cat == cat)
-    save(temp, file = paste0("cat",cat,".RData",sep=""))
+for(catnum in cats){
+    cat <- subset(dat,dat$cat == catnum)
+    save(cat, file = paste0("cat",catnum,".RData",sep=""))
 }
 
 

### 2015-10-25 fithomo.R
make building blocks

@@ -1,0 +1,8 @@
+##fit time homogeneous hmm
+
+fithomo3s <- fithomo(3,cat)
+fithomo4s <- fithomo(4,cat)
+fithomo5s <- fithomo(35cat)
+fithomo6s <- fithomo(6,cat)
+
+save(list(c(fithomo3s,fithomo4s,fithomo5s,fithomo6s)),file="hmmhomo.RData")

### 2015-10-25 fitsin.R

@@ -1,0 +1,8 @@
+##fit time dependent hmm (sin)
+
+fitsin3s <- fitsin(3,cat)
+fitsin4s <- fitsin(4,cat)
+fitsin5s <- fitsin(35cat)
+fitsin6s <- fitsin(6,cat)
+
+save(list(c(fitsin3s,fitsin4s,fitsin5s,fitsin6s)),file="hmmsin.RData")

### 2015-10-25 Fitting_and_Simulation_code.R
more blocks

@@ -110,7 +110,7 @@ fitblock <- function(state,cat,seed=2830){
   model <- depmix(LogDist~1,
                   data=cat,
                   nstate=state,
-                  transition=~factor(period),
+                  transition=~factor(Block),
                   family=gaussian())
   set.seed(seed)
   fitmodel <- fit(model)
@@ -121,7 +121,7 @@ simblock <- function(state,cat,fit){
   model <- depmix(LogDist~1,
                   data=cat,
                   nstate=state,
-                  transition=~factor(period),
+                  transition=~factor(Block),
                   family=gaussian())
   model<-setpars(model,getpars(fit))
   sim <- simhmm(model)

### 2015-10-25 fitblock.R

@@ -1,0 +1,8 @@
+##fit time dependent hmm (block)
+
+fitblock3s <- fitblock(3,cat)
+fitblock4s <- fitblock(4,cat)
+fitblock5s <- fitblock(35cat)
+fitblock6s <- fitblock(6,cat)
+
+save(list(c(fitblock3s,fitblock4s,fitblock5s,fitblock6s)),file="hmmblock.RData")

### 2015-10-25 fitquad.R

@@ -1,0 +1,8 @@
+##fit time dependent hmm (quad)
+
+fitquad3s <- fitquad(3,cat)
+fitquad4s <- fitquad(4,cat)
+fitquad5s <- fitquad(35cat)
+fitquad6s <- fitquad(6,cat)
+
+save(list(c(fitquad3s,fitquad4s,fitquad5s,fitquad6s)),file="hmmquad.RData")

### 2015-10-26 paper3.Rnw
revamped introduction

@@
 \usepackage{natbib}
+\usepackage[utf8]{inputenc} % for accented characters
+%% stuff for editing
+\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
+%% to suppress notes & comments: \usepackage[final]{changes}
+\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
 \bibliographystyle{chicago}
@@
 
-Animal ecologists are interested in factors that determine behavioural sequences of movement patterns (i.e.
-when animals move in a certain pattern, and under what circumstances they switch to alternative movement pattern).
-It is extremely difficult to observe, distinguish and classify the underlying behavioural modes/states exhibited by animals.
-Thus, the only observable information are the behavioural results from their location records.
-Over limited time scales, the path of a  moving individual can often be characterized by biased/correlated random walks \citep{okubo_diffusion_1980,turchin_quantitative_1998,okubo_diffusion_2001}, however; they fail to capture the movement patterns over longer periods due to change in movement behaviour \citep{firle_influence_1998,morales_scaling_2002}.
+%% BMB: note that it is *much* easier to keep track of changes in Github,
+%% which works on the line level, if you don't use line-wrap ...
 
-In recent years, animal ecologists have proposed the multiphasic movement framework, which considers combinations of complex unobserved functions of internal states and external (ecological and environmental) factors in an attempt to capture the tendency of an individual to stay in one behaivoural state for an extended time period and switching to other behavioural states \citep{firle_influence_1998,nathan_movement_2008}.
-Behavioural states of an animal in movement studies via biotelemetry are often unobserved directly, however, the multiphasic movement framework provides a mean to interpret, characterize and distinguish movement patterns with specific functions.
-Models such as finite mixture model (FMM) \citep{tracey_mapping_2012} and hidden Markov model (HMM) \citep{van_de_kerk_hidden_2014, patterson2009classifying, langrock_flexible_2012} are able both to group observed movements into behavioural states and characterize movement hehaviour within the states.
-In this paper, hidden Markov models and extensions are the model of choice but they are one choice out of a universe of possibilities.
+Given a sequence of animal movements observed via direct observation
+or telemetry, movement models aim to find a parsimonious description
+that can be used to understand past movements as well as predict
+future movements. Ecologists have long considered the effects of
+individual-level covariates (sex, age, nutritional status) and
+environmental covariates (habitat type, location of predators or prey)
+on movement \todo{citations}. More recently, modelers have developed
+\emph{hidden Markov models} (HMMs)
+\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012}
+-- formalized in ecological terms as the ``multiphasic movement
+framework'' \citep{fryxell_multiple_2008} -- that consider the effects
+of organisms' \emph{internal} states; in particular, HMMs model animal
+movement as though individual animals' movement behavior is determined
+by which of a discrete set of unobserved movement states
+(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
+Conditional on the state occupied by an individual, HMMs typically
+assume that animals follow a standard correlated walk model
+\citep{okubo_diffusion_1980,turchin_quantitative_1998}.
 
-Finite mixture models (a special case of hidden Markov model with constant probabilities between all states) are designed to accommodate a particular form of heterogeneity in the movement pattern; that is, the movement pattern may consist of clusters/ states/ components each having a distinct distribution.
-Although FMMs are a useful tool, they do not handle one of the key features of movement date, sequential dependence.
-FMMs do not account for serial dependency in the behavioural results thus, unable to make inference on underlying behavioural sequence.
-Markov models are useful for this scenario, allowing sequential dependence relationships in the movements.
-The hidden Markov model, a discrete-time time-series model that allows serial dependence and focuses on observations that are regularly spaced in time, has been used to study animal movement and space use patterns \citep{franke_analysis_2004,franke_prediction_2006,patterson_classifying_2009}.
-Hidden Markov models provide a better fit than FMM (latent-state models for unobserved states), and a transition component \citep{morales_extracting_2004} explaining the changes in behavioural sequence \citep{fryxell_multiple_2008}.
-The hidden Markov framework captures the observational aspect of describing the underlying behavioural sequence and the behavioural result, however, the transitions estimated by hidden Markov models are homogeneous, which unnatural realistically and biologically.
-Thus, to date, there is no existing framework that fits these two things together.
+Ever-increasing capabilities of remote
+sensors are making movement data available over an
+ever-wider range of time scales, at both higher resolution (e.g. hourly data
+from GPS collars vs. daily or weekly fixes for radio or VHF
+collars) and longer extent (e.g. from a few days to significant
+fractions of a year, or longer).  When analyzing such
+remote-sensing data, ecologists will more
+often have to account for temporal variability in movement
+behaviour at diurnal and seasonal scales that were previously
+not captured in the data.
 
-In many applications, the hidden states often have some significant characteristics so that the observed data can be reasonably grouped into distinguishable clusters.
-For example, two-state models are often used with dichotomous features \citep{schliehe-diecks_application_2012, mckellar_using_2014,langrock_flexible_2012, fryxell_multiple_2008}, and sometimes ( very few examples) three states  \citep{dean2012behavioural, morales_extracting_2004,franke_prediction_2006, kerk2015hidden}.
-Beyond three states are difficult to interpret biologically given the behavioural states are unobserved directly, and will require a lot more parameters and data to model.
-Higher quality and quantity of data are available over the years, thus the restriction of limited data can be overlooked.
-Model sections by goodness of fit are identifying numerous clusters suggesting potential behavioural states, beyond our understanding and interpretations biologically \citep{van_de_kerk_hidden_2014}.
-Thus, the question arises whether (1) it is reasonable to have numerous states beyond biological distinguishable interpretation and their meaning, (2) over predicting number of clusters under the absents of casual factors, and (3) a framework that can handle these assumptions.
+HMMs have typically been used to model movements over short time
+scales, where the probability of switching between movement states stays
+relatively constant; changes in behavioural mode switching based on
+the local environment can be accounted for incorporating environmental
+covariates in the HMM \citep{patterson_classifying_2009}, or by more
+\emph{ad hoc} comparisons between inferred state and environmental
+conditions \citep{fryxell_multiple_2008}. \todo{I \emph{think} Fryxell
+  et al use ad hoc rather than formal methods for their environment
+  comparison? check this ...}  \cite{schliehe-diecks_application_2012}
+consider changes in switching behaviour over time scales of six-hour
+observation periods\todo{check?}, but
+in general ecologists have used other tools to describe behavioural
+changes over longer (diurnal, seasonal, or ontogenetic) time scales.
+\todo{Fleming et al Am Nat 2014?  Gurarie?}
 
-Endangered spices such as Florida panthers (Puma Concolor Coryi) revealed multiphasic movement when fitting hidden semi-Markov models (extension of hidden Markov models) \citep{van_de_kerk_hidden_2014}.
-Three state models were used but greater than three states suggest better fits but were difficult to interpret biologically.
-In this paper, we will investigate the movement patterns of the Florida panther and use time covariates to model strong diurnal cycle observed in Florida panthers that was not previously accounted for and see if (1) temporal heterogeneity can reduce the number of states to an appropriate and biologically interpretable level, (2) readjusting the complexity levels of the hidden Markov framework can capture the behavioural inhomogeneous transition patterns, (3) capture the movement patterns over long period of time and able to produce biologically reasonable movements.
-The ultimate goal is to see if increasing the complexity in the hidden Markov framework can decrease the overall model complexity by reducing number of states.
+For movement behaviours that change on a very fast time scale, such
+that movement behaviours at successive observations are independent,
+\emph{finite mixture models} (FMM) -- a special case of HMMs where the
+probability of state occupancy is independent of the previous state --
+can adequately describe movement. \todo{are there animal movement
+  studies that use FMMs?}  When movement varies over long time scales
+(relative to the time between observations) with little short-term
+persistence or correlation, movement could be well represented by FMMs
+where the occupancy probabilities changed systematically
+\todo{smoothly? deterministically?} over time.  Thus FMMs and HMMs,
+with or without temporal variation in the occupancy or switching
+probabilities, form a useful family of models for capturing
+changes in movement behaviour over a range of time scales.
+
+Our primary goal in this paper is to introduce the use of
+HMMs with temporally varying switching probabilities -- in particular,
+switching probabilities that follow a diurnal cycle -- for modeling
+animal movement recorded over long time scales.  We take data from 
+\cite{van_de_kerk_hidden_2014}, who used temporally homogeneous
+HMMs \todo{we will clarify below that they actually used HSMMs} 
+to analyze hourly movement data
+from 18 Florida panthers (\emph{Puma concolor coryi}).
+
+\cite{van_de_kerk_hidden_2014} found that the best-fitting HMMs
+according to BIC incorporated a surprisingly large number of hidden
+behavioural states (as many as 6 for individuals with a large amount
+of available data); for practical reasons, they restricted their
+detailed analysis to models with only 3 underlying states.  In
+contrast, most studies using HMM have decided \emph{a priori} on a
+number of underlying states, typically using either two
+\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
+  fryxell_multiple_2008}, or three states
+\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,van_de_kerk_hidden_2014}.
+As \citeauthor{van_de_kerk_hidden_2014} comment, behavioural
+repertoires with more than three distinct states are difficult to
+interpret, one possible reason that other authors have not used
+\citeauthor{van_de_kerk_hidden_2014}'s model-based approach to
+identifying the number of behavioural modes.
+
+Our second goal, therefore, is to explore whether
+\citeauthor{van_de_kerk_hidden_2014}'s results might be driven at
+least in part by structural problems with the HMM, i.e.  the
+assumption of temporal homogeneity in behaviour.  When large data sets
+are available, information-theoretic model selection methods will
+typically choose complex, highly parameterized models; when there is
+only one way in which models can become more complex (e.g. by
+increasing the number of behavioural modes), complexity that is
+present in the data but not accounted for in the model (e.g. spatial
+or temporal heterogeneity) can be misidentified as other forms of
+complexity.  We predict that increasing volumes of data will
+increasingly lead researchers who are accustomed to fitting relatively
+small models to sparse data into such traps.  We examine whether
+allowing for diurnal variation in the Florida panther data leads to
+selection of models with smaller numbers of behavioural modes; we also
+fit models to simulated data with varying numbers of behavioural modes
+and degrees of temporal heterogeneity to test our conjecture that
+heterogeneity can be misidentified as behavioural complexity.
 
@@
 
-GPS collars were fitted to 18 Florida Panther (Puma Concolor Coryi) in 2005-2012 by Florida Fish and Wildlife and Conservation Commission (FWC) staff using trained hounds and houndsmen.
-Locations used in this study were the hourly GPS coordinates then decomposed into step lengths (in meters) and turning angles (in radians).
+GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
+Fish and Wildlife and Conservation Commission (FWC) staff using
+trained hounds and houndsmen. Locations used in this study were the
+hourly GPS coordinates then decomposed into step lengths (in meters)
+and turning angles (in radians).
 
-Hidden semi-Markov models (HSMM), an extension of HMM that permits explicit modelling of dwell times (ie.
-Time spent in a specific state) using alternative distributions \citep{langrock_flexible_2012} were used to reveal multiphasic movement of Florida panthers \citep{van_de_kerk_hidden_2014}.
-Of 18 Florida panthers GPS-tracked in hourly intervals, 13 panthers were modelled independently with 2 to 6 states time homogeneous HSMM model with alternative distributions for step lengths (log-Normal, gamma, and Weibull), turning angle (von Mises and wrapped Cauchy) and dwell times (Poisson and negative binomial).
-Weibull and wrapped Cauchy were the choice of distributions chosen via Akaike information criterion (AIC) among the alternative distributions for step length and turning angle respectively.
-Furthermore, Bayesian information criterion (BIC) were used to evaluate goodness of fit between models with different number of states.
-BIC values of these models decrease as the number of states increased from three to six states suggesting the six states model was favoured statistically, on the other hand, it was hard to interpret the six states biologically.
-Despite the flexibily to model dwell time, it did not affect results, thus, in this paper, we will use HMM instead of HSMM.
-Furthermore, the state occupancies based on the Viterbi algorithm, which predicts the most likely sequence of movement states based on observed movement results \citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict future movement patterns and cannot capture which movement modes panthers are likely to occupy depending on time of day from the time homogeneous transitions (ie, the transition between all states are constant through time).
+Hidden semi-Markov models (HSMM), an extension of HMM that permits
+explicit modelling of dwell times (ie. Time spent in a specific state)
+using alternative distributions \citep{langrock_flexible_2012} were
+used to reveal multiphasic movement of Florida panthers
+\citep{van_de_kerk_hidden_2014}. Of 18 Florida panthers GPS-tracked in
+hourly intervals, 13 panthers were modelled independently with 2 to 6
+states time homogeneous HSMM model with alternative distributions for
+step lengths (log-Normal, gamma, and Weibull), turning angle (von
+Mises and wrapped Cauchy) and dwell times (Poisson and negative
+binomial). Weibull and wrapped Cauchy were the choice of distributions
+chosen via Akaike information criterion (AIC) among the alternative
+distributions for step length and turning angle
+respectively. Furthermore, Bayesian information criterion (BIC) were
+used to evaluate goodness of fit between models with different number
+of states. BIC values of these models decrease as the number of states
+increased from three to six states suggesting the six states model was
+favoured statistically, on the other hand, it was hard to interpret
+the six states biologically. Despite the flexibility to model dwell
+time, it did not affect results, thus, in this paper, we will use HMM
+instead of HSMM. Furthermore, the state occupancies based on the
+Viterbi algorithm, which predicts the most likely sequence of movement
+states based on observed movement results
+\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
+future movement patterns and cannot capture which movement modes
+panthers are likely to occupy depending on time of day from the time
+homogeneous transitions (ie, the transition between all states are
+constant through time).
 
@@
 We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{van_de_kerk_hidden_2014} for HMM fitting.
-BIC was used for model selection because we were interested in explaning the data generation process by identifying the 'true' number of underlying movement states.
+BIC was used for model selection because we were interested in explaining the data generation process by identifying the 'true' number of underlying movement states.
 
-We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by Morale et al.
+We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by Morales et al.
 (2004).
@@
 HMM is a simple straight forward framework, but HMM extensions of adding a small amount of complexity can complicate the model greatly and reveal deeper understanding.
-For example, \citet{van_de_kerk_hidden_2014} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitant distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.
+For example, \citet{van_de_kerk_hidden_2014} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitat distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.
 
-We have presented a relatively simple concept yet litted-studied extension by covariate dependent HMM can overestimate the number of true states in covariate dependent HMM (in our case, time-dependent transitions).
+We have presented a relatively simple concept yet little-studied extension by covariate dependent HMM can overestimate the number of true states in covariate dependent HMM (in our case, time-dependent transitions).
 HMM with more than three states are not biologically interpretable and often discarded, but we cannot ignore that increasing complexity (number of states) states can increase goodness of fit statistically.

### 2015-10-26 fitsim.R
remove files that do not belong

@@ -1,8 +1,0 @@
-### fit and sim cat
-
-
-hmmfit(cat)
-
-hourly(cat)
-
-hmmsim()

### 2015-10-26 fitblock.R
don't need a script of functions

@@ -1,5 +1,19 @@
 ##fit time dependent hmm (block)
 
+library(depmixS4)
+
+fitblock <- function(state,cat,seed=2830){
+    model <- depmix(LogDist~1,
+                    data=cat,
+                    nstate=state,
+                    transition=~factor(Block),
+                    family=gaussian())
+    set.seed(seed)
+    fitmodel <- fit(model)
+    return(fitmodel)
+}
+
+
 fitblock3s <- fitblock(3,cat)
 fitblock4s <- fitblock(4,cat)
 fitblock5s <- fitblock(35cat)

### 2015-10-26 fithomo.R

@@ -1,4 +1,16 @@
 ##fit time homogeneous hmm
+library(depmixS4)
+
+fithomo <- function(state,cat){
+    model <- depmix(LogDist~1,
+                    data=cat,
+                    nstate=state,
+                    transition=~1,
+                    family=gaussian())
+    fitmodel <- fit(model)
+    return(fitmodel)
+}
+
 
 fithomo3s <- fithomo(3,cat)
 fithomo4s <- fithomo(4,cat)

### 2015-10-26 fitquad.R

@@ -1,4 +1,18 @@
 ##fit time dependent hmm (quad)
+library(depmixS4)
+
+fitquad <- function(state,cat,seed=2830){
+    model <- depmix(LogDist~1,
+                    data=cat,
+                    nstate=state,
+                    transition=~I(Time/24)+I((Time/24)^2),
+                    family=gaussian())
+    set.seed(seed)
+    fitmodel <- fit(model)
+    return(fitmodel)
+}
+
+
 
 fitquad3s <- fitquad(3,cat)
 fitquad4s <- fitquad(4,cat)

### 2015-10-26 fitsin.R

@@ -1,5 +1,19 @@
 ##fit time dependent hmm (sin)
 
+library(depmixS4)
+
+fitsin <- function(state,cat,seed=2830){
+    model <- depmix(LogDist~1,
+                    data=cat,
+                    nstate=state,
+                    transition=~cos(2*pi*Time/24)+ sin(2*pi*Time/24),
+                    family=gaussian())
+    set.seed(seed)
+    fitmodel <- fit(model)
+    return(fitmodel)
+}
+
+
 fitsin3s <- fitsin(3,cat)
 fitsin4s <- fitsin(4,cat)
 fitsin5s <- fitsin(35cat)

### 2015-10-26 pantherDataFrame.R
need rowwise transmute

@@ -11,7 +11,7 @@ block <- function(t){
 }
 
 ## creating suitable dataframe for depmixS4
-dat <- (dat %>% transmute(
+dat <- (dat %>% rowwise() %>% transmute(
     cat = animal_id,
     Sex = Sex,
     Time = as.numeric(Time) - 1, ##as.numeric conversion

### 2015-10-26 fitblock.R
fix typo

@@ -16,7 +16,7 @@ fitblock <- function(state,cat,seed=2830){
 
 fitblock3s <- fitblock(3,cat)
 fitblock4s <- fitblock(4,cat)
-fitblock5s <- fitblock(35cat)
+fitblock5s <- fitblock(5,cat)
 fitblock6s <- fitblock(6,cat)
 
 save(list(c(fitblock3s,fitblock4s,fitblock5s,fitblock6s)),file="hmmblock.RData")

### 2015-10-26 fithomo.R

@@ -14,7 +14,7 @@ fithomo <- function(state,cat){
 
 fithomo3s <- fithomo(3,cat)
 fithomo4s <- fithomo(4,cat)
-fithomo5s <- fithomo(35cat)
+fithomo5s <- fithomo(5,cat)
 fithomo6s <- fithomo(6,cat)
 
 save(list(c(fithomo3s,fithomo4s,fithomo5s,fithomo6s)),file="hmmhomo.RData")

### 2015-10-26 fitquad.R

@@ -16,7 +16,7 @@ fitquad <- function(state,cat,seed=2830){
 
 fitquad3s <- fitquad(3,cat)
 fitquad4s <- fitquad(4,cat)
-fitquad5s <- fitquad(35cat)
+fitquad5s <- fitquad(5,cat)
 fitquad6s <- fitquad(6,cat)
 
 save(list(c(fitquad3s,fitquad4s,fitquad5s,fitquad6s)),file="hmmquad.RData")

### 2015-10-26 fitsin.R

@@ -16,7 +16,7 @@ fitsin <- function(state,cat,seed=2830){
 
 fitsin3s <- fitsin(3,cat)
 fitsin4s <- fitsin(4,cat)
-fitsin5s <- fitsin(35cat)
+fitsin5s <- fitsin(5,cat)
 fitsin6s <- fitsin(6,cat)
 
 save(list(c(fitsin3s,fitsin4s,fitsin5s,fitsin6s)),file="hmmsin.RData")

### 2015-10-26 bioplots.R
all the pieces are there to generate BICplots but seed problem

@@ -1,0 +1,52 @@
+library(dplyr)
+library(plyr)
+library(ggplot2)
+
+fitlist <- list(fitfmm3,fitfmm4,fitfmm5,fitfmm6,
+               fitfmmsin3,fitfmmsin4,fitfmmsin5,fitfmmsin6,
+               fithomo3s,fithomo4s,fithomo5s,fithomo6s,
+               fithourly3,fithourly4,
+               fitblock3s,fitblock4s,fitblock5s,fitblock6s,
+               fitquad3s,fit130quad4s,fitquad5s,fitquad6s,
+                fitsin3s,fit130sin4s,fitsin5s,fitsin6s)
+
+sumdf <- function(lst){
+    BIC <- ldply(lst,BIC)
+    nstates <-ldply(lst,nstates)
+    para <- ldply(lst,npar)
+    model <- c('FMM','FMM','FMM','FMM',
+               'FMM + THsin','FMM + THsin','FMM + THsin','FMM + THsin',
+               'HMM','HMM','HMM','HMM',
+               'HMM + THhourly','HMM + THhourly',
+               'HMM + THblock','HMM + THblock','HMM + THblock','HMM + THblock',
+               'HMM + THquad','HMM + THquad','HMM + THquad','HMM + THquad',
+               'HMM + THsin','HMM + THsin','HMM + THsin','HMM + THsin')
+
+    type <- c('FMM','FMM','FMM','FMM',
+              'FMM + TH','FMM + TH','FMM + TH','FMM + TH',
+              'HMM','HMM','HMM','HMM',
+              'HMM + TH','HMM + TH',
+              'HMM + TH','HMM + TH','HMM + TH','HMM + TH',
+              'HMM + TH','HMM + TH','HMM + TH','HMM + TH',
+              'HMM + TH','HMM + TH','HMM + TH','HMM + TH')
+
+    deltaBIC <- BIC-min(BIC)
+    temp <- data.frame(BICS=BIC$V1,deltaBIC=deltaBIC$V1,nstates=nstates$V1,parameters=para$V1,model,type)
+    return(temp)
+}
+
+
+bicplot <- function(df){
+    temp <- ggplot(df, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
+        scale_y_continuous()+
+            scale_size_continuous( name="Num of Para") +
+                geom_point(aes(size=parameters))+
[…]

### 2015-10-26 fitblock.R

@@ -19,4 +19,4 @@ fitblock4s <- fitblock(4,cat)
 fitblock5s <- fitblock(5,cat)
 fitblock6s <- fitblock(6,cat)
 
-save(list(c(fitblock3s,fitblock4s,fitblock5s,fitblock6s)),file="hmmblock.RData")
+save(fitblock3s,fitblock4s,fitblock5s,fitblock6s,file="hmmblock.RData")

### 2015-10-26 fitfmm.R

@@ -1,0 +1,20 @@
+library(depmixS4)
+
+fitmix <- function(state,cat,seed=2830){
+    model <- mix(LogDist~1,
+                 data=cat,
+                 prior=~1,
+                 nstate=state,
+                 family=gaussian(),
+                 initdata=cat)
+    set.seed(seed)
+    fitmodel <- fit(model)
+    return(fitmodel)
+}
+
+fitfmm3 <- fitmix(3,cat)
+fitfmm4 <- fitmix(4,cat)
+fitfmm5 <- fitmix(5,cat)
+fitfmm6 <- fitmix(6,cat)
+
+save(fitfmm3,fitfmm4,fitfmm5,fitfmm6,file="fitfmm.RData")

### 2015-10-26 fithomo.R

@@ -17,4 +17,4 @@ fithomo4s <- fithomo(4,cat)
 fithomo5s <- fithomo(5,cat)
 fithomo6s <- fithomo(6,cat)
 
-save(list(c(fithomo3s,fithomo4s,fithomo5s,fithomo6s)),file="hmmhomo.RData")
+save(fithomo3s,fithomo4s,fithomo5s,fithomo6s,file="hmmhomo.RData")

### 2015-10-26 fithourly.R

@@ -1,0 +1,17 @@
+library(depmixS4)
+
+fithourly <- function(state,cat,seed=2830){
+    model <- depmix(LogDist~1,
+                    data=cat,
+                    nstate=state,
+                    transition=~factor(Time),
+                    family=gaussian())
+    set.seed(seed)
+    fitmodel <- fit(model)
+    return(fitmodel)
+}
+
+fithourly3 <- fithourly(3,cat)
+fithourly4 <- fithourly(4,cat)
+
+save(fithourly3,fithourly4,file="fithourly.RData")

### 2015-10-26 fitquad.R

@@ -19,4 +19,4 @@ fitquad4s <- fitquad(4,cat)
 fitquad5s <- fitquad(5,cat)
 fitquad6s <- fitquad(6,cat)
 
-save(list(c(fitquad3s,fitquad4s,fitquad5s,fitquad6s)),file="hmmquad.RData")
+save(fitquad3s,fitquad4s,fitquad5s,fitquad6s,file="hmmquad.RData")

### 2015-10-26 fitsin.R

@@ -19,4 +19,4 @@ fitsin4s <- fitsin(4,cat)
 fitsin5s <- fitsin(5,cat)
 fitsin6s <- fitsin(6,cat)
 
-save(list(c(fitsin3s,fitsin4s,fitsin5s,fitsin6s)),file="hmmsin.RData")
+save(fitsin3s,fitsin4s,fitsin5s,fitsin6s,file="hmmsin.RData")

### 2015-11-06 paper3.Rnw
fixed minor stuff in Rnw

@@
 %% stuff for editing
-\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
+\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout] %{changes}
 %% to suppress notes & comments: \usepackage[final]{changes}
@@
 
-\keywords{abc}
+\keywords{Hidden Markov, temporal heterogeneity}
 
@@
 environmental covariates (habitat type, location of predators or prey)
-on movement \todo{citations}. More recently, modelers have developed
+on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
+More recently, modelers have developed
 \emph{hidden Markov models} (HMMs)
@@
 consider changes in switching behaviour over time scales of six-hour
-observation periods\todo{check?}, but
+observation periods, but
 in general ecologists have used other tools to describe behavioural
@@
 can adequately describe movement. \todo{are there animal movement
-  studies that use FMMs?}  When movement varies over long time scales
+  studies that use FMMs?} 
+%% \citep{tracey_mapping_2012}  
+When movement varies over long time scales
 (relative to the time between observations) with little short-term
@@
 animal movement recorded over long time scales.  We take data from 
-\cite{van_de_kerk_hidden_2014}, who used temporally homogeneous
-HMMs \todo{we will clarify below that they actually used HSMMs} 
-to analyze hourly movement data
-from 18 Florida panthers (\emph{Puma concolor coryi}).
+\cite{van_de_kerk_hidden_2014}, who used temporally homogeneous hidden semi-Markov
+models, HSMMs (an extension of HMMs that allows flexible modelling of dwell time)
 
-\cite{van_de_kerk_hidden_2014} found that the best-fitting HMMs
+%%\todo{we will clarify below that they actually used HSMMs} 
+%%to analyze hourly movement data
+%%from 18 Florida panthers (\emph{Puma concolor coryi}).
+
+\cite{van_de_kerk_hidden_2014} found that the best-fitting HSMMs
 according to BIC incorporated a surprisingly large number of hidden
@@
   fryxell_multiple_2008}, or three states
-\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,van_de_kerk_hidden_2014}.
-As \citeauthor{van_de_kerk_hidden_2014} comment, behavioural
+\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
+van_de_kerk_hidden_2014}.As \citeauthor{van_de_kerk_hidden_2014} comment, behavioural
 repertoires with more than three distinct states are difficult to
@@
 \end{enumerate}
-The likelihood above is the likelihood of the observed sequence given a particular hidden sequence.
-In order to calculate the likelihood of the observed sequence given the HMM, we need to sum over all possible hidden sequences.
-This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and estimate parameters numerically \citep{zucchini_hidden_2009}.
+The likelihood above is the likelihood of the observed sequence given a particular
+hidden sequence. In order to calculate the likelihood of the observed sequence given the
+HMM, we need to sum over all possible hidden sequences. This direct method is computati
+nally inefficient, but there exist more efficient ways to computation the likelihood and
+estimate parameters numerically \citep{zucchini_hidden_2009}.
 

### 2015-11-06 paper3.Rnw
fixing issues

@@
 assume that animals follow a standard correlated walk model
-\citep{okubo_diffusion_1980,turchin_quantitative_1998}.
+\citep{okubo_diffusion_1980,turchin1998quantitative}.
 
@@
 animal movement recorded over long time scales.  We take data from 
-\cite{van_de_kerk_hidden_2014}, who used temporally homogeneous hidden semi-Markov
+\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
 models, HSMMs (an extension of HMMs that allows flexible modelling of dwell time)
@@
 
-\cite{van_de_kerk_hidden_2014} found that the best-fitting HSMMs
+\cite{kerk2015hidden} found that the best-fitting HSMMs
 according to BIC incorporated a surprisingly large number of hidden
@@
 \citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
-van_de_kerk_hidden_2014}.As \citeauthor{van_de_kerk_hidden_2014} comment, behavioural
+kerk2015hidden}.As \citeauthor{kerk2015hidden} comment, behavioural
 repertoires with more than three distinct states are difficult to
 interpret, one possible reason that other authors have not used
-\citeauthor{van_de_kerk_hidden_2014}'s model-based approach to
+\citeauthor{kerk2015hidden}'s model-based approach to
 identifying the number of behavioural modes.
@@
 Our second goal, therefore, is to explore whether
-\citeauthor{van_de_kerk_hidden_2014}'s results might be driven at
+\citeauthor{kerk2015hidden}'s results might be driven at
 least in part by structural problems with the HMM, i.e.  the
@@
 used to reveal multiphasic movement of Florida panthers
-\citep{van_de_kerk_hidden_2014}. Of 18 Florida panthers GPS-tracked in
+\citep{kerk2015hidden}. Of 18 Florida panthers GPS-tracked in
 hourly intervals, 13 panthers were modelled independently with 2 to 6
@@
 
-We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{van_de_kerk_hidden_2014} for HMM fitting.
+We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.
 BIC was used for model selection because we were interested in explaining the data generation process by identifying the 'true' number of underlying movement states.
@@
 
-The package depmixS4 \citep{visser2012package} for R
+The package depmixS4 \citep{visser2010depmixs4} for R
 (R Development Core Team 2011) was used for fitting covariate dependent transition HMM
@@
 HMM is a simple straight forward framework, but HMM extensions of adding a small amount of complexity can complicate the model greatly and reveal deeper understanding.
-For example, \citet{van_de_kerk_hidden_2014} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitat distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.
+For example, \citet{kerk2015hidden} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitat distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.
 

### 2015-11-06 Fitting_and_Simulation_code.R

@@ -4,12 +4,13 @@ library(dplyr)
 source("mikesim.R")
 
 ###Fit and simulate time-homogeneous HMMs
-fithomo <- function(state,cat){
+fithomo <- function(state,cat,seed=2830){
   model <- depmix(LogDist~1,
                   data=cat,
                   nstate=state,
                   transition=~1,
                   family=gaussian())
+  set.seed(seed)
   fitmodel <- fit(model)
   return(fitmodel)
 }

### 2015-11-06 pantherDataFrame.R

@@ -21,17 +21,12 @@ dat <- (dat %>% rowwise() %>% transmute(
   )
 )
 
-tempdat <- (dat %>% 
-    group_by(cat) %>% 
-      count(cat) %>% 
-        filter(n>9000)
-)
+cats <- c(1,2,14,15)
+
 
 dat <- (dat %>% 
-          filter(cat %in% tempdat$cat)
+          filter(cat %in% cats)
 )
-
-cats <- unique(dat$cat)
 
 for(catnum in cats){
     cat <- subset(dat,dat$cat == catnum)

### 2015-11-13 paper3.Rnw
have to make reproducible plots now

@@
 \emph{ad hoc} comparisons between inferred state and environmental
-conditions \citep{fryxell_multiple_2008}. \todo{I \emph{think} Fryxell
-  et al use ad hoc rather than formal methods for their environment
-  comparison? check this ...}  \cite{schliehe-diecks_application_2012}
+conditions \citep{fryxell_multiple_2008}.\cite{schliehe-diecks_application_2012}
 consider changes in switching behaviour over time scales of six-hour

### 2015-11-13 seedfind3.R

@@ -1,0 +1,115 @@
+##fit and sim functions
+library(depmixS4)
+library(dplyr)
+###Fit and simulate time-homogeneous HMMs
+fithomo <- function(state,cat,seed=2830){
+  model <- depmix(LogDist~1,
+                  data=cat,
+                  nstate=state,
+                  transition=~1,
+                  family=gaussian())
+  set.seed(seed)
+  fitmodel <- fit(model)
+  return(fitmodel)
+}
+
+#########################################################################
+
+## Fit and Sim of Hourly-transition HMMs
+
+fithourly <- function(state,cat,seed=2830){
+  model <- depmix(LogDist~1,
+                  data=cat,
+                  nstate=state,
+                  transition=~factor(Time),
+                  family=gaussian())
+  set.seed(seed)
+  fitmodel <- fit(model)
+  return(fitmodel)
+}
+
+########################################################################
+
+##fit and simulate time-dependent transition HMMs (sin)
+
+fitsin <- function(state,cat,seed=2830){
+  model <- depmix(LogDist~1,
+                  data=cat,
+                  nstate=state,
+                  transition=~cos(2*pi*Time/24)+ sin(2*pi*Time/24),
+                  family=gaussian())
+  set.seed(seed)
+  fitmodel <- fit(model)
+  return(fitmodel)
+}
+#######################################################################
+
+##fit and simulate time-dependent transition HMMs (quadratic)
+
+fitquad <- function(state,cat,seed=2830){
+  model <- depmix(LogDist~1,
+                  data=cat,
+                  nstate=state,
+                  transition=~I(Time/24)+I((Time/24)^2),
+                  family=gaussian())
+  set.seed(seed)
+  fitmodel <- fit(model)
+  return(fitmodel)
+}
+
+###########################################
+
+##fit and simulate time-dependent transition HMMs (block)
+
+fitblock <- function(state,cat,seed=2830){
+  model <- depmix(LogDist~1,
+                  data=cat,
+                  nstate=state,
+                  transition=~factor(Block),
+                  family=gaussian())
+  set.seed(seed)
+  fitmodel <- fit(model)
+  return(fitmodel)
+}
+
+###########################################################
+
+##fit and simulate FMMs
+
[…]

### 2015-11-13 fitblock.R
fithomo and fitsin is ready for all cats

@@ -2,7 +2,8 @@
 
 library(depmixS4)
 
-fitblock <- function(state,cat,seed=2830){
+fitblock <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
     model <- depmix(LogDist~1,
                     data=cat,
                     nstate=state,

### 2015-11-13 fitfmm.R

@@ -1,6 +1,7 @@
 library(depmixS4)
 
-fitmix <- function(state,cat,seed=2830){
+fitmix <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
     model <- mix(LogDist~1,
                  data=cat,
                  prior=~1,

### 2015-11-13 fitfmmsin.R

@@ -1,6 +1,7 @@
 library(depmixS4)
 
-fitmixsin <- function(state,cat,seed=2830){
+fitmixsin <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
     model <- mix(LogDist~1,
                  data=cat,
                  prior=~cos(2*pi*Time/24)+ sin(2*pi*Time/24),

### 2015-11-13 fithomo.R

@@ -1,7 +1,8 @@
 ##fit time homogeneous hmm
 library(depmixS4)
 
-fithomo <- function(state,cat){
+fithomo <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
     model <- depmix(LogDist~1,
                     data=cat,
                     nstate=state,
@@ -12,9 +13,9 @@ fithomo <- function(state,cat){
 }
 
 
-fithomo3s <- fithomo(3,cat)
-fithomo4s <- fithomo(4,cat)
-fithomo5s <- fithomo(5,cat)
-fithomo6s <- fithomo(6,cat)
+fithomo3s <- fithomo(3,cat,3)
+fithomo4s <- fithomo(4,cat,2)
+fithomo5s <- fithomo(5,cat,1)
+fithomo6s <- fithomo(6,cat,1)
 
 save(fithomo3s,fithomo4s,fithomo5s,fithomo6s,file="hmmhomo.RData")

### 2015-11-13 fithourly.R

@@ -1,6 +1,7 @@
 library(depmixS4)
 
-fithourly <- function(state,cat,seed=2830){
+fithourly <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
     model <- depmix(LogDist~1,
                     data=cat,
                     nstate=state,

### 2015-11-13 fitquad.R

@@ -1,7 +1,8 @@
 ##fit time dependent hmm (quad)
 library(depmixS4)
 
-fitquad <- function(state,cat,seed=2830){
+fitquad <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
     model <- depmix(LogDist~1,
                     data=cat,
                     nstate=state,

### 2015-11-13 fitsin.R

@@ -2,7 +2,8 @@
 
 library(depmixS4)
 
-fitsin <- function(state,cat,seed=2830){
+fitsin <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
     model <- depmix(LogDist~1,
                     data=cat,
                     nstate=state,
@@ -14,9 +15,9 @@ fitsin <- function(state,cat,seed=2830){
 }
 
 
-fitsin3s <- fitsin(3,cat)
-fitsin4s <- fitsin(4,cat)
-fitsin5s <- fitsin(5,cat)
-fitsin6s <- fitsin(6,cat)
+fitsin3s <- fitsin(3,cat,seed=4)
+fitsin4s <- fitsin(4,cat,seed=4)
+fitsin5s <- fitsin(5,cat,seed=4)
+fitsin6s <- fitsin(6,cat,seed=4)
 
 save(fitsin3s,fitsin4s,fitsin5s,fitsin6s,file="hmmsin.RData")

### 2015-11-13 pantherDataFrame.R

@@ -15,11 +15,13 @@ dat <- (dat %>% rowwise() %>% transmute(
     cat = animal_id,
     Sex = Sex,
     Time = as.numeric(Time) - 1, ##as.numeric conversion
-    Distance = Steplength.m. + 0.01, # log hack
+    Distance = Steplength.m.,
     LogDist = log10(Distance),
     Block = block(Time)
   )
 )
+
+dat$LogDist[is.infinite(dat$LogDist)] <- NA
 
 cats <- c(1,2,14,15)
 
@@ -27,6 +29,7 @@ cats <- c(1,2,14,15)
 dat <- (dat %>% 
           filter(cat %in% cats)
 )
+
 
 for(catnum in cats){
     cat <- subset(dat,dat$cat == catnum)

### 2015-11-13 seedfind3.R

@@ -9,7 +9,7 @@ fithomo <- function(state,cat,seed=2830){
                   transition=~1,
                   family=gaussian())
   set.seed(seed)
-  fitmodel <- fit(model)
+  fitmodel <- fit(model,verbose=TRUE,emcontrol=em.control(maxit=500))
   return(fitmodel)
 }
 
@@ -102,14 +102,18 @@ fitmixsin <- function(state,cat,seed=2830){
   return(fitmodel)
 }
 
-
+##########################################################
 cat1 <- subset(dat,dat$cat == 1 )
 cat2 <- subset(dat,dat$cat == 2 )
 cat14 <- subset(dat,dat$cat == 14 )
 cat15 <- subset(dat,dat$cat == 15 )
 
 
-hh <- fithomo(3,cat1,seed=3)
-hh <- fithomo(3,cat2,seed=3)
-hh <- fithomo(3,cat14,seed=220)
-hh <- fithomo(3,cat15,seed=3)
+###########################
+
+
+
+hh <- fitsin(5,cat1,seed=2)
+hh <- fitsin(5,cat2,seed=2)
+hh <- fitsin(5,cat14,seed=4)
+hh <- fitsin(5,cat15,seed=4)

### 2015-11-13 fitblock.R
all fitting complete, now have to plot BICs

@@ -15,9 +15,9 @@ fitblock <- function(state,cat,seed=NULL){
 }
 
 
-fitblock3s <- fitblock(3,cat)
-fitblock4s <- fitblock(4,cat)
-fitblock5s <- fitblock(5,cat)
-fitblock6s <- fitblock(6,cat)
+fitblock3s <- fitblock(3,cat,seed=3)
+fitblock4s <- fitblock(4,cat,seed=3)
+fitblock5s <- fitblock(5,cat,seed=3)
+fitblock6s <- fitblock(6,cat,seed=3)
 
 save(fitblock3s,fitblock4s,fitblock5s,fitblock6s,file="hmmblock.RData")

### 2015-11-13 fitfmmsin.R

@@ -13,9 +13,9 @@ fitmixsin <- function(state,cat,seed=NULL){
     return(fitmodel)
 }
 
-fitfmmsin3 <- fitmixsin(3,cat)
-fitfmmsin4 <- fitmixsin(4,cat)
-fitfmmsin5 <- fitmixsin(5,cat)
-fitfmmsin6 <- fitmixsin(6,cat)
+fitfmmsin3 <- fitmixsin(3,cat,seed=3)
+fitfmmsin4 <- fitmixsin(4,cat,seed=1)
+fitfmmsin5 <- fitmixsin(5,cat,seed=1)
+fitfmmsin6 <- fitmixsin(6,cat.seed=1)
 
 save(fitfmmsin3,fitfmmsin4,fitfmmsin5,fitfmmsin6,file="fitfmmsin.RData")

### 2015-11-13 fithourly.R

@@ -12,7 +12,7 @@ fithourly <- function(state,cat,seed=NULL){
     return(fitmodel)
 }
 
-fithourly3 <- fithourly(3,cat)
-fithourly4 <- fithourly(4,cat)
+fithourly3 <- fithourly(3,cat,seed=1)
+fithourly4 <- fithourly(4,cat,seed=1)
 
 save(fithourly3,fithourly4,file="fithourly.RData")

### 2015-11-13 fitmixseeds.R

@@ -1,0 +1,23 @@
+##fitmix seeds
+
+hh <- fitmix(3,cat14,seed=64)
+hh <- fitmix(3,cat15,seed=3)
+hh <- fitmix(3,cat1,seed=43)
+hh <- fitmix(3,cat2,seed=562)
+
+hh <- fitmix(4,cat14,seed=4)
+hh <- fitmix(4,cat15,seed=1)
+hh <- fitmix(4,cat1,seed=9)
+hh <- fitmix(4,cat2,seed=265)
+
+hh <- fitmix(5,cat14,seed=54)
+hh <- fitmix(5,cat15,seed=1)
+hh <- fitmix(5,cat1,seed=3)
+hh <- fitmix(5,cat2,seed=11)
+
+hh <- fitmix(6,cat14,seed=16)
+hh <- fitmix(6,cat15,seed=1)
+hh <- fitmix(6,cat1,seed=1)
+hh <- fitmix(6,cat2,seed=48)
+
+

### 2015-11-13 fitquad.R

@@ -15,9 +15,9 @@ fitquad <- function(state,cat,seed=NULL){
 
 
 
-fitquad3s <- fitquad(3,cat)
-fitquad4s <- fitquad(4,cat)
-fitquad5s <- fitquad(5,cat)
-fitquad6s <- fitquad(6,cat)
+fitquad3s <- fitquad(3,cat,seed=3)
+fitquad4s <- fitquad(4,cat,seed=3)
+fitquad5s <- fitquad(5,cat,seed=3)
+fitquad6s <- fitquad(6,cat,seed=3)
 
 save(fitquad3s,fitquad4s,fitquad5s,fitquad6s,file="hmmquad.RData")

### 2015-11-13 seedfind3.R

@@ -1,7 +1,49 @@
 ##fit and sim functions
 library(depmixS4)
 library(dplyr)
-###Fit and simulate time-homogeneous HMMs
+
+# setup -----------------
+
+dat <- read.csv('ArchivePantherData.csv')
+
+block <- function(t){
+  if(t %in% 7:16)return("block2")
+  if(t %in% 17:20)return("block3")
+  return("block1")
+}
+
+## creating suitable dataframe for depmixS4
+dat <- (dat %>% rowwise() %>% transmute(
+  cat = animal_id,
+  Sex = Sex,
+  Time = as.numeric(Time) - 1, ##as.numeric conversion
+  Distance = Steplength.m.,
+  LogDist = log10(Distance),
+  Block = block(Time)
+)
+)
+
+dat$LogDist[is.infinite(dat$LogDist)] <- NA
+
+cats <- c(1,2,14,15)
+
+
+dat <- (dat %>% 
+          filter(cat %in% cats)
+)
+
+# subsetting ---------------
+cat1 <- subset(dat,dat$cat == 1 )
+cat2 <- subset(dat,dat$cat == 2 )
+cat14 <- subset(dat,dat$cat == 14 )
+cat15 <- subset(dat,dat$cat == 15 )
+
+#######################
+
+
+
+
+#Fit time-homogeneous HMMs --------------
 fithomo <- function(state,cat,seed=2830){
   model <- depmix(LogDist~1,
                   data=cat,
@@ -13,9 +55,7 @@ fithomo <- function(state,cat,seed=2830){
   return(fitmodel)
 }
 
-#########################################################################
-
-## Fit and Sim of Hourly-transition HMMs
+## Fit and Sim of Hourly-transition HMMs ------
 
 fithourly <- function(state,cat,seed=2830){
   model <- depmix(LogDist~1,
@@ -28,9 +68,8 @@ fithourly <- function(state,cat,seed=2830){
   return(fitmodel)
 }
 
-########################################################################
 
-##fit and simulate time-dependent transition HMMs (sin)
+##fit and simulate time-dependent transition HMMs (sin) -----
 
 fitsin <- function(state,cat,seed=2830){
   model <- depmix(LogDist~1,
@@ -42,9 +81,8 @@ fitsin <- function(state,cat,seed=2830){
   fitmodel <- fit(model)
   return(fitmodel)
 }
[…]

### 2015-12-11 paper3.Rnw
miror edits

@@
 probability of state occupancy is independent of the previous state --
-can adequately describe movement. \todo{are there animal movement
-  studies that use FMMs?} 
-%% \citep{tracey_mapping_2012}  
+can adequately describe movement \citep{tracey_mapping_2012}.  
 When movement varies over long time scales
@@
 persistence or correlation, movement could be well represented by FMMs
-where the occupancy probabilities changed systematically
-\todo{smoothly? deterministically?} over time.  Thus FMMs and HMMs,
+where the occupancy probabilities changed deterministically over time.  Thus FMMs and HMMs,
 with or without temporal variation in the occupancy or switching
@@
 hidden sequence. In order to calculate the likelihood of the observed sequence given the
-HMM, we need to sum over all possible hidden sequences. This direct method is computati
-nally inefficient, but there exist more efficient ways to computation the likelihood and
+HMM, we need to sum over all possible hidden sequences. This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
 estimate parameters numerically \citep{zucchini_hidden_2009}.
@@
 
-We simulated simple covariate dependent transitions and fitted with covariate independent transitions to confirm if the covariate independent transition HMM over fit the true number of states from the covariate dependent transition HMM.
-From simulation, it is likely to over estimate the number of states by fitting time homogenous transition HMMs if the true transition is time inhomogeneous.
-See Appendix.
-
 \begin{knitrout}
@@
 
-As a compliment, we also added FMM and FMM with temporal heterogeneities to compare the temporal effects in goodness of fit.
+As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
 Looking at figure 2 time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC.

### 2015-12-21 paper3.Rnw
comment out \usepackage{changes} properly
doublespacing
changes in the paper

@@
 %% stuff for editing
-\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout] %{changes}
+%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
 %% to suppress notes & comments: \usepackage[final]{changes}
 \usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
+\usepackage{setspace}
 \bibliographystyle{chicago}
@@
 
+\doublespacing
 <<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
@@
 
-\section{Motivation of Hidden Markov Extensions}
+\section{Methods}
 
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
-\includegraphics[width=\maxwidth]{figures/BICred_plot-1} \caption[this is the first figure]{this is the first figure}\label{fig:BICred_plot}
+\includegraphics[width=\maxwidth]{figures/cat1BICPLOT} \caption{Model Reduction Process via Individual based Model Adjusted BIC }\label{fig:BICred_plot}
 \end{figure}
@@
 
-Adding temporal heterogeneity complexity in the transitions for 3 to 6 state HSMM was a challenge and the number of parameters increase dramatically, creating overfitting problems.
+Adding temporal heterogeneity in the transitions for 3 to 6 state HSMM was a challenge because the number of parameters and model complexity increase dramatically, thus creating overfitting problems.
 Model reduction was used to balance out the increase in parameters and see if there are evidence suggesting adding temporal heterogeneity complexity can reduce the number of states to a statistically and biologically interpretable level.
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/130all} \caption{this is second figure}
+\includegraphics[width=\maxwidth]{figures/cat1BIC} \caption{Overall Adjusted BIC Comparisons between different HMM models}
 \end{figure}
@@
 As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
-Looking at figure 2 time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC.
+Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC.
 The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC.
@@
 Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.
-
-Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC.
-The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1,2,3,4,5,6,21,22,23,24}$, $b_{2}$ for $ t={7,8,9,10,11,12,13,14,15,16}$, and $b_{3}$ for $t={17,18,19,20}$) and it reduced the number of states by 1 but had no effect on the females.
-The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.
-We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.
 
@@
 
-Lastly,the observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity.
+Lastly, the observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity.
 FMM has no autocorrelation at all.
@@
 The biggest issue with multiple states is identifying them biologically.
-The states itself have no fixed meaning, though it suggest distinct characteristics present in the data.
 We really do not know if there Florida panthers really moves according to hidden states, and even if they do, we don't know the true number nor able to observe them.
-Three distinct movement states are ideal and biologically interpretable for Florida panthers: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state.
+Three distinct movement states seems biologically interpretable for Florida panthers: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state.
 Long step lengths are easy to capture in all models, where as short step lengths are hard to capture given the GPS errors and unstable movements patterns suggesting multiple states associate with short step lengths.
-Thus, lower number of short step length states with higher variance makes more biologically than multiple distinguishable short step length states with low variance.
+Thus, lower number of short step length states with higher variance can be equalivent or more realistic than multiple distinguishable short step length states with low variance.
 
@@
 
-We have presented a relatively simple concept yet little-studied extension by covariate dependent HMM can overestimate the number of true states in covariate dependent HMM (in our case, time-dependent transitions).
+We have presented a relatively simple concept yet little-studied extension by covariate dependent HMM can overfit the number of true states in covariate dependent HMM (in our case, time-dependent transitions).
 HMM with more than three states are not biologically interpretable and often discarded, but we cannot ignore that increasing complexity (number of states) states can increase goodness of fit statistically.
@@
 must be overcome in order to maximize its potential.
-The current model is based on the reduced univariate step length response emission without fitting turning angles.Although turning angles are strongly correlated with step length does not effect the overall prediction on number of states (see Appendix), movement trajectories predictions still require both step length and turning angles.
-Furthermore, modelling temporal heterogeneity only reduce the number of predicted states by 1 resulting in 4/5 states might suggest (1) other factors that might also contribute to over predicting number of states, or (2) there really are 1 or 2 more different movement characteristics that are biologically unknown.
+The current model is based on the reduced univariate step length response emission without fitting turning angles.
+Although turning angles are strongly correlated with step length, it does not effect the overall prediction on number of states.
+But movement trajectories predictions, and spatial models still require both step length and turning angles.
+Furthermore, modelling temporal heterogeneity only reduce the number of predicted states by 1 resulting in 4/5 states might suggest (1) other factors such as spatial components that might also contribute to over predicting number of states, or (2) there really are 1 or 2 more different movement characteristics that are biologically unknown.
 Despite the challenges, our study provides an important extension in HMM modeling that is not in the seen before and a real life application that it exist and applicable in animal movement ecology.

### 2015-12-21 cat1simplots.R

@@ -1,7 +1,8 @@
 ## creating simdf 
 library(reshape2)
+source("plotsimfunctions.R")
 #load("cat1alldat.RData") my windows machine
-load("fitcat1.RData")
+#load("fitcat1.RData")
 tempdf <- data.frame(obs = cat$LogDist,
                      fmm5 = simfmm5$obs,
                      fmmsin4 = simfmmsin4$obs,

### 2015-12-21 plotsimfunctions.R

@@ -2,6 +2,7 @@ library(dplyr)
 library(plyr)
 library(ggplot2)
 library(depmixS4)
+library(tikzDevice)
 
 fitlist <- list(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,
                 fitfmmsin3,fitfmmsin4,fitfmmsin5,fitfmmsin6,
@@ -41,9 +42,10 @@ bicplot <- function(df){
     scale_y_continuous()+
     scale_size_continuous( name="Num of Parameters") +
     geom_point(aes(size=parameters))+
-    labs(y=expression("$\\Delta$ BIC"),
+    labs(y=expression(paste(Delta, "BIC")),
          x="Number of States") +
-    geom_line(aes(colour=model),size=0.5) +theme_bw()
+    geom_line(aes(colour=model),size=0.5) +theme_bw() +
+    ggtitle("Adjusted BIC Model Comparisons")
   return(temp)
 }
 

### 2015-12-26 allmodBIC.R
minor changes

@@ -1,0 +1,248 @@
+library(dplyr)
+library(plyr)
+library(ggplot2)
+library(depmixS4)
+library(circular)
+
+
+# VM setup ----
+setClass("vonMises", contains="response")
+## library(VGAM)  ## this may mess up multinomial()
+
+
+setGeneric("vonMises", function(y, pstart = NULL, fixed = NULL, ...) 
+  standardGeneric("vonMises"))
+
+setMethod("vonMises", 
+          signature(y = "ANY"), 
+          function(y, pstart = NULL, fixed = NULL, ...) {
+            y <- matrix(y, length(y))
+            x <- matrix(1) 
+            parameters <- list()
+            npar <- 2
+            if(is.null(fixed)) fixed <- as.logical(rep(0, npar))
+            if(!is.null(pstart)) {
+              if(length(pstart) != npar) stop("length of 'pstart' must be ", npar)
+              parameters$mu <- pstart[1]
+              parameters$kappa <- pstart[2]
+            }
+            mod <- new("vonMises", parameters = parameters, fixed = fixed,
+                       x = x, y = y, npar = npar)
+            mod
+          }
+)
+
+setMethod("dens","vonMises",
+          function(object,log=FALSE) {
+            dvonmises(object@y, mu = predict(object), 
+                      kappa = object@parameters$kappa, 
+                      log = log)
+          }
+)
+
+setMethod("getpars","response",
+          function(object,which="pars",...) {
+            switch(which,
+                   "pars" = {
+                     parameters <- numeric()
+                     parameters <- unlist(object@parameters)
+                     pars <- parameters
+                   },
+                   "fixed" = {
+                     pars <- object@fixed
+                   }
+            )
+            return(pars)
+          }
+)
+
+setMethod("setpars","vonMises",
+          function(object, values, which="pars", ...) {
+            npar <- npar(object)
+            if(length(values)!=npar) stop("length of 'values' must be",npar)
+            # determine whether parameters or fixed constraints are being set
+            nms <- names(object@parameters)
+            switch(which,
+                   "pars"= {
+                     object@parameters$mu <- values[1]
[…]

### 2015-12-26 weibullVM.R

@@ -107,7 +107,6 @@ setMethod("fit", "vonMises",
 setClass("weibull", contains="response")
 library(MASS)
 library(stats)
-data(speed)
 setGeneric("weibull", function(y, pstart = NULL, fixed = NULL, ...) 
   standardGeneric("weibull"))
 
@@ -196,7 +195,6 @@ setMethod("fit", "weibull",
 )
 
 # depmix model setup 3 states ----
-load("cat1.RData")
 dist <- pmax(cat$Distance,1e-3)
 rModels <- list()
 rModels[[1]] <- list()
@@ -223,8 +221,7 @@ inMod <- transInit(~ 1, ns = 3, pstart = rep(1/3, 3),
                    data = data.frame(1))
 mod3 <- makeDepmix(response = rModels, transition = transition,
                    prior=inMod,homogeneous = FALSE)
-fm3 <- fit(mod3, verbose = TRUE, emc=em.control(rand=FALSE))
-summary(fm3)
+WVMHMM3s <- fit(mod3, verbose = TRUE, emc=em.control(rand=FALSE))
 
 # HMM WVM 4 states ----
 rModels <- list()
@@ -258,8 +255,7 @@ inMod <- transInit(~ 1, ns = 4, pstart = rep(1/4, 4),
                    data = data.frame(1))
 mod4 <- makeDepmix(response = rModels, transition = transition,
                    prior=inMod,homogeneous = FALSE)
-fm4 <- fit(mod4, verbose = TRUE, emc=em.control(rand=FALSE))
-summary(fm4)
+WVMHMM4s <- fit(mod4, verbose = TRUE, emc=em.control(rand=FALSE))
 
 # HMM WVM 5 states ----
 
@@ -300,8 +296,7 @@ inMod <- transInit(~ 1, ns = 5, pstart = rep(1/5, 5),
                    data = data.frame(1))
 mod5 <- makeDepmix(response = rModels, transition = transition,
                    prior=inMod,homogeneous = FALSE)
-fm5 <- fit(mod5, verbose = TRUE, emc=em.control(rand=FALSE))
-summary(fm5)
+WVMHMM5s <- fit(mod5, verbose = TRUE, emc=em.control(rand=FALSE))
 
 # HMM WVM 6 states ----
 
@@ -354,7 +349,6 @@ summary(WVMHMM6s)
 
 
 # depmix model setup 3 states time het ----
-load("cat1.RData")
[…]

### 2015-12-26 paper3.Rnw

@@
 \begin{abstract}
+
 Clustering time-series data into discrete groups can improve
@@
 autocorrelation (persistence) in group occupancy can obscure such
-signals. We use latent-state and hidden Markov models (HMMs), two
-standard clustering techniques, to model high-resolution hourly
-movement data from Florida panthers. Allowing for temporal
-heterogeneity in transition probabilities, a straightforward but
-rarely explored model extension, resolves previous HMM modeling issues
+signals. We use finite mixture and hidden Markov models, two standard
+clustering techniques, to model high-resolution hourly movement data
+from Florida panthers. Allowing for temporal heterogeneity in
+transition probabilities, a straightforward but rarely explored model
+extension, resolves some shortcomings of current modeling frameworks
 and clarifies the behavioural patterns of panthers.
-...\end{abstract}
 
-\keywords{Hidden Markov, temporal heterogeneity}
+\end{abstract}
+
+\keywords{hidden Markov model, animal movement, temporal autocorrelation, temporal heterogeneity}
 
@@
 
-Given a sequence of animal movements observed via direct observation
-or telemetry, movement models aim to find a parsimonious description
+Given a sequence of animal movements observed directly 
+or via telemetry, movement models aim to find a parsimonious description
 that can be used to understand past movements as well as predict
@@
 \emph{hidden Markov models} (HMMs)
-\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012}
--- formalized in ecological terms as the ``multiphasic movement
-framework'' \citep{fryxell_multiple_2008} -- that consider the effects
+\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} %
+--- used in animal ecology under the rubric of the ``multiphasic movement
+framework'' \citep{fryxell_multiple_2008} --- that consider the effects
 of organisms' \emph{internal} states; in particular, HMMs model animal
@@
 HMMs have typically been used to model movements over short time
-scales, where the probability of switching between movement states stays
-relatively constant; changes in behavioural mode switching based on
+scales, where the probability of switching between movement states is
+approximately constant. Changes in behavioural mode switching 
+probabilities based on
 the local environment can be accounted for incorporating environmental
 covariates in the HMM \citep{patterson_classifying_2009}, or by more
-\emph{ad hoc} comparisons between inferred state and environmental
-conditions \citep{fryxell_multiple_2008}.\cite{schliehe-diecks_application_2012}
-consider changes in switching behaviour over time scales of six-hour
-observation periods, but
-in general ecologists have used other tools to describe behavioural
+\emph{ad hoc} comparisons between inferred states and environmental
+conditions \citep{fryxell_multiple_2008}. 
+\cite{schliehe-diecks_application_2012}
+consider temporal trends in switching behaviour over the
+time scales of a six-hour
+observation period, but
+in general ecologists have turned to other tools to describe behavioural
 changes over longer (diurnal, seasonal, or ontogenetic) time scales.
@@
 For movement behaviours that change on a very fast time scale, such
-that movement behaviours at successive observations are independent,
-\emph{finite mixture models} (FMM) -- a special case of HMMs where the
-probability of state occupancy is independent of the previous state --
+that movement behaviours at successive observations are effectively independent,
+\emph{finite mixture models} (FMM) --- which can be considered
+a special case of HMMs where the
+probability of state occupancy is independent of the previous state ---
 can adequately describe movement \citep{tracey_mapping_2012}.  
@@
 persistence or correlation, movement could be well represented by FMMs
-where the occupancy probabilities changed deterministically over time.  Thus FMMs and HMMs,
+where the occupancy probabilities change deterministically over time.  Thus FMMs and HMMs,
 with or without temporal variation in the occupancy or switching
@@
 switching probabilities that follow a diurnal cycle -- for modeling
-animal movement recorded over long time scales.  We take data from 
+animal movement recorded over long time scales.  We re-analyze data from 
 \cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
-models, HSMMs (an extension of HMMs that allows flexible modelling of dwell time)
+models (HSMMs: an extension of HMMs that allow
+flexible modelling of the distribution of \emph{dwell times}, the 
+lengths of consecutive occupancy of a behavioural state) to
+analyze the movement of Florida panthers.
 
@@
 
-\bibliography{paper.bib}
+\bibliography{paper}
 

### 2015-12-28 paper3.Rnw
more tweaks to methods

@@
 \cite{kerk2015hidden} found that the best-fitting HSMMs
-according to BIC incorporated a surprisingly large number of hidden
+incorporated a surprisingly large number of hidden
 behavioural states (as many as 6 for individuals with a large amount
@@
 
+\subsection{Data and previous analyses}
+
+\todo{what do we think of this subsection title?}
+
 GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
 Fish and Wildlife and Conservation Commission (FWC) staff using
-trained hounds and houndsmen. Locations used in this study were the
-hourly GPS coordinates then decomposed into step lengths (in meters)
-and turning angles (in radians).
+trained hounds and houndsmen. Of these panthers, 13 had 
+sufficient data to be used by \cite{kerk2015hidden}. Here we focus
+on the four cats with the most data, in part because our goal is to understand
+the issues that arise when simple models are fitted to large data sets, and 
+in part because the general trend in telemetry studies is to larger data sets \todo{give cat IDs - these are shown in Fig S2a of vdK.
+Give threshold amount of data; justify/describe how we decided.
+Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?}
+As is typical in studies of animal movement, 
+we took first differences of the data by decomposing
+contiguous sequences of hourly GPS coordinates into 
+successive step lengths (in meters) and turning angles (in radians)
+\citep{turchin1998quantitative,kerk2015hidden}.
 
-Hidden semi-Markov models (HSMM), an extension of HMM that permits
-explicit modelling of dwell times (ie. Time spent in a specific state)
-using alternative distributions \citep{langrock_flexible_2012} were
-used to reveal multiphasic movement of Florida panthers
-\citep{kerk2015hidden}. Of 18 Florida panthers GPS-tracked in
-hourly intervals, 13 panthers were modelled independently with 2 to 6
-states time homogeneous HSMM model with alternative distributions for
-step lengths (log-Normal, gamma, and Weibull), turning angle (von
-Mises and wrapped Cauchy) and dwell times (Poisson and negative
-binomial). Weibull and wrapped Cauchy were the choice of distributions
-chosen via Akaike information criterion (AIC) among the alternative
-distributions for step length and turning angle
-respectively. Furthermore, Bayesian information criterion (BIC) were
-used to evaluate goodness of fit between models with different number
-of states. BIC values of these models decrease as the number of states
-increased from three to six states suggesting the six states model was
-favoured statistically, on the other hand, it was hard to interpret
-the six states biologically. Despite the flexibility to model dwell
-time, it did not affect results, thus, in this paper, we will use HMM
-instead of HSMM. Furthermore, the state occupancies based on the
+\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
+an extension of HMM that permits explicit modelling of dwell times 
+\citep{langrock_flexible_2012}, considering both Poisson and
+negative binomial distributions.  As shown 
+\todo{I was going to say ``but not emphasized'', but that seems snarky}
+by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
+the estimated shape parameter of the negative binomial dwell time
+distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
+intervals were not calculated), implying that a geometric distribution
+(i.e., negative binomial with shape=1) might be adequate. In turn,
+this suggests that we might not lose much accuracy by reverting to 
+a simpler HMM framework, which makes precisely this assumption.
+\todo{could say here (or maybe better in the conclusions) that
+revisiting HSMM in this context would be worthwhile future work}
+
+\cite{kerk2015hidden} considered time-homogeneous models with 
+a variety of candidate distributions %
+--- log-Normal, Gamma, and Weibull distributions for step lengths
+and von Mises and wrapped Cauchy distributions for the turning angle --- %
+concluding on the basis of the Akaike information criterion (AIC) 
+that Weibull step length and wrapped Cauchy turning angle distributions
+were best.
+
+Our analysis again aims for simplicity and qualitative
+conclusions rather than for picking the very best
+predictive model; we focus on models that treat each step as
+a univariate, log-Normally distributed observation, glossing
+over both the relatively subtle distinctions in shape between
+the three candidate step-length distributions and the effects
+of considering multivariate (step length plus turning angle)
+observations. (Note that most movement analyses, including
+\cite{kerk2015hidden}, are only partially multivariate, 
+treating step length and turning angle at a particular time
+as independent observations and neglecting possible correlations
+between them.)  However, we do briefly compare all three
+step-length distributions, with and without incorporating
+a von Mises-distributed turning angle in the model (Figure~1).
+
+\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
+to test the relative penalized goodness of fit for
+models ranging from 2 to 6 latent behavioural modes.
+\todo{not sure about terminology here (should be consistent about
+what we call the modes/latent states/whatever}
+In general, BIC values decreased as the number of states
+increased from 
+three to six states,
+\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
+suggesting that the six-state model was
+favoured statistically. On the other hand, it was hard to interpret
+the six states biologically.
+We follow their lead in using BIC-optimality (i.e., minimum
+BIC across a family of models) as the criterion for identifying
+the best model, because we are interested in explaining the 
+data generation process by identifying the 'true' number 
+of underlying movement states.  Using BIC also simplifies \ldots
+\todo{cite S. Richards Ecology paper?  it's easier to see if BIC
+identifies the true model than to see if AIC finds the model with
+the minimum K-L distance; admit that tapering effect sizes do
+happen here; claim similar results with AIC, although
+generally higher number of states identified}
+
+\todo{This description of the V. algorithm is not coherent; need to give
+just a \emph{brief} definition here (since we are going to use it) and criticize/
+explain shortcomings later}
+
+Furthermore, the state occupancies based on the
 Viterbi algorithm, which predicts the most likely sequence of movement
@@
 constant through time).
+
+\subsection{Model description}
 
@@
 We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.
-BIC was used for model selection because we were interested in explaining the data generation process by identifying the 'true' number of underlying movement states.
 
-We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by Morales et al.
-(2004).
+We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by \cite{morales_extracting_2004}.
 For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:
 
-$\pi_{ij} = \frac{\mathrm{exp}(\eta_{ij}(t))}{1+\sum_{j=2}^{n}\mathrm{exp}(\eta_{ij}(t))},$ for $j={2,...,n}$
+$\pi_{ij} = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right),$ for $j={2,...,n}$
 
@@
 
-$\eta_{ij}(t)= b_{ij1}+b_{ij2}\mathrm{cos}(\frac{2\pi t}{24})+b_{ij3}\mathrm{sin}(\frac{2\pi t}{24})$
+$\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})$
 
@@
 We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches.
-First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\bigtriangleup$B = BIC - min(BIC)).
+First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\Delta$ BIC = BIC - min(BIC)).
 Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity.
@@
 As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
-Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC.
-The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC.
+Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC.
+The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC.
 Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level.

### 2015-12-29 cat14simplots.R
all plots are done, now fix up Makefile

@@ -2,7 +2,7 @@
 library(reshape2)
 source('plotsimfunctions.R')
 #load("cat14alldat.RData") my windows machine
-load("fitcat14.RData")
+#load("fitcat14.RData")
 tempdf <- data.frame(obs = cat$LogDist,
                      fmm4 = simfmm4$obs,
                      fmmsin3 = simfmmsin3$obs,
@@ -16,3 +16,11 @@ tempdf <- data.frame(obs = cat$LogDist,
 print(bicplot(sumdat))
 print(avgplot(tempdf))
 print(acfplot(tempdf))
+print(msd45(fitsin4s,fithomo5s))
+print(avgplot(vitcat(cat,fithomo3s)))
+print(avgplot(vitcat(cat,fitsin3s)))
+print(avgplot(vitcat5(cat,fithomo5s)))
+print(avgplot(vitcat5(cat,fitsin5s)))
+print(acfplot(vitcat(cat,fithomo3s))+ggtitle("Viterbi vs Observed ACF"))
+print(acfplot(vitcat5(cat,fitsin5s))+ggtitle("Viterbi vs Observed ACF"))
+

### 2015-12-29 cat15simplots.R

@@ -2,7 +2,7 @@
 library(reshape2)
 source('plotsimfunctions.R')
 #load("cat15alldat.RData") my windows machine
-load("fitcat15.RData")
+#load("fitcat15.RData")
 tempdf <- data.frame(obs = cat$LogDist,
                      fmm3 = simfmm3$obs,
                      fmmsin4 = simfmmsin4$obs,
@@ -16,3 +16,11 @@ tempdf <- data.frame(obs = cat$LogDist,
 print(bicplot(sumdat))
 print(avgplot(tempdf))
 print(acfplot(tempdf))
+print(msd56(fitsin5s,fithomo6s))
+print(avgplot(vitcat(cat,fithomo3s)))
+print(avgplot(vitcat(cat,fitsin3s)))
+print(avgplot(vitcat5(cat,fithomo5s)))
+print(avgplot(vitcat5(cat,fitsin5s)))
+print(acfplot(vitcat(cat,fithomo3s))+ggtitle("Viterbi vs Observed ACF"))
+print(acfplot(vitcat5(cat,fitsin5s))+ggtitle("Viterbi vs Observed ACF"))
+

### 2015-12-29 cat1simplots.R

@@ -16,3 +16,6 @@ tempdf <- data.frame(obs = cat$LogDist,
 print(bicplot(sumdat))
 print(avgplot(tempdf))
 print(acfplot(tempdf))
+print(msd56(fitsin5s,fithomo6s))
+print(avgplot(vitcat(cat,fithomo3s)))
+print(acfplot(vitcat(cat,fithomo3s))+ggtitle("Viterbi vs Observed ACF"))

### 2015-12-29 cat2simplots.R

@@ -2,7 +2,7 @@
 library(reshape2)
 source('plotsimfunctions.R')
 #load("cat2alldat.RData") my windows machine
-load("fitcat2.RData")
+#load("fitcat2.RData")
 tempdf <- data.frame(obs = cat$LogDist,
                      fmm4 = simfmm4$obs,
                      fmmsin4 = simfmmsin4$obs,
@@ -16,3 +16,6 @@ tempdf <- data.frame(obs = cat$LogDist,
 print(bicplot(sumdat))
 print(avgplot(tempdf))
 print(acfplot(tempdf))
+print(msd45(fitsin4s,fithomo5s))
+print(avgplot(vitcat(cat,fithomo3s)))
+print(acfplot(vitcat(cat,fithomo3s))+ggtitle("Viterbi vs Observed ACF"))

### 2015-12-29 plotsimfunctions.R

@@ -76,3 +76,87 @@ acfplot <- function(df) {
     ggtitle("ACF plot of BIC Selected Models")
   return(tempplot)
 }
+
+msd56 <- function(mod5,mod6) {
+  temp5 <- tail(getpars(mod5),10)
+  temp6 <- tail(getpars(mod6),12)
+  
+  dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
+                     log_sd=temp5[c(2,4,6,8,10)], states="5") 
+  dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
+                     log_sd=temp6[c(2,4,6,8,10,12)],states="6") 
+  
+  dat1 <- dat1[order(dat1$log_mean),]
+  dat2 <- dat2[order(dat2$log_mean),]
+  
+  dat <- rbind(dat1,dat2)
+  tempplot <- ggplot(dat,aes(log_mean,log_sd,color=states))+
+    geom_point()+geom_line() +theme_bw() +
+    ggtitle("State Parameters")
+  return(tempplot)
+}
+
+
+msd45 <- function(mod4,mod5) {
+  temp4 <- tail(getpars(mod4),8)
+  temp5 <- tail(getpars(mod5),10)
+  
+  dat1 <- data.frame(log_mean=temp4[c(1,3,5,7)],
+                     log_sd=temp4[c(2,4,6,8)], states="4") 
+  dat2 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
+                     log_sd=temp5[c(2,4,6,8,10)],states="5") 
+  
+  dat1 <- dat1[order(dat1$log_mean),]
+  dat2 <- dat2[order(dat2$log_mean),]
+  
+  dat <- rbind(dat1,dat2)
+  tempplot <- ggplot(dat,aes(log_mean,log_sd,color=states))+
+    geom_point()+geom_line() +theme_bw() +
+    ggtitle("State Parameters")
+  return(tempplot)
+}
+
+vitcat <- function(catt,mod){
+  params <- tail(getpars(mod),6)
+  vitdist <- numeric(ntimes(mod))
+  states <- viterbi(mod)$state
+  for(i in 1:ntimes(mod)){
+    if(states[i] == 1){
+      vitdist[i] <- rnorm(1,params[1],params[2])
+    }
+    if(states[i] == 2){
+      vitdist[i] <- rnorm(1,params[3],params[4])
+    }
[…]

### 2015-12-29 sample.cat1seeds.R
testing workflow

@@ -40,9 +40,6 @@ fitblock6s <- fitblock(6,cat,seed=2830)
 fithourly3 <- fithourly(3,cat,seed=1)
 fithourly4 <- fithourly(4,cat,seed=1)
 
-source('mikesim.R')
-source("simfunctions.R")
-
 simfmm5 <- simmix(5,cat,fitfmm5s)
 simfmmsin4 <- simmixsin(4,cat,fitfmmsin4)
 simhomo6 <- simhomo(6,cat,fithomo6s)
@@ -50,5 +47,3 @@ simhourly3 <- simhourly(3,cat,fithourly3)
 simblock4 <- simblock(4,cat,fitblock4s)
 simquad5 <- simquad(5,cat,fitquad5s)
 simsin5 <- simsin(5,cat,fitsin5s)
-
-save.image(file="fitcat1.RData")

### 2015-12-29 sample.R
updated sample test

@@ -1,0 +1,1 @@
+cat <- head(cat,100)

### 2015-12-29 sample.cat1seeds.R

@@ -6,6 +6,8 @@ fitfmm4s <- fitmix(4,cat,seed=9)
 fitfmm5s <- fitmix(5,cat,seed=3)
 fitfmm6s <- fitmix(6,cat,seed=1)
 
+quit()
+
 #FMM sin ----
 fitfmmsin3 <- fitmixsin(3,cat,seed=3)
 fitfmmsin4 <- fitmixsin(4,cat,seed=1)

### 2015-12-29 sample.R
it worked, now more cleaning

@@ -1,1 +1,1 @@
-cat <- head(cat,100)
+cat <- head(cat,750)

### 2015-12-29 sample.cat1seeds.R

@@ -6,8 +6,6 @@ fitfmm4s <- fitmix(4,cat,seed=9)
 fitfmm5s <- fitmix(5,cat,seed=3)
 fitfmm6s <- fitmix(6,cat,seed=1)
 
-quit()
-
 #FMM sin ----
 fitfmmsin3 <- fitmixsin(3,cat,seed=3)
 fitfmmsin4 <- fitmixsin(4,cat,seed=1)

### 2015-12-29 paper3.Rnw
Please remove and reclone the repo, the repo is clean enough I think

@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
-\includegraphics[width=\maxwidth]{figures/cat1BICPLOT} \caption{Model Reduction Process via Individual based Model Adjusted BIC }\label{fig:BICred_plot}
+\includegraphics[width=\maxwidth]{figures/cat1REDplot} \caption{Model Reduction Process via Individual based Model Adjusted BIC }\label{fig:BICred_plot}
 \end{figure}
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/130avg} \caption{Average step length by time of day}
+\includegraphics[width=\maxwidth]{figures/cat1avg} \caption{Average step length by time of day}
 \end{figure}
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/130acf} \caption{ACF }
+\includegraphics[width=\maxwidth]{figures/cat1acf} \caption{ACF }
 \end{figure}
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/msd_plot-1} 
+\includegraphics[width=\maxwidth]{figures/cat1states} 
 \end{figure}

### 2015-12-29 BICS.R

@@ -1,88 +1,0 @@
-library(depmixS4)
-library(ggplot2)
-library(dplyr)
-library(plyr)
-library(reshape2)
-
-
-loglik <- ldply(list(fit130homo3,fit130homo4,fit130homo5,fit130homo6,
-                     fit130hourly3,fit130hourly4,
-                     fit130block3,fit130block4,fit130block5,fit130block6,
-                     fit130quad3,fit130quad4,fit130quad5,fit130quad6,
-                     fit130sin3,fit130sin4,fit130sin5,fit130sin6),logLik)
-
-
-loglik <- loglik - min(loglik)
-state <- c(3,4,5,6,3,4,3,4,5,6,3,4,5,6,3,4,5,6)
-type <- c('Homogeneous','Homogeneous','Homogeneous','Homogeneous','Hourly','Hourly',
-          'Block','Block','Block','Block','Quadratic','Quadratic','Quadratic','Quadratic','Sinusoidal',
-          'Sinusoidal','Sinusoidal','Sinusoidal')
-bics <- ldply(list(fit130homo3,fit130homo4,fit130homo5,fit130homo6,
-                   fit130hourly3,fit130hourly4,
-                   fit130block3,fit130block4,fit130block5,fit130block6,
-                   fit130quad3,fit130quad4,fit130quad5,fit130quad6,
-                   fit130sin3,fit130sin4,fit130sin5,fit130sin6),BIC)
-lengthpars <- function(model){
-  return(length(getpars(model)))
-}
-
-parameters <- ldply(list(fit130homo3,fit130homo4,fit130homo5,fit130homo6,
-                         fit130hourly3,fit130hourly4,
-                         fit130block3,fit130block4,fit130block5,fit130block6,
-                         fit130quad3,fit130quad4,fit130quad5,fit130quad6,
-                         fit130sin3,fit130sin4,fit130sin5,fit130sin6),lengthpars)
-bics <- bics - min(bics)
-simdf <- data.frame(loglik,state,bics,type,parameters)
-
-bicsred <- ldply(list(fit130homo3,fit130homo4,fit130homo5,fit130homo6,
-                   fm3,fm4,fm5,fm6),BIC)
-
[…]

### 2015-12-29 Fitting_and_Simulation_code.R

@@ -1,188 +1,0 @@
-##fit and sim functions
-library(depmixS4)
-library(dplyr)
-source("mikesim.R")
-
-###Fit and simulate time-homogeneous HMMs
-fithomo <- function(state,cat,seed=2830){
-  model <- depmix(LogDist~1,
-                  data=cat,
-                  nstate=state,
-                  transition=~1,
-                  family=gaussian())
-  set.seed(seed)
-  fitmodel <- fit(model)
-  return(fitmodel)
-}
-
-simhomo <- function(state,cat,fit){
-  model <- depmix(LogDist~1,
-                  data=cat,
-                  nstate=state,
-                  transition=~1,
-                  family=gaussian())
-  model<-setpars(model,getpars(fit))
-  sim <- simhmm(model)
-  df <- data.frame(obs= sim@response[[1]][[1]]@y,states=sim@states)
-  return(df)
-}
-
-#########################################################################
-
-## Fit and Sim of Hourly-transition HMMs
-
-fithourly <- function(state,cat,seed=2830){
-  model <- depmix(LogDist~1,
-                  data=cat,
-                  nstate=state,
-                  transition=~factor(Time),
-                  family=gaussian())
-  set.seed(seed)
-  fitmodel <- fit(model)
-  return(fitmodel)
-}
-
-simhourly <- function(state,cat,fit){
-  model <- depmix(LogDist~1,
-                  data=cat,
-                  nstate=state,
-                  transition=~factor(Time),
-                  family=gaussian())
-  model<-setpars(model,getpars(fit))
-  sim <- simhmm(model)
-  df <- data.frame(obs= sim@response[[1]][[1]]@y,states=sim@states)
-  return(df)}
-########################################################################
-
-##fit and simulate time-dependent transition HMMs (sin)
-
-fitsin <- function(state,cat,seed=2830){
-  model <- depmix(LogDist~1,
-                  data=cat,
-                  nstate=state,
-                  transition=~cos(2*pi*Time/24)+ sin(2*pi*Time/24),
-                  family=gaussian())
-  set.seed(seed)
-  fitmodel <- fit(model)
-  return(fitmodel)
-}
-
-simsin <- function(state,cat,fit){
-  model <- depmix(LogDist~1,
-                  data=cat,
-                  nstate=state,
-                  transition=~cos((2*pi*Time)/24)+ sin((2*pi*Time)/24),
-                  family=gaussian())
[…]

### 2015-12-29 bicplots.R

@@ -1,52 +1,0 @@
-library(dplyr)
-library(plyr)
-library(ggplot2)
-library(depmixS4)
-
-fitlist <- list(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,
-                fitfmmsin3,fitfmmsin4,fitfmmsin5,fitfmmsin6,
-                fithomo3s,fithomo4s,fithomo5s,fithomo6s,
-                fithourly3,fithourly4,
-                fitblock3s,fitblock4s,fitblock5s,fitblock6s,
-                fitquad3s,fitquad4s,fitquad5s,fitquad6s,
-                fitsin3s,fitsin4s,fitsin5s,fitsin6s)
-
-sumdf <- function(lst){
-  BIC <- ldply(lst,BIC)
-  nstates <-ldply(lst,nstates)
-  para <- ldply(lst,npar)
-  model <- c('FMM','FMM','FMM','FMM',
-             'FMM + THsin','FMM + THsin','FMM + THsin','FMM + THsin',
-             'HMM','HMM','HMM','HMM',
-             'HMM + THhourly','HMM + THhourly',
-             'HMM + THblock','HMM + THblock','HMM + THblock','HMM + THblock',
-             'HMM + THquad','HMM + THquad','HMM + THquad','HMM + THquad',
-             'HMM + THsin','HMM + THsin','HMM + THsin','HMM + THsin')
-  
-  type <- c('FMM','FMM','FMM','FMM',
-            'FMM + TH','FMM + TH','FMM + TH','FMM + TH',
-            'HMM','HMM','HMM','HMM',
-            'HMM + TH','HMM + TH',
-            'HMM + TH','HMM + TH','HMM + TH','HMM + TH',
-            'HMM + TH','HMM + TH','HMM + TH','HMM + TH',
-            'HMM + TH','HMM + TH','HMM + TH','HMM + TH')
-  
-  deltaBIC <- BIC-min(BIC)
-  temp <- data.frame(BICS=BIC$V1,deltaBIC=deltaBIC$V1,nstates=nstates$V1,parameters=para$V1,model,type)
-  return(temp)
-}
-
-bicplot <- function(df){
-  temp <- ggplot(df, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
-    scale_y_continuous()+
-    scale_size_continuous( name="Num of Parameters") +
[…]

### 2015-12-29 fitblock.R

@@ -1,23 +1,0 @@
-##fit time dependent hmm (block)
-
-library(depmixS4)
-
-fitblock <- function(state,cat,seed=NULL){
-  if(!is.null(seed))set.seed(seed)
-    model <- depmix(LogDist~1,
-                    data=cat,
-                    nstate=state,
-                    transition=~factor(Block),
-                    family=gaussian())
-    set.seed(seed)
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-
-
-fitblock3s <- fitblock(3,cat,seed=2830)
-fitblock4s <- fitblock(4,cat,seed=2830)
-fitblock5s <- fitblock(5,cat,seed=2830)
-fitblock6s <- fitblock(6,cat,seed=2830)
-
-save(fitblock3s,fitblock4s,fitblock5s,fitblock6s,file="hmmblock.RData")

### 2015-12-29 fitfmm.R

@@ -1,57 +1,0 @@
-library(depmixS4)
-library(dplyr)
-
-# Fitting FMM function ------
-fitmix <- function(state,cat,seed=2830){
-  model <- mix(LogDist~1,
-               data=cat,
-               prior=~1,
-               nstate=state,
-               family=gaussian(),
-               initdata=cat)
-  set.seed(seed)
-  fitmodel <- fit(model)
-  return(fitmodel)
-}
-
-cat14 <- dat %>% filter(cat==14)
-
-# Fitting FMM to cat14 ------
-fitfmm3s <- fitmix(3,cat14,seed=64)
-fitfmm4s <- fitmix(4,cat14,seed=4)
-fitfmm5s <- fitmix(5,cat14,seed=54)
-fitfmm6s <- fitmix(6,cat14,seed=16)
-
-save(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,file="fmmcat14.RData")
-
-cat15 <- dat %>% filter(cat==15)
-
-# Fitting FMM to cat15 ------
-fitfmm3s <- fitmix(3,cat15,seed=3)
-fitfmm4s <- fitmix(4,cat15,seed=1)
-fitfmm5s <- fitmix(5,cat15,seed=1)
-fitfmm6s <- fitmix(6,cat15,seed=1)
-
-save(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,file="fmmcat15.RData")
-
-cat1 <- dat %>% filter(cat==1)
-
-# Fitting FMM to cat1 ------
-fitfmm3s <- fitmix(3,cat1,seed=43)
-fitfmm4s <- fitmix(4,cat1,seed=9)
-fitfmm5s <- fitmix(5,cat1,seed=3)
-fitfmm6s <- fitmix(6,cat1,seed=1)
-
-save(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,file="fmmcat1.RData")
-
-# Fitting FMM to cat2 ------
-
-cat2 <- dat %>% filter(cat==2)
-
-fitfmm3s <- fitmix(3,cat2,seed=562)
-fitfmm4s <- fitmix(4,cat2,seed=265)
-fitfmm5s <- fitmix(5,cat2,seed=11)
-fitfmm6s <- fitmix(6,cat2,seed=48)
-
-save(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,file="fmmcat2.RData")
-

### 2015-12-29 fitfmmsin.R

@@ -1,21 +1,0 @@
-library(depmixS4)
-
-fitmixsin <- function(state,cat,seed=NULL){
-  if(!is.null(seed))set.seed(seed)
-    model <- mix(LogDist~1,
-                 data=cat,
-                 prior=~cos(2*pi*Time/24)+ sin(2*pi*Time/24),
-                 nstate=state,
-                 family=gaussian(),
-                 initdata=cat)
-    set.seed(seed)
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-
-fitfmmsin3 <- fitmixsin(3,cat,seed=3)
-fitfmmsin4 <- fitmixsin(4,cat,seed=1)
-fitfmmsin5 <- fitmixsin(5,cat,seed=1)
-fitfmmsin6 <- fitmixsin(6,cat,seed=1)
-
-save(fitfmmsin3,fitfmmsin4,fitfmmsin5,fitfmmsin6,file="fitfmmsin.RData")

### 2015-12-29 fithomo.R

@@ -1,21 +1,0 @@
-##fit time homogeneous hmm
-library(depmixS4)
-
-fithomo <- function(state,cat,seed=NULL){
-  if(!is.null(seed))set.seed(seed)
-    model <- depmix(LogDist~1,
-                    data=cat,
-                    nstate=state,
-                    transition=~1,
-                    family=gaussian())
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-
-
-fithomo3s <- fithomo(3,cat,3)
-fithomo4s <- fithomo(4,cat,2)
-fithomo5s <- fithomo(5,cat,1)
-fithomo6s <- fithomo(6,cat,1)
-
-save(fithomo3s,fithomo4s,fithomo5s,fithomo6s,file="hmmhomo.RData")

### 2015-12-29 fithourly.R

@@ -1,18 +1,0 @@
-library(depmixS4)
-
-fithourly <- function(state,cat,seed=NULL){
-  if(!is.null(seed))set.seed(seed)
-    model <- depmix(LogDist~1,
-                    data=cat,
-                    nstate=state,
-                    transition=~factor(Time),
-                    family=gaussian())
-    set.seed(seed)
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-
-fithourly3 <- fithourly(3,cat,seed=1)
-fithourly4 <- fithourly(4,cat,seed=1)
-
-save(fithourly3,fithourly4,file="fithourly.RData")

### 2015-12-29 fitquad.R

@@ -1,23 +1,0 @@
-##fit time dependent hmm (quad)
-library(depmixS4)
-
-fitquad <- function(state,cat,seed=NULL){
-  if(!is.null(seed))set.seed(seed)
-    model <- depmix(LogDist~1,
-                    data=cat,
-                    nstate=state,
-                    transition=~I(Time/24)+I((Time/24)^2),
-                    family=gaussian())
-    set.seed(seed)
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-
-
-
-fitquad3s <- fitquad(3,cat,seed=3030)
-fitquad4s <- fitquad(4,cat,seed=3030)
-fitquad5s <- fitquad(5,cat,seed=3030)
-fitquad6s <- fitquad(6,cat,seed=3030)
-
-save(fitquad3s,fitquad4s,fitquad5s,fitquad6s,file="hmmquad.RData")

### 2015-12-29 fitsin.R

@@ -1,23 +1,0 @@
-##fit time dependent hmm (sin)
-
-library(depmixS4)
-
-fitsin <- function(state,cat,seed=NULL){
-  if(!is.null(seed))set.seed(seed)
-    model <- depmix(LogDist~1,
-                    data=cat,
-                    nstate=state,
-                    transition=~cos(2*pi*Time/24)+ sin(2*pi*Time/24),
-                    family=gaussian())
-    set.seed(seed)
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-
-
-fitsin3s <- fitsin(3,cat,seed=2830)
-fitsin4s <- fitsin(4,cat,seed=2830)
-fitsin5s <- fitsin(5,cat,seed=2830)
-fitsin6s <- fitsin(6,cat,seed=2830)
-
-save(fitsin3s,fitsin4s,fitsin5s,fitsin6s,file="hmmsin.RData")

### 2015-12-29 newfit.R

@@ -1,28 +1,0 @@
-library(depmixS4)
-library(dplyr)
-source("mikesim.R")
-
-mod <- function(cat,state){
-    model <- depmix(LogDist~1,
-                    data=cat,
-                    nstate=state,
-                    transition=~1,
-                    family=gaussian()
-                    )
-    return(model)
-}
-
-###Fit and simulate time-homogeneous HMMs
-fithomo <- function(cat,state,model=mod(cat,state)){
-    model <- mod
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-
-fitsin <- function(cat,state,model=mod(cat,state)){
-    model <- update(mod,.transition =
-                            ~cos(2*pi*Time/24)+ sin(2*pi*Time/24))
-    fitmodel <- fit(model)
-    return(fitmodel)
-}
-

### 2015-12-29 plotfuns.R

@@ -1,87 +1,0 @@
-library(dplyr)
-library(plyr)
-library(ggplot2)
-library(reshape2)
-
-sumdf <- function(lst){
-  BIC <- ldply(lst,BIC)
-  nstates <-ldply(lst,nstates)
-  para <- ldply(lst,npar)
-  model <- c('LSM','LSM','LSM','LSM',
-             'LSM + THsin','LSM + THsin','LSM + THsin','LSM + THsin',
-             'HMM','HMM','HMM','HMM',
-             'HMM + THhourly','HMM + THhourly',
-             'HMM + THblock','HMM + THblock','HMM + THblock','HMM + THblock',
-             'HMM + THquad','HMM + THquad','HMM + THquad','HMM + THquad',
-             'HMM + THsin','HMM + THsin','HMM + THsin','HMM + THsin')
-  
-  type <- c('LSM','LSM','LSM','LSM',
-            'LSM + TH','LSM + TH','LSM + TH','LSM + TH',
-            'HMM','HMM','HMM','HMM',
-            'HMM + TH','HMM + TH',
-            'HMM + TH','HMM + TH','HMM + TH','HMM + TH',
-            'HMM + TH','HMM + TH','HMM + TH','HMM + TH',
-            'HMM + TH','HMM + TH','HMM + TH','HMM + TH')
-  
-  deltaBIC <- BIC-min(BIC)
-  temp <- data.frame(BICS=BIC$V1,deltaBIC=deltaBIC$V1,nstates=nstates$V1,parameters=para$V1,model,type)
-  return(temp)
-}
-
-bicplot <- function(df){
-  temp <- ggplot(df, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
-    scale_y_continuous()+
-    scale_size_continuous(expression(paste("    # of \n parameters"))) + 
-    geom_point(aes(size=parameters))+ 
-    labs(y=expression(paste(Delta,'BIC'))) + 
-    geom_line(aes(colour=model),size=0.5) +theme_bw()
-  return(temp)
-}
-
-bicplot2 <- function(df,t1, t2=null){
-  tempdf <- df %>% filter((type == t1) | (type==t2)) %>% mutate(deltabic = BICS - min(BICS)) 
-  temp <- ggplot(tempdf, aes(x=nstates,y=deltabic,colour=model))+
-    scale_y_continuous()+
[…]

### 2015-12-29 someplotingthing.R

@@ -1,155 +1,0 @@
-source('plotfuns.R')
-
-sumdf130 <- sumdf(fit130)
-
-cat130simdf <- data.frame(cat130simdf,tempbest,tempdat2)
-
-bicplot(sumdf130) + labs(x="Number of States") + scale_colour_discrete()
-bicplot2(sumdf130,'HMM','HMM + TH') + labs(x="Number of States") + scale_colour_discrete() +facet_wrap(~ type,ncol=2)
-bicplot1(sumdf130,'FMM') + labs(x="Number of States") + 
-  scale_colour_discrete() +facet_wrap(~ type,ncol=2)
-
-
-head(sumdf130)
-
-bicplot(sumdf131) + labs(x = "Number of States")
-bicplot(sumdf94) + labs(x = "Number of States")
-bicplot(sumdf48) + labs(x = "Number of States")
-
-acfplot(cat130simdf%>%select(c(time,obs,vitobs)))
-
-avgplot((cat130simdf%>%select(c(time,obs,vitobs))))
-
-temp130 <- cat130simdf %>% mutate(resvit=obs-vitobs,
-                                  resmean=obs-mean,
-                                  resvitfmm=obs-vitobs.1,
-                                  vitobsFMM=vitobs.1)
-
-acfplot(temp130%>%select(c(time,obs,resvit,resmean,resvitfmm)))
-
-
-acfplot(cat130simdf%>%select(c(time,FMM3,FMMsin6,HMM6,HMMsin5,obs)))
-acfplot(cat130simdf%>%select(c(time,FMM3,obs)))
-acfplot(cat130simdf%>%select(c(time,FMMsin4,obs)))
-acfplot(cat130simdf%>%select(c(time,HMM6,obs)))
-acfplot(cat130simdf%>%select(c(time,HMMsin5,obs)))
-
-
-avgplot(cat130simdf%>%select(c(time,FMM3,FMMsin6,HMM6,HMMsin5,obs)))+labs(y=expression("log Step Length"),x=expression("Time of Day"))+ggtitle("Average Step Length by Time of Day")
-avgplot(cat130simdf%>%select(c(time,HMMsin5,obs,HMMhourly4)))+labs(y=expression("log Step Length"),x=expression("Time of Day"))
-
-##vit acf and avg
[…]

### 2015-12-29 paper3.Rnw
continuing tweaks on methods

@@
 \usepackage{natbib}
+\usepackage{amsmath}
 \usepackage[utf8]{inputenc} % for accented characters
@@
 complexity.  We predict that increasing volumes of data will
-increasingly lead researchers who are accustomed to fitting relatively
+increasingly lead researchers who are accustomed to fitting 
 small models to sparse data into such traps.  We examine whether
@@
 treating step length and turning angle at a particular time
-as independent observations and neglecting possible correlations
+as independent observations while neglecting possible correlations
 between them.)  However, we do briefly compare all three
@@
 suggesting that the six-state model was
-favoured statistically. On the other hand, it was hard to interpret
-the six states biologically.
-We follow their lead in using BIC-optimality (i.e., minimum
+favoured statistically; however,
+the authors used 3-state models in most
+of their analyses for ease of biological interpretation.
+
+We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
 BIC across a family of models) as the criterion for identifying
@@
 data generation process by identifying the 'true' number 
-of underlying movement states.  Using BIC also simplifies \ldots
-\todo{cite S. Richards Ecology paper?  it's easier to see if BIC
-identifies the true model than to see if AIC finds the model with
-the minimum K-L distance; admit that tapering effect sizes do
-happen here; claim similar results with AIC, although
-generally higher number of states identified}
+of underlying movement states.  Using BIC also simplifies 
+evaluation of model selection procedures; 
+it is more convenient to test whether our model selection
+procedure has selected the model used to simulate the data,
+rather than testing whether it has selected the model with
+the minimal Kullback-Leibler distance
+\citep{Richards2005}. We recognize that 
+ecologists will often be interested in maximizing predictive
+accuracy rather than selecting a true model, and that as
+usual in ecological systems the true model will be far more
+complicated than any candidate model \citep{BurnhamAnderson1998},
+but we suspect that the qualitative conclusions stated here
+for BIC-optimality will carry over to analyses using AIC instead.
 
@@
 
-In a HMM, the joint likelihood of emissions $\textbf{Y} =$ (\boldmath{$ y_{1},...,y_{T}$} \unboldmath) and an hidden states sequence $\textbf{Z}, z_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters \boldmath{$\theta$} \unboldmath and covariates $\textbf{X}_{1:T} =$ (\boldmath{$x_{1},...x_{T}$} \unboldmath), can be written as: \\
+\newcommand{\obs}{\mathbf{y}}
+\newcommand{\cov}{\mathbf{x}}
+\newcommand{\state}{z}
 
-$P(\textbf{Y}_{1:T},\textbf{Z}_{1:T}\mid$\boldmath{$\theta$}\unboldmath$,\textbf{X}_{1:T})=P(z_{1}\mid $\boldmath{$x_{1}$}\unboldmath$)P($\boldmath{$y_{1}$}\unboldmath$\mid z_{1},$\boldmath{$x_{1}$}\unboldmath$) \\ \prod\limits_{k=2}^{T}P(z_{k} \mid z_{k-1},$\boldmath{$x_{k}$}\unboldmath$)P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$ \\
+In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $\mathbf Y = \obs_{1},..., \obs_{T}$ and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters {$\boldsymbol \theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
+\begin{equation}
+\begin{split}
+P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
+P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
+&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
+\end{split}
+\label{eq:HMMlik}
+\end{equation}
 
-where we have the following three distinct components:
+The model contains three distinct components:
 
-\begin{enumerate}
-\item Initial Probability: $P(z_{1} = i \mid $\boldmath{$x_{1}$}\unboldmath$)P($\boldmath{$y_{1}$}\unboldmath $\mid z_{1},$ \boldmath{$x_{1}$}\unboldmath $)$, the probability of state $i$ at time $t=1$ where covariate \boldmath{$x_{1}$} \unboldmath and the vector of observation density \boldmath{$y_{1}$} \unboldmath condition on covariates \boldmath{$x_{1}$} \unboldmath at state $z_{1}$.
+\begin{description}
+\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$, the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$ and the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ at state $\state_{1}$.
 
-\item Transition Probability: $P(z_{k}=j \mid z_{k-1}=i,$\boldmath{$x_{k}$}\unboldmath$)$, the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $z_{k}$ at time $t=k$.
+\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$, the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\state_{k}$ at time $t=k$.
 
-\item Emission Probability: $P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$, a vector of observation density \boldmath{$y_{k}$} \unboldmath condition on covariates \boldmath{$x_{k}$} \unboldmath at state $z_{k}$ at time $t=k$.
-\end{enumerate}
-The likelihood above is the likelihood of the observed sequence given a particular
-hidden sequence. In order to calculate the likelihood of the observed sequence given the
-HMM, we need to sum over all possible hidden sequences. This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
+\item[Emission probability] $P({\boldmath y_{k}} | \state_{k},\cov_{k})$, a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
+\end{description}
+
+Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
+given (conditional on) a particular
+hidden sequence. 
+In order to calculate the overall, unconditional (or marginal) 
+likelihood of the 
+observed sequence, we need to average over all possible hidden sequences. 
+This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
 estimate parameters numerically \citep{zucchini_hidden_2009}.
@@
 
-We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by \cite{morales_extracting_2004}.
+We modeled three types of transition complexity instead of time homogeneous transition; extension of ``Switch with covariates" by \cite{morales_extracting_2004}.
 For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:
-
-$\pi_{ij} = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right),$ for $j={2,...,n}$
-
-$\pi_{i1} = 1 - \sum_{j=2}^{n}\pi_{ij}$
+\begin{equation}
+\begin{split}
+\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
+\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
+\end{split}
+\end{equation}
 
@@
 \begin{enumerate}
-\item Multiple block transition:  The transitioning probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
-
-$\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm}b_{m,t}$
-
+\item Multiple block transition:  The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
+$$
+\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm}b_{m,t}
+$$
 where $a_{ijm}$ are parameters, and $b_{m,t}$ is a binary indicator (1 to the time block at the corresponding time $t$, and 0 otherwise).  
-
 \item Quadratic functional transition with a time covariate:
-
-$\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}$
-
+$$
+\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}
+$$
 \item Sinusoidal functional transition with a time covariate:
-
-$\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})$
-
+$$
+\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})
+$$
 \end{enumerate}
@@
 Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC.
-The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1,2,3,4,5,6,21,22,23,24}$, $b_{2}$ for $ t={7,8,9,10,11,12,13,14,15,16}$, and $b_{3}$ for $t={17,18,19,20}$) and it reduced the number of states by 1 but had no effect on the females.
+The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females.
 The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.

### 2015-12-29 paper3.Rnw
more on methods

@@
 \begin{document}
+\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}
+
 %\SweaveOpts{concordance=TRUE}
@@
 approximately constant. Changes in behavioural mode switching 
+\todo{do we want to call these transition probabilities (throughout)?}
 probabilities based on
@@
 \citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
-kerk2015hidden}.As \citeauthor{kerk2015hidden} comment, behavioural
+kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
 repertoires with more than three distinct states are difficult to
@@
 least in part by structural problems with the HMM, i.e. the
-assumption of temporal homogeneity in behaviour.  When large data sets
+assumption of temporally homogeneous behaviour.  When large data sets
 are available, information-theoretic model selection methods will
@@
 GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
-Fish and Wildlife and Conservation Commission (FWC) staff using
-trained hounds and houndsmen. Of these panthers, 13 had 
-sufficient data to be used by \cite{kerk2015hidden}. Here we focus
+Fish and Wildlife and Conservation Commission staff using
+trained hounds and houndsmen. Of these animals, 13 had 
+sufficient data to be used by \cite{kerk2015hidden}.
+\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
+Here we focus
 on the four cats with the most data, in part because our goal is to understand
 the issues that arise when simple models are fitted to large data sets, and 
-in part because the general trend in telemetry studies is to larger data sets \todo{give cat IDs - these are shown in Fig S2a of vdK.
+in part because the general trend in telemetry studies is toward larger data sets.
+\todo{give cat IDs - these are shown in Fig S2a of vdK.
 Give threshold amount of data; justify/describe how we decided.
@@
 \citep{langrock_flexible_2012}, considering both Poisson and
-negative binomial distributions.  As shown 
+negative binomial distributions for dwell times.  As shown 
 \todo{I was going to say ``but not emphasized'', but that seems snarky}
@@
 of their analyses for ease of biological interpretation.
-
 We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
@@
 the best model, because we are interested in explaining the 
-data generation process by identifying the 'true' number 
+data generation process by identifying the `true' number 
 of underlying movement states.  Using BIC also simplifies 
 evaluation of model selection procedures; 
-it is more convenient to test whether our model selection
+it is easier to test whether our model selection
 procedure has selected the model used to simulate the data,
@@
 usual in ecological systems the true model will be far more
-complicated than any candidate model \citep{BurnhamAnderson1998},
-but we suspect that the qualitative conclusions stated here
+complicated than any candidate model \citep{BurnhamAnderson1998};
+we suspect that the qualitative conclusions stated here
 for BIC-optimality will carry over to analyses using AIC instead.
 
-\todo{This description of the V. algorithm is not coherent; need to give
+\todo{The following description of the V. algorithm is not coherent; need to give
 just a \emph{brief} definition here (since we are going to use it) and criticize/
@@
 P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
-P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
+P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \times \\
 &  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
@@
 \begin{description}
-\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$, the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$ and the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ at state $\state_{1}$.
+\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$; the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ and state $\state_{1}$.
 
-\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$, the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\state_{k}$ at time $t=k$.
+\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\cov_{k}$ at time $t=k$.
 
-\item[Emission probability] $P({\boldmath y_{k}} | \state_{k},\cov_{k})$, a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
+\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
 \end{description}
@@
 observed sequence, we need to average over all possible hidden sequences. 
-This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
-estimate parameters numerically \citep{zucchini_hidden_2009}.
+While naive approaches to this averaging process there are several
+more efficient algorithms for computing the marginal likelihood and
+numerically estimating parameters \citep{zucchini_hidden_2009};
+we use the implementations in the {\tt depmixS4} package for R
+\citep{visser2010depmixs4,R}.
 
 We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.
+\todo{where do we actually focus on these three states?
+Think this could probably be introduced in the Discussion, if we need it}
 
-We modeled three types of transition complexity instead of time homogeneous transition; extension of ``Switch with covariates" by \cite{morales_extracting_2004}.
-For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:
+For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
+However, this matrix needs to respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
+Therefore, as is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is unnecessary (i.e.
+we have only $n \times (n-1)$ distinct parameters; we nevertheless index $j$ from 2 to $n$ for notational clarity):
 \begin{equation}
@@
 
-where $\eta_{ij}(t)$ can be modelled as the following: 
-\begin{enumerate}
-\item Multiple block transition:  The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
+We considered three different models for diurnal variation in behaviour; extension of ``Switch with covariates" by \cite{morales_extracting_2004}. \todo{second clause isn't clear; expand?}.
+
+\begin{description}
+\item[Multiple block transition] Here we assume piecewise-constant transition probabilities.
+The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
 $$
-\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm}b_{m,t}
+\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
 $$
-where $a_{ijm}$ are parameters, and $b_{m,t}$ is a binary indicator (1 to the time block at the corresponding time $t$, and 0 otherwise).  
-\item Quadratic functional transition with a time covariate:
+where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta or  indicator variable (1 for the time block at the corresponding time $t$, and 0 otherwise).  
+\item[Quadratic transition model] We assume the elements of the linear predictor
+are quadratic functions of hour, with an additional diurnal continuity constraint that $\eta_{ij}(0)=\eta_{ij}(24)$ \todo{am I right/do we actually impose this constraint? give a bit more detail ...}
 $$
@@
 $$
-\item Sinusoidal functional transition with a time covariate:
+\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, \todo{apparently true from eqs here, but not true if we constrain the quadratic model??} but automatically satisfies the diurnal continuity constraint.
 $$
-\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})
+\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
 $$
-\end{enumerate}
+\end{description}
+
+\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.}
+
+\subsection{Model evaluation}
+
+\todo{Give more details about parameters, structure of simulations, etc.}
+
+The {\tt depmixS4} package
+was used for fitting covariate dependent transition HMM
+models, simulating states and step lengths using the
+estimated parameters, and for estimating states with the Viterbi
+algorithm.
+
+\todo{describe the Viterbi algorithm here, and the autocorrelation function
+computation - everything that appears in the results should be described
+explicitly here}
 
 We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches.
-First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\Delta$ BIC = BIC - min(BIC)).
+First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\dbic$ = BIC - min(BIC)).
 Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity.
@@
 
-The package depmixS4 \citep{visser2010depmixs4} for R
-(R Development Core Team 2011) was used for fitting covariate dependent transition HMM
-models, simulating states and step lengths using the
-estimated parameters, and for estimating states with the Viterbi
-algorithm.
+\todo{give at least a hint of why/what each of these three 
+approaches ($\dbic$, hourly averages, ACF) will show}
 
@@
 
-As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
-Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC.
+As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
+\todo{if FMMs are here they should be in methods} Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC.
 The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC.

### 2015-12-30 paper3.Rnw
need to do Ben's notes one by one, Makefile, make sure to gitroot

@@
 approximately constant. Changes in behavioural mode switching 
-\todo{do we want to call these transition probabilities (throughout)?}
+\todo{do we want to call these transition probabilities (throughout)? ML: I prefer to call it this because it is consistent with explaning HMM parts.}
 probabilities based on
@@
 
-\todo{what do we think of this subsection title?}
+\todo{what do we think of this subsection title? ML:sounds good to me}
 
@@
 Give threshold amount of data; justify/describe how we decided.
-Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?}
+Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?
+ML: Yes, I have all the plots ready in the figures folder.
+(Note, the public catID is different from what Madelon used, so we have to be careful)}
 As is typical in studies of animal movement, 
@@
 \todo{could say here (or maybe better in the conclusions) that
-revisiting HSMM in this context would be worthwhile future work}
+revisiting HSMM in this context would be worthwhile future work. 
+ML: conclusion is good, I still need to add viterbi paragraph}
 
@@
 \todo{not sure about terminology here (should be consistent about
-what we call the modes/latent states/whatever}
+what we call the modes/latent states/whatever) ML: What do you think? Are ecologist more comfortable with modes/ latent states?}
 In general, BIC values decreased as the number of states
@@
 three to six states,
-\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
+\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent. ML: numeric}
 suggesting that the six-state model was

### 2016-01-11 paper3.Rnw
need to rescale plots, working on todo, half done

@@
 library("ggplot2")
+library("depmixS4")
+library("dplyr")
+library("reshape2")
+library("plyr")
 @
@@
 HMMs have typically been used to model movements over short time
-scales, where the probability of switching between movement states is
-approximately constant. Changes in behavioural mode switching 
-\todo{do we want to call these transition probabilities (throughout)? ML: I prefer to call it this because it is consistent with explaning HMM parts.}
+scales, where the probability of transitioning between movement states is
+approximately constant. Changes in behavioural mode transition 
+\todo{do we want to call these transition probabilities (throughout)? ML: DONE}
 probabilities based on
@@
 \cite{schliehe-diecks_application_2012}
-consider temporal trends in switching behaviour over the
+consider temporal trends in transitioning behaviour over the
 time scales of a six-hour
@@
 where the occupancy probabilities change deterministically over time.  Thus FMMs and HMMs,
-with or without temporal variation in the occupancy or switching
+with or without temporal variation in the occupancy or transitioning
 probabilities, form a useful family of models for capturing
@@
 Our primary goal in this paper is to introduce the use of
-HMMs with temporally varying switching probabilities -- in particular,
-switching probabilities that follow a diurnal cycle -- for modeling
+HMMs with temporally varying transitioning probabilities -- in particular,
+transitioning probabilities that follow a diurnal cycle -- for modeling
 animal movement recorded over long time scales.  We re-analyze data from 
@@
 \citep{turchin1998quantitative,kerk2015hidden}.
+
+\begin{table}[]
+\centering
+\caption{My caption}
+\label{Cat-ID}
+\begin{tabular}{ccc}
+ (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
+  130 & 1 & 10286    \\
+  131& 2  & 1111    \\
+ 48& 14  & 12222    \\
+ 94& 15  & 9999  
+\end{tabular}
+\end{table}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
-\includegraphics[width=\maxwidth]{figures/cat1REDplot} \caption{Model Reduction Process via Individual based Model Adjusted BIC }\label{fig:BICred_plot}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\begin{figure}
+<<All mod BIC plot, warning=FALSE, echo=FALSE, message=FALSE>>=
+load("cat1allmod.RData")
+ggplot(allmodsumdf, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ model,ncol=4)+
+    scale_y_continuous()+
+    scale_size_continuous( name="Num of Parameters") +
+    #    geom_point(aes(size=parameters))+
+    labs(y=expression(paste(Delta, "BIC")),
+         x="Number of States")  +
+    geom_line(aes(colour=model),size=0.5) +theme_bw() +
+    ggtitle("Adjusted BIC Model Comparisons")
+
+@
+\caption{Model Reduction Process via Individual based Model Adjusted BIC }
+\label{fig:BICred_plot}
 \end{figure}
-
 \end{knitrout}
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/cat1BIC} \caption{Overall Adjusted BIC Comparisons between different HMM models}
+
+<<Adjusted BIC comparisons, warning=FALSE, echo=FALSE, message=FALSE>>=
+load("cat1sumdat.RData")
+ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
+    scale_y_continuous()+
+    scale_size_continuous( name="Num of Parameters") +
+    geom_point(aes(size=parameters))+
+    labs(y=expression(paste(Delta, "BIC")),
+         x="Number of States") +
+    geom_line(aes(colour=model),size=0.5) +theme_bw() +
+    ggtitle("Adjusted BIC Model Comparisons")
+
+@
+
+\caption{Overall Adjusted BIC Comparisons between different HMM models}
 \end{figure}
-
 \end{knitrout}
@@
 The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females.
-The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.
+The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transitioning pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.
 We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/cat1avg} \caption{Average step length by time of day}
+
+<<avg step length by time, message=FALSE, echo=FALSE, warning=FALSE>>=
+  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
+  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
+  avgdf <- cbind(simdf,obsdf)
+  avgdf2 <- melt(avgdf,'time')
+ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
+    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
+    scale_x_continuous(breaks=0:23) + theme_bw() + 
+    ggtitle("Average Step Lengths at Different Hours of the Day")
+
+@
+
+\caption{Average step length by time of day}
 \end{figure}
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/cat1acf} \caption{ACF }
+<<acf plot, message=FALSE, echo=FALSE, warning=FALSE>>=
+acffun <- function(sim){
+  a <- acf(sim,na.action=na.pass,plot=FALSE)
+  df <- data.frame(ACF=c(a$acf),lag=a$lag)
+  return(df)
+}
+
+  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
+  ggplot(temp,aes(lag,ACF,colour=.id))+
+    geom_point()+geom_line() +theme_bw() +
+    ggtitle("ACF plot of BIC Selected Models")
+
+@
+\caption{ACF }
 \end{figure}
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figures/cat1states} 
+<<state plots, warning=FALSE, message=FALSE, echo=FALSE>>=
+temp5 <- tail(getpars(fitsin5s),10)
+temp6 <- tail(getpars(fithomo6s),12)
+dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
+                     log_sd=temp5[c(2,4,6,8,10)], states="5") 
+dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
+                     log_sd=temp6[c(2,4,6,8,10,12)],states="6") 
+dat1 <- dat1[order(dat1$log_mean),]
+dat2 <- dat2[order(dat2$log_mean),]
+
+dat <- rbind(dat1,dat2)
+ggplot(dat,aes(log_mean,log_sd,color=states))+
+    geom_point()+geom_line() +theme_bw() +
+    ggtitle("State Parameters")
+
+@
+\caption{States}
 \end{figure}

### 2016-01-13 paper3.Rnw
updates

@@
 scales, where the probability of transitioning between movement states is
-approximately constant. Changes in behavioural mode transition 
+approximately constant. Changes in latent/hidden behavioural state/mode transition 
 \todo{do we want to call these transition probabilities (throughout)? ML: DONE}
@@
 \citeauthor{kerk2015hidden}'s model-based approach to
-identifying the number of behavioural modes.
+identifying the number of latent states.
 
@@
 only one way in which models can become more complex (e.g. by
-increasing the number of behavioural modes), complexity that is
+increasing the number of latent states), complexity that is
 present in the data but not accounted for in the model (e.g. spatial
@@
 allowing for diurnal variation in the Florida panther data leads to
-selection of models with smaller numbers of behavioural modes; we also
-fit models to simulated data with varying numbers of behavioural modes
+selection of models with smaller numbers of latent states; we also
+fit models to simulated data with varying numbers of latent states
 and degrees of temporal heterogeneity to test our conjecture that
@@
 \centering
-\caption{My caption}
+\caption{Cat-ID}
 \label{Cat-ID}
@@
   130 & 1 & 10286    \\
-  131& 2  & 1111    \\
- 48& 14  & 12222    \\
- 94& 15  & 9999  
+  131 & 2  & 9458    \\
+ 48& 14  & 14645    \\
+ 94& 15  & 10250  
 \end{tabular}
@@
 to test the relative penalized goodness of fit for
-models ranging from 2 to 6 latent behavioural modes.
+models ranging from 2 to 6 latent states.
 \todo{not sure about terminology here (should be consistent about
-what we call the modes/latent states/whatever) ML: What do you think? Are ecologist more comfortable with modes/ latent states?}
+what we call the modes/latent states/whatever) ML: latent states}
 In general, BIC values decreased as the number of states
 increased from 
-three to six states,
+3 to 6 states,
 \todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent. ML: numeric}
-suggesting that the six-state model was
+suggesting that the 6-state model was
 favoured statistically; however,
@@
 \citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
-future movement patterns and cannot capture which movement modes
+future movement patterns and cannot capture which latent states
 panthers are likely to occupy depending on time of day from the time
@@
 
-We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.
+We considered 3 (biologically interpretable number of states identified: (1) resting state, characterized by short step length; (2) Moderately active state characterized by intermediate step lengths, and (3) Traveling state, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.
 \todo{where do we actually focus on these three states?

### 2016-01-14 paper3.Rnw
missing parameter chart I think

@@
 
-Furthermore, the state occupancies based on the
-Viterbi algorithm, which predicts the most likely sequence of movement
-states based on observed movement results
-\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
-future movement patterns and cannot capture which latent states
-panthers are likely to occupy depending on time of day from the time
-homogeneous transitions (ie, the transition between all states are
-constant through time).
-
 \subsection{Model description}
@@
 
-For some kinds of model predictions, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of estimating the states by the Viterbi algorithm (the tradition state prediction method) and can easily overlook more complicated models.
-The Viterbi algorithm predicts the most likely sequence of movement states based on the observations which allows the time homogenous transition model's prediction to capture the temporal heterogeneity from the observations.
+For some kinds of model predictions, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of predicting the states by the Viterbi algorithm (the tradition state prediction method) and can easily overlook more complicated models.
+The Viterbi algorithm solves/predicts for the most likely/probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}.
+This approach allows the time homogenous transition model's prediction to capture the temporal heterogeneity from the observations.
+It is useful to predict missing data in the observation sequence but fail to future movements beyond the observations sequence.
 In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.
-See Appendix.
 

### 2016-01-18 paper3.Rnw
just need to write something about parameters and then I'm done

@@
 just a \emph{brief} definition here (since we are going to use it) and criticize/
-explain shortcomings later}
+explain shortcomings later. ML: I moved the whole previous paragraph to the end where we actually talk about viterbi}
 
@@
 
-We considered 3 (biologically interpretable number of states identified: (1) resting state, characterized by short step length; (2) Moderately active state characterized by intermediate step lengths, and (3) Traveling state, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.
 \todo{where do we actually focus on these three states?
-Think this could probably be introduced in the Discussion, if we need it}
+Think this could probably be introduced in the Discussion, if we need it.
+ML: I took it out and gave a half a line in model evaluation subsection}
 
@@
 However, this matrix needs to respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
+The FMM is a special case of HMM where; it can be modelled in the HMM framework by setting the transition probabilities:
+
+\begin{equation}
+\begin{split}
+\pi_{ij} = \pi_{i*j},     \forall i 
+\end{split}
+\end{equation}
+
 Therefore, as is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is unnecessary (i.e.
@@
 
-We considered three different models for diurnal variation in behaviour; extension of ``Switch with covariates" by \cite{morales_extracting_2004}. \todo{second clause isn't clear; expand?}.
+We considered three different transitions models for diurnal variation in behaviour; extension of ``Switch with covariates" by \cite{morales_extracting_2004} to account for temporal hetergeneitiy \todo{second clause isn't clear; expand?
+ML: okay now?
+}.
 
@@
 
-We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches.
+We assessed the fit of 3 to 6 state, time homogeneous and inhomogeneous transition HMMs to four of the thirteen Florida panthers with the most data using three approaches.
 First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\dbic$ = BIC - min(BIC)).
@@
 Lastly, autocorrelation comparison between the observed and simulated data.
+Increasing the number of latent states of HMM, and incorportating temporal transition heterogeneity of FMM and HMM, can both (independently) improve goodness of fit in their own ways, but it is unclear the tradeoff between the two methods and the combination of both is an underused model extension.
 
 \todo{give at least a hint of why/what each of these three 
-approaches ($\dbic$, hourly averages, ACF) will show}
+approaches ($\dbic$, hourly averages, ACF) will show. ML: What do you think?}
 

### 2016-01-18 paper3.Rnw
BB's turn

@@
 
+Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorportating temportal heterogeneity.
+
+\begin{table}[]
+\centering
+\caption{Number of Parameters for each model type}
+\label{Cat-ID}
+\begin{tabular}{cccc}
+ Type & Model  & Number of States & Number of Parameters  \\
+  FMM & FMM & 3 & 9    \\
+  FMM & FMM & 4 & 12    \\
+  FMM & FMM & 5 & 15   \\
+  FMM & FMM & 6 & 18  \\
+  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
+  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
+  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
+  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
+  HMM & HMM & 3 & 18    \\
+  HMM & HMM & 4 & 28    \\
+  HMM & HMM & 5 & 40   \\
+  HMM & HMM & 6 & 54  \\
+  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
+  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
+  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
+  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
+  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
+  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
+  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
+  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
+  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
+  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
+  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
+  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
+  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
+  HMM + TH & HMM + TH (Hourly)  & 4 & 396
+
+\end{tabular}
+\end{table}
+
 \todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.}
@@
 computation - everything that appears in the results should be described
-explicitly here}
+explicitly here....ML: at the bottom}
 
@@
 
+<<viterbi, warning=FALSE, message=FALSE, echo=FALSE >>=
+load("cat1.RData")
+vitcat <- function(catt,mod){
+  params <- tail(getpars(mod),6)
+  vitdist <- numeric(ntimes(mod))
+  states <- viterbi(mod)$state
+  for(i in 1:ntimes(mod)){
+    if(states[i] == 1){
+      vitdist[i] <- rnorm(1,params[1],params[2])
+    }
+    if(states[i] == 2){
+      vitdist[i] <- rnorm(1,params[3],params[4])
+    }
+    if(states[i] == 3){
+      vitdist[i] <- rnorm(1,params[5],params[6])
+    }
+  }
+  dat <- data.frame(viterbi=vitdist,time=catt$Time, obs=catt$LogDist)
+  return(dat)
+}
+
+temp <- vitcat(cat,fithomo3s) %>% dplyr:::select(-time)%>%ldply(.,acffun)
+ggplot(temp,aes(lag,ACF,colour=.id))+
+geom_point()+geom_line() +theme_bw() +
+ggtitle("ACF plot of BIC Selected Models")
+
+@
+
 Although the time inhomogeneous transition offer an

### 2016-01-28 paper3.Rnw
BMB beginning of updates

@@
 \bibliographystyle{chicago}
-\title{ HMM}
+\title{Incorporating periodic variability in hidden Markov models for animal movement}
 \author{Michael Li and Benjamin M. Bolker }
@@
 prediction as well as providing insight into the nature of underlying,
-unobservable states of the system. However, temporal heterogeneity and
-autocorrelation (persistence) in group occupancy can obscure such
-signals. We use finite mixture and hidden Markov models, two standard
-clustering techniques, to model high-resolution hourly movement data
-from Florida panthers. Allowing for temporal heterogeneity in
-transition probabilities, a straightforward but rarely explored model
-extension, resolves some shortcomings of current modeling frameworks
-and clarifies the behavioural patterns of panthers.
+unobservable states of the system. However, temporal variation in the
+probability that individuals occupy or move between groups can obscure
+such signals. We use finite mixture and hidden Markov models, two
+standard clustering techniques, to model high-resolution hourly
+movement data from Florida panthers. Allowing for temporal
+heterogeneity in transition probabilities, a straightforward but
+rarely explored model extension, resolves some shortcomings of current
+modeling frameworks and clarifies the behavioural patterns of
+panthers. More generally, incorporating previously neglected
+structure in statistical models can allow researchers to more
+accurately identify models of appropriate complexity.
 
@@
 
-\keywords{hidden Markov model, animal movement, temporal autocorrelation, temporal heterogeneity}
+\keywords{hidden Markov model, animal movement, temporal autocorrelation, temporal heterogeneity, Florida panther}
 
@@
 approximately constant. Changes in latent/hidden behavioural state/mode transition 
-\todo{do we want to call these transition probabilities (throughout)? ML: DONE}
 probabilities based on
@@
 \cite{schliehe-diecks_application_2012}
-consider temporal trends in transitioning behaviour over the
+consider temporal trends in behavioural transitions over the
 time scales of a six-hour
@@
 
-For movement behaviours that change on a very fast time scale, such
-that movement behaviours at successive observations are effectively independent,
+For movement behaviours that change on a fast time scale, such
+that movement behaviours recorded
+at successive observations are effectively independent,
 \emph{finite mixture models} (FMM) --- which can be considered
@@
 persistence or correlation, movement could be well represented by FMMs
-where the occupancy probabilities change deterministically over time.  Thus FMMs and HMMs,
-with or without temporal variation in the occupancy or transitioning
+where the occupancy probabilities change deterministically over time.
+Thus FMMs and HMMs,
+with or without temporal variation in the occupancy or transition
 probabilities, form a useful family of models for capturing
 changes in movement behaviour over a range of time scales.
+\todo{do we want a figure showing this?}
 
 Our primary goal in this paper is to introduce the use of
-HMMs with temporally varying transitioning probabilities -- in particular,
-transitioning probabilities that follow a diurnal cycle -- for modeling
-animal movement recorded over long time scales.  We re-analyze data from 
+HMMs with temporally varying transition probabilities -- in particular,
+transition probabilities that follow a diurnal cycle -- for modeling
+animal movement recorded over long time scales.  
+In addition to simulation-based examples,
+we also re-analyze data from 
 \cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
@@
 lengths of consecutive occupancy of a behavioural state) to
-analyze the movement of Florida panthers.
+describe the movement and putative underlying
+behavioural states of Florida panthers.
 
@@
 detailed analysis to models with only 3 underlying states.  In
-contrast, most studies using HMM have decided \emph{a priori} on a
-number of underlying states, typically using either two
-\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
-  fryxell_multiple_2008}, or three states
-\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
-kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
+contrast, most studies using HMM have chosen the
+number of underlying states \emph{a priori}, typically using either two
+\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
+\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
 repertoires with more than three distinct states are difficult to
-interpret, one possible reason that other authors have not used
+interpret, one possible reason that other authors have not adopted
 \citeauthor{kerk2015hidden}'s model-based approach to
@@
 \subsection{Data and previous analyses}
-
-\todo{what do we think of this subsection title? ML:sounds good to me}
 
@@
 The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females.
-The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transitioning pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.
+The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transition pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.
 We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.

### 2016-02-01 paper3.Rnw
tweaking paper, working on figs

@@
 \usepackage{amsmath}
+\usepackage{alltt}
 \usepackage[utf8]{inputenc} % for accented characters
@@
 <<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
-library("ggplot2")
+library("ggplot2"); theme_set(theme_bw())
+scale_colour_discrete <- function(...,palette="Set1")
+    scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Set1")
+    scale_fill_brewer(...,palette=palette)
+zmargin <- theme(panel.margin=grid::unit(0,"lines"))
+library("knitr")
+opts_chunk$set(fig.width=6,fig.height=4)
 library("depmixS4")
+library("plyr")
 library("dplyr")
 library("reshape2")
-library("plyr")
 @
@@
 describe the movement and putative underlying
-behavioural states of Florida panthers.
+behavioural states of Florida panthers (\emph{Puma concolor coryi}).
 
@@
 Here we focus
-on the four cats with the most data, in part because our goal is to understand
+on the four cats with the most data (all with approximately 10,000-15,000 observations: see Table~1 in Supplementary Material), 
+in part because our goal is to understand
 the issues that arise when simple models are fitted to large data sets, and 
-in part because the general trend in telemetry studies is toward larger data sets.
-\todo{give cat IDs - these are shown in Fig S2a of vdK.
-Give threshold amount of data; justify/describe how we decided.
-Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?
-ML: Yes, I have all the plots ready in the figures folder.
-(Note, the public catID is different from what Madelon used, so we have to be careful)}
-As is typical in studies of animal movement, 
+in part because the general trend in telemetry studies is toward larger data sets.  As is typical in studies of animal movement, 
 we took first differences of the data by decomposing
@@
 
-\begin{table}[]
-\centering
-\caption{Cat-ID}
-\label{Cat-ID}
-\begin{tabular}{ccc}
- (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
-  130 & 1 & 10286    \\
-  131 & 2  & 9458    \\
- 48& 14  & 14645    \\
- 94& 15  & 10250  
-\end{tabular}
-\end{table}
-
 \cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
@@
 negative binomial distributions for dwell times.  As shown 
-\todo{I was going to say ``but not emphasized'', but that seems snarky}
 by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
@@
 were best.  
-
-Our analysis again aims for simplicity and qualitative
+Since our analysis aims for simplicity and qualitative
 conclusions rather than for picking the very best
-predictive model; we focus on models that treat each step as
+predictive model, we focus on models that treat each step as
 a univariate, log-Normally distributed observation, glossing
-over both the relatively subtle distinctions in shape between
+over both the differences in shape between
 the three candidate step-length distributions and the effects
 of considering multivariate (step length plus turning angle)
-observations. (Note that most movement analyses, including
+observations.  However, we do briefly compare log-Normal and Weibull
+step-length distributions\todo{have we run Gamma as well?}, with and without 
+a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).\todo{W/VM/HMM (blue/solid/left panel) should correspond to vdK's results, but that has $\dbic=0$ for 5 states rather than 6.
+Is that right?
+Is that HMM/HSMM diff?}
+(Note that most movement analyses, including
 \cite{kerk2015hidden}, are only partially multivariate, 
@@
 as independent observations while neglecting possible correlations
-between them.)  However, we do briefly compare all three
-step-length distributions, with and without incorporating
-a von Mises-distributed turning angle in the model (Figure~1).
+between them.)
 
@@
 models ranging from 2 to 6 latent states.
-\todo{not sure about terminology here (should be consistent about
-what we call the modes/latent states/whatever) ML: latent states}
 In general, BIC values decreased as the number of states
@@
 3 to 6 states,
-\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent. ML: numeric}
 suggesting that the 6-state model was
@@
 of their analyses for ease of biological interpretation.
-We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
+We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
 BIC across a family of models) as the criterion for identifying
 the best model, because we are interested in explaining the 
-data generation process by identifying the `true' number 
+data generation process by identifying the ``true'' number 
 of underlying movement states.  Using BIC also simplifies 
@@
 for BIC-optimality will carry over to analyses using AIC instead.
-
-\todo{The following description of the V. algorithm is not coherent; need to give
-just a \emph{brief} definition here (since we are going to use it) and criticize/
-explain shortcomings later. ML: I moved the whole previous paragraph to the end where we actually talk about viterbi}
 
@@
 observed sequence, we need to average over all possible hidden sequences. 
-While naive approaches to this averaging process there are several
-more efficient algorithms for computing the marginal likelihood and
+There are several efficient algorithms for computing the marginal likelihood and
 numerically estimating parameters \citep{zucchini_hidden_2009};
-we use the implementations in the {\tt depmixS4} package for R
+we use those implemented in the \texttt{depmixS4} package for R
 \citep{visser2010depmixs4,R}.
 
-\todo{where do we actually focus on these three states?
-Think this could probably be introduced in the Discussion, if we need it.
-ML: I took it out and gave a half a line in model evaluation subsection}
+For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
+The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy.
+It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.
+\todo{is this right?}
 
-For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
-However, this matrix needs to respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
-The FMM is a special case of HMM where; it can be modelled in the HMM framework by setting the transition probabilities:
-
-\begin{equation}
-\begin{split}
-\pi_{ij} = \pi_{i*j},     \forall i 
-\end{split}
-\end{equation}
-
-Therefore, as is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is unnecessary (i.e.
+In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
+As is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without loss of generality (i.e.
 we have only $n \times (n-1)$ distinct parameters; we nevertheless index $j$ from 2 to $n$ for notational clarity):
@@
 
-We considered three different transitions models for diurnal variation in behaviour; extension of ``Switch with covariates" by \cite{morales_extracting_2004} to account for temporal hetergeneitiy \todo{second clause isn't clear; expand?
-ML: okay now?
-}.
+We considered three different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of allowing for a covariate-dependent transition matrix.
 
@@
 \item[Quadratic transition model] We assume the elements of the linear predictor
-are quadratic functions of hour, with an additional diurnal continuity constraint that $\eta_{ij}(0)=\eta_{ij}(24)$ \todo{am I right/do we actually impose this constraint? give a bit more detail ...}
+are quadratic functions of hour.
+Note that the quadratic model is not diurnally continuous, i.e.
+there is no constraint forcing $\eta_{ij}(0)=\eta_{ij}(1)$; imposing a diurnal continuity constraint would collapse the model to a constant.
 $$
@@
 $$
-\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, \todo{apparently true from eqs here, but not true if we constrain the quadratic model??} but automatically satisfies the diurnal continuity constraint.
+\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
 $$
@@
 
-Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorportating temportal heterogeneity.
+Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorportating temporal heterogeneity.
 
@@
   HMM + TH & HMM + TH (Hourly)  & 4 & 396    
-
 \end{tabular}
@@
 
-\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.}
+\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.  Make the table into a figure!}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\begin{figure}
-<<All mod BIC plot, warning=FALSE, echo=FALSE, message=FALSE>>=
+<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#130.
+$\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
 load("cat1allmod.RData")
-ggplot(allmodsumdf, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ model,ncol=4)+
-    scale_y_continuous()+
-    scale_size_continuous( name="Num of Parameters") +
+df2 <- transform(allmodsumdf,
+            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
+            turn_angle=ifelse(grepl("VM",model),"yes","no"),
+            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
+ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
+    geom_line()+
+    facet_wrap(~ homog)+
+    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
+    ## scale_size_continuous( name="# parameters") +
     #    geom_point(aes(size=parameters))+
-    labs(y=expression(paste(Delta, "BIC")),
-         x="Number of States")  +
-    geom_line(aes(colour=model),size=0.5) +theme_bw() +
-    ggtitle("Adjusted BIC Model Comparisons")
-
+    scale_linetype(name="von Mises\nturning angle?")+
+    labs(y=expression(Delta*"BIC"),
+         x="Number of States")+
+    geom_hline(yintercept=0,colour="gray")+
+    zmargin
 @
-\caption{Model Reduction Process via Individual based Model Adjusted BIC }
-\label{fig:BICred_plot}
-\end{figure}
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\begin{figure}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-<<Adjusted BIC comparisons, warning=FALSE, echo=FALSE, message=FALSE>>=
+<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
 load("cat1sumdat.RData")
 ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
-    scale_y_continuous()+
     scale_size_continuous( name="Num of Parameters") +
     geom_point(aes(size=parameters))+
+    scale_colour_brewer(palette="Dark2")+
     labs(y=expression(paste(Delta, "BIC")),
          x="Number of States") +
-    geom_line(aes(colour=model),size=0.5) +theme_bw() +
-    ggtitle("Adjusted BIC Model Comparisons")
-
+    geom_line() + zmargin
 @
-
-\caption{Overall Adjusted BIC Comparisons between different HMM models}
-\end{figure}
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\begin{figure}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-<<avg step length by time, message=FALSE, echo=FALSE, warning=FALSE>>=
+<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
   simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
@@
     geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
-    scale_x_continuous(breaks=0:23) + theme_bw() + 
-    ggtitle("Average Step Lengths at Different Hours of the Day")
-
+    scale_x_continuous(breaks=0:23)
 @
-
-\caption{Average step length by time of day}
-\end{figure}
-
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\begin{figure}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-<<acf plot, message=FALSE, echo=FALSE, warning=FALSE>>=
+<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
 acffun <- function(sim){
@@
   ggplot(temp,aes(lag,ACF,colour=.id))+
-    geom_point()+geom_line() +theme_bw() +
-    ggtitle("ACF plot of BIC Selected Models")
-
+    geom_point()+geom_line()
 @
-\caption{ACF }
-\end{figure}
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\begin{figure}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-<<state plots, warning=FALSE, message=FALSE, echo=FALSE>>=
+<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
 temp5 <- tail(getpars(fitsin5s),10)
@@
 ggplot(dat,aes(log_mean,log_sd,color=states))+
-    geom_point()+geom_line() +theme_bw() +
-    ggtitle("State Parameters")
-
+    geom_point()+geom_line()
 @
-\caption{States}
-\end{figure}
-\end{knitrout}
 
@@
 
+\appendix
+\section{Supplementary material}
+\begin{table}[]
+\centering
+\caption{Cat-ID}
+\label{Cat-ID}
+\begin{tabular}{ccc}
+  (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
+  130 & 1   & 10286    \\
+  131 & 2   & 9458    \\
+  48  & 14  & 14645    \\
+  94  & 15  & 10250  
+\end{tabular}
+\end{table}
+
 \bibliography{paper}

### 2016-02-11 paper3.Rnw
bits and pieces on ms

@@
 of organisms' \emph{internal} states; in particular, HMMs model animal
-movement as though individual animals' movement behavior is determined
+movement as though individual animals' movement behaviour is determined
 by which of a discrete set of unobserved movement states
@@
 It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.
-\todo{is this right?}
+\todo{BMB: is this right?}
 
@@
 
-We considered three different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of allowing for a covariate-dependent transition matrix.
+We considered three different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
 
@@
 $$
-where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta or  indicator variable (1 for the time block at the corresponding time $t$, and 0 otherwise).  
+where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).  
 \item[Quadratic transition model] We assume the elements of the linear predictor
 are quadratic functions of hour.
-Note that the quadratic model is not diurnally continuous, i.e.
-there is no constraint forcing $\eta_{ij}(0)=\eta_{ij}(1)$; imposing a diurnal continuity constraint would collapse the model to a constant.
+The quadratic model is not diurnally continuous, i.e.
+there is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity constraint would collapse the model to a constant.
 $$
-\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}
+\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
 $$
@@
 
-Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorportating temporal heterogeneity.
+Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity.
 
@@
 \centering
-\caption{Number of Parameters for each model type}
+\caption{Number of Parameters for each model type. \textbf{BMB: supplementary material?}
 \label{Cat-ID}
@@
 
-The {\tt depmixS4} package
-was used for fitting covariate dependent transition HMM
-models, simulating states and step lengths using the
-estimated parameters, and for estimating states with the Viterbi
+We used the {\tt depmixS4} package
+to fit covariate-dependent transition HMMs, 
+simulate states and step lengths using the
+estimated parameters, and estimate the most likely
+states with the Viterbi
 algorithm.
@@
 
-We assessed the fit of 3 to 6 state, time homogeneous and inhomogeneous transition HMMs to four of the thirteen Florida panthers with the most data using three approaches.
-First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\dbic$ = BIC - min(BIC)).
-Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity.
-Lastly, autocorrelation comparison between the observed and simulated data.
-Increasing the number of latent states of HMM, and incorportating temporal transition heterogeneity of FMM and HMM, can both (independently) improve goodness of fit in their own ways, but it is unclear the tradeoff between the two methods and the combination of both is an underused model extension.
+We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations).
+(1) BIC relative to the optimal-BIC model within each type of transition complexity\todo{BMB: not sure what that means.
+Clarify?} ($\dbic$ = BIC - min(BIC)).
+(2) Computed average step-length by hour of day for the observed data and for data simulated from the models.
+(3) Computed temporal autocorrelation function (ACF) for the observed data and for data simulated from the models.
+For the second and third metrics (hourly average and ACF), we simulated data from the models because the metrics are hard to compute from first principles for HMMs with temporally varying transition probabilities.
 
@@
 Long step lengths are easy to capture in all models, where as short step lengths are hard to capture given the GPS errors and unstable movements patterns suggesting multiple states associate with short step lengths.
-Thus, lower number of short step length states with higher variance can be equalivent or more realistic than multiple distinguishable short step length states with low variance.
+Thus, lower number of short step length states with higher variance can be equivalent or more realistic than multiple distinguishable short step length states with low variance.
 
@@
 The Viterbi algorithm solves/predicts for the most likely/probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}.
-This approach allows the time homogenous transition model's prediction to capture the temporal heterogeneity from the observations.
+This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations.
 It is useful to predict missing data in the observation sequence but fail to future movements beyond the observations sequence.
@@
 \begin{tabular}{ccc}
-  (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
+  (van de Kerk 2015) & (IR@UF)  & Number of Observations  \\
   130 & 1   & 10286    \\

### 2016-02-29 paper3.Rnw
compiling problem, need to rewrite results for the paper
minor edits, cat14 has a small fitting problem

@@
 
-\begin{table}[]
-\centering
-\caption{Number of Parameters for each model type. \textbf{BMB: supplementary material?}
-\label{Cat-ID}
-\begin{tabular}{cccc}
- Type & Model  & Number of States & Number of Parameters  \\
-  FMM & FMM & 3 & 9    \\
-  FMM & FMM & 4 & 12    \\
-  FMM & FMM & 5 & 15   \\
-  FMM & FMM & 6 & 18  \\
-  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
-  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
-  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
-  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
-  HMM & HMM & 3 & 18    \\
-  HMM & HMM & 4 & 28    \\
-  HMM & HMM & 5 & 40   \\
-  HMM & HMM & 6 & 54  \\
-  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
-  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
-  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
-  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
-  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
-  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
-  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
-  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
-  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
-  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
-  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
-  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
-  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
-  HMM + TH & HMM + TH (Hourly)  & 4 & 396    
-\end{tabular}
-\end{table}
-
 \todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.  Make the table into a figure!}
@@
 
-Adding temporal heterogeneity in the transitions for 3 to 6 state HSMM was a challenge because the number of parameters and model complexity increase dramatically, thus creating overfitting problems.
-Model reduction was used to balance out the increase in parameters and see if there are evidence suggesting adding temporal heterogeneity complexity can reduce the number of states to a statistically and biologically interpretable level.
-Without loss of generality, we assessed time homogeneous transition HMM models instead of HSMM and made sure each model reduction does not change the BIC selection of number of states, which HMM heavily depends on.
-Goodness of fit will decrease, but the information on the number of states remains the same.
+Figure 1 shows the BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.
+Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.
 
@@
 
+\begin{table}[]
+\centering
+\caption{Number of Parameters for each model type. \textbf{BMB: supplementary material?}
+\label{Cat-ID}
+\begin{tabular}{cccc}
+ Type & Model  & Number of States & Number of Parameters  \\
+  FMM & FMM & 3 & 9    \\
+  FMM & FMM & 4 & 12    \\
+  FMM & FMM & 5 & 15   \\
+  FMM & FMM & 6 & 18  \\
+  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
+  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
+  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
+  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
+  HMM & HMM & 3 & 18    \\
+  HMM & HMM & 4 & 28    \\
+  HMM & HMM & 5 & 40   \\
+  HMM & HMM & 6 & 54  \\
+  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
+  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
+  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
+  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
+  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
+  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
+  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
+  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
+  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
+  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
+  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
+  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
+  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
+  HMM + TH & HMM + TH (Hourly)  & 4 & 396    
+\end{tabular}
+\end{table}
+
+<<BICred_plot cat131,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#131.
+$\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
+load("cat2allmod.RData")
+df2 <- transform(allmodsumdf,
+            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
+            turn_angle=ifelse(grepl("VM",model),"yes","no"),
+            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
+ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
+    geom_line()+
+    facet_wrap(~ homog)+
+    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
+    ## scale_size_continuous( name="# parameters") +
+    #    geom_point(aes(size=parameters))+
+    scale_linetype(name="von Mises\nturning angle?")+
+    labs(y=expression(Delta*"BIC"),
+         x="Number of States")+
+    geom_hline(yintercept=0,colour="gray")+
+    zmargin
+@
+
 \bibliography{paper}

### 2016-02-29 paper3.Rnw
maybe save everything to RDS?

@@
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
-load("cat2allmod.RData")
+allmodsumdf <- readRDS("cat2allmod.RDS")
 df2 <- transform(allmodsumdf,

### 2016-02-29 paper3.Rnw
need to figure out what is wrong with WNMHMM5 for cat14

@@
 As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
-\todo{if FMMs are here they should be in methods} Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC.
+Looking at Figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC.
 The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC.
@@
 
+<<BICred_plot cat94,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#94.
+$\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
+allmodsumdf <- readRDS("cat15allmod.RDS")
+df2 <- transform(allmodsumdf,
+            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
+            turn_angle=ifelse(grepl("VM",model),"yes","no"),
+            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
+ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
+    geom_line()+
+    facet_wrap(~ homog)+
+    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
+    ## scale_size_continuous( name="# parameters") +
+    #    geom_point(aes(size=parameters))+
+    scale_linetype(name="von Mises\nturning angle?")+
+    labs(y=expression(Delta*"BIC"),
+         x="Number of States")+
+    geom_hline(yintercept=0,colour="gray")+
+    zmargin
+@
+
 \bibliography{paper}

### 2016-02-29 paper3.Rnw
latex compiling problem

@@
 library("reshape2")
+library("gridExtra")
 @
@@
 
-Figure 1 shows the BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.
+Figure 1 shows the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results.
 Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.
@@
 As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
-Looking at Figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC.
-The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC.
-Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level.
-Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.
-
-Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC.
-The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females.
-The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transition pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters.
-We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.
+Looking at Figure 2, models with temporal heterogenity is better than models without in both FMM and HMM frameworks, but time homogenous HMM models is better than FMM with sinusoidal temporal heterogeneity.
+Temporal heterogenity is good with good models such as the quadratic and sinuosidal.
+Multiple block models is good for cats with distinguishable diurnal patterns and hourly model is too complicated.
 
@@
 
-Next, we used the fitted models and simulate the same number of observations as the observed movement data and compare the average distance by time with the observed data.
+Figure 3 shows a clear diurnal pattern from the average hourly step lengths from the observed data.
 Models that assume temporal homogeneity (FMM and HMM) fail to capture the diurnal activity cycle.
-Models incorporating temporal heterogeneity can capture the observed patterns in the hourly average step.
+Models incorporating temporal heterogeneity can capture the observed patterns without the Viterbi algorithm.
 
@@
 
-Lastly, the observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity.
-FMM has no autocorrelation at all.
-HMM without temporal heterogeneity can capture the first few lags of the ACF but is unable to capture lags beyond 7.
-FMM with temporal heterogeneity can capture lags beyond 12, but not short-term correlation.
-HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data.
+The observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity.
+HMM without temporal heterogeneity can capture short-term autocorrelation (without Viterbi algorithm) but is unable to capture lags beyond 7.
+FMM with temporal heterogeneity can capture lags beyond 12, but not short-term autocorrelations.
+HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data whereas the FMM completely missed the autocorrelations.
 
@@
 dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
-                     log_sd=temp5[c(2,4,6,8,10)], states="5") 
+                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM + TH(sin)") 
 dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
-                     log_sd=temp6[c(2,4,6,8,10,12)],states="6") 
+                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM") 
 dat1 <- dat1[order(dat1$log_mean),]
@@
 dat <- rbind(dat1,dat2)
-ggplot(dat,aes(log_mean,log_sd,color=states))+
-    geom_point()+geom_line()
+sd1 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
+    geom_point()+geom_line() + ggtitle("Cat 1")
+@
+
+<<state_plots cat2, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
+load("cat2sumdat.RData")
+temp5 <- tail(getpars(fithomo5s),10)
+temp4 <- tail(getpars(fitsin4s),8)
+dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
+                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
+dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
+                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
+dat1 <- dat1[order(dat1$log_mean),]
+dat2 <- dat2[order(dat2$log_mean),]
+
+dat <- rbind(dat2,dat1)
+sd2 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
+    geom_point()+geom_line() + ggtitle("Cat 2")
+@
+
+<<state_plots cat14, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
+load("cat14sumdat.RData")
+temp5 <- tail(getpars(fithomo5s),10)
+temp4 <- tail(getpars(fitsin4s),8)
+dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
+                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
+dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
+                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
+dat1 <- dat1[order(dat1$log_mean),]
+dat2 <- dat2[order(dat2$log_mean),]
+
+dat <- rbind(dat2,dat1)
+sd14 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
+    geom_point()+geom_line() + ggtitle("Cat 14")
+@
+
+<<state_plots cat15, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
+load("cat15sumdat.RData")
+temp5 <- tail(getpars(fitsin5s),10)
+temp6 <- tail(getpars(fithomo6s),12)
+dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
+                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
+dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
+                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM + TH(sin)") 
+dat1 <- dat1[order(dat1$log_mean),]
+dat2 <- dat2[order(dat2$log_mean),]
+
+dat <- rbind(dat1,dat2)
+sd15 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
+    geom_point()+geom_line() + ggtitle("Cat 15")
+
+grid.arrange(sd1,sd2,sd14,sd15)
 @
@@
 In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.
-
-<<viterbi, warning=FALSE, message=FALSE, echo=FALSE >>=
-load("cat1.RData")
-vitcat <- function(catt,mod){
-  params <- tail(getpars(mod),6)
-  vitdist <- numeric(ntimes(mod))
-  states <- viterbi(mod)$state
-  for(i in 1:ntimes(mod)){
-    if(states[i] == 1){
-      vitdist[i] <- rnorm(1,params[1],params[2])
-    }
-    if(states[i] == 2){
-      vitdist[i] <- rnorm(1,params[3],params[4])
-    }
-    if(states[i] == 3){
-      vitdist[i] <- rnorm(1,params[5],params[6])
-    }
-  }
-  dat <- data.frame(viterbi=vitdist,time=catt$Time, obs=catt$LogDist)
-  return(dat)
-}
-
-temp <- vitcat(cat,fithomo3s) %>% dplyr:::select(-time)%>%ldply(.,acffun)
-ggplot(temp,aes(lag,ACF,colour=.id))+
-geom_point()+geom_line() +theme_bw() +
-ggtitle("ACF plot of BIC Selected Models")
-
-@
 
@@
 \centering
-\caption{Number of Parameters for each model type. \textbf{BMB: supplementary material?}
+\caption{Number of Parameters for each model type. 
 \label{Cat-ID}
@@
 
+<<viterbi, warning=FALSE, message=FALSE, echo=FALSE >>=
+load("cat1.RData")
+vitcat <- function(catt,mod){
+  params <- tail(getpars(mod),6)
+  vitdist <- numeric(ntimes(mod))
+  states <- viterbi(mod)$state
+  for(i in 1:ntimes(mod)){
+    if(states[i] == 1){
+      vitdist[i] <- rnorm(1,params[1],params[2])
+    }
+    if(states[i] == 2){
+      vitdist[i] <- rnorm(1,params[3],params[4])
+    }
+    if(states[i] == 3){
+      vitdist[i] <- rnorm(1,params[5],params[6])
+    }
+  }
+  dat <- data.frame(viterbi=vitdist,time=catt$Time, obs=catt$LogDist)
+  return(dat)
+}
+
+temp <- vitcat(cat,fithomo3s) %>% dplyr:::select(-time)%>%ldply(.,acffun)
+ggplot(temp,aes(lag,ACF,colour=.id))+
+geom_point()+geom_line() +theme_bw() +
+ggtitle("ACF plot of BIC Selected Models")
+
+@
+
+<<adj_BIC_comparisons cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+load("cat2sumdat.RData")
+ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
+    scale_size_continuous( name="Num of Parameters") +
+    geom_point(aes(size=parameters))+
+    scale_colour_brewer(palette="Dark2")+
+    labs(y=expression(paste(Delta, "BIC")),
+         x="Number of States") +
+    geom_line() + zmargin
+@
+
+<<adj_BIC_comparisons cat14, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+load("cat14sumdat.RData")
+ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
+    scale_size_continuous( name="Num of Parameters") +
+    geom_point(aes(size=parameters))+
+    scale_colour_brewer(palette="Dark2")+
+    labs(y=expression(paste(Delta, "BIC")),
+         x="Number of States") +
+    geom_line() + zmargin
+@
+
+<<adj_BIC_comparisons cat15, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+load("cat15sumdat.RData")
+ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
+    scale_size_continuous( name="Num of Parameters") +
+    geom_point(aes(size=parameters))+
+    scale_colour_brewer(palette="Dark2")+
+    labs(y=expression(paste(Delta, "BIC")),
+         x="Number of States") +
+    geom_line() + zmargin
+@
+
+<<avg_step_length_by_time cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
+load("cat2sumdat.RData")
+  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
+  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
+  avgdf <- cbind(simdf,obsdf)
+  avgdf2 <- melt(avgdf,'time')
+ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
+    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
+    scale_x_continuous(breaks=0:23)
+@
+
+<<avg_step_length_by_time cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
+load("cat14sumdat.RData")
+  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
+  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
+  avgdf <- cbind(simdf,obsdf)
+  avgdf2 <- melt(avgdf,'time')
+ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
+    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
+    scale_x_continuous(breaks=0:23)
+@
+
+<<avg_step_length_by_time cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
+load("cat15sumdat.RData")
+  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
+  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
+  avgdf <- cbind(simdf,obsdf)
+  avgdf2 <- melt(avgdf,'time')
+ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
+    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
+    scale_x_continuous(breaks=0:23)
+@
+
+<<acf_plot cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
+load('cat2sumdat.RData')
+acffun <- function(sim){
+  a <- acf(sim,na.action=na.pass,plot=FALSE)
+  df <- data.frame(ACF=c(a$acf),lag=a$lag)
+  return(df)
+}
+
+  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
+  ggplot(temp,aes(lag,ACF,colour=.id))+
+    geom_point()+geom_line()
+@
+
+<<acf_plot cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
+load("cat14sumdat.RData")
+acffun <- function(sim){
+  a <- acf(sim,na.action=na.pass,plot=FALSE)
+  df <- data.frame(ACF=c(a$acf),lag=a$lag)
+  return(df)
+}
+
+  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
+  ggplot(temp,aes(lag,ACF,colour=.id))+
+    geom_point()+geom_line()
+@
+
+<<acf_plot cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
+load("cat15sumdat.RData")
+acffun <- function(sim){
+  a <- acf(sim,na.action=na.pass,plot=FALSE)
+  df <- data.frame(ACF=c(a$acf),lag=a$lag)
+  return(df)
+}
+
+  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
+  ggplot(temp,aes(lag,ACF,colour=.id))+
+    geom_point()+geom_line()
+@
 \bibliography{paper}

### 2016-02-29 paper3.Rnw
fix table caption
lengend change?

@@
 dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
-                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
+                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM + TH(sin)") 
 dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
-                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM + TH(sin)") 
+                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM") 
 dat1 <- dat1[order(dat1$log_mean),]
@@
 \centering
-\caption{Number of Parameters for each model type. 
+\caption{Number of Parameters for each model type.}
 \label{Cat-ID}

### 2016-02-29 paper3.Rnw
back to you ben!

@@
 in general ecologists have turned to other tools to describe behavioural
-changes over longer (diurnal, seasonal, or ontogenetic) time scales.
-\todo{Fleming et al Am Nat 2014?  Gurarie?}
+changes over longer (diurnal, seasonal, or ontogenetic) time scales \citep{gurarie2009novel}.
 
@@
 changes in movement behaviour over a range of time scales.
-\todo{do we want a figure showing this?}
 
@@
 sufficient data to be used by \cite{kerk2015hidden}.
-\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
 Here we focus
@@
 a simpler HMM framework, which makes precisely this assumption.
-\todo{could say here (or maybe better in the conclusions) that
-revisiting HSMM in this context would be worthwhile future work. 
-ML: conclusion is good, I still need to add viterbi paragraph}
 
@@
 observations.  However, we do briefly compare log-Normal and Weibull
-step-length distributions\todo{have we run Gamma as well?}, with and without 
-a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).\todo{W/VM/HMM (blue/solid/left panel) should correspond to vdK's results, but that has $\dbic=0$ for 5 states rather than 6.
-Is that right?
-Is that HMM/HSMM diff?}
+step-length distributions, with and without 
+a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).
 (Note that most movement analyses, including
@@
 It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.
-\todo{BMB: is this right?}
 
@@
 Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity.
-
-\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.  Make the table into a figure!}
+See supplementary material Table 1.
 
 \subsection{Model evaluation}
-
-\todo{Give more details about parameters, structure of simulations, etc.}
 
@@
 
-\todo{describe the Viterbi algorithm here, and the autocorrelation function
-computation - everything that appears in the results should be described
-explicitly here....ML: at the bottom}
-
 We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations).
-(1) BIC relative to the optimal-BIC model within each type of transition complexity\todo{BMB: not sure what that means.
-Clarify?} ($\dbic$ = BIC - min(BIC)).
-(2) Computed average step-length by hour of day for the observed data and for data simulated from the models.
-(3) Computed temporal autocorrelation function (ACF) for the observed data and for data simulated from the models.
+(1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)).
+(2) Compare average step-length by hour of day for the observed data and for data simulated from the models.
+(3) Compare temporal autocorrelations for the observed data and for data simulated from the models.
 For the second and third metrics (hourly average and ACF), we simulated data from the models because the metrics are hard to compute from first principles for HMMs with temporally varying transition probabilities.
-
-\todo{give at least a hint of why/what each of these three 
-approaches ($\dbic$, hourly averages, ACF) will show. ML: What do you think?}
 
@@
 
-<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#130.
+<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
@@
 
-As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.
+As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions (24-hour multiple block transition model) to compare the temporal effects in goodness of fit.
 Looking at Figure 2, models with temporal heterogenity is better than models without in both FMM and HMM frameworks, but time homogenous HMM models is better than FMM with sinusoidal temporal heterogeneity.
@@
 
-<<state_plots cat2, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
+<<state_plots cat2, warning=FALSE, message=FALSE, echo=FALSE>>=
 load("cat2sumdat.RData")
@@
 
-<<state_plots cat14, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
+<<state_plots cat14, warning=FALSE, message=FALSE, echo=FALSE>>=
 load("cat14sumdat.RData")
@@
 
-<<state_plots cat15, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
+<<state_plots cat15, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="State Parameter Estimates">>=
 load("cat15sumdat.RData")
@@
 We really do not know if there Florida panthers really moves according to hidden states, and even if they do, we don't know the true number nor able to observe them.
-Three distinct movement states seems biologically interpretable for Florida panthers: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state.
+Three distinct movement states seems biologically interpretable for Florida panthers according \citet{kerk2015hidden}: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state.
 Long step lengths are easy to capture in all models, where as short step lengths are hard to capture given the GPS errors and unstable movements patterns suggesting multiple states associate with short step lengths.
@@
 
-\appendix
+\newpage
 \section{Supplementary material}
+
 \begin{table}[]
@@
 
-<<BICred_plot cat131,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#131.
+<<BICred_plot cat131,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#2.
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
@@
 
-<<BICred_plot cat94,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#94.
+<<BICred_plot cat94,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#15.
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
@@
 load("cat1.RData")
+load("cat1sumdat.RData")
 vitcat <- function(catt,mod){

### 2016-02-29 paper3.Rnw
refactor Figure 5 (cat states)

@@
 
-<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
-temp5 <- tail(getpars(fitsin5s),10)
-temp6 <- tail(getpars(fithomo6s),12)
-dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
-                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM + TH(sin)") 
-dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
-                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM") 
-dat1 <- dat1[order(dat1$log_mean),]
-dat2 <- dat2[order(dat2$log_mean),]
-
-dat <- rbind(dat1,dat2)
-sd1 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
-    geom_point()+geom_line() + ggtitle("Cat 1")
-@
-
-<<state_plots cat2, warning=FALSE, message=FALSE, echo=FALSE>>=
-load("cat2sumdat.RData")
-temp5 <- tail(getpars(fithomo5s),10)
-temp4 <- tail(getpars(fitsin4s),8)
-dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
-                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
-dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
-                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
-dat1 <- dat1[order(dat1$log_mean),]
-dat2 <- dat2[order(dat2$log_mean),]
-
-dat <- rbind(dat2,dat1)
-sd2 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
-    geom_point()+geom_line() + ggtitle("Cat 2")
-@
-
-<<state_plots cat14, warning=FALSE, message=FALSE, echo=FALSE>>=
-load("cat14sumdat.RData")
-temp5 <- tail(getpars(fithomo5s),10)
-temp4 <- tail(getpars(fitsin4s),8)
-dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
-                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
-dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
-                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
-dat1 <- dat1[order(dat1$log_mean),]
-dat2 <- dat2[order(dat2$log_mean),]
-
-dat <- rbind(dat2,dat1)
-sd14 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
-    geom_point()+geom_line() + ggtitle("Cat 14")
-@
-
-<<state_plots cat15, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="State Parameter Estimates">>=
-load("cat15sumdat.RData")
-temp5 <- tail(getpars(fitsin5s),10)
-temp6 <- tail(getpars(fithomo6s),12)
-dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
-                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM + TH(sin)") 
-dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
-                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM") 
-dat1 <- dat1[order(dat1$log_mean),]
-dat2 <- dat2[order(dat2$log_mean),]
-
-dat <- rbind(dat1,dat2)
-sd15 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
-    geom_point()+geom_line() + ggtitle("Cat 15")
-
-grid.arrange(sd1,sd2,sd14,sd15)
+<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters">>=
+## translate single fit to data frame of params
+trans_pars <- function(fit,npar,Model=NULL) {
+    x <- tail(getpars(fit),npar)
+    d <- data.frame(log_mean=x[names(x)!="sd"],
+                    log_sd=x[names(x)=="sd"])
+    if (!is.null(Model)) d$Model <- Model
+    d <- arrange(d,log_mean)
+    return(d)
+}
+## get pars for best-fit TH(sin) and HMM models for each cat
+get_cat_pars <- function(catnum) {
+    load(paste0("cat",catnum,"sumdat.RData"))
+    ## figure out
+    w <- with(sumdat,which(deltaBIC==min(deltaBIC[model=="HMM + THsin"])))
+    optstates <- sumdat$nstates[w]
+    ## ASSUME that best fit is het-sin, corresponding homo has +1 states
+    sinfit <- get(paste0("fitsin",optstates,"s"))
+    homofit <- get(paste0("fithomo",optstates+1,"s"))
+    d <- rbind(trans_pars(sinfit,optstates*2,Model="HMM+TH(sin)"),
+               trans_pars(homofit,(optstates+1)*2,Model="HMM"))
+    return(d)
+}
+all_cat_pars <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars,.id="catID")
+ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
+    geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)
 @

### 2016-03-02 paper3.Rnw
working on paper

@@
 standard clustering techniques, to model high-resolution hourly
-movement data from Florida panthers. Allowing for temporal
+movement data from Florida panthers (\emph{Puma concolor coryi}).
+Allowing for temporal
 heterogeneity in transition probabilities, a straightforward but
@@
 \citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
-\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
+\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, and as
+we discuss further below, behavioural
 repertoires with more than three distinct states are difficult to
-interpret, one possible reason that other authors have not adopted
+interpret --- one possible reason that other authors have not adopted
 \citeauthor{kerk2015hidden}'s model-based approach to
@@
 \citeauthor{kerk2015hidden}'s results might be driven at
-least in part by structural problems with the HMM, i.e. the
+least in part by structural problems with their statistical
+model, i.e. the
 assumption of temporally homogeneous behaviour.  When large data sets
@@
 this suggests that we might not lose much accuracy by reverting to 
-a simpler HMM framework, which makes precisely this assumption.
+a simpler HMM framework, which corresponds
+to making precisely this assumption.
 
@@
 data generation process by identifying the ``true'' number 
-of underlying movement states.  Using BIC also simplifies 
+of underlying movement states.  
+\todo{it occurs to me that we haven't really included any
+information about simulation testing in this version!
+We should probably add a short section; if don't, then we
+should delete following sentence}
+Using BIC also simplifies 
 evaluation of model selection procedures; 
@@
 P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
-P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \times \\
+P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
 &  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
@@
 
+The emissions $\obs_i$ are boldfaced to denote that we may have a vector of observations at each time point (e.g., step length and turning angle).
 The model contains three distinct components:
@@
 \begin{description}
-\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$; the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ and state $\state_{1}$.
+\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$, times the vector of observations $\obs_{1}$ conditioned on covariates $\cov_{1}$ and state $\state_{1}$.
 
@@
 
-\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
+\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observations $\obs_{k}$ conditioned on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
 \end{description}
+
+\todo{maybe give the formula for number of parameters (constrained) for each model type here?}
 
@@
 
-We considered three different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
+We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
 
 \begin{description}
+\item[Hourly model] \todo{describe hourly model.
+Explain that it's included for comparative purposes, i.e.
+not really practical.
+Explain that we only fitted up to 4 states (because too many parameters)}
 \item[Multiple block transition] Here we assume piecewise-constant transition probabilities.
@@
 
-Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity.
-See supplementary material Table 1.
+Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines\todo{citation?}, could be useful directions for future exploration.
+
+Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
 
@@
 We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations).
-(1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)).
-(2) Compare average step-length by hour of day for the observed data and for data simulated from the models.
-(3) Compare temporal autocorrelations for the observed data and for data simulated from the models.
-For the second and third metrics (hourly average and ACF), we simulated data from the models because the metrics are hard to compute from first principles for HMMs with temporally varying transition probabilities.
+(1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
+fit of each model type.
+(2) Comparing average step-length by hour of day for the observed data and for data simulated from the models shows how well a particular class of models can capture the diurnal variation in behaviour.
+(3) Comparing temporal autocorrelations for the observed data and for data simulated from the models shows how well a particular class of models
+captures serial correlation at both short and long scales. 
+We used simulations to predict hourly step length and ACF 
+because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
+\todo{need to explain here, if we haven't elsewhere, that using the Viterbi algorithm to simulate double-counts the data ...}
 
@@
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
-load("cat1allmod.RData")
+L <- load("cat1allmod.RData")
 df2 <- transform(allmodsumdf,
@@
 
-Figure 1 shows the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results.
-Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.
+Figure~\ref{fig:BICred_plot} shows the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results.
+Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.\todo{do we need to run some models with 7 states to show that $\dbic$ eventually increases?}
 
-<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
-load("cat1sumdat.RData")
+<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models",fig.width=8>>=
+L <- load("cat1sumdat.RData")
 ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
-    scale_size_continuous( name="Num of Parameters") +
+    scale_size_continuous( name="# of parameters") +
+    scale_x_continuous(expand=c(0,0.5),breaks=2:6)+
     geom_point(aes(size=parameters))+
@@
 
-As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions (24-hour multiple block transition model) to compare the temporal effects in goodness of fit.
-Looking at Figure 2, models with temporal heterogenity is better than models without in both FMM and HMM frameworks, but time homogenous HMM models is better than FMM with sinusoidal temporal heterogeneity.
-Temporal heterogenity is good with good models such as the quadratic and sinuosidal.
-Multiple block models is good for cats with distinguishable diurnal patterns and hourly model is too complicated.
+As a complement, we also fitted FMM; FMM with sinusoidal variation in state occupancy probabilities; and HMM with hourly transitions (24-hour multiple block transition model) to compare the temporal effects in goodness of fit.
+Figure~\ref{fig:adj_BIC_comparisons} shows that models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMM models are better than FMM with sinusoidal temporal heterogeneity.
+Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
+The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.
+Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
+However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
+normally be interpreted as an enormous difference in goodness of fit,
+between the sinusoidal and quadratic models (both of which have 90 
+parameters).
 
-<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
-  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
-  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
-  avgdf <- cbind(simdf,obsdf)
-  avgdf2 <- melt(avgdf,'time')
+<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day.
+\\textbf{todo: leave out hmm quad for clarity?
+fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such.
+Maybe set obs as a wide, pointless, semitransparent line?})",fig.width=8>>=
+simdf <- simdat %>% dplyr::select(-obs) %>%
+    group_by(time) %>% summarise_each(funs(mean))
+obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
+    filter(!is.na(obs)) %>% group_by(time) %>%
+    summarise_each(funs(mean))
+
+avgdf2 <- melt(simdf,'time')
 ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
-    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
-    scale_x_continuous(breaks=0:23)
+    geom_point() +geom_line()+ xlab("Time of Day") +
+    ylab("Step lengths (log10)")+
+    scale_x_continuous(breaks=0:23)+
+    geom_line(data=obsdf,
+              aes(y=obs),lwd=2.5,alpha=0.3,colour="black")
 @
 
-Figure 3 shows a clear diurnal pattern from the average hourly step lengths from the observed data.
-Models that assume temporal homogeneity (FMM and HMM) fail to capture the diurnal activity cycle.
-Models incorporating temporal heterogeneity can capture the observed patterns without the Viterbi algorithm.
+Figure~\ref{fig:avg_step_length_by_time} shows a clear diurnal pattern from the average hourly step lengths from the observed data.
+As expected, temporally homogeneous models (FMM and HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle.
+All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns.
+However, the block model does markedly worse than the other temporal models, and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
 
-<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
+<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF \\textbf{fix similarly to Fig 3}">>=
 acffun <- function(sim){
@@
 }
-
-  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
+temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)
   ggplot(temp,aes(lag,ACF,colour=.id))+
@@
 
-The observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity.
+The observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity (Figure~\ref{fig:acf_plot}).
 HMM without temporal heterogeneity can capture short-term autocorrelation (without Viterbi algorithm) but is unable to capture lags beyond 7.
@@
 
-<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters">>=
+<<get_cat_pars, warning=FALSE, message=FALSE, echo=FALSE, cache=TRUE>>=
 ## translate single fit to data frame of params
@@
 get_cat_pars <- function(catnum) {
-    load(paste0("cat",catnum,"sumdat.RData"))
+    L <- load(paste0("cat",catnum,"sumdat.RData"))
     ## figure out
@@
 all_cat_pars <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars,.id="catID")
+@
+
+<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters">>=
 ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
@@
 @
+
+TODO: add ellipses?
 
@@
 This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations.
-It is useful to predict missing data in the observation sequence but fail to future movements beyond the observations sequence.
+It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it cannot predict future movements beyond the observation sequence.
 In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.
 
-Although the time inhomogeneous transition offer an
+Although the time inhomogeneous transitions offer an
 objective approach to (1) reduce the number of states to a biologically interpretable level, (2) captures the observational autocorrelation and temporal pattern, (3) reduce the overall complexity as illustrated by our study, several challenges
@@
 \centering
-\caption{Number of Parameters for each model type.}
-\label{Cat-ID}
+\caption{Number of Parameters for each model type.
+\textbf{todo: add general formula for each model type, e.g.
+FMM = $S$ (occupancies) + $2S$ (emissions: mean and std dev for each state); HMM = $S(S-1)+2S=S^2+S$.
+Be clear about whether we're talking about restricted (constrained) or unconstrained.}
+}
+\label{npartab}
 \begin{tabular}{cccc}
@@
 
-<<BICred_plot cat131,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#2.
+<<BICred_plot_cat131,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#2.
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
@@
 
-<<BICred_plot cat94,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#15.
+<<BICred_plot_cat94,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#15.
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
@@
 
-temp <- vitcat(cat,fithomo3s) %>% dplyr:::select(-time)%>%ldply(.,acffun)
+temp <- vitcat(cat,fithomo3s) %>% dplyr::select(-time)%>%ldply(.,acffun)
 ggplot(temp,aes(lag,ACF,colour=.id))+
@@
 
-<<adj_BIC_comparisons cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+<<adj_BIC_comparisons_cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons between different HMM models">>=
 load("cat2sumdat.RData")
@@
 
-<<adj_BIC_comparisons cat14, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+<<adj_BIC_comparisons_cat14, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
 load("cat14sumdat.RData")
@@
 
-<<adj_BIC_comparisons cat15, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+<<adj_BIC_comparisons_cat15, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
 load("cat15sumdat.RData")
@@
 
-<<avg_step_length_by_time cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
+<<avg_step_length_by_time_cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
 load("cat2sumdat.RData")
-  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
-  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
+  simdf <- simdat %>% dplyr::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
+  obsdf <- simdat %>% dplyr::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr::select(obs)
   avgdf <- cbind(simdf,obsdf)
@@
 
-<<avg_step_length_by_time cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
+<<avg_step_length_by_time_cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
 load("cat14sumdat.RData")
-  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
-  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
+  simdf <- simdat %>% dplyr::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
+  obsdf <- simdat %>% dplyr::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr::select(obs)
   avgdf <- cbind(simdf,obsdf)
@@
 
-<<avg_step_length_by_time cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
+<<avg_step_length_by_time_cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
 load("cat15sumdat.RData")
-  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
-  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
+  simdf <- simdat %>% dplyr::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
+  obsdf <- simdat %>% dplyr::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr::select(obs)
   avgdf <- cbind(simdf,obsdf)
@@
 
-<<acf_plot cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
+<<acf_plot_cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
 load('cat2sumdat.RData')
@@
 
-  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
+  temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)
   ggplot(temp,aes(lag,ACF,colour=.id))+
@@
 
-<<acf_plot cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
+<<acf_plot_cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
 load("cat14sumdat.RData")
@@
 
-<<acf_plot cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
+<<acf_plot_cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
 load("cat15sumdat.RData")

### 2016-03-02 paper3.Rnw
up to figure 5

@@
 L <- load("cat1sumdat.RData")
-ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
+(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+
+    facet_wrap(~ type,ncol=4)+
     scale_size_continuous( name="# of parameters") +
@@
          x="Number of States") +
-    geom_line() + zmargin
+    geom_line() + zmargin)
 @
@@
 fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such.
-Maybe set obs as a wide, pointless, semitransparent line?})",fig.width=8>>=
+Opinion on this way of handling obs data (need to include in legend)?})",fig.width=8>>=
 simdf <- simdat %>% dplyr::select(-obs) %>%
@@
 Figure~\ref{fig:avg_step_length_by_time} shows a clear diurnal pattern from the average hourly step lengths from the observed data.
-As expected, temporally homogeneous models (FMM and HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle.
+As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle.
 All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns.
@@
 
-The observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity (Figure~\ref{fig:acf_plot}).
-HMM without temporal heterogeneity can capture short-term autocorrelation (without Viterbi algorithm) but is unable to capture lags beyond 7.
-FMM with temporal heterogeneity can capture lags beyond 12, but not short-term autocorrelations.
-HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data whereas the FMM completely missed the autocorrelations.
+Like the diurnal pattern shown in Figure~\ref{fig:avg_step_length_by_time},
+the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure~\ref{fig:acf_plot}) shows the need to incorporate temporal heterogeneity in the model even without any formal statistical modeling.
+Because there are an unlimited number of potential complexities that can be added to movement models (e.g.
+spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.
+In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects.
+HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
+Temporally homogeneous FMM, by definition, produces neither short- nor long-term autocorrelation.
+FMM without temporal heterogeneity, although it captures the diurnal pattern well, underpredicts the degree of short-term autocorrelation.
 
@@
 
-<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters">>=
-ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
-    geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)
+<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale}">>=
+source("bhull.R")
+g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
+    geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
+g0 +
+    geom_encircle(data=subset(all_cat_pars,
+                              (catID=="15" & log_mean<1.5) |
+                              (catID!="15" & log_mean<1.8)),
+                  colour="black",linetype=2,expand=0.05)
 @
 
-TODO: add ellipses?
-
+\todo{maybe do Viterbi plots for mean state occupancy by hour for HMM vs HMM/TH models? Still trying to figure out what's really going on here ...}
 The biggest issue with multiple states is identifying them biologically.
@@
 ggtitle("ACF plot of BIC Selected Models")
-
 @
 
-<<adj_BIC_comparisons_cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons between different HMM models">>=
+\todo{does it make any sense to compress figs~\ref{adj_BIC-comparisons_cat2} etc.
+into a single plot (via facetgrid or multi-lines)?
+Same idea for all repeated plots; also, update to use percent-+-percent syntax}
+
+<<adj_BIC_comparisons_cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons between different HMM models",fig.width=8>>=
 load("cat2sumdat.RData")
-ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
-    scale_size_continuous( name="Num of Parameters") +
-    geom_point(aes(size=parameters))+
-    scale_colour_brewer(palette="Dark2")+
-    labs(y=expression(paste(Delta, "BIC")),
-         x="Number of States") +
-    geom_line() + zmargin
+adj_BIC_plot %+% sumdat
 @
 
-<<adj_BIC_comparisons_cat14, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+<<adj_BIC_comparisons_cat14, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models",fig.width=8>>=
 load("cat14sumdat.RData")
-ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
-    scale_size_continuous( name="Num of Parameters") +
-    geom_point(aes(size=parameters))+
-    scale_colour_brewer(palette="Dark2")+
-    labs(y=expression(paste(Delta, "BIC")),
-         x="Number of States") +
-    geom_line() + zmargin
+adj_BIC_plot %+% sumdat
 @
 
-<<adj_BIC_comparisons_cat15, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
+<<adj_BIC_comparisons_cat15, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models",fig.width=8>>=
 load("cat15sumdat.RData")
-ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
-    scale_size_continuous( name="Num of Parameters") +
-    geom_point(aes(size=parameters))+
-    scale_colour_brewer(palette="Dark2")+
-    labs(y=expression(paste(Delta, "BIC")),
-         x="Number of States") +
-    geom_line() + zmargin
+adj_BIC_plot %+% sumdat
 @

### 2016-03-14 paper3.Rnw
update encircle

@@
 library("gridExtra")
+if (!require("ggalt")) {
+    cat("installing ggalt from BB repo ...\n")
+    library("devtools")
+    install_github("bbolker/ggalt")
+    library("ggalt")
+}
 @
@@
 <<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale}">>=
-source("bhull.R")
 g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
-g0 +
+(g0c <- g0 +
     geom_encircle(data=subset(all_cat_pars,
@@
                               (catID!="15" & log_mean<1.8)),
-                  colour="black",linetype=2,expand=0.05)
+                  colour="black",linetype=2,expand=0.05))
 @

### 2016-04-11 paper3.Rnw
small changes

@@
     scale_size_continuous( name="# of parameters") +
-    scale_x_continuous(expand=c(0,0.5),breaks=2:6)+
+    scale_x_continuous(expand=c(0,0.5),breaks=2:7)+
     geom_point(aes(size=parameters))+
@@
  Type & Model  & Number of States & Number of Parameters  \\
-  FMM & FMM & 3 & 9    \\
-  FMM & FMM & 4 & 12    \\
-  FMM & FMM & 5 & 15   \\
-  FMM & FMM & 6 & 18  \\
-  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
-  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
-  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
-  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
-  HMM & HMM & 3 & 18    \\
-  HMM & HMM & 4 & 28    \\
-  HMM & HMM & 5 & 40   \\
-  HMM & HMM & 6 & 54  \\
-  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
-  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
-  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
-  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
-  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
-  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
-  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
-  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
-  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
-  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
-  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
-  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
-  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
-  HMM + TH & HMM + TH (Hourly)  & 4 & 396    
+  FMM & FMM & 3 & 8    \\
+  FMM & FMM & 4 & 11    \\
+  FMM & FMM & 5 & 14   \\
+  FMM & FMM & 6 & 17  \\
+  FMM & FMM & 7 & 20  \\
+  FMM + TH & FMM + TH (Sin)  & 3 & 12    \\
+  FMM + TH & FMM + TH (Sin)  & 4 & 17    \\
+  FMM + TH & FMM + TH (Sin)  & 5 & 22    \\
+  FMM + TH & FMM + TH (Sin)  & 6 & 27    \\
+  FMM + TH & FMM + TH (Sin)  & 7 & 32    \\
+  HMM & HMM & 3 & 14    \\
+  HMM & HMM & 4 & 23    \\
+  HMM & HMM & 5 & 24   \\
+  HMM & HMM & 6 & 47  \\
+  HMM & HMM & 7 & 62  \\
+  HMM + TH & HMM + TH (Block)  & 3 & 26    \\
+  HMM + TH & HMM + TH (Block)  & 4 & 47    \\
+  HMM + TH & HMM + TH (Block)  & 5 & 74    \\
+  HMM + TH & HMM + TH (Block)  & 6 & 107    \\
+  HMM + TH & HMM + TH (Block)  & 7 & 146    \\
+  HMM + TH & HMM + TH (Quad)  & 3 & 26    \\
+  HMM + TH & HMM + TH (Quad)  & 4 & 47    \\
+  HMM + TH & HMM + TH (Quad)  & 5 & 74    \\
+  HMM + TH & HMM + TH (Quad)  & 6 & 107    \\
+  HMM + TH & HMM + TH (Quad)  & 7 & 146    \\
+  HMM + TH & HMM + TH (Sin)  & 3 & 26    \\
+  HMM + TH & HMM + TH (Sin)  & 4 & 47    \\
+  HMM + TH & HMM + TH (Sin)  & 5 & 74    \\
+  HMM + TH & HMM + TH (Sin)  & 6 & 107    \\
+  HMM + TH & HMM + TH (Sin)  & 7 & 146    \\
+  HMM + TH & HMM + TH (Hourly)  & 3 & 152    \\
+  HMM + TH & HMM + TH (Hourly)  & 4 & 299    
 \end{tabular}

### 2016-04-11 cat1seeds.R

@@ -5,7 +5,7 @@ fitfmm3s <- fitmix(3,cat,seed=43)
 fitfmm4s <- fitmix(4,cat,seed=9)
 fitfmm5s <- fitmix(5,cat,seed=3)
 fitfmm6s <- fitmix(6,cat,seed=1)
-fithmm7s <- fitmix(7,cat,seed=1)
+fitfmm7s <- fitmix(7,cat,seed=1)
 
 #FMM sin ----
 fitfmmsin3 <- fitmixsin(3,cat,seed=3)

### 2016-04-11 summary.R

@@ -2,33 +2,33 @@ library(depmixS4)
 library(dplyr)
 library(plyr)
 
-fitlist <- list(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,
-                fitfmmsin3,fitfmmsin4,fitfmmsin5,fitfmmsin6,
-                fithomo3s,fithomo4s,fithomo5s,fithomo6s,
+fitlist <- list(fitfmm3s,fitfmm4s,fitfmm5s,fitfmm6s,fitfmm7s,
+                fitfmmsin3,fitfmmsin4,fitfmmsin5,fitfmmsin6,fitfmmsin7,
+                fithomo3s,fithomo4s,fithomo5s,fithomo6s,fithomo7s,
                 fithourly3,fithourly4,
-                fitblock3s,fitblock4s,fitblock5s,fitblock6s,
-                fitquad3s,fitquad4s,fitquad5s,fitquad6s,
-                fitsin3s,fitsin4s,fitsin5s,fitsin6s)
+                fitblock3s,fitblock4s,fitblock5s,fitblock6s,fitblock7s,
+                fitquad3s,fitquad4s,fitquad5s,fitquad6s,fitquad7s,
+                fitsin3s,fitsin4s,fitsin5s,fitsin6s,fitsin7s)
 
 sumdf <- function(lst){
   BIC <- ldply(lst,BIC)
   nstates <-ldply(lst,nstates)
-  para <- ldply(lst,npar)
-  model <- c('FMM','FMM','FMM','FMM',
-             'FMM + THsin','FMM + THsin','FMM + THsin','FMM + THsin',
-             'HMM','HMM','HMM','HMM',
+  para <- ldply(lst,freepars)
+  model <- c('FMM','FMM','FMM','FMM','FMM',
+             'FMM + THsin','FMM + THsin','FMM + THsin','FMM + THsin','FMM + THsin',
+             'HMM','HMM','HMM','HMM','HMM',
              'HMM + THhourly','HMM + THhourly',
-             'HMM + THblock','HMM + THblock','HMM + THblock','HMM + THblock',
-             'HMM + THquad','HMM + THquad','HMM + THquad','HMM + THquad',
-             'HMM + THsin','HMM + THsin','HMM + THsin','HMM + THsin')
+             'HMM + THblock','HMM + THblock','HMM + THblock','HMM + THblock','HMM + THblock',
[…]

### 2016-04-13 cat2seeds.R
uploading cat2 data, still fitting cat2allmod.RData

@@ -54,7 +54,7 @@ simhomo5 <- simhomo(5,cat,fithomo5s)
 simhourly3 <- simhourly(3,cat,fithourly3)
 simblock4 <- simblock(4,cat,fitblock4s)
 simquad4 <- simquad(4,cat,fitquad4s)
-simsin4 <- simsin(4,cat,fitsin4s)
+simsin5 <- simsin(5,cat,fitsin5s)
 
 simdat <- data.frame(obs = cat$LogDist,
                      fmm4 = simfmm4$obs,
@@ -63,7 +63,7 @@ simdat <- data.frame(obs = cat$LogDist,
                      hmmhourly3 = simhourly3$obs,
                      hmmblock4 = simblock4$obs,
                      hmmquad4 = simquad4$obs,
-                     hmmsin4 = simsin4$obs,
+                     hmmsin5 = simsin5$obs,
                      time = cat$Time)
 
 

### 2016-04-13 Weibull_het.R
all the meta data are complete, I owe you a parameter formula and then I am done, back to you

@@ -317,7 +317,7 @@ inMod <- transInit(~ 1, ns = 7, pstart = rep(1/7, 7),
                    data = data.frame(1))
 ws7 <- makeDepmix(response = rModels, transition = transition,
                   prior=inMod,homogeneous = FALSE)
-Wsin7 <- fit(ws7, verbose = TRUE, emc=em.control(rand=FALSE, maxit=460))
+Wsin7 <- fit(ws7, verbose = TRUE, emc=em.control(rand=FALSE, maxit=280))
 summary(Wsin7)
 
 

### 2016-04-13 paper3.Rnw
adding formulas for constraint parameter calculation

@@
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
-allmodsumdf <- readRDS("cat2allmod.RDS")
+load("cat2allmod.RData")
 df2 <- transform(allmodsumdf,
@@
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
-allmodsumdf <- readRDS("cat15allmod.RDS")
+load("cat15allmod.RData")
 df2 <- transform(allmodsumdf,

### 2016-06-07 simtest.R
simple simtest, going to do a full simulation soon

@@ -13,8 +13,8 @@ system.time(mod <- depmix(y~1
 )
 getpars(mod)
 
-set.seed(unlist(strsplit(input_files,"[.]"))[1])
-
+# set.seed(unlist(strsplit(input_files,"[.]"))[1])
+set.seed(101)
 randpars <- sample(1:4,length(getpars(mod))-6,replace=TRUE)
 newmod <- setpars(mod,c(0.5,0.5,randpars,0,2,4,2))
 newmod

### 2016-06-07 guelph_talk/guelph_panther.Rnw
working on Guelph talk

\documentclass[english]{beamer}
\definecolor{links}{HTML}{2A1B81}
\hypersetup{colorlinks,linkcolor=,urlcolor=links}
\usepackage{natbib}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{color}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{animate}
\usepackage{arydshln}
\usepackage[normalem]{ulem}
\newcommand{\colemph}[1]{{\color{red} \emph{#1}}}
%\bibliographystyle{shortreflist}
\bibliographystyle{notitle}

\usetheme{Frankfurt}
\usecolortheme{dove}
\setbeamercovered{transparent}
\setbeamercolor{description item}{fg=blue}
\newcommand{\citepx}[1]{{\small \citep{#1}}}
\newcommand{\citex}[1]{{\small \cite{#1}}}
% \renewcommand{\descriptionlabel}[1]{\hspace{\labelsep}\textbf{#1}}
\usepackage{babel}

% "Beamer infolines outer theme with miniframe bullets only for the current section"
% (http://tex.stackexchange.com/a/45152/3323)
%\useoutertheme[subsection=false]{miniframes}
\useoutertheme[subsection=false]{miniframes}
%% \setbeamertemplate{mini frame in other subsection}{}
%% \usepackage{etoolbox}
\makeatletter
\beamer@compressfalse
%% \patchcmd{\insertnavigation}{\hskip-1.875ex plus-1fill}{}{}{}
%% \patchcmd{\sectionentry}{\beamer@section@set@min@width}{}{}{}
%% \patchcmd{\sectionentry}{\hskip1.875ex plus 1fill}{}{}{}
%% \patchcmd{\sectionentry}{\hyperlink{Navigation#3}{{\usebeamertemplate{section in head/foot shaded}}}}{}{}{}
%% \patchcmd{\slideentry}{\beamer@ypos=#2\relax}{}{}{}
%% \patchcmd{\fakeslideentry}{\beamer@ypos=#2\relax}{}{}{}
\makeatother

\begin{document}

\newcommand{\rzero}{{\mathcal R}_0}
\newcommand{\rzeroi}[1]{{\mathcal R}_{0,#1}}

\makeatletter
\def\newblock{\beamer@newblock}
\makeatother

<<setup,echo=FALSE,message=FALSE>>=
library(knitr)
opts_chunk$set(echo=FALSE,error=FALSE,
               fig.align="center",out.width="0.8\\textwidth",
               fig.width=6,fig.height=4)
library(deSolve)
library(reshape)
library(lattice)
library(latticeExtra)
library(emdbook)
library(nlme)
library(bbmle)
library(tikzDevice)
## http://yihui.name/en/2011/04/produce-authentic-math-formulas-in-r-graphics/
options(tikzMetricPackages = c( "\\usepackage[utf8]{inputenc}",
                "\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}",
                "\\usepackage{amssymb}"))

##options(tikzMetricPackages = c("\\usepackage[utf8]{inputenc}",
##    "\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}",
##    "\\usepackage{amssymb}"))
library(ggplot2)
theme_set(theme_bw())
library("ggplot2"); theme_set(theme_classic())
library("MASS")
## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally")
library("ggstance")
library(scales) ## for squish()
library(gridExtra) ## for grid.arrange()
library(mgcv)
library(plyr)   ## for mutate()
library(psyphy) ## for restricted-link models
library(reshape2)
library(splines)  ## for vir asymptote profile
library(deSolve)
library(Hmisc)
library(gsubfn)
@

\title[Animal movement]{Model complexity and model choice for animal movement models}
\author[Ben Bolker]{Ben~Bolker, McMaster University \\
  {\tiny Departments of Mathematics \& Statistics and Biology}}
\institute{Guelph Biomathematics \& Biostatistics Symposium}
\date{9 June 2016}
% \pgfdeclareimage[height=0.5cm]{uflogo}{letterhdwm}
% \logo{\pgfuseimage{uflogo}}
\AtBeginSubsection[]{
  \frame<beamer>{ 
     \frametitle{Outline}   
     \tableofcontents[currentsection,currentsubsection] 
   }
 }

\begin{frame}
\titlepage
\end{frame}
% \beamerdefaultoverlayspecification{<+->}

\begin{frame}
\frametitle{Outline}
\tableofcontents{}
\end{frame}

\begin{frame}
\frametitle{Acknowledgements}
\begin{description}
\item[People] Michael Li, Madelon van de Kerk, Dave Onorato
\item[Support] NSERC Discovery grant
\end{description}

\end{frame}

\begin{frame}
\frametitle{To do}
\begin{itemize}
\item pix!
\item general REFS: Turchin, Morales et al, Langrock ... ?
\end{itemize}
\end{frame}

\section{Animal movement}
\subsection{}

\begin{frame}
\frametitle{Animal movement: data}

\begin{itemize}
\item observations:  \\
e.g. mass mark-recapture, longitudinal density,
direct observation, telemetry (VHF, GPS)
\item most methods provide a sequence of times and locations for each individual
\item summaries:
\begin{itemize}
\item home range \\
(convex hull, kernel density estimate, etc.)
\item root-mean-squared displacement
\item step length and turning angle
\end{itemize}
\item covariates: \\
e.g. habitat map, individual characteristics (sex, age, weight \ldots)
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Animal movement: questions}
\begin{itemize}
\item simple description
\item how do animals' movements change  
as a function of their (internal or external) environment?  \\
what does that tell us about their biology?
\item how might animals' distributions, etc. change when conditions 
(density, habitat, ...) change?
\end{itemize}
\end{frame}

\section[Panthers]{Florida panthers}
\subsection{}

\begin{frame}
\frametitle{Biological/conservation issues}

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item Florida panther: \emph{Felix ...}
    \item severely reduced habitat, endangered
    \item small, isolated population
    \item currently recovering
    \end{itemize}
  \end{column}
\end{columns}

\end{frame}

\begin{frame}
\frametitle{Panther movement questions}

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item abc
    \end{itemize}
  \end{column}
\end{columns}

\end{frame}

\begin{frame}
\frametitle{Panther movement data}

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item panthers tracked, captured
    \item GPS collars
    \item 18 males (13 male, 5 female, 1-15 years old)
    \item 3200 panther days, hourly/bihourly; 49000 locations
    \item ?? per panther
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    \includegraphics[trim={10cm 0 10cm 0},clip,width=\textwidth]{pix/panther_collar.jpg}
  \end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{example movement tracks}
FIXME: picture here!
(animation?? map background??)
\end{frame}

\section[HMM]{Hidden Markov models}
\subsection{}

\begin{frame}
\frametitle{Hidden Markov models}

\begin{itemize}
\item finite mixture model with temporal dependence
\item discrete time steps
\item discrete latent state; \emph{transition matrix}
\item observations from \emph{emission distributions} \\
(continuous or discrete, univariate or multivariate)
\end{itemize}

\includegraphics[width=\textwidth]{pix/hmm.jpg}
\end{frame}

\begin{frame}
\frametitle{Hidden Markov models (cont.)}

\begin{itemize}
\item \emph{forward-backward algorithm} for estimating parameters
\item \emph{Viterbi algorithm} for estimating most probable state sequences
\item {\tt depmixS4} package \citepx{visser2010depmixs4}
(also {\tt moveHMM} \citepx{michelot_movehmm_2016}, {\tt move.HMM} \citepx{augustine_hmm})
\end{itemize}

\end{frame}

\section[Basic analysis]{Basic analysis (vdK et al.)}
\subsection{}

\begin{frame}
\frametitle{State distributions}

\includegraphics[trim={0 6.5cm 0 0}, clip, width=\textwidth]{pix/jae_fig2.png}

\end{frame}

\begin{frame}
\frametitle{Parameter estimates}

\includegraphics[width=\textwidth]{pix/jae_fig3.png}

\end{frame}

\begin{frame}
\frametitle{Transition parameters}

picture/table here of transition parameters (network diagram??)
\end{frame}

\begin{frame}
\frametitle{Diurnal variation}

\includegraphics[trim={0 8cm 0 0}, clip, width=\textwidth]{pix/jae_fig4.png}

\end{frame}

\begin{frame}
\frametitle{what can we conclude so far?}

\textbf{good news}

\begin{itemize}
\item basic biology: males move faster, farther
\item three states are identifiable, sensible
\item ...
\end{itemize}

\pause
\textbf{bad news}
\begin{itemize}
\item diurnal variation in Viterbi results - but it's not in the model!
\item estimates of model complexity are too high
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Model complexity}

\includegraphics[height=\textheight]{pix/jae_fig1.png}

\end{frame}

\section[Diurnal model]{Incorporating diurnal variation}
\subsection{}

\begin{frame}
\frametitle{Broadening the model}

Attempting to fix these problems:
\begin{itemize}
\item extend the model to allow covariates
\item specifically, allow for diurnal variation
\begin{itemize}
\item simplify model (log-Normal step length only)
\item \emph{fixed} state-specific emissions parameters\\
(step length mean and std dev)
\item time-varying transition parameters
\item also try \emph{finite mixture models} \\
(independent occupancy)
\end{itemize}
\item how much does this help?
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Temporal models}

figure showing alternative temporal models
\end{frame}

\begin{frame}
\frametitle{Comparison}

Figure 1 from paper
\end{frame}

\begin{frame}
\frametitle{Goodness of fit}

Figure 2 from paper
\end{frame}

\begin{frame}
\frametitle{Temporal patterns}
Figures 3 (and 4?) from paper
\end{frame}

\begin{frame}
\frametitle{Model complexity}
Figure showing BIC plots for all cats tried
\end{frame}

\begin{frame}
\frametitle{Model complexity}
Simulation results
\end{frame}

\section{Broader issues/outlook}
\subsection{}

\begin{frame}
\frametitle{Big data and small models}

\begin{itemize}
\item simple model families +  \\
  model misspecification $\to$ \\
  overparameterization

\end{itemize}
\end{frame}

\begin{frame}
\frametitle{an aside on AIC vs BIC}

\end{frame}

\begin{frame}
\frametitle{Tools needed}

\begin{itemize}
\item cross-validation \citepx{wenger_assessing_2012}
\item diagnostic plots
\item score tests?
\end{itemize}

\end{frame}

\begin{frame}
\begin{multicols}{2}
\frametitle{References}

% \tiny
\fontsize{3pt}{5pt}\selectfont
\bibliography{../paper}
\end{multicols}
\end{frame}

\end{document}

### 2016-06-08 guelph_talk/guelph_panther.Rnw
guelph talk ...

@@
 
+\newcommand{\fixme}[1]{{\color{red} \textbf{#1}}}
 \usetheme{Frankfurt}
@@
 \begin{description}
-\item[People] Michael Li, Madelon van de Kerk, Dave Onorato
-\item[Support] NSERC Discovery grant
+\item[People] \textbf{Michael Li}, \textbf{Madelon van de Kerk}, Dave Onorato, Madan Oli
+\item[Agencies] US Fish and Wildlife, US Geological Survey, US National Park Service
+\item[Funding] NSERC Discovery grant, NSF IGERT program
 \end{description}
@@
 \item covariates: \\
-e.g. habitat map, individual characteristics (sex, age, weight \ldots)
+e.g. habitat map,  \\
+individual characteristics (sex, age, weight \ldots)
 \end{itemize}
@@
   \end{column}
+  \begin{column}{0.5\textwidth}
+    (cute panther pictures here)
+  \end{column}
 \end{columns}
@@
     \begin{itemize}
-    \item abc
+    \item how does panther movement vary by sex and life history stage \\
+      (juvenile, adult, mom with kittens \ldots)
+    \item how does movement mediate exposure to threats \\
+      (intraspecific aggression, roadkill) ?
+    \item can we predict the effects of future changes in 
+      population density / population structure / habitat?
     \end{itemize}
+  \end{column}
+  \begin{column}{0.5\textwidth}
+    (more cute pictures here ... or roadkill?)
   \end{column}
@@
 \frametitle{example movement tracks}
-FIXME: picture here!
-(animation?? map background??)
+\includegraphics[width=\textwidth]{pix/track1.pdf}
 \end{frame}
@@
 (continuous or discrete, univariate or multivariate)
+\item \textbf{multiphasic movement} \citepx{fryxell_multiple_2008,langrock_flexible_2012}
 \end{itemize}
@@
 
+\newcommand{\obs}{\mathbf{y}}
+\newcommand{\cov}{\mathbf{x}}
+\newcommand{\state}{z}
+
 \begin{frame}
 \frametitle{Hidden Markov models (cont.)}
+
+\fixme{notation, underbraces}
+
+\begin{equation}
+\begin{split}
+P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
+P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
+&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
+\end{split}
+\label{eq:HMMlik}
+\end{equation}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Hidden Markov models (part 3)}
 
@@
 \item {\tt depmixS4} package \citepx{visser2010depmixs4}
-(also {\tt moveHMM} \citepx{michelot_movehmm_2016}, {\tt move.HMM} \citepx{augustine_hmm})
+(also {\tt moveHMM} \citepx{michelot_movehmm_2016})
+\item hidden \emph{semi-Markov} models: allow for non-geometric \emph{dwell distributions} \citepx{langrock_hidden_2011-1,augustine_hmm}: {\tt move.HMM}
 \end{itemize}
@@
 
-\section[Basic analysis]{Basic analysis (vdK et al.)}
+\section[Basic analysis]{Basic analysis \citepx{kerk2015hidden}}
 \subsection{}
@@
 \includegraphics[width=\textwidth]{pix/jae_fig3.png}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Tracks with Viterbi estimates}
+
+\includegraphics[width=\textwidth]{pix/track1C.pdf}
 
@@
 
-\includegraphics[height=\textheight]{pix/jae_fig1.png}
+\includegraphics[height=0.8\textheight]{pix/BICplot.pdf}
 

### 2016-06-08 simtest.R
ready to run simulations

@@ -13,8 +13,8 @@ system.time(mod <- depmix(y~1
 )
 getpars(mod)
 
-# set.seed(unlist(strsplit(input_files,"[.]"))[1])
-set.seed(101)
+set.seed(unlist(strsplit(input_files,"[.]"))[1])
+
 randpars <- sample(1:4,length(getpars(mod))-6,replace=TRUE)
 newmod <- setpars(mod,c(0.5,0.5,randpars,0,2,4,2))
 newmod
@@ -69,3 +69,49 @@ print(summary(fit1))
 print(summary(fit2))
 print(summary(fit3))
 print(summary(fit4))
+
+dat <- data.frame(seed=seed
+                  , twoS_LL = logLik(fit1)
+                  , twoS_AIC = AIC(fit1)
+                  , twoS_BIC = BIC(fit1)
+                  , twoS_mu1 = getpars(fit1)[7]
+                  , twoS_mu2 = getpars(fit1)[9]
+                  , twoS_sd1 = getpars(fit1)[8]
+                  , twoS_sd2 = getpars(fit1)[10]
+                  , threeS_LL = logLik(fit2)
+                  , threeS_AIC = AIC(fit2)
+                  , threeS_BIC = BIC(fit2)
+                  , threeS_mu1 = getpars(fit2)[13]
+                  , threeS_mu2 = getpars(fit2)[15]
+                  , threeS_mu3 = getpars(fit2)[17]
+                  , threeS_sd1 = getpars(fit2)[14]
+                  , threeS_sd2 = getpars(fit2)[16]
+                  , threeS_sd3 = getpars(fit2)[18]
+                  , fourS_LL = logLik(fit3)
+                  , fourS_AIC = AIC(fit3)
+                  , fourS_BIC = BIC(fit3)
+                  , fourS_mu1 = getpars(fit3)[21]
+                  , fourS_mu2 = getpars(fit3)[23]
+                  , fourS_mu3 = getpars(fit3)[25]
+                  , fourS_mu4 = getpars(fit3)[27]
+                  , fourS_sd1 = getpars(fit3)[22]
+                  , fourS_sd2 = getpars(fit3)[24]
+                  , fourS_sd3 = getpars(fit3)[26]
+                  , fourS_sd4 = getpars(fit3)[28]
+                  , fiveS_LL = logLik(fit4)
+                  , fiveS_AIC = AIC(fit4)
+                  , fiveS_BIC = BIC(fit4)
[…]

### 2016-06-08 simtest.R
simulation complete

@@ -13,7 +13,8 @@ system.time(mod <- depmix(y~1
 )
 getpars(mod)
 
-set.seed(unlist(strsplit(input_files,"[.]"))[1])
+seed = unlist(strsplit(input_files,"[.]"))[1]
+set.seed(seed)
 
 randpars <- sample(1:4,length(getpars(mod))-6,replace=TRUE)
 newmod <- setpars(mod,c(0.5,0.5,randpars,0,2,4,2))
@@ -115,3 +116,5 @@ dat <- data.frame(seed=seed
 
 rownames(dat) <- NULL
 saveRDS(dat, file=paste("sim",seed,"RDS",sep="."))
+
+# rdsave(seed)

### 2016-06-08 plotsimtest.R
AIC and BIC plots

@@ -1,0 +1,38 @@
+library(ggplot2)
+library(reshape2)
+library(dplyr)
+
+dat <- readRDS("simtest.RDS")
+
+findS <- function(Smin,S2,S3,S4,S5){
+  if(Smin==S2){return(2)}
+  if(Smin==S3){return(3)}
+  if(Smin==S4){return(4)}
+  if(Smin==S5){return(5)}
+}
+
+datABICS <- (dat 
+  %>% rowwise()
+  %>% mutate(minBIC = min(twoS_BIC,threeS_BIC,fourS_BIC,fiveS_BIC)
+             , minAIC = min(twoS_AIC,threeS_AIC,fourS_AIC,fiveS_AIC))
+  %>% mutate(minBICStates = findS(minBIC,twoS_BIC,threeS_BIC,fourS_BIC,fiveS_BIC)
+             , minAICStates = findS(minAIC,twoS_AIC,threeS_AIC,fourS_AIC,fiveS_AIC))
+  %>% select(c(seed,minAICStates,minBICStates))
+)
+
+print(table(datABICS$minAICStates))
+print(table(datABICS$minBICStates))
+
+AICdat <- datABICS %>% count(minAICStates) %>% transmute(States=minAICStates,n=n,type="AIC")
+BICdat <- datABICS %>% count(minBICStates) %>% transmute(States=minBICStates,n=n,type="BIC")
+
+ICdat <- rbind(AICdat,BICdat)
+
+g1 <- (ggplot(ICdat,aes(x=States,y=n,color=type,group=type))
+       + geom_point()
+       + geom_line()
+       + theme_bw()
+       + ylab("Counts")
+       + xlab("Number of States")
+       + ggtitle("Simulation Results: Fitting HMMs to Two-State temphet HMM")
+) 

### 2016-06-08 guelph_talk/guelph_panther.Rnw
still working on guelph talk

@@
 ##    "\\usepackage{amssymb}"))
-library(ggplot2)
-theme_set(theme_bw())
-library("ggplot2"); theme_set(theme_classic())
+library("ggplot2"); theme_set(theme_bw())
+library(RColorBrewer)
+zmargin <- theme(panel.margin=grid::unit(0,"lines"))
+
 library("MASS")
@@
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
-library("gridExtra")
-library("plyr")  ## for ldply()
-library("dplyr") ## for full_join()
-library("reshape2")  ## for melt()
-library("GGally")
-library("ggstance")
+library(gridExtra)
+library(plyr)  ## for ldply()
+library(dplyr) ## for full_join()
+library(reshape2)  ## for melt()
+library(tidyr)
+library(GGally)
+library(ggstance)
 library(scales) ## for squish()
@@
 library(gsubfn)
+@
+
+<<cat1sumdat,echo=FALSE,cache=TRUE>>=
+L <- load("../cat1sumdat.RData")
+ss <- sumdat %>% filter(grepl("HMM",type)) %>%
+    mutate(deltaBIC=deltaBIC-min(deltaBIC),
+           model=factor(model,
+              levels=c("HMM","HMM + THsin","HMM + THquad",
+                       "HMM + THblock", "HMM + THhourly"),
+              labels=c("homogeneous","sinusoid","quadratic",
+                       "block","hourly")))
 @
@@
 
+\begin{columns}
+\begin{column}{0.5\textwidth}
 \begin{itemize}
@@
 \item most methods provide a sequence of times and locations for each individual
+\end{itemize}
+\end{column}
+\begin{column}{0.5\textwidth}
+(sequence pix)
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\frametitle{}
+\begin{columns}
+\begin{column}{0.5\textwidth}
+\begin{itemize}
 \item summaries:
@@
 \end{itemize}
-
+\end{column}
+\begin{column}{0.5\textwidth}
+(step length/turning angle pix)
+\end{column}
+\end{columns}
 \end{frame}
@@
     \begin{itemize}
-    \item Florida panther: \emph{Felix ...}
-    \item severely reduced habitat, endangered
+    \item Florida panther: \emph{Puma concolor coryi}
+    \item endangered subspecies
+    \item severely reduced habitat
     \item small, isolated population
@@
   \begin{column}{0.5\textwidth}
-    (cute panther pictures here)
+    \includegraphics[width=0.8\textwidth]{pix/map1.jpg} \\
+      {\tiny \href{http://www.peer.org/campaigns/wildlife-protection/florida-panther/ecological-needs.html}{www.peer.org}} \\
+     \includegraphics[width=0.9\textwidth]{pix/kitten1.jpg}
   \end{column}
 \end{columns}
-
 \end{frame}
@@
     \begin{itemize}
-    \item how does panther movement vary by sex and life history stage \\
+    \item movement variation by sex and life history stage \\
       (juvenile, adult, mom with kittens \ldots)
-    \item how does movement mediate exposure to threats \\
+    \item effects of movement on threats \\
       (intraspecific aggression, roadkill) ?
-    \item can we predict the effects of future changes in 
-      population density / population structure / habitat?
+    \item predicting the effects of future changes in 
+      population density / population structure / habitat
     \end{itemize}
@@
   \begin{column}{0.5\textwidth}
-    (more cute pictures here ... or roadkill?)
+    \includegraphics[width=0.8\textwidth]{pix/momkitten1.jpg} \\
+    \includegraphics[width=0.7\textwidth]{pix/roadkill.jpg}
   \end{column}
@@
 
-\fixme{notation, underbraces}
+\textbf{state:}
+\begin{equation*}
+\begin{split}
+  S_t & \sim \textrm{Multinomial}(S_{t-1},\mu_{S,t}) \\
+\mu_{S,t} & = \textrm{multi-logistic}(\mathbf{X}_{S,t} \boldsymbol \beta_{S} ) 
+\end{split}
+\end{equation*}
 
-\begin{equation}
+\textbf{emission:}
+\begin{equation*}
 \begin{split}
-P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
-P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
-&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
+\mathbf{Z}_t & \sim \left\{ \textrm{Dist}_1(\mu_{Z_1,S_t}),
+\dots, \textrm{Dist}_n(\mu_{Z_n,S_t}) \right\}
+ \\
+\mu_{Z_i,S_t} & = g^{-1}\left(\mathbf{X}_{Z_i,t} \boldsymbol \beta_{Z_i,S_t} \right)
 \end{split}
-\label{eq:HMMlik}
-\end{equation}
+\end{equation*}
 
@@
 
-\section[Diurnal model]{Incorporating diurnal variation}
+\begin{frame}
+\frametitle{Model complexity (Manx shearwaters, \citex{dean_behavioural_2013})}
+
+<<shearwater>>=
+sheardat <- read.table("shearwater_loglik.dat")
+names(sheardat) <- c("n","loglik")
+npar <- function(n) {
+    ## 2 emissions parameters per state;
+    ## n*(n-1) transition parameters
+    2*n+n*(n-1)
+}
+## don't know how many total data points;
+## ~130 individuals tagged; 10-minute intervals for 2 days = 250 points
+##
+nobs <- 30000
+sheardat2 <- mutate(sheardat,
+       npar=npar(n),
+       dloglik=loglik-min(loglik),
+       AIC=2*npar-2*loglik,
+       dAIC=AIC-min(AIC),
+       BIC=log(nobs)*npar-2*loglik,
+       dBIC=BIC-min(BIC)) %>%
+    select(-c(AIC,BIC)) %>%
+    mutate(type=ifelse(dBIC==0,"min BIC",
+                ifelse(dAIC==0,"min AIC","")),
+           type=factor(type,levels=c("","min AIC","min BIC")))
+
+ggplot(sheardat2,aes(n,dloglik))+
+    geom_line(colour="gray")+
+    geom_point(size=5,aes(colour=type))+
+    xlab("number of latent states")+
+    ylab(expression(Delta~"negative log-likelihood"))+
+    scale_colour_manual(name="",values=c("gray","blue","red"))
+## min BIC ~ 7; min AIC ~ 10
+@ 
+\end{frame}
+
+\section[Diurnal model]{Incorporating diurnal variation \citepx{li_incorporating_2015}}
 \subsection{}
@@
 \begin{frame}
-\frametitle{Comparison}
+\frametitle{Temporal patterns (step length)}
 
-Figure 1 from paper
+<<timeplot,out.width="0.9\\textwidth">>=
+simdf <- simdat %>% dplyr::select(-obs) %>%
+    group_by(time) %>% summarise_each(funs(mean))
+obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
+    filter(!is.na(obs)) %>% group_by(time) %>%
+    summarise_each(funs(mean))
+combdf <- rbind(gather(simdf,model,steplen,-time),
+          data_frame(time=obsdf$time,model="obs",steplen=obsdf$obs))
+mnames <- c("hmm6","hmmblock4","hmmquad5",
+            "hmmsin5","hmmhourly3","obs")
+mnames2 <- c("homogeneous (n=6)","block (n=4)",
+             "quadratic (n=5)","sin (n=5)","hourly (n=3)","observed")
+combdf2 <- combdf %>%
+    filter(model %in% mnames) %>%
+    mutate(model=factor(model,levels=mnames,labels=mnames2))
+colvec <- c(brewer.pal(5,"Dark2"),"black")
+lwdvec <- rep(c(1,4),c(5,1))
+alphavec <- rep(c(1,0.2),c(5,1))
+ggplot(combdf2,aes(x=time,y=steplen,colour=model, lwd=model, alpha=model)) + 
+    geom_line()+ xlab("Time of day") +
+    ylab(expression("Mean step length"~(log[10]*m)))+
+    scale_colour_manual(values=colvec)+
+    scale_alpha_manual(values=alphavec)+
+    scale_size_manual(values=lwdvec)
+@
+
 \end{frame}
@@
 \begin{frame}
-\frametitle{Goodness of fit}
+\frametitle{Temporal patterns (autocorrelation)}
+<<acfplot,cache=TRUE,out.width="0.9\\textwidth">>=
+acffun <- function(sim){
+  a <- acf(sim,na.action=na.pass,plot=FALSE)
+  df <- data.frame(ACF=c(a$acf),lag=a$lag)
+  return(df)
+}
+temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun,.id="model") %>%
+    filter(model %in% mnames) %>%
+    mutate(model=factor(model,levels=mnames,labels=mnames2))
 
-Figure 2 from paper
+ggplot(temp,
+       aes(x=lag,y=ACF,colour=model, lwd=model, alpha=model))+ 
+    geom_line()+ xlab("Time of day") +
+    ylab("Autocorrelation") +
+    scale_colour_manual(values=colvec)+
+    scale_alpha_manual(values=alphavec)+
+    scale_size_manual(values=lwdvec)+
+    geom_hline(yintercept=0,lty=2)
+@
+
 \end{frame}
@@
 \begin{frame}
-\frametitle{Temporal patterns}
-Figures 3 (and 4?) from paper
+\frametitle{Goodness of fit/model complexity}
+<<adj_BIC,out.width="0.9\\textwidth">>=
+(adj_BIC_plot <- ggplot(ss, aes(x=nstates,y=deltaBIC,colour=model))+
+    scale_size_continuous( name="# of parameters") +
+    scale_x_continuous(expand=c(0,0.5),breaks=2:7)+
+    geom_point(aes(size=parameters))+
+    scale_colour_brewer(palette="Dark2")+
+    labs(y=expression(paste(Delta, "BIC")),
+         x="number of latent states")+
+    geom_line() + zmargin)
+@
+
 \end{frame}
@@
 \subsection{}
+
+\begin{frame}
+\frametitle{Animal movement: open challenges}
+\begin{itemize}
+\item Cognition
+\item Intraspecific interaction/collective movement
+\item Continuous-time movement models
+\item Edges, barriers, and corridors
+\item Efficient (big-data) approaches
+\end{itemize}
+\end{frame}
 

### 2016-06-08 guelph_talk/guelph_panther.Rnw
talk cont'd

@@
 % \logo{\pgfuseimage{uflogo}}
-\AtBeginSubsection[]{
+\AtBeginSection[]{
   \frame<beamer>{ 
@@
 
-\end{frame}
-
-\begin{frame}
-\frametitle{To do}
-\begin{itemize}
-\item pix!
-\item general REFS: Turchin, Morales et al, Langrock ... ?
-\end{itemize}
 \end{frame}
@@
 \item three states are identifiable, sensible
-\item ...
+\item dwell distributions approximately geometric \\
+(HSMM $\to$ HMM)
 \end{itemize}
@@
 \item Cognition
-\item Intraspecific interaction/collective movement
-\item Continuous-time movement models
-\item Edges, barriers, and corridors
-\item Efficient (big-data) approaches
+\item Intraspecific interaction/collective movement \citepx{delgado_statistical_2014}
+\item Continuous-time movement models \citepx{calabrese_ctmm_2016}
+\item Edges, barriers, and corridors \citepx{beyer_you_2016}
+\item Efficient (big-data) approaches \citepx{brillinger_three_2008}
+\item Putting it all together ...
 \end{itemize}
@@
   overparameterization
-
+\item protocols and tools for model checking \citepx{potts_generalized_2014};
+score tests?
+\item flexible computational frameworks
 \end{itemize}
@@
 
+\begin{columns}
+\begin{column}{0.5\textwidth}
+\begin{itemize}
+\item ``should I use AIC or BIC? I heard that AIC is inconsistent ...''
+\item complexity penalty = 2 (AIC) vs $\log(n)$ (BIC)
+\item best prediction vs. model identification \citepx{yang_can_2005}
+\item \emph{effect size spectrum}: \\
+tapering or discrete?
+\end{itemize}
+\end{column}
+\begin{column}{0.5\textwidth}
+<<effectsize,echo=FALSE,fig.height=5,fig.width=4>>=
+d <- data.frame(type=rep(c("tapering","discrete"),each=10),
+                n=rep(1:10,2),
+                val=c(0.6^(0:9),rep(c(1,0),c(3,7))))
+ggplot(d,aes(n,val,fill=type))+
+    geom_bar(stat="identity",position="identity",alpha=0.6)+
+    theme(axis.ticks=element_blank(),
+          axis.text.x=element_blank(),
+          axis.text.y=element_blank())+
+    labs(x="effect",y="effect size")+
+    scale_fill_brewer(palette="Set2")
+@ 
+\end{column}
+\end{columns}
 \end{frame}

### 2016-06-08 plotsimtest.R
dwell time plot

@@ -36,3 +36,45 @@ g1 <- (ggplot(ICdat,aes(x=States,y=n,color=type,group=type))
        + xlab("Number of States")
        + ggtitle("Simulation Results: Fitting HMMs to Two-State temphet HMM")
 ) 
+
+#### fit and sim cat1sinhmm5 
+# system("make catsdat.Rout")
+# system("make cat1.df.Rout")
+# load(".cat1.df.RData")
+# source("fitfunctions.R")
+# source("mikesim.R")
+# source("simfunctions.R")
+# fitsin5s <- fitsin(5,cat,seed=2830)
+# simsin5 <- simsin(5,cat,fitsin5s)
+# 
+# dwelldf <- data.frame(hmmsin5obs=simsin5$obs,hmmsin5states=simsin5$states
+#                       ,time = cat$Time, count=1)
+
+load("cat1sinhmm5.RData")
+
+for(i in 1:10283){ # hacks, not bothering with the last "group"
+  j <- 1
+  while(j>0){
+    if(dwelldf[i,2] == dwelldf[i+j,2]){
+      dwelldf[i,4] <- dwelldf[i,4]+1
+      j = j+1}
+    if(dwelldf[i,2] != dwelldf[i+j,2]){
+      j = -1}
+    }
+}
+
+dwelldf2 <- (dwelldf
+  %>% transmute(States = hmmsin5states
+                , time=time
+                , count=count)
+  %>% group_by(States,time)
+  %>% summarise(averageDT=mean(count))
+)
+
+g2 <- (ggplot(dwelldf2,aes(x=time,y=averageDT,group=States,color=factor(States)))
+  + geom_point()
+  + geom_line()
+  + theme_bw()
+)
+
+summary(fitsin5s)

### 2016-06-08 guelph_talk/guelph_panther.Rnw
... more ...

@@
 library(tikzDevice)
+library(plotrix)
 ## http://yihui.name/en/2011/04/produce-authentic-math-formulas-in-r-graphics/
@@
 \begin{description}
-\item[People] \textbf{Michael Li}, \textbf{Madelon van de Kerk}, Dave Onorato, Madan Oli
-\item[Agencies] US Fish and Wildlife, US Geological Survey, US National Park Service
+\item[People] \textbf{Michael Li}, \textbf{Madelon van de Kerk}, \\
+Dave Onorato, Madan Oli
+\item[Agencies] US Fish and Wildlife Service, US Geological Survey, US National Park Service
 \item[Funding] NSERC Discovery grant, NSF IGERT program
@@
 \begin{column}{0.5\textwidth}
-(sequence pix)
+<<arcpic,echo=FALSE,fig.width=3,fig.height=3>>=
+x <- c(0,-0.5,0.5,0)
+y <- c(0,0.5,1,1.5)
+par(mar=c(1,1,1,1),cex=1.5,xpd=NA)
+plot(x,y,axes=FALSE,ann=FALSE,pch=16)
+eps <- 0.04
+arrows(x[-4]+c(-1,1,-1)*eps,y[-4]+eps,x[-1]+c(1,-1,1)*eps,y[-1]-eps,lwd=2)
+draw.arc(x[2],y[2],radius=0.2,angle1=-pi/6,angle2=pi/10,col=4)
+text(x[2]+0.28,y[2],expression(theta[1]),col=4)
+draw.arc(x[3],y[3],radius=0.2,angle1=pi-pi/6,angle2=pi+pi/8,col=4)
+text(x[3]-0.28,y[3],expression(theta[2]),col=4)
+text((x[2]+x[3])/2,(y[2]+y[3])/2,pos=3,expression(s[1]),col=2)
+text((x[3]+x[4])/2,(y[3]+y[4])/2,pos=3,expression(s[2]),col=2)
+posvec <- c(4,2,4,3)
+for(i in 1:4)
+    text(x[i], y[i], substitute(t[j],
+                                 list(j = i)),pos=posvec[i])
+@ 
 \end{column}
@@
 \begin{column}{0.5\textwidth}
-(step length/turning angle pix)
+<<getdat,echo=FALSE,cache=TRUE,warning=FALSE>>=
+load("../allcats.RData")  ## all data
+d1 <- subset(dat,cat==1)
+g1 <- ggplot(d1,aes(x=LogDist))+
+    geom_histogram(binwidth=0.1)+
+    labs(x=expression("step length"~(log[10]*m)))
+g2 <- ggplot(d1,aes(x=Turningangle))+
+    geom_histogram(binwidth=pi/16,alpha=0.2,colour="black",lwd=0.2)+
+    coord_polar()+labs(x="",y="")+
+    scale_x_continuous(breaks=c(-1,-0.5,0,0.5)*pi,
+                       labels=c(180,270,0,90))
+print(g1)
+@
+
+<<g2plot,warning=FALSE>>=
+print(g2)
+@ 
 \end{column}
@@
   \begin{column}{0.5\textwidth}
-    \includegraphics[trim={10cm 0 10cm 0},clip,width=\textwidth]{pix/panther_collar.jpg}
+    \includegraphics[trim={10cm 5cm 10cm 0},clip,width=\textwidth]{pix/panther_collar.jpg}
   \end{column}
@@
 
-\includegraphics[trim={0 6.5cm 0 0}, clip, width=\textwidth]{pix/jae_fig2.png}
+\includegraphics[width=\textwidth]{pix/distpars.pdf}
 
@@
 
-\includegraphics[width=\textwidth]{pix/jae_fig3.png}
+\includegraphics[width=\textwidth]{pix/parplot.pdf}
 
@@
 
-\begin{frame}
-\frametitle{Transition parameters}
+%% \begin{frame}
+%% \frametitle{Transition parameters}
 
-picture/table here of transition parameters (network diagram??)
-\end{frame}
+%% picture/table here of transition parameters (network diagram??)
+%% \end{frame}
 
@@
 \begin{frame}
-\frametitle{Model complexity}
+\frametitle{Model complexity (bad news)}
 
@@
 \begin{frame}
-\frametitle{Broadening the model}
+\frametitle{Expanding the model}
 
@@
 \item extend the model to allow covariates
+\pause
 \item specifically, allow for diurnal variation
@@
 \frametitle{Temporal models}
-
-figure showing alternative temporal models
+<<tempmodel>>=
+bpar <- c(1,0.2,0.8,0.3)
+bseg <- c(10,15,20,24)
+hourpars <- c(0,  0.5, 0.8, 0.9, 0.87, 0.8,
+              0.3,0.4, 0.3, 0.5, 0.4,  0.7,
+              0.9,0.8, 0.7, 0.6, 0.7,  0.6, 0.7,
+              0.5,0.1,0.3,0.4,0.5)
+dd <- data.frame(time=seq(0,to=23.9,length=101)) %>%
+    mutate(sin=((1+sin(2*pi*time/24))/2),
+           quad=1-(time/24)^2,
+           block=ifelse(time<bseg[1],bpar[1],
+                   ifelse(time<bseg[2],bpar[2],
+                      ifelse(time<bseg[3],bpar[3],bpar[4]))),
+           hourly=hourpars[floor(time)+1]) %>%
+    gather(model,val,-time)
+ggplot(dd,aes(time,val,colour=model))+geom_line(size=1.5)+
+    scale_x_continuous(limits=c(0,24),breaks=seq(0,24,by=6))+
+    theme(axis.ticks.y=element_blank(),
+          axis.text.y=element_blank())+
+    labs(y="")
+@ 
 \end{frame}
@@
 \begin{frame}
-\frametitle{Model complexity}
-Figure showing BIC plots for all cats tried
-\end{frame}
-
-\begin{frame}
-\frametitle{Model complexity}
-Simulation results
+\frametitle{Diurnal model: conclusions}
+\begin{itemize}
+\item diurnal structure greatly improves fit ($\Delta \textrm{BIC} \approx 500$)
+\item slightly improves latent-state issue ($n=6 \to 5$)
+\item lots left to do!
+\begin{itemize}
+\item seasonal variation
+\item incorporate habitat, home range behaviour
+\item etc. etc. etc.
+\end{itemize}
+\end{itemize}
 \end{frame}
@@
 \subsection{}
-
-\begin{frame}
-\frametitle{Animal movement: open challenges}
-\begin{itemize}
-\item Cognition
-\item Intraspecific interaction/collective movement \citepx{delgado_statistical_2014}
-\item Continuous-time movement models \citepx{calabrese_ctmm_2016}
-\item Edges, barriers, and corridors \citepx{beyer_you_2016}
-\item Efficient (big-data) approaches \citepx{brillinger_three_2008}
-\item Putting it all together ...
-\end{itemize}
-\end{frame}
 
@@
   overparameterization
-\item protocols and tools for model checking \citepx{potts_generalized_2014};
-score tests?
-\item flexible computational frameworks
+\item Gelman: ``Sample sizes are never large'':
+(\href{http://andrewgelman.com/2005/07/31/n_is_never_larg/}{blog post})
+\begin{quote}
+$N$ is never enough because if it were ``enough'' you'd already be on to the next problem for which you need more data.
+\end{quote}
 \end{itemize}
@@
 \begin{frame}
-\frametitle{an aside on AIC vs BIC}
+\frametitle{An aside on AIC vs BIC}
 
@@
 \begin{frame}
+\frametitle{Animal movement: open challenges}
+\begin{itemize}
+\item Cognition/memory \citepx{bracis_memory_2015}
+\item Intraspecific interaction/collective movement \citepx{delgado_statistical_2014}
+\item Continuous-time movement models \citepx{calabrese_ctmm_2016}
+\item Edges, barriers, and corridors \citepx{beyer_you_2016}
+\item Efficient (big-data) approaches \citepx{brillinger_three_2008}
+\item Putting it all together ...
+\end{itemize}
+\end{frame}
+
+\begin{frame}
 \frametitle{Tools needed}
@@
 \item cross-validation \citepx{wenger_assessing_2012}
-\item diagnostic plots
-\item score tests?
+\item protocols and tools for model checking \citepx{potts_generalized_2014};
+score tests?
+\item flexible computational frameworks \\
+(ecologists can't afford consultants/ \\
+there are too many species out there)
 \end{itemize}

### 2016-06-09 guelph_talk/guelph_panther.Rnw
almost final version?

@@
            model=factor(model,
-              levels=c("HMM","HMM + THsin","HMM + THquad",
+              levels=c("HMM","HMM + THquad","HMM + THsin",
                        "HMM + THblock", "HMM + THhourly"),
-              labels=c("homogeneous","sinusoid","quadratic",
+              labels=c("homogeneous","quadratic","sinusoid",
                        "block","hourly")))
@@
 \begin{frame}
+\frametitle{Model complexity: simulation}
+<<modsim>>=
+mnames3 <- c("hmm","hmmquad",
+            "hmmsin","hmmhourly")
+mnames4 <- c("homogeneous",
+             "quadratic","sin","hourly")
+simdf2 <- readRDS("../simdf.RDS") %>%
+    filter(!type %in% c("fmm","fmmsin")) %>%
+    group_by(type) %>%
+    mutate(dBIC=BIC-min(BIC)) %>% ungroup() %>%
+    mutate(type=factor(type,levels=mnames3,labels=mnames4))
+ggplot(simdf2,aes(states,dBIC,colour=type))+
+    geom_point(size=2)+geom_line()+
+    scale_colour_brewer(palette="Dark2")+
+    labs(y=expression(paste(Delta, "BIC")),
+         x="number of latent states")
+@
+\end{frame}
+
+\begin{frame}
 \frametitle{Diurnal model: conclusions}
@@
 \frametitle{Animal movement: open challenges}
+\begin{columns}
+\begin{column}{0.5\textwidth}
 \begin{itemize}
@@
 \end{itemize}
+\end{column}
+\begin{column}{0.5\textwidth}
+\includegraphics[width=\textwidth]{pix/kitchensink.jpg}
+\end{column}
+\end{columns}
 \end{frame}

### 2016-06-14 guelph_talk/guelph_panther.Rnw
more on guelph talk
gitignore, guelph tweaks

@@
 \usepackage[T1]{fontenc}
-\usepackage[latin1]{inputenc}
+\usepackage[utf8]{inputenc}
 \usepackage{amsmath}
@@
 \item[People] \textbf{Michael Li}, \textbf{Madelon van de Kerk}, \\
-Dave Onorato, Madan Oli
+Dave Onorato, Madan Oli; many unnamed field biologists
 \item[Agencies] US Fish and Wildlife Service, US Geological Survey, US National Park Service
@@
 \begin{frame}
+\includegraphics[width=\textwidth]{pix/panther2.jpg}
+
+\url{http://tinyurl.com/panthermoves}; 
+{\small
+  \url{http://www.slideshare.net/bbolker/model-complexity-and-model-choice-for-animal-movement-models}}
+
+\end{frame}
+
+\begin{frame}
 \begin{multicols}{2}
@@
 % \tiny
-\fontsize{3pt}{5pt}\selectfont
+\fontsize{5pt}{7pt}\selectfont
 \bibliography{../paper}

### 2016-06-15 dataframe.R
testing new pipeline

@@ -1,0 +1,34 @@
+library(depmixS4)
+library(dplyr)
+
+files <- commandArgs(trailingOnly = TRUE)
+catid <- unlist(strsplit(files[2],split="[.]"))[1]
+rawdat <- read.csv(files[1])
+
+block <- function(t){
+  if(t %in% 7:16)return("block2")
+  if(t %in% 17:20)return("block3")
+  return("block1")
+}
+
+## creating suitable dataframe for depmixS4
+rawdf <- (rawdat 
+  %>% rowwise() 
+  %>% transmute(cat = animal_id
+    , Sex = Sex
+    , Time = as.numeric(Time) - 1 ##as.numeric conversion
+    , Distance = Steplength.m.
+    , LogDist = log10(Distance)
+    , Turningangle = Turningangle.rad.
+    , Block = block(Time)
+    )
+)
+
+rawdf$LogDist[is.infinite(rawdf$LogDist)] <- NA
+
+catnum <- c(catid)
+
+dat <- rawdf %>% filter(cat %in% catnum)
+
+
+saveRDS(dat,file=paste("cat",catid,"RDS",sep="."))

### 2016-06-15 fmm.R

@@ -1,0 +1,93 @@
+library(depmixS4)
+library(dplyr)
+library(plyr)
+
+files <- commandArgs(trailingOnly = TRUE)
+catid <- unlist(strsplit(files[1],split="[.]"))[2]
+dat <- readRDS(files[1])
+source(files[2])
+
+fitmix <- function(state,cat,seed=2830){
+  model <- mix(LogDist~1,
+               data=cat,
+               prior=~1,
+               nstate=state,
+               family=gaussian(),
+               initdata=cat)
+  set.seed(seed)
+  fitmodel <- fit(model,emcontrol=em.control(maxit=100))
+  return(fitmodel)
+}
+
+simmix <- function(state,cat,fit){
+  model <- mix(LogDist~1,
+               data=cat,
+               prior=~1,
+               nstate=state,
+               family=gaussian(),
+               initdata=cat)
+  model<-setpars(model,getpars(fit))
+  sim <- simulate(model)
+  df <- data.frame(obs= sim@response[[1]][[1]]@y,states=sim@states)
+  return(df)
+}
+
+sumdf <- function(lst){
+  BIC <- ldply(lst,BIC)
+  nstates <-ldply(lst,nstates)
+  para <- ldply(lst,freepars)
+  model <- c('FMM','FMM','FMM','FMM','FMM')
+  type <- c('FMM','FMM','FMM','FMM','FMM')
+  temp <- data.frame(BICS=BIC$V1,nstates=nstates$V1,parameters=para$V1,model,type)
+  return(temp)
+}
+
+if(catid == 1){
+  fitfmm3s <- fitmix(3,dat,seed=43)
+  fitfmm4s <- fitmix(4,dat,seed=9)
+  fitfmm5s <- fitmix(5,dat,seed=3)
+  fitfmm6s <- fitmix(6,dat,seed=1)
+  fitfmm7s <- fitmix(7,dat,seed=1)
+}
+
+if(catid == 2){
+  fitfmm3s <- fitmix(3,cat,seed=562)
+  fitfmm4s <- fitmix(4,cat,seed=265)
[…]

### 2016-06-15 fmm.R
change iteration and paste error

@@ -15,7 +15,7 @@ fitmix <- function(state,cat,seed=2830){
                family=gaussian(),
                initdata=cat)
   set.seed(seed)
-  fitmodel <- fit(model,emcontrol=em.control(maxit=100))
+  fitmodel <- fit(model,emcontrol=em.control(maxit=500))
   return(fitmodel)
 }
 
@@ -89,5 +89,5 @@ simdf <- data.frame(fmm3S=simmix(3,dat,fitfmm3s)[2]
   , fmm7obs=simmix(7,dat,fitfmm7s)[1]
 )
 
-saveRDS(catsum,file="cat",catid,"fmm","sum","RDS",sep=".")
-saveRDS(simdf,file="cat",catid,"fmm","sim","RDS",sep=".")
+saveRDS(catsum,file=paste("cat",catid,"fmm","sum","RDS",sep="."))
+saveRDS(simdf,file=paste("cat",catid,"fmm","sim","RDS",sep="."))

### 2016-06-15 fmmsin.R
lets run fmmsin

@@ -1,0 +1,68 @@
+library(depmixS4)
+library(dplyr)
+library(plyr)
+
+files <- commandArgs(trailingOnly = TRUE)
+catid <- unlist(strsplit(files[1],split="[.]"))[2]
+dat <- readRDS(files[1])
+source(files[2])
+
+fitmixsin <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
+  model <- mix(LogDist~1,
+               data=cat,
+               prior=~cos(2*pi*Time/24)+ sin(2*pi*Time/24),
+               nstate=state,
+               family=gaussian(),
+               initdata=cat)
+  set.seed(seed)
+  fitmodel <- fit(model,emcontrol=em.control(maxit=500))
+  return(fitmodel)
+}
+
+simmixsin <- function(state,cat,fit){
+  model <- mix(LogDist~1,
+               data=cat,
+               prior=~cos((2*pi*Time)/24)+ sin((2*pi*Time)/24),
+               nstate=state,
+               family=gaussian(),
+               initdata=cat)
+  model<-setpars(model,getpars(fit))
+  sim <- simulate(model)
+  df <- data.frame(obs= sim@response[[1]][[1]]@y,states=sim@states)
+  return(df)
+}
+
+sumdf <- function(lst){
+  BIC <- ldply(lst,BIC)
+  nstates <-ldply(lst,nstates)
+  para <- ldply(lst,freepars)
+  model <- c('FMM + THsin','FMM + THsin','FMM + THsin','FMM + THsin','FMM + THsin')
+  type <- c('FMM','FMM','FMM','FMM','FMM')
+  temp <- data.frame(BICS=BIC$V1,nstates=nstates$V1,parameters=para$V1,model,type)
+  return(temp)
+}
+
+fitfmmsin3 <- fitmixsin(3,dat,seed=3)
+fitfmmsin4 <- fitmixsin(4,dat,seed=1)
+fitfmmsin5 <- fitmixsin(5,dat,seed=1)
+fitfmmsin6 <- fitmixsin(6,dat,seed=1)
+fitfmmsin7 <- fitmixsin(7,dat,seed=1)
+
+fitlist <- list(fitfmmsin3s,fitfmmsin4s,fitfmmsin5s,fitfmmsin6s,fitfmmsin7s)
+
+catsum <- sumdf(fitlist)
[…]

### 2016-06-15 fmmsin.R
testing hmm and fix fmmsin error

@@ -43,11 +43,11 @@ sumdf <- function(lst){
   return(temp)
 }
 
-fitfmmsin3 <- fitmixsin(3,dat,seed=3)
-fitfmmsin4 <- fitmixsin(4,dat,seed=1)
-fitfmmsin5 <- fitmixsin(5,dat,seed=1)
-fitfmmsin6 <- fitmixsin(6,dat,seed=1)
-fitfmmsin7 <- fitmixsin(7,dat,seed=1)
+fitfmmsin3s <- fitmixsin(3,dat,seed=3)
+fitfmmsin4s <- fitmixsin(4,dat,seed=1)
+fitfmmsin5s <- fitmixsin(5,dat,seed=1)
+fitfmmsin6s <- fitmixsin(6,dat,seed=1)
+fitfmmsin7s <- fitmixsin(7,dat,seed=1)
 
 fitlist <- list(fitfmmsin3s,fitfmmsin4s,fitfmmsin5s,fitfmmsin6s,fitfmmsin7s)
 

### 2016-06-15 hmm.R

@@ -1,0 +1,89 @@
+library(depmixS4)
+library(dplyr)
+library(plyr)
+
+files <- commandArgs(trailingOnly = TRUE)
+catid <- unlist(strsplit(files[1],split="[.]"))[2]
+dat <- readRDS(files[1])
+source(files[2])
+
+fithmm <- function(state,cat,seed=NULL){
+  if(!is.null(seed))set.seed(seed)
+  model <- depmix(LogDist~1,
+                  data=cat,
+                  nstate=state,
+                  transition=~1,
+                  family=gaussian())
+  fitmodel <- fit(model,emcontrol=em.control(maxit=iter))
+  return(fitmodel)
+}
+
+simhmm <- function(state,cat,fit){
+  model <- depmix(LogDist~1,
+                  data=cat,
+                  nstate=state,
+                  transition=~1,
+                  family=gaussian())
+  model<-setpars(model,getpars(fit))
+  sim <- simhmm(model)
+  df <- data.frame(obs= sim@response[[1]][[1]]@y,states=sim@states)
+  return(df)
+}
+
+sumdf <- function(lst){
+  BIC <- ldply(lst,BIC)
+  nstates <-ldply(lst,nstates)
+  para <- ldply(lst,freepars)
+  model <- c('HMM','HMM','HMM','HMM','HMM')
+  type <- c('HMM','HMM','HMM','HMM','HMM')
+  temp <- data.frame(BICS=BIC$V1,nstates=nstates$V1,parameters=para$V1,model,type)
+  return(temp)
+}
+if(catid == 1){
+  fithmm3s <- fithmm(3,cat,3)
+  fithmm4s <- fithmm(4,cat,2)
+  fithmm5s <- fithmm(5,cat,1)
+  fithmm6s <- fithmm(6,cat,1)
+  fithmm7s <- fithmm(7,cat,1)
+}
+
+if(catid == 2){
+  fithmm3s <- fithomo(3,cat,2830)
+  fithmm4s <- fithomo(4,cat,2)
+  fithmm5s <- fithomo(5,cat,1)
+  fithmm6s <- fithomo(6,cat,1)
+  fithmm7s <- fithomo(7,cat,1)
+}
+
+if(catid == 14){
+  fithmm3s <- fithmm(3,cat,3030)
+  fithmm4s <- fithmm(4,cat,2)
[…]

### 2016-06-15 hmm.R
name error

@@ -40,35 +40,35 @@ sumdf <- function(lst){
   return(temp)
 }
 if(catid == 1){
-  fithmm3s <- fithmm(3,cat,3)
-  fithmm4s <- fithmm(4,cat,2)
-  fithmm5s <- fithmm(5,cat,1)
-  fithmm6s <- fithmm(6,cat,1)
-  fithmm7s <- fithmm(7,cat,1)
+  fithmm3s <- fithmm(3,dat,3)
+  fithmm4s <- fithmm(4,dat,2)
+  fithmm5s <- fithmm(5,dat,1)
+  fithmm6s <- fithmm(6,dat,1)
+  fithmm7s <- fithmm(7,dat,1)
 }
 
 if(catid == 2){
-  fithmm3s <- fithomo(3,cat,2830)
-  fithmm4s <- fithomo(4,cat,2)
-  fithmm5s <- fithomo(5,cat,1)
-  fithmm6s <- fithomo(6,cat,1)
-  fithmm7s <- fithomo(7,cat,1)
+  fithmm3s <- fithomo(3,dat,2830)
+  fithmm4s <- fithomo(4,dat,2)
+  fithmm5s <- fithomo(5,dat,1)
+  fithmm6s <- fithomo(6,dat,1)
+  fithmm7s <- fithomo(7,dat,1)
 }
 
 if(catid == 14){
-  fithmm3s <- fithmm(3,cat,3030)
-  fithmm4s <- fithmm(4,cat,2)
-  fithmm5s <- fithmm(5,cat,1)
-  fithmm6s <- fithmm(6,cat,1)
-  fithmm7s <- fithmm(7,cat,7)
+  fithmm3s <- fithmm(3,dat,3030)
+  fithmm4s <- fithmm(4,dat,2)
+  fithmm5s <- fithmm(5,dat,1)
+  fithmm6s <- fithmm(6,dat,1)
+  fithmm7s <- fithmm(7,dat,7)
 }
 
 if(catid == 15){
-  fithmm3s <- fithmm(3,cat,3030)
-  fithmm4s <- fithmm(4,cat,2)
-  fithmm5s <- fithmm(5,cat,1)
-  fithmm6s <- fithmm(6,cat,1)
-  fithmm7s <- fithmm(7,cat,1)
+  fithmm3s <- fithmm(3,dat,3030)
+  fithmm4s <- fithmm(4,dat,2)
[…]

### 2016-06-15 hmm.R
iteration error

@@ -14,7 +14,7 @@ fithmm <- function(state,cat,seed=NULL){
                   nstate=state,
                   transition=~1,
                   family=gaussian())
-  fitmodel <- fit(model,emcontrol=em.control(maxit=iter))
+  fitmodel <- fit(model,emcontrol=em.control(maxit=500))
   return(fitmodel)
 }
 

### 2016-06-22 simtest.R
running simulations

@@ -1,4 +1,5 @@
 library(depmixS4)
+library(plyr)
 
 t <- rep(0:23,500)
 y <- rep(1,12000)
@@ -23,98 +24,102 @@ sim <- simhmm(newmod)
 df <- data.frame(obs= sim@response[[1]][[1]]@y,states=sim@states,time=t)
 hist(df$obs)
 
-system.time(simmod1 <- depmix(obs~1
+system.time(hmm2 <- depmix(obs~1
   , data=df
   , transition=~1
   , nstate=2
   , family=gaussian())
 )
 
-system.time(fit1 <- fit(simmod1,verbose=FALSE))
+system.time(hmm2s <- fit(hmm2,verbose=FALSE))
 
-system.time(simmod2 <- depmix(obs~1
+system.time(hmm3 <- depmix(obs~1
   , data=df
   , transition=~1
   , nstate=3
   , family=gaussian())
 )
 
-system.time(fit2 <- fit(simmod2,verbose=FALSE))
+system.time(hmm3s <- fit(hmm3,verbose=FALSE))
 
-system.time(simmod3 <- depmix(obs~1
+system.time(hmm4 <- depmix(obs~1
   , data=df
   , transition=~1
   , nstate=4
   , family=gaussian())
 )
 
-system.time(fit3 <- fit(simmod3,verbose=FALSE))
+system.time(hmm4s <- fit(hmm4,verbose=FALSE))
 
 
 
-system.time(simmod4 <- depmix(obs~1
+system.time(hmm5 <- depmix(obs~1
   , data=df
   , transition=~1
   , nstate=5
   , family=gaussian())
 )
 
-system.time(fit4 <- fit(simmod4,verbose=FALSE))
+system.time(hmm5s <- fit(hmm5,verbose=FALSE))
 
-print(BIC(fit1))
-print(BIC(fit2))
-print(BIC(fit3))
-print(BIC(fit4))
+system.time(hmmsin2 <- depmix(obs~1
+                           , data=df
+                           , transition=~cos((2*pi*t)/24)+ sin((2*pi*t)/24)
+                           , nstate=2
+                           , family=gaussian())
+)
 
-print(summary(fit1))
-print(summary(fit2))
-print(summary(fit3))
-print(summary(fit4))
+system.time(hmmsin2s <- fit(hmmsin2,verbose=FALSE))
 
-dat <- data.frame(seed=seed
-                  , twoS_LL = logLik(fit1)
[…]

### 2016-06-22 paper3.Rnw
working on paper

@@
 \documentclass{article}
-\usepackage{natbib}
+\usepackage[sort]{natbib}
 \usepackage{amsmath}
@@
 Clustering time-series data into discrete groups can improve
-prediction as well as providing insight into the nature of underlying,
+prediction and provide insight into the nature of underlying,
 unobservable states of the system. However, temporal variation in the
-probability that individuals occupy or move between groups can obscure
-such signals. We use finite mixture and hidden Markov models, two
+rates at which individuals move between groups can obscure
+such signals. We use finite mixture and hidden Markov models (HMMs), two
 standard clustering techniques, to model high-resolution hourly
@@
 heterogeneity in transition probabilities, a straightforward but
-rarely explored model extension, resolves some shortcomings of current
-modeling frameworks and clarifies the behavioural patterns of
-panthers. More generally, incorporating previously neglected
-structure in statistical models can allow researchers to more
-accurately identify models of appropriate complexity.
+rarely extension of the standard HMM framework, 
+resolves some shortcomings of current
+models and clarifies the behavioural patterns of
+panthers. More generally, we point out that model misspecification
+(omitting important sources of variation) can lead to overfitting,
+and as a corollary that incorporating previously neglected
+structure in statistical models can allow more
+accurate assessment of appropriate model complexity.
 
@@
 
+\todo{ethics statement? numbers vs words?}
 %% BMB: note that it is *much* easier to keep track of changes in Github,
@@
 
-Given a sequence of animal movements observed directly 
-or via telemetry, movement models aim to find a parsimonious description
-that can be used to understand past movements as well as predict
+Given a sequence of animal movements, movement models aim to find a parsimonious description
+that can be used to understand past movements and predict
 future movements. Ecologists have long considered the effects of
@@
 of organisms' \emph{internal} states; in particular, HMMs model animal
-movement as though individual animals' movement behaviour is determined
+movement as though individual animals' movement behaviour 
+at particular times is determined
 by which of a discrete set of unobserved movement states
@@
 fractions of a year, or longer).  When analyzing such
-remote-sensing data, ecologists will more
+long-term data, ecologists will more
 often have to account for temporal variability in movement
@@
 approximately constant. Changes in latent/hidden behavioural state/mode transition 
+% BMB: pick one (perhaps include a parenthetical statement/definition
+% saying that we treat these as a synonyms.  Just stringing them together
+% with slashes is awkward)
 probabilities based on
@@
 covariates in the HMM \citep{patterson_classifying_2009}, or by more
-\emph{ad hoc} comparisons between inferred states and environmental
+direct comparisons between inferred states and environmental
 conditions \citep{fryxell_multiple_2008}. 
@@
 incorporated a surprisingly large number of hidden
-behavioural states (as many as 6 for individuals with a large amount
-of available data); for practical reasons, they restricted their
-detailed analysis to models with only 3 underlying states.  In
+behavioural states (as many as six for individuals with a large amount
+of available data); for reasons of computational 
+practicality and biological interpretability, they restricted their
+detailed analysis to models with only three underlying states.  In
 contrast, most studies using HMM have chosen the
@@
 \citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
-\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, and as
+\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
+In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
+states, but like \cite{kerk2015hidden} they chose to consider only 
+models with three states (their criterion for choosing ...\todo{check \cite{zhao_knee_2008}!}).
+As \cite{kerk2015hidden} comment, and as
 we discuss further below, behavioural
@@
 Our second goal, therefore, is to explore whether
-\citeauthor{kerk2015hidden}'s results might be driven at
+\citeauthor{kerk2015hidden}'s results on optimal
+model complexity might be driven at
 least in part by structural problems with their statistical
 model, i.e. the
-assumption of temporally homogeneous behaviour.  When large data sets
-are available, information-theoretic model selection methods will
+assumption of temporally homogeneous behaviour.  For large data sets, 
+information-theoretic model selection methods will
 typically choose complex, highly parameterized models; when there is
@@
 distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
-intervals were not calculated), implying that a geometric distribution
+intervals were not given), implying that a geometric distribution
 (i.e., negative binomial with shape=1) might be adequate. In turn,
@@
 the three candidate step-length distributions and the effects
-of considering multivariate (step length plus turning angle)
+of considering multivariate (i.e., step length plus turning angle)
 observations.  However, we do briefly compare log-Normal and Weibull
@@
 treating step length and turning angle at a particular time
-as independent observations while neglecting possible correlations
-between them.)
+as multivariate observations for the purpose of HMM
+analysis but neglecting possible correlations
+between the two measures.)
 
@@
 complicated than any candidate model \citep{BurnhamAnderson1998};
-we suspect that the qualitative conclusions stated here
+we believe that the qualitative conclusions stated here
 for BIC-optimality will carry over to analyses using AIC instead.
@@
 numerically estimating parameters \citep{zucchini_hidden_2009};
-we use those implemented in the \texttt{depmixS4} package for R
+we used those implemented in the \texttt{depmixS4} package for R
 \citep{visser2010depmixs4,R}.
@@
 As is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without loss of generality (i.e.
-we have only $n \times (n-1)$ distinct parameters; we nevertheless index $j$ from 2 to $n$ for notational clarity):
+we have only $n \times (n-1)$ distinct parameters; we index $j$ from 2 to $n$ for notational clarity):
 \begin{equation}

### 2016-06-22 paper3.Rnw
working on paper

@@
 
-\todo{ethics statement? numbers vs words?}
+\todo{ethics statement? numbers vs words? BIC-knee stuff; incorporate \cite{potts_generalized_2014}}
+
 %% BMB: note that it is *much* easier to keep track of changes in Github,
@@
 at successive observations are effectively independent,
-\emph{finite mixture models} (FMM) --- which can be considered
+\emph{finite mixture models} (FMMs) --- which can be considered
 a special case of HMMs where the
@@
 
-We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations).
+We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data (\todo{fill in value} $> ??
+$ observations).
 (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
@@
 because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
-\todo{need to explain here, if we haven't elsewhere, that using the Viterbi algorithm to simulate double-counts the data ...}
+\todo{need to note here that using the Viterbi algorithm to simulate double-counts the data ...}
 
@@
 
-Figure~\ref{fig:BICred_plot} shows the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results.
-Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.\todo{do we need to run some models with 7 states to show that $\dbic$ eventually increases?}
+Figure~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results. 
+\todo{is it \textbf{exactly} consistent? Weibull/vM/homog gets 5, not 6? Can we explain?}
+Furthermore, the BIC-optimal number of states does not depend on the model within the time-homogenous or time-heterogeneous frameworks, but time-heterogeneous models have the BIC-optimal number of states overall.\todo{??
+clarify?
+do you mean time-het models have lower overall BIC (not shown in this graph)?}
 
-<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models",fig.width=8>>=
+<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1.
+\\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
 L <- load("cat1sumdat.RData")
@@
 
-As a complement, we also fitted FMM; FMM with sinusoidal variation in state occupancy probabilities; and HMM with hourly transitions (24-hour multiple block transition model) to compare the temporal effects in goodness of fit.
-Figure~\ref{fig:adj_BIC_comparisons} shows that models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMM models are better than FMM with sinusoidal temporal heterogeneity.
+As a complement, we also fitted FMM and FMM with sinusoidal variation in state occupancy probabilities to compare the temporal effects in goodness of fit.
+\todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
+As a reminder, FMMs assume that the latent state in each time step is 
+\emph{independent} of the latent state at the previous time step; 
+time-varying FMMs can accurately describe movement when behaviour can
+change on a short time scale, but the average propensity for different
+behaviours changes over time.
+
+Figure~\ref{fig:adj_BIC_comparisons} shows that models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity.
 Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
@@
 fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such.
-Opinion on this way of handling obs data (need to include in legend)?})",fig.width=8>>=
+Handle obs data as in Guelph talk.
+Make FMM lines dashed use manual linetype scale)})",fig.width=8>>=
 simdf <- simdat %>% dplyr::select(-obs) %>%
@@
 All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns.
-However, the block model does markedly worse than the other temporal models, and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
+However, the block model does markedly worse than the other temporal models (changing the block definitions might help), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
 
@@
 Like the diurnal pattern shown in Figure~\ref{fig:avg_step_length_by_time},
-the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure~\ref{fig:acf_plot}) shows the need to incorporate temporal heterogeneity in the model even without any formal statistical modeling.
-Because there are an unlimited number of potential complexities that can be added to movement models (e.g.
+the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure~\ref{fig:acf_plot}) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
+Because there are a huge number of potential complexities that can be added to movement models (e.g.
 spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.
@@
 
-<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale}">>=
+<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale; need a real/more complete figure caption here; emphasize that dashes are for emphasis only; rename models (time-homog/heterog)}">>=
 g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
@@
 
+\todo{in figure, maybe replace/also show HMM restricted to 3 states?}
 \todo{maybe do Viterbi plots for mean state occupancy by hour for HMM vs HMM/TH models? Still trying to figure out what's really going on here ...}
-The biggest issue with multiple states is identifying them biologically.
-We really do not know if there Florida panthers really moves according to hidden states, and even if they do, we don't know the true number nor able to observe them.
-Three distinct movement states seems biologically interpretable for Florida panthers according \citet{kerk2015hidden}: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state.
-Long step lengths are easy to capture in all models, where as short step lengths are hard to capture given the GPS errors and unstable movements patterns suggesting multiple states associate with short step lengths.
-Thus, lower number of short step length states with higher variance can be equivalent or more realistic than multiple distinguishable short step length states with low variance.
+
+The hardest problem with multiple latent states is interpreting them
+biologically. We have no way of knowing what panthers are actually
+thinking (it is certainly more complex than being in one of a small
+number of discrete latent states); 
+we don't know the ``true'' number of latent states,
+nor are we able to observe them directly, although
+incorporating additional direct observations of 
+behaviour (if available) can at least partially
+address this problem \citep{fryxell_multiple_2008}.
+Three distinct
+movement states seem biologically interpretable for Florida panthers
+according to \citet{kerk2015hidden}: Short step length suggests resting
+states, intermediate step length a foraging state, and long
+step length a traveling state. 
+\todo{do we need this?}
+Figure~\ref{state_plots} compares the estimated parameter values
+for several cats
+(mean and standard deviation of the step length in each state) between
+the time-homogeneous and time-heterogeneous models. 
+In general, the states with longer mean step lengths are relatively
+similar between model classes. For cats 14 and 15, the states with
+the longest or next-longest mean step lengths have similar means
+and standard deviations; for cats 1 and 2, three long-step states
+in the homogeneous HMM appear to divide two long-step states in 
+the heterogeneous HMM. For short-step states, the heterogeneous HMM
+tends to identify a high-variance state, while the homogeneous HMM
+picks up states with very short step lengths (questionable in any
+case because we have not taken any special efforts to account for
+GPS error).
+
+Thus,
+lower number of short step length states with higher variance can be
+equivalent or more realistic than multiple distinguishable short step
+length states with low variance.
+\todo{clarify this/do we still need this??}
 
@@
 
-Identifying behavioural states based on some set of observations is a common methodological problem in behavioural ecology; finding solutions that are both biologically and statistically interpretable and defensible is challenging.
-Models that suggest a higher number of states fit the data well but are hard to interpret biologically; models that are biologically interpretable are not statistically supported.
-While the HMM framework has been applied in many research areas, very few studies have considered modeling with more than 3 states.
-HMM is a simple straight forward framework, but HMM extensions of adding a small amount of complexity can complicate the model greatly and reveal deeper understanding.
-For example, \citet{kerk2015hidden} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitat distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.
+Identifying behavioural states based on observations is a
+common methodological problem in behavioural ecology; finding
+solutions that are both biologically and statistically interpretable
+and defensible is challenging. Models that suggest a higher number of
+states fit the data well but are hard to interpret biologically,
+while 
+models that are biologically interpretable are not always statistically
+supported. While the HMM framework has been applied in many research
+areas, very few studies have considered modeling with more than 3
+states. HMM is a simple, straightforward framework, but HMM extensions
+of adding a small amount of complexity can complicate the model
+greatly and reveal deeper understanding. For example,
+\citet{kerk2015hidden} used 2 to 6 movement states HSMM to model
+Florida panthers, \citet{morales_extracting_2004} used habitat
+distance dependent transition HMM to model elk, and outside of
+ecology, \citet{raffa_multivariate_2015} used mixed effect transition
+and emission HMM in a smoking study.
 
-We have presented a relatively simple concept yet little-studied extension by covariate dependent HMM can overfit the number of true states in covariate dependent HMM (in our case, time-dependent transitions).
-HMM with more than three states are not biologically interpretable and often discarded, but we cannot ignore that increasing complexity (number of states) states can increase goodness of fit statistically.
-Our analyses revealed modeling temporal heterogeneity in HMM transition can reduce the number of states estimated in time homogeneous transitions HMM movement of Florida panthers and reduce the overall model complexity.
+We have presented a relatively simple concept yet little-studied
+extension by covariate dependent HMM can overfit the number of true
+states in covariate dependent HMM (in our case, time-dependent
+transitions). HMM with more than three states are not biologically
+interpretable and often discarded, but we cannot ignore that
+increasing complexity (number of states) states can increase goodness
+of fit statistically. Our analyses revealed modeling temporal
+heterogeneity in HMM transition can reduce the number of states
+estimated in time homogeneous transitions HMM movement of Florida
+panthers and reduce the overall model complexity.
 
-For some kinds of model predictions, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of predicting the states by the Viterbi algorithm (the tradition state prediction method) and can easily overlook more complicated models.
-The Viterbi algorithm solves/predicts for the most likely/probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}.
-This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations.
-It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it cannot predict future movements beyond the observation sequence.
-In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.
+For some kinds of model predictions, the time homogeneous transition
+model can perform as good as the time inhomogeneous transition models
+in terms of predicting the states by the Viterbi algorithm (the
+tradition state prediction method) and can easily overlook more
+complicated models. The Viterbi algorithm solves/predicts for the most
+likely/probable sequence of movement states based on the observations
+\citep{zucchini_hidden_2009,langrock_flexible_2012}. This approach
+allows the time homogeneous transition model's prediction to capture
+the temporal heterogeneity from the observations. It is useful to
+predict missing data in the observation sequence, but because it is
+conditional on the observed values, it cannot predict future movements
+beyond the observation sequence. In general, the time homogeneous
+transition model gives the broader scope of movement pattern where as
+the time inhomogeneous transitions models gives a more realistic
+movement structure and future predictions.
 
-Although the time inhomogeneous transitions offer an
-objective approach to (1) reduce the number of states to a biologically interpretable level, (2) captures the observational autocorrelation and temporal pattern, (3) reduce the overall complexity as illustrated by our study, several challenges
-must be overcome in order to maximize its potential.
-The current model is based on the reduced univariate step length response emission without fitting turning angles.
-Although turning angles are strongly correlated with step length, it does not effect the overall prediction on number of states.
-But movement trajectories predictions, and spatial models still require both step length and turning angles.
-Furthermore, modelling temporal heterogeneity only reduce the number of predicted states by 1 resulting in 4/5 states might suggest (1) other factors such as spatial components that might also contribute to over predicting number of states, or (2) there really are 1 or 2 more different movement characteristics that are biologically unknown.
-Despite the challenges, our study provides an important extension in HMM modeling that is not in the seen before and a real life application that it exist and applicable in animal movement ecology.
+Although the time inhomogeneous transitions offer an objective
+approach to (1) reduce the number of states to a biologically
+interpretable level, (2) captures the observational autocorrelation
+and temporal pattern, (3) reduce the overall complexity as illustrated
+by our study, several challenges must be overcome in order to maximize
+its potential. The current model is based on the reduced univariate
+step length response emission without fitting turning angles. Although
+turning angles are strongly correlated with step length, it does not
+effect the overall prediction on number of states. But movement
+trajectories predictions, and spatial models still require both step
+length and turning angles. Furthermore, modelling temporal
+heterogeneity only reduce the number of predicted states by 1
+resulting in 4/5 states might suggest (1) other factors such as
+spatial components that might also contribute to over predicting
+number of states, or (2) there really are 1 or 2 more different
+movement characteristics that are biologically unknown. Despite the
+challenges, our study provides an important extension in HMM modeling
+that is not in the seen before and a real life application that it
+exist and applicable in animal movement ecology.
 
@@
             distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
-            turn_angle=ifelse(grepl("VM",model),"yes","no"),
+            turn_angle=ifelse(grepl("VM",model),"step length+\nturning angle",
+                              "turning angle only"),
             homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
 ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
-    geom_line()+
+    geom_line(aes(size=distrib))+
     facet_wrap(~ homog)+
@@
     #    geom_point(aes(size=parameters))+
-    scale_linetype(name="von Mises\nturning angle?")+
+    ## scale_linetype(name="von Mises\nturning angle?")+
+    scale_linetype(name="model")+
+    scale_size_manual(name="step length\ndistribution",values=c(0.75,1.5))+
     labs(y=expression(Delta*"BIC"),

### 2016-06-22 sim_results/plotsimtest.R
finish simluation results

@@ -1,0 +1,80 @@
+library(ggplot2)
+library(reshape2)
+library(dplyr)
+
+simfiles <- list.files(pattern="RDS")
+
+BICdat <- data.frame()
+for(i in simfiles){
+  seednum <-unlist(strsplit(i,split="[.]"))[2]
+  tempdat <- readRDS(i)
+  BICdf <- tempdat %>% mutate(seed=seednum) %>% group_by(model) %>% slice(which.min(BIC))
+  BICdat <- rbind(BICdat,BICdf)
+}
+
+AICdat <- data.frame()
+for(i in simfiles){
+  seednum <-unlist(strsplit(i,split="[.]"))[2]
+  tempdat <- readRDS(i)
+  AICdf <- tempdat %>% mutate(seed=seednum) %>% group_by(model) %>% slice(which.min(AIC))
+  AICdat <- rbind(AICdat,AICdf)
+}
+
+AICdat2 <- AICdat %>% ungroup() %>% select(-c(LL,AIC,BIC,parameters)) %>% group_by(model) %>% count(nstates,model) %>% mutate(IC = "AIC")
+BICdat2 <- BICdat %>% ungroup() %>% select(-c(LL,AIC,BIC,parameters)) %>% group_by(model) %>% count(nstates,model) %>% mutate(IC = "BIC")
+
+ICdat <- rbind(AICdat2,BICdat2)
+
+g1 <- (ggplot(ICdat,aes(x=nstates,y=n,color=model,group=model))
+       + geom_point()
+       + geom_line()
+       + theme_bw()
+       + facet_grid(.~IC)
+       + ylab("Counts")
+       + xlab("Number of States")
+       + ggtitle("Simulation Results: Fitting HMMs to Two-State temphet HMM")
+) 
+
+g1
+# 
+# #### fit and sim cat1sinhmm5 
+# # system("make catsdat.Rout")
+# # system("make cat1.df.Rout")
+# # load(".cat1.df.RData")
+# # source("fitfunctions.R")
+# # source("mikesim.R")
+# # source("simfunctions.R")
+# # fitsin5s <- fitsin(5,cat,seed=2830)
+# # simsin5 <- simsin(5,cat,fitsin5s)
+# # 
+# # dwelldf <- data.frame(hmmsin5obs=simsin5$obs,hmmsin5states=simsin5$states
+# #                       ,time = cat$Time, count=1)
+# 
+# load("cat1sinhmm5.RData")
+# 
[…]

### 2016-07-05 paper3.Rnw
temp push

@@
 increased from 
-3 to 6 states,
-suggesting that the 6-state model was
+three to six states,
+suggesting that the six-state model was
 favoured statistically; however,
-the authors used 3-state models in most
+the authors used three-state models in most
 of their analyses for ease of biological interpretation.
@@
 We should probably add a short section; if don't, then we
-should delete following sentence}
+should delete following sentence }
 Using BIC also simplifies 
@@
 
-\todo{maybe give the formula for number of parameters (constrained) for each model type here?}
+\todo{maybe give the formula for number of parameters (constrained) for each model type here?...ML: Sum(I+T+E), I=Prior=n(s-1) where n is the number of parameters for the prior complexity, T = ns(s-1), E=2s}
 
@@
 
-We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data (\todo{fill in value} $> ??
-$ observations).
+We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($> 9000 $ observations).
 (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
@@
 $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
-L <- load("cat1allmod.RData")
-df2 <- transform(allmodsumdf,
-            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
-            turn_angle=ifelse(grepl("VM",model),"yes","no"),
-            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
+#L <- load("cat1allmod.RData")
+allmodsumdf2 <- readRDS("cat.1.df.RDS")$multimoddf
+df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
+            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
+            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
+            homog=ifelse(grepl("TH",type),"time-heterogeneous HMM","homogeneous HMM"))
 ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
@@
 
-Figure~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results. 
+Figure~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles)
 \todo{is it \textbf{exactly} consistent? Weibull/vM/homog gets 5, not 6? Can we explain?}
-Furthermore, the BIC-optimal number of states does not depend on the model within the time-homogenous or time-heterogeneous frameworks, but time-heterogeneous models have the BIC-optimal number of states overall.\todo{??
-clarify?
-do you mean time-het models have lower overall BIC (not shown in this graph)?}
 
@@
 \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
-L <- load("cat1sumdat.RData")
-(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+
+#L <- load("cat1sumdat.RData")
+sumdat <- transform(readRDS("cat.1.df.RDS")$HMMsumdf,
+                    mod = ifelse(grepl("HMM",model),"no","yes")
+)
+linetypev <- c(rep(2,2),rep(1,5))
+(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model,linetype=model))+
     facet_wrap(~ type,ncol=4)+
@@
     geom_point(aes(size=parameters))+
+    scale_linetype_manual(values=linetypev)+
     scale_colour_brewer(palette="Dark2")+

### 2016-07-05 models/weibullVMsin.R

@@ -460,8 +460,8 @@ sumdf <- function(lst){
   BIC <- ldply(lst,BIC)
   nstates <-ldply(lst,nstates)
   para <- ldply(lst,freepars)
-  model <- c('HMM WBVM','HMM WBVM','HMM WBVM','HMM WBVM','HMM WBVM')
-  type <- c('HMM','HMM','HMM','HMM','HMM')
+  model <- c('HMM WBVMsin','HMM WBVMsin','HMM WBVMsin','HMM WBVMsin','HMM WBVMsin')
+  type <- c('HMM + TH','HMM + TH','HMM + TH','HMM + TH','HMM + TH')
   temp <- data.frame(BICS=BIC$V1,nstates=nstates$V1,parameters=para$V1,model,type)
   return(temp)
 }

### 2016-07-05 paper3.Rnw
main paper pictures are done, still need to fix the appendix pictures

@@
 library("gridExtra")
+library("RColorBrewer")
 if (!require("ggalt")) {
@@
 }
+# library("ggalt")
 @
@@
 #L <- load("cat1allmod.RData")
-allmodsumdf2 <- readRDS("cat.1.df.RDS")$multimoddf
+allmodsumdf2 <- readRDS("summary_stats/cat.1.df.RDS")$multimoddf
 df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
@@
 #L <- load("cat1sumdat.RData")
-sumdat <- transform(readRDS("cat.1.df.RDS")$HMMsumdf,
+sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
                     mod = ifelse(grepl("HMM",model),"no","yes")
@@
 linetypev <- c(rep(2,2),rep(1,5))
-(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model,linetype=model))+
-    facet_wrap(~ type,ncol=4)+
-    scale_size_continuous( name="# of parameters") +
-    scale_x_continuous(expand=c(0,0.5),breaks=2:7)+
-    geom_point(aes(size=parameters))+
-    scale_linetype_manual(values=linetypev)+
-    scale_colour_brewer(palette="Dark2")+
-    labs(y=expression(paste(Delta, "BIC")),
-         x="Number of States") +
-    geom_line() + zmargin)
+(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
+                                    , y=deltaBIC
+                                    , colour=model
+                                    , linetype=model))
+  + facet_wrap(~ type,ncol=4)
+  + scale_size_continuous( name="# of parameters")
+  + scale_x_continuous(expand=c(0,0.5),breaks=2:7)
+  + geom_point(aes(size=parameters))
+  + scale_linetype_manual(values=linetypev)
+  + scale_colour_brewer(palette="Dark2")
+  + labs(y=expression(paste(Delta, "BIC"))
+         , x="Number of States") 
+  + geom_line() 
+  + zmargin)
 @
@@
 Make FMM lines dashed use manual linetype scale)})",fig.width=8>>=
-simdf <- simdat %>% dplyr::select(-obs) %>%
-    group_by(time) %>% summarise_each(funs(mean))
-obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
-    filter(!is.na(obs)) %>% group_by(time) %>%
-    summarise_each(funs(mean))
+# simdf <- simdat %>% dplyr::select(-obs) %>%
+#     group_by(time) %>% summarise_each(funs(mean))
+# obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
+#     filter(!is.na(obs)) %>% group_by(time) %>%
+#     summarise_each(funs(mean))
+# 
+# avgdf2 <- melt(simdf,'time')
+avgdf <- readRDS("summary_stats/cat.1.df.RDS")$avgplotdf
+obsdf <- (readRDS("summary_stats/cat.1.RDS") 
+  %>% dplyr::select(c(LogDist,Time)) 
+  %>% filter(!is.na(LogDist)) 
+  %>% group_by(Time) 
+  %>% summarise_each(funs(mean))
+  %>% transmute(time = Time, variable = "obs", value=LogDist)
+)
 
-avgdf2 <- melt(simdf,'time')
-ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
-    geom_point() +geom_line()+ xlab("Time of Day") +
-    ylab("Step lengths (log10)")+
-    scale_x_continuous(breaks=0:23)+
-    geom_line(data=obsdf,
-              aes(y=obs),lwd=2.5,alpha=0.3,colour="black")
+mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
+            "hmmquad5","hmmsin5","obs")
+mnames2 <- c("fmm-homogeneous  (n=5)"
+             , "fmm-sin (n=4)"
+             , "homogeneous (n=6)"
+             , "block (n=4)"
+             , "hourly (n=3)"
+             , "quadratic (n=5)"
+             , "sin (n=5)"
+             , "observed")
+
+combdf <- (rbind(avgdf,obsdf)
+           %>% transmute(time=time,model=variable,steplen=value)
+           %>% filter(model %in% mnames)
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+)
+linetypev <- rep(c(2,1),c(2,6))
+colvec <- c(brewer.pal(7,"Dark2"),"black")
+lwdvec <- rep(c(1,4),c(7,1))
+alphavec <- rep(c(1,0.2),c(7,1))
+(ggplot(combdf,aes(x=time,y=steplen,colour=model,lwd=model,alpha=model,linetype=model)) 
+  + geom_line()
+  + xlab("Time of Day") 
+  + ylab(expression('Mean step length'~(log[10]*m)))
+  + scale_color_manual(values=colvec)
+  + scale_alpha_manual(values=alphavec)
+  + scale_size_manual(values=lwdvec)
+  + scale_linetype_manual(values=linetypev)
+)
 @
@@
 }
-temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)
-  ggplot(temp,aes(lag,ACF,colour=.id))+
-    geom_point()+geom_line()
+# temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)
+
+acfdf <- (readRDS("summary_stats/cat.1.df.RDS")$acfdf 
+          %>% mutate(model = .id)
+)
+(ggplot(acfdf,aes(x=lag,y=ACF,colour=model,lwd=model,alpha=model,linetype=model)) 
+  + geom_line()
+  + xlab("Lag") 
+  + ylab("Autocorrelation")
+  + scale_color_manual(values=colvec)
+  + scale_alpha_manual(values=alphavec)
+  + scale_size_manual(values=lwdvec)
+  + scale_linetype_manual(values=linetypev)
+)
 @

### 2016-07-07 paper3.Rnw
included viterbi inside the plots

@@
 sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
-                    mod = ifelse(grepl("HMM",model),"no","yes")
+                    mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
 )
 linetypev <- c(rep(2,2),rep(1,5))
-(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
+(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                     , y=deltaBIC
                                     , colour=model
-                                    , linetype=model))
-  + facet_wrap(~ type,ncol=4)
-  + scale_size_continuous( name="# of parameters")
-  + scale_x_continuous(expand=c(0,0.5),breaks=2:7)
-  + geom_point(aes(size=parameters))
+                                    , linetype=model
+                                    , group=model))
+  + facet_wrap(~mod,ncol = 2,scales = "free_x")
+  + scale_size_continuous(name="# of states")
+  + scale_x_continuous(breaks=seq(0,300,by=10))
+  + geom_point(aes(shape=factor(nstates),size=1.5,alpha=nstates))
   + scale_linetype_manual(values=linetypev)
@@
 )
+vit <- readRDS("summary_stats/cat.1.vit.RDS")
+vitdf <- (vit %>% dplyr::select(c(viterbi,time))
+  %>% group_by(time)
+  %>% summarise_each(funs(mean))
+  %>% mutate(model="homogeneous (n=3)")
+)
 
@@
 )
-linetypev <- rep(c(2,1),c(2,6))
-colvec <- c(brewer.pal(7,"Dark2"),"black")
-lwdvec <- rep(c(1,4),c(7,1))
-alphavec <- rep(c(1,0.2),c(7,1))
+linetypev <- c(rep(c(2,1),c(2,6)),1)
+colvec <- c(c(brewer.pal(7,"Dark2"),"black"),"black")
+lwdvec <- c(rep(c(1,4),c(7,1)),4)
+alphavec <- c(rep(c(1,0.2),c(7,1)),1)
 (ggplot(combdf,aes(x=time,y=steplen,colour=model,lwd=model,alpha=model,linetype=model)) 
@@
   + scale_linetype_manual(values=linetypev)
-)
+) + geom_point(data=vitdf,aes(y=viterbi),lwd=2.5,alpha=1,colour="black",shape=118)
 @
@@
 }
-# temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)
+vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
+)
 
@@
   + scale_linetype_manual(values=linetypev)
-)
+) + geom_point(data=vitacf,lwd=2.5,alpha=1,colour="black",shape=118)
 @

### 2016-07-07 paper3.Rnw
failing to compile, I don't know why

@@
 library("RColorBrewer")
-if (!require("ggalt")) {
-    cat("installing ggalt from BB repo ...\n")
-    library("devtools")
-    install_github("bbolker/ggalt")
-    library("ggalt")
-}
+# if (!require("ggalt")) {
+#     cat("installing ggalt from BB repo ...\n")
+#     library("devtools")
+#     install_github("bbolker/ggalt")
+#     library("ggalt")
+# }
 # library("ggalt")
@@
 
+Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
+The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components:
+
 We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
@@
 Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines\todo{citation?}, could be useful directions for future exploration.
-
-Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
 
@@
 <<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale; need a real/more complete figure caption here; emphasize that dashes are for emphasis only; rename models (time-homog/heterog)}">>=
-g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
-    geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
-(g0c <- g0 +
-    geom_encircle(data=subset(all_cat_pars,
-                              (catID=="15" & log_mean<1.5) |
-                              (catID!="15" & log_mean<1.8)),
-                  colour="black",linetype=2,expand=0.05))
+# g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
+#     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
+# (g0c <- g0 +
+#     geom_encircle(data=subset(all_cat_pars,
+#                               (catID=="15" & log_mean<1.5) |
+#                               (catID!="15" & log_mean<1.8)),
+#                   colour="black",linetype=2,expand=0.05))
 @
@@
 
-<<viterbi, warning=FALSE, message=FALSE, echo=FALSE >>=
-load("cat1.RData")
-load("cat1sumdat.RData")
-vitcat <- function(catt,mod){
-  params <- tail(getpars(mod),6)
-  vitdist <- numeric(ntimes(mod))
-  states <- viterbi(mod)$state
-  for(i in 1:ntimes(mod)){
-    if(states[i] == 1){
-      vitdist[i] <- rnorm(1,params[1],params[2])
-    }
-    if(states[i] == 2){
-      vitdist[i] <- rnorm(1,params[3],params[4])
-    }
-    if(states[i] == 3){
-      vitdist[i] <- rnorm(1,params[5],params[6])
-    }
-  }
-  dat <- data.frame(viterbi=vitdist,time=catt$Time, obs=catt$LogDist)
-  return(dat)
-}
-
-temp <- vitcat(cat,fithomo3s) %>% dplyr::select(-time)%>%ldply(.,acffun)
-ggplot(temp,aes(lag,ACF,colour=.id))+
-geom_point()+geom_line() +theme_bw() +
-ggtitle("ACF plot of BIC Selected Models")
-@
-
 \todo{does it make any sense to compress figs~\ref{adj_BIC-comparisons_cat2} etc.

### 2016-07-11 paper3.Rnw
lots of change, almost done

@@
 
-Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
-The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components:
-
 We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
@@
 
+Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
+The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components.
+Let $n$ be the number of hidden states and $k_{I}, k_{t}, k,{e}$ be the parameters of the prior distribution, transition function and emission distributions.
+Then the number of free parameters of an HMM is:
+
+K = $$ k_{i} * (n-1) + k_{t}*n*(n-1) + k_{e}*n $$
+
 \subsection{Model evaluation}
@@
 captures serial correlation at both short and long scales.
+
+\subsection{BICS}
+Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
+The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components.
+Let $n$ be the number of hidden states and $k_{I}, k_{t}, k,{e}$ be the parameters of the prior distribution, transition function and emission distributions.
+Then the number of free parameters of an HMM is:
+
+K = $$ k_{i} * (n-1) + k_{t}*n*(n-1) + k_{e}*n $$ ...
+
+\subsection{Comparing observed data against simulation and Viterbi predictions}
+
 We used simulations to predict hourly step length and ACF 
@@
 sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
-                    mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
+                    mod = ifelse(grepl("HMM",model),"no","yes")
 )
 linetypev <- c(rep(2,2),rep(1,5))
-(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
+(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                     , y=deltaBIC
                                     , colour=model
-                                    , linetype=model
-                                    , group=model))
-  + facet_wrap(~mod,ncol = 2,scales = "free_x")
-  + scale_size_continuous(name="# of states")
-  + scale_x_continuous(breaks=seq(0,300,by=10))
-  + geom_point(aes(shape=factor(nstates),size=1.5,alpha=nstates))
+                                    , linetype=model))
++ facet_wrap(~ type,ncol=4)
++ scale_size_continuous( name="# of parameters")
++ scale_x_continuous(expand=c(0,0.5),breaks=2:7)
++ geom_point(aes(size=parameters))
 + scale_linetype_manual(values=linetypev)
@@
 + zmargin)
+
+### I want this picture but it is too big or something
+
+# sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
+#                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
+# )
+# linetypev <- c(rep(2,2),rep(1,5))
+# (adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
+#                                     , y=deltaBIC
+#                                     , colour=model
+#                                     , linetype=model
+#                                     , group=model))
+#   + facet_wrap(~mod,ncol = 2,scales = "free_x")
+#   + scale_size_continuous(name="# of states")
+#   + scale_x_continuous(breaks=seq(0,300,by=10))
+#   + geom_point(aes(shape=factor(nstates),size=1.5,alpha=nstates))
+#   + scale_linetype_manual(values=linetypev)
+#   + scale_colour_brewer(palette="Dark2")
+#   + labs(y=expression(paste(Delta, "BIC"))
+#          , x="Number of States") 
+#   + geom_line()
+#   + zmargin)
 @

### 2016-07-11 paper3.Rnw
done for now

@@
 \begin{description}
-\item[Hourly model] \todo{describe hourly model.
-Explain that it's included for comparative purposes, i.e.
-not really practical.
-Explain that we only fitted up to 4 states (because too many parameters)}
 \item[Multiple block transition] Here we assume piecewise-constant transition probabilities.
@@
 $$
+
+\item[Hourly model] Lastly, we extended the multi-block approach and assign  a different transition matrix for every hour of the day.
+This model is included for comparative purposes due to the large number of parameters in the model which makes it not really practical.
+We only fitted up to four states using the hourly model.
 \end{description}
@@
 Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines\todo{citation?}, could be useful directions for future exploration.
-
-Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
-The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components.
-Let $n$ be the number of hidden states and $k_{I}, k_{t}, k,{e}$ be the parameters of the prior distribution, transition function and emission distributions.
-Then the number of free parameters of an HMM is:
-
-K = $$ k_{i} * (n-1) + k_{t}*n*(n-1) + k_{e}*n $$
 
@@
 
-\subsection{BICS}
-Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).
-The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components.
-Let $n$ be the number of hidden states and $k_{I}, k_{t}, k,{e}$ be the parameters of the prior distribution, transition function and emission distributions.
+Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity.
+The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components.
+Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the parameters of the prior distribution, transition function and emission distributions.
 Then the number of free parameters of an HMM is:
 
-K = $$ k_{i} * (n-1) + k_{t}*n*(n-1) + k_{e}*n $$ ...
+\begin{equation}
+\underbrace{k_{i}*(n-1)}_{\text{Initial}} 
+{+} \underbrace{k_{t}*n*(n-1)}_{\text{Transition}}
+{+} \underbrace{ k_{e}*n}_{\text{Emission}}
+\label{eq:numparam}
+\end{equation}
 
-\subsection{Comparing observed data against simulation and Viterbi predictions}
+When increasing in number of states, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas, HMMs with temporal heterogenity will not by Eq.~\ref{eq:numparam}.
+When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
 
@@
 because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
-\todo{need to note here that using the Viterbi algorithm to simulate double-counts the data ...}
+We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions \emph{double count} the observed data.
+The Viterbi algorithm solves/predicts for the most probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}.
+It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not relibly evaluate goodness of fit for the different structral complexities of HMM models.
 
 \section{Results}
+
+<<sim_results, warning=FALSE,echo=FALSE,message=FALSE>>=
+ICdat <- readRDS("./sim_results/ICdat.rds")
+aa <- readRDS("./sim_results/sim.101.RDS")
+
+ICdat <- mutate(ICdat,
+                model=factor(model,
+                         levels=c("HMM","HMMsin"),
+                         labels=c("homogeneous","heterogeneous"))) %>% filter(IC=="BIC")
+simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,
+                            linetype=IC,shape=IC))+
+                        geom_point(size=3) +
+                        geom_line() +
+                        scale_colour_brewer(palette="Set1")+
+                        labs(y="counts",x="# latent states")
+print(simplot)
+
+@
+
+We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.
+Figure~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can over estiate the number of states via BIC-optimal approach.
 
@@
             turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
-            homog=ifelse(grepl("TH",type),"time-heterogeneous HMM","homogeneous HMM"))
+            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
 ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
@@
 
-As a complement, we also fitted FMM and FMM with sinusoidal variation in state occupancy probabilities to compare the temporal effects in goodness of fit.
+As a complement, we also fitted FMM and FMM with sinusoidal variation in state occupancy probabilities to compare the temporal effects in goodness of fit (dashed lines).
 \todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
@@
 However, the block model does markedly worse than the other temporal models (changing the block definitions might help), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
+We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points).
 
@@
 
-Thus,
-lower number of short step length states with higher variance can be
-equivalent or more realistic than multiple distinguishable short step
-length states with low variance.
-\todo{clarify this/do we still need this??}
-
 \section{Discussion}
@@
 tradition state prediction method) and can easily overlook more
-complicated models. The Viterbi algorithm solves/predicts for the most
-likely/probable sequence of movement states based on the observations
-\citep{zucchini_hidden_2009,langrock_flexible_2012}. This approach
-allows the time homogeneous transition model's prediction to capture
-the temporal heterogeneity from the observations. It is useful to
-predict missing data in the observation sequence, but because it is
-conditional on the observed values, it cannot predict future movements
-beyond the observation sequence. In general, the time homogeneous
+complicated models.
+This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations.
+The limitation of the Viterbi appoach is that it cannot predict future movements beyond the observation sequence.
+In general, the time homogeneous
 transition model gives the broader scope of movement pattern where as

### 2016-07-11 paper3.Rnw
Figure num problem fix

@@
 
-<<sim_results, warning=FALSE,echo=FALSE,message=FALSE>>=
+<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Simulation test for BIC-optimal approach",fig.width=6,fig.height=4>>=
 ICdat <- readRDS("./sim_results/ICdat.rds")

### 2016-07-12 paper3.Rnw
working on paper, leftover isec talk changes

@@
 opts_chunk$set(fig.width=6,fig.height=4)
+## fix complaints about "incompatible color definition" from xcolor ...
+## http://tex.stackexchange.com/questions/148188/knitr-xcolor-incompatible-color-definition
+knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
 library("depmixS4")
@@
 # }
-# library("ggalt")
+library("ggalt")
 @
@@
 heterogeneity in transition probabilities, a straightforward but
-rarely extension of the standard HMM framework, 
+little-used extension of the standard HMM framework, 
 resolves some shortcomings of current
@@
 
-Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines\todo{citation?}, could be useful directions for future exploration.
+Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could be useful directions for future exploration.
 
@@
 The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components.
-Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the parameters of the prior distribution, transition function and emission distributions.
+Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$.
 Then the number of free parameters of an HMM is:
@@
 \begin{equation}
-\underbrace{k_{i}*(n-1)}_{\text{Initial}} 
-{+} \underbrace{k_{t}*n*(n-1)}_{\text{Transition}}
-{+} \underbrace{ k_{e}*n}_{\text{Emission}}
+\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
+{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
+{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
 \label{eq:numparam}
@@
 
-When increasing in number of states, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas, HMMs with temporal heterogenity will not by Eq.~\ref{eq:numparam}.
+As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas HMMs with temporal heterogenity will increase quadratically (Eq.~\ref{eq:numparam}).
 When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
@@
 because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
-We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions \emph{double count} the observed data.
-The Viterbi algorithm solves/predicts for the most probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}.
-It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not relibly evaluate goodness of fit for the different structral complexities of HMM models.
+We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions, which use the most probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}, double-count the observed data.
+It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not reliably evaluate goodness of fit for the different structural complexities of HMM models.
+\todo{think about/check out \textbf{pseudo-residuals}}
 
@@
 
+\todo{The description of the simulations should be in Methods, not Results; here we should describe the results of the simulations only}
 We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.
-Figure~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can over estiate the number of states via BIC-optimal approach.
+Figure~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach.
 
@@
 <<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale; need a real/more complete figure caption here; emphasize that dashes are for emphasis only; rename models (time-homog/heterog)}">>=
-# g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
-#     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
-# (g0c <- g0 +
-#     geom_encircle(data=subset(all_cat_pars,
-#                               (catID=="15" & log_mean<1.5) |
-#                               (catID!="15" & log_mean<1.8)),
-#                   colour="black",linetype=2,expand=0.05))
+g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
+     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
+ (g0c <- g0 +
+     geom_encircle(data=subset(all_cat_pars,
+                               (catID=="15" & log_mean<1.5) |
+                               (catID!="15" & log_mean<1.8)),
+                   colour="black",linetype=2,expand=0.05))
 @
@@
 \todo{do we need this?}
-Figure~\ref{state_plots} compares the estimated parameter values
+Figure~\ref{fig:state_plots} compares the estimated parameter values
 for several cats
@@
 
-Identifying behavioural states based on observations is a
+Identifying behavioural states based solely on
+observations of movement behaviour is a
 common methodological problem in behavioural ecology; finding
-solutions that are both biologically and statistically interpretable
-and defensible is challenging. Models that suggest a higher number of
-states fit the data well but are hard to interpret biologically,
-while 
-models that are biologically interpretable are not always statistically
+solutions that are both biologically interpretable
+and statistically defensible is challenging. 
+Models with large numbers of
+states fit the data well (even when model complexity
+is penalized) but are hard to interpret biologically,
+while models with fewer states
+that are biologically interpretable are not always statistically
 supported. While the HMM framework has been applied in many research
-areas, very few studies have considered modeling with more than 3
-states. HMM is a simple, straightforward framework, but HMM extensions
+areas, few studies have considered modeling with more than 3
+states. \todo{what is the rest of this paragraph supposed to get across?}
+HMM is a simple, straightforward framework, but HMM extensions
 of adding a small amount of complexity can complicate the model
@@
 
-We have presented a relatively simple concept yet little-studied
+We have presented a relatively simple yet little-studied
 extension by covariate dependent HMM can overfit the number of true
@@
 For some kinds of model predictions, the time homogeneous transition
-model can perform as good as the time inhomogeneous transition models
+model can perform as well as the time inhomogeneous transition models
 in terms of predicting the states by the Viterbi algorithm (the
 tradition state prediction method) and can easily overlook more
-complicated models.
-This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations.
-The limitation of the Viterbi appoach is that it cannot predict future movements beyond the observation sequence.
-In general, the time homogeneous
-transition model gives the broader scope of movement pattern where as
-the time inhomogeneous transitions models gives a more realistic
-movement structure and future predictions.
+complicated models. This approach allows the time homogeneous
+transition model's prediction to capture the temporal heterogeneity
+from the observations. The limitation of the Viterbi appoach is that
+it cannot predict future movements beyond the observation sequence. In
+general, the time homogeneous transition model gives the broader scope
+of movement pattern where as the time inhomogeneous transitions models
+gives a more realistic movement structure and future predictions.
 
-Although the time inhomogeneous transitions offer an objective
-approach to (1) reduce the number of states to a biologically
-interpretable level, (2) captures the observational autocorrelation
-and temporal pattern, (3) reduce the overall complexity as illustrated
-by our study, several challenges must be overcome in order to maximize
-its potential. The current model is based on the reduced univariate
+Models with time-inhomogeneous transitions appear to
+offer a simple way to (1) reduce the selected number of states to a biologically
+interpretable level, (2) capture observed diurnal and autocorrelation
+patterns, (3) provide fitted models that are better (lower BIC)
+and, overall, less complex than time-homogeneous models.
+However, several challenges must be overcome in order to maximize
+their potential. The current model is based on the reduced univariate
 step length response emission without fitting turning angles. Although
 turning angles are strongly correlated with step length, it does not
-effect the overall prediction on number of states. But movement
+affect the overall prediction on number of states. But movement
 trajectories predictions, and spatial models still require both step
@@
 challenges, our study provides an important extension in HMM modeling
-that is not in the seen before and a real life application that it
-exist and applicable in animal movement ecology.
+that has not been seen before and a real life-application to
+animal movement ecology.
 
@@
 
-\todo{does it make any sense to compress figs~\ref{adj_BIC-comparisons_cat2} etc.
+\todo{does it make any sense to compress figs~\ref{fig:adj_BIC_comparisons_cat2} etc.
 into a single plot (via facetgrid or multi-lines)?

### 2016-07-12 paper3.Rnw
ggalt tweaking

@@
 library("RColorBrewer")
-# if (!require("ggalt")) {
+if (!require("ggalt") || length(find("geom_encircle"))==0) {
 #     cat("installing ggalt from BB repo ...\n")
-#     library("devtools")
-#     install_github("bbolker/ggalt")
-#     library("ggalt")
-# }
+    try(detach("package:ggalt"),silent=TRUE)
+    devtools::install_github("hrbrmstr/ggalt")
     library("ggalt")
+}
 @

### 2016-07-15 paper3.Rnw
update figure 3 ...

@@
 linetypev <- c(rep(2,2),rep(1,5))
-(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
-                                    , y=deltaBIC
-                                    , colour=model
-                                    , linetype=model))
-+ facet_wrap(~ type,ncol=4)
-+ scale_size_continuous( name="# of parameters")
-+ scale_x_continuous(expand=c(0,0.5),breaks=2:7)
-+ geom_point(aes(size=parameters))
-+ scale_linetype_manual(values=linetypev)
-+ scale_colour_brewer(palette="Dark2")
-+ labs(y=expression(paste(Delta, "BIC"))
-       , x="Number of States") 
-+ geom_line() 
-+ zmargin)
+## adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
+##                                     , y=deltaBIC
+##                                     , colour=model
+##                                     , linetype=model))
+## + facet_wrap(~ type,ncol=4)
+## + scale_size_continuous( name="# of parameters")
+## + scale_x_continuous(expand=c(0,0.5)) ## ,breaks=2:7)
+## + geom_point(aes(size=parameters))
+## + scale_linetype_manual(values=linetypev)
+## + scale_colour_brewer(palette="Dark2")
+## + scale_shape_discrete(name="# of states")
+## + scale_alpha_discrete(name="# of states")    
+## + labs(y=expression(paste(Delta, "BIC"))
+##        , x="Number of States") 
+## + geom_line() 
+## + zmargin
 
@@
 
-# sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
-#                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
-# )
-# linetypev <- c(rep(2,2),rep(1,5))
-# (adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
-#                                     , y=deltaBIC
-#                                     , colour=model
-#                                     , linetype=model
-#                                     , group=model))
-#   + facet_wrap(~mod,ncol = 2,scales = "free_x")
-#   + scale_size_continuous(name="# of states")
-#   + scale_x_continuous(breaks=seq(0,300,by=10))
-#   + geom_point(aes(shape=factor(nstates),size=1.5,alpha=nstates))
-#   + scale_linetype_manual(values=linetypev)
-#   + scale_colour_brewer(palette="Dark2")
-#   + labs(y=expression(paste(Delta, "BIC"))
-#          , x="Number of States") 
-#   + geom_line()
-#   + zmargin)
+sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
+                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
+)
+linetypev <- c(rep(2,2),rep(1,5))
+(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
+                                  , y=deltaBIC
+                                  , colour=model
+                                  , linetype=model
+                                  , group=model))
+   + facet_wrap(~mod,ncol = 2,scales = "free_x")
+   + scale_size_continuous(name="# of states")
+   + scale_x_continuous(breaks=seq(0,300,by=30))
+   + geom_point(aes(shape=factor(nstates),alpha=nstates))
+   + scale_linetype_manual(values=linetypev)
+   + scale_colour_brewer(palette="Dark2")
+   + scale_shape_discrete(name="Number of states")
+   + scale_alpha_continuous(name="Number of states")
+   + labs(y=expression(paste(Delta, "BIC"))
+          , x="Number of parameters") 
+   + geom_line()
+   + zmargin)
 @
@@
   + ylab(expression('Mean step length'~(log[10]*m)))
-  + scale_color_manual(values=colvec)
+  + scale_colour_manual(values=colvec)
   + scale_alpha_manual(values=alphavec)
@@
   + scale_linetype_manual(values=linetypev)
-) + geom_point(data=vitdf,aes(y=viterbi),lwd=2.5,alpha=1,colour="black",shape=118)
+) + geom_point(data=vitdf,aes(y=viterbi),
+               lwd=2.5,alpha=1,colour="black",shape=118,show.legend=FALSE)
+## adding geom_point() with vitdf screws up the legend ...
 @
@@
 )
-(ggplot(acfdf,aes(x=lag,y=ACF,colour=model,lwd=model,alpha=model,linetype=model)) 
+(ggplot(acfdf,aes(x=lag,y=ACF,colour=model,
+                  lwd=model,alpha=model,linetype=model)) 
   + geom_line()
@@
   + ylab("Autocorrelation")
-  + scale_color_manual(values=colvec)
+  + scale_colour_manual(values=colvec)
   + scale_alpha_manual(values=alphavec)
@@
   + scale_linetype_manual(values=linetypev)
-) + geom_point(data=vitacf,lwd=2.5,alpha=1,colour="black",shape=118)
+) + geom_point(data=vitacf,lwd=2.5,alpha=1,colour="black",shape=118,
+             )
+## vitacf messes up legend ...
 @

### 2016-07-21 paper3.Rnw
discussion kind of finish

@@
 
-Identifying behavioural states based solely on
-observations of movement behaviour is a
-common methodological problem in behavioural ecology; finding
-solutions that are both biologically interpretable
-and statistically defensible is challenging. 
-Models with large numbers of
-states fit the data well (even when model complexity
-is penalized) but are hard to interpret biologically,
-while models with fewer states
-that are biologically interpretable are not always statistically
-supported. While the HMM framework has been applied in many research
-areas, few studies have considered modeling with more than 3
-states. \todo{what is the rest of this paragraph supposed to get across?}
-HMM is a simple, straightforward framework, but HMM extensions
-of adding a small amount of complexity can complicate the model
-greatly and reveal deeper understanding. For example,
-\citet{kerk2015hidden} used 2 to 6 movement states HSMM to model
-Florida panthers, \citet{morales_extracting_2004} used habitat
-distance dependent transition HMM to model elk, and outside of
-ecology, \citet{raffa_multivariate_2015} used mixed effect transition
-and emission HMM in a smoking study.
+HMMs are widely used and flexible to model these type of movement data.
+With the increasing volumes and higher resolution data that are available, traditional HMMs will overfit with big data with large number of states that may not be feasible.
+Models with large numbers of states fit the data well (even when model complexity is penalized) but are hard to interpret biologically, while models with fewer states that are biologically interpretable are not always statistically supported.
+We need to work harder to make these models appropriate complex that can be both biologically interpretable and statistically defensible.
 
@@
 states in covariate dependent HMM (in our case, time-dependent
-transitions). HMM with more than three states are not biologically
+transitions) with simulations and with Florida panther data. HMM with more than three states are not biologically
 interpretable and often discarded, but we cannot ignore that
 increasing complexity (number of states) states can increase goodness
-of fit statistically. Our analyses revealed modeling temporal
-heterogeneity in HMM transition can reduce the number of states
-estimated in time homogeneous transitions HMM movement of Florida
-panthers and reduce the overall model complexity.
+of fit statistically.
+Our analyses revealed modelling with time-inhomogeneous transitions appear to offer a simple way to (1) reduce the selected number of states to a biologically interpretable level, (2) capture observed diurnal and autocorrelation patterns, (3) provide fitted models that are better (lower BIC)
+and, overall, less complex than time-homogeneous models.
 
-For some kinds of model predictions, the time homogeneous transition
-model can perform as well as the time inhomogeneous transition models
-in terms of predicting the states by the Viterbi algorithm (the
-tradition state prediction method) and can easily overlook more
-complicated models. This approach allows the time homogeneous
-transition model's prediction to capture the temporal heterogeneity
-from the observations. The limitation of the Viterbi appoach is that
-it cannot predict future movements beyond the observation sequence. In
-general, the time homogeneous transition model gives the broader scope
-of movement pattern where as the time inhomogeneous transitions models
-gives a more realistic movement structure and future predictions.
-
-Models with time-inhomogeneous transitions appear to
-offer a simple way to (1) reduce the selected number of states to a biologically
-interpretable level, (2) capture observed diurnal and autocorrelation
-patterns, (3) provide fitted models that are better (lower BIC)
-and, overall, less complex than time-homogeneous models.
-However, several challenges must be overcome in order to maximize
-their potential. The current model is based on the reduced univariate
-step length response emission without fitting turning angles. Although
-turning angles are strongly correlated with step length, it does not
-affect the overall prediction on number of states. But movement
-trajectories predictions, and spatial models still require both step
-length and turning angles. Furthermore, modelling temporal
-heterogeneity only reduce the number of predicted states by 1
-resulting in 4/5 states might suggest (1) other factors such as
-spatial components that might also contribute to over predicting
-number of states, or (2) there really are 1 or 2 more different
-movement characteristics that are biologically unknown. Despite the
-challenges, our study provides an important extension in HMM modeling
-that has not been seen before and a real life-application to
-animal movement ecology.
+Incorportating temporal heterogenity in animal movement is definitely the right direction, but there are much more work to be done.
+From the ecological point of view, our model lacks association of behavioural and environmental features that can potentially improve goodness of fit, predictions, and reduce the number of states.
+It is really hard to optimize and allocate/distribute parameters when increasing model complexity by adding more predictors or adding more states, especially in HMMs where they are state dependent and may have non-linear relation between states and parameters.
+More diagnostics and tests are needed to identify features, patterns and mismatches in the data and to validate model fitting performance and predictability.
 

### 2016-07-24 paper3.Rnw
update discussion

@@
 \usepackage{alltt}
+\usepackage{hyperref}
 \usepackage[utf8]{inputenc} % for accented characters
@@
 
-HMMs are widely used and flexible to model these type of movement data.
-With the increasing volumes and higher resolution data that are available, traditional HMMs will overfit with big data with large number of states that may not be feasible.
-Models with large numbers of states fit the data well (even when model complexity is penalized) but are hard to interpret biologically, while models with fewer states that are biologically interpretable are not always statistically supported.
-We need to work harder to make these models appropriate complex that can be both biologically interpretable and statistically defensible.
+%% notes from blackboard
+%% - HMMs are widely used; need to work harder to make them appropriately complex (*)
+%% - recapitulate problem: big data overfits/large number of states is problematic (*)
+%% - adding temphet goes in the right direction (i.e. sims work, model is better/lower BIC, fewer states, less complex/fewer parameters overall)
+%% - GOALS: identifying states and/or association of behaviour with environment
+%%   *or* prediction
+%% - need for more diagnostics (e.g. Potts et al 2014); identify mismatches
+%% - more flexible models
 
-We have presented a relatively simple yet little-studied
-extension by covariate dependent HMM can overfit the number of true
-states in covariate dependent HMM (in our case, time-dependent
-transitions) with simulations and with Florida panther data. HMM with more than three states are not biologically
-interpretable and often discarded, but we cannot ignore that
-increasing complexity (number of states) states can increase goodness
-of fit statistically.
-Our analyses revealed modelling with time-inhomogeneous transitions appear to offer a simple way to (1) reduce the selected number of states to a biologically interpretable level, (2) capture observed diurnal and autocorrelation patterns, (3) provide fitted models that are better (lower BIC)
-and, overall, less complex than time-homogeneous models.
+HMMs are a widely used and flexible tool for modeling animal movement
+behaviour; we need to work harder to make sure they are both appropriately
+complex and biologically interpretable. 
+With the increasing volumes of movement data available,
+ecologists who naively use traditional homogeneous HMMs
+and standard information-theoretic criteria
+to estimate the number of behavioural states will generally
+overfit their data, in the sense of ``discovering'' large
+number of states that are difficult to interpret biologically.
 
-Incorportating temporal heterogenity in animal movement is definitely the right direction, but there are much more work to be done.
-From the ecological point of view, our model lacks association of behavioural and environmental features that can potentially improve goodness of fit, predictions, and reduce the number of states.
-It is really hard to optimize and allocate/distribute parameters when increasing model complexity by adding more predictors or adding more states, especially in HMMs where they are state dependent and may have non-linear relation between states and parameters.
-More diagnostics and tests are needed to identify features, patterns and mismatches in the data and to validate model fitting performance and predictability.
+We have presented a relatively simple but little-used extension
+(time-dependent transitions) that partly resolves the problem.
+Time-dependent transitions appear to offer a simple way to (1) reduce
+the selected number of states closer to a biologically interpretable
+level; (2) capture observed diurnal and autocorrelation patterns in a
+predictive model; (3) improve overall model fit (i.e., lower BIC) and
+reduce the level of complexity (number of parameters) of the most
+parsimonious models. Simple simulations where the true number of
+states is known, and transitions among states vary over time, confirm
+that using BIC with homogeneous HMMs overestimates the number of
+behavioural states, while time-dependent HMMs correctly estimate
+the number.
+
+\textbf{BMB: I think we want one (or two?) paragraph(s) here that revisit(s)
+the deeper statistical issues of 
+\begin{enumerate}
+\item that different approaches are needed for prediction vs. state identification. If we're
+aiming for state identification it might be OK to pre-specify the
+number of states and, \emph{post hoc}, match (Viterbi-based) state
+estimates with environmental variation (à la Fryxell et al.). 
+If we have specific
+environmental covariates in mind, though, it might be better to
+fit a covariate-dependent model so that we can explicitly model
+the switching process \ldots  If we want to do prediction, we have
+to include the covariates, because Viterbi can only estimate
+state occupancy if observed movements are available \ldots
+\item prediction (of movements) vs estimation of number of states also
+represent two different questions; \citep{celeux_deviance_2006} (I haven't
+read it!) and \href{http://www.modulad.fr/archives/JPFenelon/Prog-JPF-2006/Celeux/CycleFenelon_Conference_Gilles_Celeux.pdf}{Celeux presentation} discuss this;
+BIC is good for prediction, but not necessarily for estimation of number
+of states.  We don't need to go into detail, but should point in this direction. Also perhaps hint a bit more above where we reference
+\citep{dean_behavioural_2013}.
+\end{enumerate}
+}
+
+Incorporating temporal heterogenity in animal movement is one step in
+the right direction, but much remains to be done. 
+Our model neglects other predictors, such as habitat type or
+location with respect to environmental features such as roads,
+that can potentially improve goodness of fit and predictions and 
+further reduce the estimated number of states. While adding
+more covariates is in principle straightforward using existing
+frameworks, including all possible biological complexities in 
+a HMM with state-dependent transitions may rapidly become intractable
+in terms of both computational time and complexity of choosing
+among possible reduced models and numbers of states.
+Better diagnostic procedures and tests are needed:
+these can both test overall
+goodness-of-fit \citep{potts_generalized_2014} and,
+more importantly, localize fitting problems to particular aspects of the
+data so that models can be constructed without needing to
+include all possible features of interest.
 

### 2016-07-27 paper3.Rnw
pictures with nice legends, paragraph shortly

@@
    + geom_point(aes(shape=factor(nstates),alpha=nstates))
-   + scale_linetype_manual(values=linetypev)
-   + scale_colour_brewer(palette="Dark2")
+   + scale_linetype_manual(values=linetypev, name="Model")
+   + scale_colour_brewer(palette="Dark2", name="Model")
    + scale_shape_discrete(name="Number of states")
@@
 
+vitdf2 <- vitdf %>% transmute(time=time,variable="viterbi",value=viterbi)
+
 mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
-            "hmmquad5","hmmsin5","obs")
+            "hmmquad5","hmmsin5","obs","viterbi")
 mnames2 <- c("fmm-homogeneous  (n=5)"
@@
              , "sin (n=5)"
-             , "observed")
+             , "observed"
+             , "viterbi homogenous (n=3)")
 
-combdf <- (rbind(avgdf,obsdf)
+combdf <- (rbind(avgdf,obsdf,vitdf2)
            %>% transmute(time=time,model=variable,steplen=value)
@@
 )
-linetypev <- c(rep(c(2,1),c(2,6)),1)
-colvec <- c(c(brewer.pal(7,"Dark2"),"black"),"black")
-lwdvec <- c(rep(c(1,4),c(7,1)),4)
-alphavec <- c(rep(c(1,0.2),c(7,1)),1)
-(ggplot(combdf,aes(x=time,y=steplen,colour=model,lwd=model,alpha=model,linetype=model)) 
+linetypev <- c(rep(c(2,1),c(2,6))
+                ,0
+               )
+colvec <- c(c(brewer.pal(7,"Dark2"),"black")
+             ,"black"
+            )
+lwdvec <- c(rep(c(1,4),c(7,1))
+             ,4
+            )
+alphavec <- c(rep(c(1,0.2),c(7,1))
+               ,1
+              )
+
+shapevec <- c(rep(32,8),118)
+
+(ggplot(combdf,aes(x=time,y=steplen,colour=model,
+                   lwd=model,
+                   alpha=model,linetype=model, shape=model))
+  + geom_point()
   + geom_line()
   + xlab("Time of Day") 
+  + scale_shape_manual(values=shapevec, name="Model")
   + ylab(expression('Mean step length'~(log[10]*m)))
-  + scale_colour_manual(values=colvec)
-  + scale_alpha_manual(values=alphavec)
-  + scale_size_manual(values=lwdvec)
-  + scale_linetype_manual(values=linetypev)
-) + geom_point(data=vitdf,aes(y=viterbi),
-               lwd=2.5,alpha=1,colour="black",shape=118,show.legend=FALSE)
+  + scale_colour_manual(values=colvec, name="Model")
+  + scale_alpha_manual(values=alphavec, name="Model")
+  + scale_size_manual(values=lwdvec, name="Model")
+  + scale_linetype_manual(values=linetypev, name="Model")
+ ) 
+# + geom_point(data=vitdf,aes(y=viterbi),
+               # lwd=2.5,alpha=1,colour="black",shape=118,show.legend=FALSE)
 ## adding geom_point() with vitdf screws up the legend ...
@@
 )
-(ggplot(acfdf,aes(x=lag,y=ACF,colour=model,
-                  lwd=model,alpha=model,linetype=model)) 
+
+combacfdf <- (rbind(acfdf,vitacf)
+           %>% transmute(lag=lag,model=model,ACF=ACF)
+           %>% filter(model %in% mnames)
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+)
+
+(ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
+                  lwd=model,alpha=model,linetype=model,shape=model))
+  + geom_point()
   + geom_line()
@@
   + ylab("Autocorrelation")
-  + scale_colour_manual(values=colvec)
-  + scale_alpha_manual(values=alphavec)
-  + scale_size_manual(values=lwdvec)
-  + scale_linetype_manual(values=linetypev)
-) + geom_point(data=vitacf,lwd=2.5,alpha=1,colour="black",shape=118,
+  + scale_shape_manual(values=shapevec, name="Model")
+  + scale_colour_manual(values=colvec, name="Model")
+  + scale_alpha_manual(values=alphavec, name="Model")
+  + scale_size_manual(values=lwdvec, name="Model")
+  + scale_linetype_manual(values=linetypev, name="Model")
 )
+
 ## vitacf messes up legend ...

### 2016-07-27 paper3.Rnw
cheesy paragraph

@@
 
-We have presented a relatively simple but little-used extension
-(time-dependent transitions) that partly resolves the problem.
-Time-dependent transitions appear to offer a simple way to (1) reduce
-the selected number of states closer to a biologically interpretable
-level; (2) capture observed diurnal and autocorrelation patterns in a
-predictive model; (3) improve overall model fit (i.e., lower BIC) and
-reduce the level of complexity (number of parameters) of the most
-parsimonious models. Simple simulations where the true number of
-states is known, and transitions among states vary over time, confirm
-that using BIC with homogeneous HMMs overestimates the number of
-behavioural states, while time-dependent HMMs correctly estimate
-the number.
-
 \textbf{BMB: I think we want one (or two?) paragraph(s) here that revisit(s)
@@
 
+On a broad spectrum, it really depends on what kind of question 
+that is being answered. On one side of the spectrum, if the goal is 
+to identify states, it might be sufficient to use a 
+simple/traditional HMM model and pre-specify the number of states and, post hoc, 
+match Viterbi-based states estimates with environmental variation 
+\citep{fryxell_multiple_2008}. On the other side of the 
+spectrum, if the goal of interest is to make predictions (out of sample), 
+it might be better to fit a covariate-dependent model so that we can explicitly
+model the switching process. In that case, fitting a covariate-dependent model is
+better for out of sample prediction because Viterbi can only estimate state 
+occupancy if observed movements are available (within sample preditions). 
+Finally, if we want to estimate the number of states, BIC is not necessarily 
+good for estimation of number of states \citep{biernacki2000assessing}, 
+but it can be useful as an approximate upper limit estimate.
+
 Incorporating temporal heterogenity in animal movement is one step in
@@
 Better diagnostic procedures and tests are needed:
-these can both test overall
-goodness-of-fit \citep{potts_generalized_2014} and,
-more importantly, localize fitting problems to particular aspects of the
+these can both test overall goodness-of-fit \citep{potts_generalized_2014} 
+and, more importantly, localize fitting problems to particular aspects of the
 data so that models can be constructed without needing to
 include all possible features of interest.
+
+\section{Conclusion}
+
+We have presented a relatively simple but little-used extension
+(time-dependent transitions) that partly resolves the problem.
+Time-dependent transitions appear to offer a simple way to (1) reduce
+the selected number of states closer to a biologically interpretable
+level; (2) capture observed diurnal and autocorrelation patterns in a
+predictive model; (3) improve overall model fit (i.e., lower BIC) and
+reduce the level of complexity (number of parameters) of the most
+parsimonious models. Simple simulations where the true number of
+states is known, and transitions among states vary over time, confirm
+that using BIC with homogeneous HMMs overestimates the number of
+behavioural states, while time-dependent HMMs correctly estimate
+the number.
 

### 2016-07-29 bmc_template/bmc_article.tex
adding tex file... gitignore didn't add the first time

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Incorporating Periodic Variability in Hidden Markov Models for Animal Movement}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}         
   email={lim88@mcmaster.ca}   % email address
]{\inits{ML}\fnm{Michael} \snm{Li}}
\author[
   addressref={aff1,aff2},
   email={bolker@mcmaster.ca}
]{\inits{BMB}\fnm{Benjamin M.} \snm{Bolker}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{                            % unique id
  \orgname{Department of Biology, McMaster University}, % university, etc
  \street{1280 Main St. West},                     %
  \postcode{L8S 4K1},                                % post or zip code
  \city{Hamilton, Ontario},                              % city
  \cny{Canada}                                    % country
}
\address[id=aff2]{
  \orgname{Department of Mathematics and Statistics, McMaster University},
  \street{1280 Main St. West},
  \postcode{L8S 4K1},
  \city{Hamilton, Ontario},
  \cny{Canada}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
 \note{Sample of title note}     % note to the article
 \note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract

%% need to break into background,

Clustering time-series data into discrete groups can improve 
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
rates at which individuals move between groups can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal heterogeneity in transition probabilities, a straightforward 
but little-used extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers. More generally, we point out that model misspecification
(omitting important sources of variation) can lead to overfitting,
and as a corollary that incorporating previously neglected
structure in statistical models can allow more
accurate assessment of appropriate model complexity.

\parttitle{Background} %if any
Text for this section.

\parttitle{Results} %if any
Text for this section.

\parttitle{Conclusion} %if any
Text for this section.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{Hidden Markov Model}
\kwd{Animal Movement}
\kwd{Temporal Autocorrelation}
\kwd{Temporal Heterogeneity}
\kwd{Florida Panther}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
% \section*{Content}
% Text and results for this section, as per the individual journal's instructions for authors. %\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}

\section*{Background}

Given a sequence of animal movements, movement models aim to find a parsimonious description
that can be used to understand past movements and predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \cite{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\cite{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} % 
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \cite{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behaviour 
at particular times is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\cite{okubo_diffusion_1980,turchin1998quantitative}.

Ever-increasing capabilities of remote
sensors are making movement data available over an
ever-wider range of time scales, at both higher resolution (e.g. hourly data
from GPS collars vs. daily or weekly fixes for radio or VHF
collars) and longer extent (e.g. from a few days to significant
fractions of a year, or longer).  When analyzing such
long-term data, ecologists will more
often have to account for temporal variability in movement
behaviour at diurnal and seasonal scales that were previously
not captured in the data.

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
% BMB: pick one (perhaps include a parenthetical statement/definition
% saying that we treat these as a synonyms.  Just stringing them together
% with slashes is awkward)
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \cite{patterson_classifying_2009}, or by more
direct comparisons between inferred states and environmental
conditions \cite{fryxell_multiple_2008}. Schliehe-Diecks et al. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales \cite{gurarie2009novel}.

For movement behaviours that change on a fast time scale, such
that movement behaviours recorded
at successive observations are effectively independent,
\emph{finite mixture models} (FMMs) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \cite{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.
Thus FMMs and HMMs,
with or without temporal variation in the occupancy or transition
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
van de Kerk et al.\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).
% 
% %%\todo{we will clarify below that they actually used HSMMs} 
% %%to analyze hourly movement data
% %%from 18 Florida panthers (\emph{Puma concolor coryi}).
% 
van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
% (their criterion for choosing ...\todo{check \cite{zhao_knee_2008}!}).
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
van de Kerk et al.'s results on optimal
model complexity might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  For large data sets, 
information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of latent states; we also
fit models to simulated data with varying numbers of latent states
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

\section*{Methods}

\subsection*{Data and previous analyses}

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by van de Kerk et al.\cite{kerk2015hidden}.
Here we focus on the four cats with the most data (all with approximately 10,000-15,000
observations: see Table~1 in Supplementary Material), 
in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets.  
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\cite{turchin1998quantitative,kerk2015hidden}.

van de Kerk et al.\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\cite{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by van de Kerk et al.\cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not given), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which corresponds
to making precisely this assumption.

van de Kerk et al.\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).
(Note that most movement analyses, including van de Kerk et al.
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

van de Kerk et al.\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
three to six states,
suggesting that the six-state model was
favoured statistically; however,
the authors used three-state models in most
of their analyses for ease of biological interpretation.
We follow van de Kerk et al.\cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  
% \todo{it occurs to me that we haven't really included any
% information about simulation testing in this version!
% We should probably add a short section; if don't, then we
% should delete following sentence }
Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\cite{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \cite{BurnhamAnderson1998};
we believe that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

\subsection*{Model description}

\newcommand{\obs}{\mathbf{y}}
\newcommand{\cov}{\mathbf{x}}
\newcommand{\state}{z}

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $ \mathbf Y = \obs_{1},..., \obs_{T}$ 
and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model
parameters \boldmath{$\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
\begin{eqnarray}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T} | \theta,\mathbf{X}_{1:T}) 
&=& P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \nonumber\\
&& \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\label{eq:HMMlik}
\end{eqnarray}

The emissions $\obs_i$ are boldfaced to denote that we may have a vector of observations at each time point (e.g., step length and turning angle).
The model contains three distinct components:

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$, times the vector of observations $\obs_{1}$ conditioned on covariates $\cov_{1}$ and state $\state_{1}$.

\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\cov_{k}$ at time $t=k$.

\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observations $\obs_{k}$ conditioned on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
\end{description}

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
There are several efficient algorithms for computing the marginal likelihood and
numerically estimating parameters \cite{zucchini_hidden_2009};
we used those implemented in the \texttt{depmixS4} package for R
\cite{visser2010depmixs4,R}.

For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy. It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.

In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
As is standard for HMMs with covariates \cite{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without loss of generality (i.e. we have only $n \times (n-1)$ distinct parameters; we index $j$ from 2 to $n$ for notational clarity):

\begin{eqnarray}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij} \nonumber
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
    Text for this section \ldots

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      Figure legend text.}
      \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go.}
      \begin{tabular}{cccc}
        \hline
           & B1  &B2   & B3\\ \hline
        A1 & 0.1 & 0.2 & 0.3\\
        A2 & ... & ..  & .\\
        A3 & ..  & .   & .\\ \hline
      \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{backmatter}
\end{document}

### 2016-07-29 bmc_template/bmc_article.tex
need to center equations

@@
 %\usepackage{amsthm,amsmath}
+\usepackage{amsmath}
 %\RequirePackage{natbib}
@@
 
+\
+
 Given a sequence of animal movements, movement models aim to find a parsimonious description
@@
 
+\
+
 Ever-increasing capabilities of remote
@@
 not captured in the data.
+
+\
 
@@
 
+\
+
 For movement behaviours that change on a fast time scale, such
@@
 changes in movement behaviour over a range of time scales.
+
+\
 
@@
 % 
+\
+
 van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
@@
 identifying the number of latent states.
+
+\
 
@@
 
+\
+
 \section*{Methods}
 
+\
+
 \subsection*{Data and previous analyses}
+
+\
 
@@
 
+\
+
 van de Kerk et al.\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
@@
 to making precisely this assumption.
+
+\
 
@@
 between the two measures.)
+
+\
 
@@
 
+\
+
 \subsection*{Model description}
+
+\
 
@@
 parameters \boldmath{$\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
-\begin{eqnarray}
-P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T} | \theta,\mathbf{X}_{1:T}) 
-&=& P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \nonumber\\
-&& \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
+
+\begin{equation}
+\begin{split}
+P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
+P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
+&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
+\end{split}
 \label{eq:HMMlik}
-\end{eqnarray}
+\end{equation}
 
@@
 
+\
+
 \begin{description}
@@
 
+\
+
 \item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\cov_{k}$ at time $t=k$.
+
+\
 
@@
 \end{description}
+
+\
 
@@
 
+\
+
 For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
@@
 It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.
+
+\
 
@@
 
-\begin{eqnarray}
+\
+
+\begin{equation}
+\begin{split}
 \pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
-\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij} \nonumber
-\end{eqnarray}
+\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
+\end{split}
+\end{equation}
+
+We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of Morales et al.\cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
+
+\
+
+\begin{description}
+\item[Multiple block transition] Here we assume piecewise-constant transition probabilities.
+The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
+
+$$
+\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
+$$
+where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).
+
+\
+
+\item[Quadratic transition model] We assume the elements of the linear predictor
+are quadratic functions of hour.
+The quadratic model is not diurnally continuous, i.e.
+there is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity constraint would collapse the model to a constant.
+$$
+\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
+$$
+
+\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
+$$
+\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
+$$
+
+\item[Hourly model] Lastly, we extended the multi-block approach and assign  a different transition matrix for every hour of the day.
+This model is included for comparative purposes due to the large number of parameters in the model which makes it not really practical.
+We only fitted up to four states using the hourly model.
+\end{description}
+
+\
+
+Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could be useful directions for future exploration.
+
+\
+
+\subsection{Model evaluation}
+
+\
+
+We used the {\tt depmixS4} package
+to fit covariate-dependent transition HMMs, 
+simulate states and step lengths using the
+estimated parameters, and estimate the most likely
+states with the Viterbi
+algorithm.
+
+\
+
+\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}
+
+We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
+HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
+with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model 
+within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall 
+goodness of fit of each model type. (2) Comparing average step-length by hour of day for the 
+observed data and for data simulated from the models shows how well a particular class of 
+models can capture the diurnal variation in behaviour. (3) Comparing temporal autocorrelations
+for the observed data and for data simulated from the models shows how well a particular class
+of models captures serial correlation at both short and long scales.
+
+\
+
+Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity.
+The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components.
+Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$.
+Then the number of free parameters of an HMM is:
+
+\begin{equation}
+\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
+{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
+{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
+\label{eq:numparam}
+\end{equation}
 

### 2016-07-29 bmc_template/bmc_article.tex
temp pushing, have to add discussion and then pictures

@@
 
+As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas HMMs with temporal heterogenity will increase quadratically (Eq.~\ref{eq:numparam}).
+When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
+
+\
+
+We used simulations to predict hourly step length and ACF 
+because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
+We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions, which use the most probable sequence of movement states based on the observations \cite{zucchini_hidden_2009,langrock_flexible_2012}, double-count the observed data.
+It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not reliably evaluate goodness of fit for the different structural complexities of HMM models.
+
+\
+
+\section{Results}
+
+\
+
+We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.
+Fig.
+~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach.
+
+\
+
+Fig.
+~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles)
+
+\
+
+As a complement, we also fitted FMM and FMM with sinusoidal variation in state occupancy probabilities to compare the temporal effects in goodness of fit (dashed lines).
+% \todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
+As a reminder, FMMs assume that the latent state in each time step is 
+\emph{independent} of the latent state at the previous time step; 
+time-varying FMMs can accurately describe movement when behaviour can
+change on a short time scale, but the average propensity for different
+behaviours changes over time.
+
+\
+
+Figure~\ref{fig:adj_BIC_comparisons} shows that models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity.
+Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
+The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.
+Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
+However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
+normally be interpreted as an enormous difference in goodness of fit,
+between the sinusoidal and quadratic models (both of which have 90 
+parameters).
+
+\
+
+Fig.
+~\ref{fig:avg_step_length_by_time} shows a clear diurnal pattern from the average hourly step lengths from the observed data.
+As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle.
+All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns.
+However, the block model does markedly worse than the other temporal models (changing the block definitions might help), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
+We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points).
+
+\
+
+Like the diurnal pattern shown in Fig. ~\ref{fig:avg_step_length_by_time},
+the strong autocorrelation of the observed step lengths at a 24-hour lag (Fig.
+~\ref{fig:acf_plot}) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
+Because there are a huge number of potential complexities that can be added to movement models (e.g.
+spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.
+In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects.
+HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
+Temporally homogeneous FMM, by definition, produces neither short- nor long-term autocorrelation.
+FMM without temporal heterogeneity, although it captures the diurnal pattern well, underpredicts the degree of short-term autocorrelation.
+
+\
+
+The hardest problem with multiple latent states is interpreting them
+biologically. We have no way of knowing what panthers are actually
+thinking (it is certainly more complex than being in one of a small
+number of discrete latent states); 
+we don't know the ``true'' number of latent states,
+nor are we able to observe them directly, although
+incorporating additional direct observations of 
+behaviour (if available) can at least partially
+address this problem \cite{fryxell_multiple_2008}.
+Three distinct
+movement states seem biologically interpretable for Florida panthers
+according to van de Kerk et al.\cite{kerk2015hidden}: Short step length suggests resting
+states, intermediate step length a foraging state, and long
+step length a traveling state. 
+% \todo{do we need this?}
+Figure~\ref{fig:state_plots} compares the estimated parameter values
+for several cats
+(mean and standard deviation of the step length in each state) between
+the time-homogeneous and time-heterogeneous models. 
+In general, the states with longer mean step lengths are relatively
+similar between model classes. For cats 14 and 15, the states with
+the longest or next-longest mean step lengths have similar means
+and standard deviations; for cats 1 and 2, three long-step states
+in the homogeneous HMM appear to divide two long-step states in 
+the heterogeneous HMM. For short-step states, the heterogeneous HMM
+tends to identify a high-variance state, while the homogeneous HMM
+picks up states with very short step lengths (questionable in any
+case because we have not taken any special efforts to account for
+GPS error).
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-08-01 bmc_template/bmc_article.tex
temp pushing

@@
 
+\
+
+\section{Discussion}
+
+\
+
+HMMs are a widely used and flexible tool for modeling animal movement
+behaviour; we need to work harder to make sure they are both appropriately
+complex and biologically interpretable. 
+With the increasing volumes of movement data available,
+ecologists who naively use traditional homogeneous HMMs
+and standard information-theoretic criteria
+to estimate the number of behavioural states will generally
+overfit their data, in the sense of ``discovering'' large
+number of states that are difficult to interpret biologically.
+
+\
+
+On a broad spectrum, it really depends on what kind of question 
+that is being answered. On one side of the spectrum, if the goal is 
+to identify states, it might be sufficient to use a 
+simple/traditional HMM model and pre-specify the number of states and, post hoc, 
+match Viterbi-based states estimates with environmental variation 
+\cite{fryxell_multiple_2008}. On the other side of the 
+spectrum, if the goal of interest is to make predictions (out of sample), 
+it might be better to fit a covariate-dependent model so that we can explicitly
+model the switching process. In that case, fitting a covariate-dependent model is
+better for out of sample prediction because Viterbi can only estimate state 
+occupancy if observed movements are available (within sample preditions). 
+Finally, if we want to estimate the number of states, BIC is not necessarily 
+good for estimation of number of states \cite{biernacki2000assessing}, 
+but it can be useful as an approximate upper limit estimate.
+
+\
+
+Incorporating temporal heterogenity in animal movement is one step in
+the right direction, but much remains to be done. 
+Our model neglects other predictors, such as habitat type or
+location with respect to environmental features such as roads,
+that can potentially improve goodness of fit and predictions and 
+further reduce the estimated number of states. While adding
+more covariates is in principle straightforward using existing
+frameworks, including all possible biological complexities in 
+a HMM with state-dependent transitions may rapidly become intractable
+in terms of both computational time and complexity of choosing
+among possible reduced models and numbers of states.
+Better diagnostic procedures and tests are needed:
+these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
+and, more importantly, localize fitting problems to particular aspects of the
+data so that models can be constructed without needing to
+include all possible features of interest.
+
+\
+
+\section{Conclusion}
+
+\
+
+We have presented a relatively simple but little-used extension
+(time-dependent transitions) that partly resolves the problem.
+Time-dependent transitions appear to offer a simple way to (1) reduce
+the selected number of states closer to a biologically interpretable
+level; (2) capture observed diurnal and autocorrelation patterns in a
+predictive model; (3) improve overall model fit (i.e., lower BIC) and
+reduce the level of complexity (number of parameters) of the most
+parsimonious models. Simple simulations where the true number of
+states is known, and transitions among states vary over time, confirm
+that using BIC with homogeneous HMMs overestimates the number of
+behavioural states, while time-dependent HMMs correctly estimate
+the number.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-08-02 bmc_template/figures.Rnw
need figure capations and fix up cat 14 and 15

\documentclass{article}

\begin{document}

<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
## fix complaints about "incompatible color definition" from xcolor ...
## http://tex.stackexchange.com/questions/148188/knitr-xcolor-incompatible-color-definition
knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
library("gridExtra")
library("RColorBrewer")
if (!require("ggalt") || length(find("geom_encircle"))==0) {
#     cat("installing ggalt from BB repo ...\n")
    try(detach("package:ggalt"),silent=TRUE)
    devtools::install_github("hrbrmstr/ggalt")
    library("ggalt")
}
@

<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Simulation test for BIC-optimal approach",fig.width=6,fig.height=4>>=
ICdat <- readRDS("../sim_results/ICdat.rds")
aa <- readRDS("../sim_results/sim.101.RDS")

ICdat <- mutate(ICdat,
                model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("homogeneous","heterogeneous"))) %>% filter(IC=="BIC")
simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,
                            linetype=IC,shape=IC))+
                        geom_point(size=3) +
                        geom_line() +
                        scale_colour_brewer(palette="Set1")+
                        labs(y="counts",x="# latent states")
print(simplot)

@

\clearpage

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name="step length\ndistribution",palette="Set1")
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

\clearpage

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
#L <- load("cat1sumdat.RData")

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.1.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

\clearpage

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=

avgdf <- readRDS("../summary_stats/cat.1.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.1.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("../summary_stats/cat.1.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

vitdf2 <- vitdf %>% transmute(time=time,variable="viterbi",value=viterbi)

mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs","viterbi")
mnames2 <- c("fmm-homogeneous  (n=5)"
             , "fmm-sin (n=4)"
             , "homogeneous (n=6)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=5)"
             , "sin (n=5)"
             , "observed"
             , "viterbi homogenous (n=3)")

combdf <- (rbind(avgdf,obsdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

shapevec <- c(rep(32,8),118)

(ggplot(combdf,aes(x=time,y=steplen,colour=model,
                   lwd=model,
                   alpha=model,linetype=model, shape=model))
  + geom_point()
  + geom_line()
  + xlab("Time of Day") 
  + scale_shape_manual(values=shapevec, name="Model")
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
 ) 
@

\clearpage

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

acfdf <- (readRDS("../summary_stats/cat.1.df.RDS")$acfdf 
          %>% mutate(model = .id)
)

combacfdf <- (rbind(acfdf,vitacf)
           %>% transmute(lag=lag,model=model,ACF=ACF)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)

(ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
                  lwd=model,alpha=model,linetype=model,shape=model))
  + geom_point()
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_shape_manual(values=shapevec, name="Model")
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
)

## vitacf messes up legend ...
@

\clearpage

<<r msdlist>>=
get_cat_pars2 <- function(catid){
  templist <- readRDS(paste("../summary_stats/cat.",catid,".df.RDS",sep=""))
  moddf <- templist$msdlist

  df <- data.frame(catID = catid
          , log_mean = c(unlist(moddf[1]),unlist(moddf[3]))
          , log_sd = c(unlist(moddf[2]),unlist(moddf[4]))
          , Model = c(rep("HMM",length(unlist(moddf[1]))),rep("HMM+TH(sin)",length(unlist(moddf[3]))))
  )
  return(df)
}

all_cat_pars2 <- ldply(lme4:::namedList(1,22,14,15),get_cat_pars2)

g0 <- ggplot(all_cat_pars2,aes(log_mean,log_sd,color=Model))+
     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
 (g0c <- g0 +
     geom_encircle(data=subset(all_cat_pars2,
                               (catID=="15" & log_mean<1.5) |
                               (catID!="15" & log_mean<1.8)),
                   colour="black",linetype=2,expand=0.05))

@

\clearpage

\section{sup}

<<BICred_plot2,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=

allmodsumdf2 <- readRDS("../summary_stats/cat.2.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name="step length\ndistribution",palette="Set1")
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

\clearpage

<<adj_BIC_comparisons2, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
#L <- load("cat1sumdat.RData")

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.2.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

\clearpage

<<avg_step_length_by_time2, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=

avgdf <- readRDS("../summary_stats/cat.2.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.2.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("../summary_stats/cat.2.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

vitdf2 <- vitdf %>% transmute(time=time,variable="viterbi",value=viterbi)

mnames <- c("fmm4","fmmsin4","hmm5","hmmblock4","hmmhourly3",
            "hmmquad4","hmmsin4","obs","viterbi")
mnames2 <- c("fmm-homogeneous  (n=4)"
             , "fmm-sin (n=4)"
             , "homogeneous (n=5)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=4)"
             , "sin (n=4)"
             , "observed"
             , "viterbi homogenous (n=3)")

combdf <- (rbind(avgdf,obsdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

shapevec <- c(rep(32,8),118)

(ggplot(combdf,aes(x=time,y=steplen,colour=model,
                   lwd=model,
                   alpha=model,linetype=model, shape=model))
  + geom_point()
  + geom_line()
  + xlab("Time of Day") 
  + scale_shape_manual(values=shapevec, name="Model")
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
 ) 
@

\clearpage

<<acf_plot2, message=FALSE, echo=FALSE, warning=FALSE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

acfdf <- (readRDS("../summary_stats/cat.2.df.RDS")$acfdf 
          %>% mutate(model = .id)
)

combacfdf <- (rbind(acfdf,vitacf)
           %>% transmute(lag=lag,model=model,ACF=ACF)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)

(ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
                  lwd=model,alpha=model,linetype=model,shape=model))
  + geom_point()
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_shape_manual(values=shapevec, name="Model")
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
)

## vitacf messes up legend ...
@

\clearpage
%cat14

<<BICred_plot14,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=

allmodsumdf2 <- readRDS("../summary_stats/cat.14.df2.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name="step length\ndistribution",palette="Set1")
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

\clearpage

<<adj_BIC_comparisons14, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
#L <- load("cat1sumdat.RData")

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.14.df2.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

\clearpage

<<avg_step_length_by_time14, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=

avgdf <- readRDS("../summary_stats/cat.14.df2.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.14.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("../summary_stats/cat.14.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

vitdf2 <- vitdf %>% transmute(time=time,variable="viterbi",value=viterbi)

mnames <- c("fmm4","fmmsin4","hmm5","hmmblock4","hmmhourly3",
            "hmmquad4","hmmsin4","obs","viterbi")
mnames2 <- c("fmm-homogeneous  (n=4)"
             , "fmm-sin (n=4)"
             , "homogeneous (n=5)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=4)"
             , "sin (n=4)"
             , "observed"
             , "viterbi homogenous (n=3)")

combdf <- (rbind(avgdf,obsdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

shapevec <- c(rep(32,8),118)

(ggplot(combdf,aes(x=time,y=steplen,colour=model,
                   lwd=model,
                   alpha=model,linetype=model, shape=model))
  + geom_point()
  + geom_line()
  + xlab("Time of Day") 
  + scale_shape_manual(values=shapevec, name="Model")
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
 ) 
@

\clearpage

<<acf_plot14, message=FALSE, echo=FALSE, warning=FALSE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

acfdf <- (readRDS("../summary_stats/cat.14.df.RDS")$acfdf 
          %>% mutate(model = .id)
)

combacfdf <- (rbind(acfdf,vitacf)
           %>% transmute(lag=lag,model=model,ACF=ACF)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)

(ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
                  lwd=model,alpha=model,linetype=model,shape=model))
  + geom_point()
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_shape_manual(values=shapevec, name="Model")
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
)

## vitacf messes up legend ...
@

\end{document}

### 2016-08-02 bmc_template/figures.Rnw
just need figure captions and then done

@@
 
-<<r msdlist>>=
+<<r msdlist,message=FALSE, echo=FALSE, warning=FALSE>>=
 get_cat_pars2 <- function(catid){
@@
 
-all_cat_pars2 <- ldply(lme4:::namedList(1,22,14,15),get_cat_pars2)
+all_cat_pars2 <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars2)
 
@@
 
-allmodsumdf2 <- readRDS("../summary_stats/cat.14.df2.RDS")$multimoddf
+allmodsumdf2 <- readRDS("../summary_stats/cat.14.df.RDS")$multimoddf
 df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
@@
 linetypev <- c(rep(2,2),rep(1,5))
-sumdat <- transform(readRDS("../summary_stats/cat.14.df2.RDS")$HMMsumdf,
+sumdat <- transform(readRDS("../summary_stats/cat.14.df.RDS")$HMMsumdf,
                      mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
@@
 
-avgdf <- readRDS("../summary_stats/cat.14.df2.RDS")$avgplotdf
+avgdf <- readRDS("../summary_stats/cat.14.df.RDS")$avgplotdf
 obsdf <- (readRDS("../summary_stats/cat.14.RDS") 
@@
 
-mnames <- c("fmm4","fmmsin4","hmm5","hmmblock4","hmmhourly3",
-            "hmmquad4","hmmsin4","obs","viterbi")
+mnames <- c("fmm4","fmmsin3","hmm5","hmmblock4","hmmhourly3",
+            "hmmquad5","hmmsin4","obs","viterbi")
 mnames2 <- c("fmm-homogeneous  (n=4)"
-             , "fmm-sin (n=4)"
+             , "fmm-sin (n=3)"
              , "homogeneous (n=5)"
@@
              , "hourly (n=3)"
-             , "quadratic (n=4)"
+             , "quadratic (n=5)"
              , "sin (n=4)"
@@
 
+\clearpage
+%cat15
+
+<<BICred_plot15,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
+
+allmodsumdf2 <- readRDS("../summary_stats/cat.15.df.RDS")$multimoddf
+df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
+            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
+            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
+            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
+(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
+    + geom_line()
+    + facet_wrap(~ homog)
+    + scale_colour_brewer(name="step length\ndistribution",palette="Set1")
+    ## scale_size_continuous( name="# parameters") +
+    #    geom_point(aes(size=parameters))+
+    + scale_linetype(name="von Mises\nturning angle?")
+    + labs(y=expression(Delta*"BIC"),
+         x="Number of States")
+    + geom_hline(yintercept=0,colour="gray")
+    + zmargin)
+@
+
+\clearpage
+
+<<adj_BIC_comparisons15, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
+#L <- load("cat1sumdat.RData")
+
+linetypev <- c(rep(2,2),rep(1,5))
+sumdat <- transform(readRDS("../summary_stats/cat.15.df.RDS")$HMMsumdf,
+                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
+)
+linetypev <- c(rep(2,2),rep(1,5))
+(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
+                                  , y=deltaBIC
+                                  , colour=model
+                                  , linetype=model
+                                  , group=model))
+   + facet_wrap(~mod,ncol = 2,scales = "free_x")
+   + scale_size_continuous(name="# of states")
+   + scale_x_continuous(breaks=seq(0,300,by=30))
+   + geom_point(aes(shape=factor(nstates),alpha=nstates))
+   + scale_linetype_manual(values=linetypev, name="Model")
+   + scale_colour_brewer(palette="Dark2", name="Model")
+   + scale_shape_discrete(name="Number of states")
+   + scale_alpha_continuous(name="Number of states")
+   + labs(y=expression(paste(Delta, "BIC"))
+          , x="Number of parameters") 
+   + geom_line()
+   + zmargin)
+@
+
+\clearpage
+
+<<avg_step_length_by_time15, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=
+
+avgdf <- readRDS("../summary_stats/cat.15.df.RDS")$avgplotdf
+obsdf <- (readRDS("../summary_stats/cat.15.RDS") 
+  %>% dplyr::select(c(LogDist,Time)) 
+  %>% filter(!is.na(LogDist)) 
+  %>% group_by(Time) 
+  %>% summarise_each(funs(mean))
+  %>% transmute(time = Time, variable = "obs", value=LogDist)
+)
+vit <- readRDS("../summary_stats/cat.15.vit.RDS")
+vitdf <- (vit %>% dplyr::select(c(viterbi,time))
+  %>% group_by(time)
+  %>% summarise_each(funs(mean))
+  %>% mutate(model="homogeneous (n=3)")
+)
+
+vitdf2 <- vitdf %>% transmute(time=time,variable="viterbi",value=viterbi)
+
+mnames <- c("fmm3","fmmsin4","hmm6","hmmblock4","hmmhourly3",
+            "hmmquad5","hmmsin5","obs","viterbi")
+mnames2 <- c("fmm-homogeneous  (n=3)"
+             , "fmm-sin (n=4)"
+             , "homogeneous (n=5)"
+             , "block (n=4)"
+             , "hourly (n=3)"
+             , "quadratic (n=5)"
+             , "sin (n=5)"
+             , "observed"
+             , "viterbi homogenous (n=3)")
+
+combdf <- (rbind(avgdf,obsdf,vitdf2)
+           %>% transmute(time=time,model=variable,steplen=value)
+           %>% filter(model %in% mnames)
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+)
+linetypev <- c(rep(c(2,1),c(2,6))
+                ,0
+               )
+colvec <- c(c(brewer.pal(7,"Dark2"),"black")
+             ,"black"
+            )
+lwdvec <- c(rep(c(1,4),c(7,1))
+             ,4
+            )
+alphavec <- c(rep(c(1,0.2),c(7,1))
+               ,1
+              )
+
+shapevec <- c(rep(32,8),118)
+
+(ggplot(combdf,aes(x=time,y=steplen,colour=model,
+                   lwd=model,
+                   alpha=model,linetype=model, shape=model))
+  + geom_point()
+  + geom_line()
+  + xlab("Time of Day") 
+  + scale_shape_manual(values=shapevec, name="Model")
+  + ylab(expression('Mean step length'~(log[10]*m)))
+  + scale_colour_manual(values=colvec, name="Model")
+  + scale_alpha_manual(values=alphavec, name="Model")
+  + scale_size_manual(values=lwdvec, name="Model")
+  + scale_linetype_manual(values=linetypev, name="Model")
+ ) 
+@
+
+\clearpage
+
+<<acf_plot15, message=FALSE, echo=FALSE, warning=FALSE>>=
+acffun <- function(sim){
+  a <- acf(sim,na.action=na.pass,plot=FALSE)
+  df <- data.frame(ACF=c(a$acf),lag=a$lag)
+  return(df)
+}
+vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
+)
+
+acfdf <- (readRDS("../summary_stats/cat.15.df.RDS")$acfdf 
+          %>% mutate(model = .id)
+)
+
+combacfdf <- (rbind(acfdf,vitacf)
+           %>% transmute(lag=lag,model=model,ACF=ACF)
+           %>% filter(model %in% mnames)
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+)
+
+(ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
+                  lwd=model,alpha=model,linetype=model,shape=model))
+  + geom_point()
+  + geom_line()
+  + xlab("Lag") 
+  + ylab("Autocorrelation")
+  + scale_shape_manual(values=shapevec, name="Model")
+  + scale_colour_manual(values=colvec, name="Model")
+  + scale_alpha_manual(values=alphavec, name="Model")
+  + scale_size_manual(values=lwdvec, name="Model")
+  + scale_linetype_manual(values=linetypev, name="Model")
+)
+
+## vitacf messes up legend ...
+@
+
 \end{document}

### 2016-08-02 bmc_template/figures.tex
just need figure captions and then done

\documentclass{article}\usepackage[]{graphicx}\usepackage{xcolor}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[Simulation test for BIC-optimal approach]{Simulation test for BIC-optimal approach}\label{fig:sim_results}
\end{figure}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/r_msdlist-1}

\end{knitrout}

\clearpage

\section{sup}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot2-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons2-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time2-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot2-1}

\end{knitrout}

\clearpage
%cat14

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot14-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons14-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time14-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot14-1}

\end{knitrout}

\clearpage
%cat15

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot15-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons15-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time15-1}

\end{knitrout}

\clearpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot15-1}

\end{knitrout}

\end{document}

### 2016-08-03 bmc_template/manuscript.tex
fig caps and ethics statement?

--- 2016-08-01 bmc_template/bmc_article.tex
@@
 \usepackage{amsmath}
+\usepackage{setspace}
+\doublespacing
+\usepackage{lineno}
+\linenumbers
 %\RequirePackage{natbib}
@@
    email={lim88@mcmaster.ca}   % email address
-]{\inits{ML}\fnm{Michael} \snm{Li}}
+]
+{\inits{ML}\fnm{Michael} \snm{Li}}
 \author[
@@
 
+% More generally, we point out that model misspecification
+% (omitting important sources of variation) can lead to overfitting,
+% and as a corollary that incorporating previously neglected
+% structure in statistical models can allow more
+% accurate assessment of appropriate model complexity.
+
+\parttitle{Background} %if any
 Clustering time-series data into discrete groups can improve 
@@
 models and clarifies the behavioural patterns of
-panthers. More generally, we point out that model misspecification
-(omitting important sources of variation) can lead to overfitting,
-and as a corollary that incorporating previously neglected
-structure in statistical models can allow more
-accurate assessment of appropriate model complexity.
-
-\parttitle{Background} %if any
-Text for this section.
+panthers.
 
 \parttitle{Results} %if any
-Text for this section.
+Simulations and Florida panthers data showed model misspecification (omitting important 
+sources of variation) can lead to overfitting and over-estimating number of behavioural states.
+Models incorporating temporal heterogeneity have lower number of states with slightly higher variation in short movement states, and able to make out of sample predictions that captures observed diurnal and autocorrelation patterns exhibited by Florida panthers.
 
 \parttitle{Conclusion} %if any
-Text for this section.
+Incorportating temporal heterogeneity reduce the selected number of behavioural states closer 
+to a biologically interpretable level, improved goodness of fit and predictability. Our 
+suggest that incorporating previously neglected structure in statistical models can
+allow more accurate assessment of appropriate model complexity.
 
@@
 \begin{equation}
-\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
+\textrm{Number of Free Parameter}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
 {+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
@@
 
+\section{Declarations}
+
 \begin{backmatter}
+
+\section*{Acknowledgements}
+  Florida Wildlife and Fisheries?
+
+\section{Data accessibility}
+Hourly step lengths and turning angles of male and female Florida pan- thers available at: http://ufdc.ufl.edu//IR00004241/00001.
+
+\section*{Author's contributions}
+    Equally contributed.
 
@@
   The authors declare that they have no competing interests.
-
-\section*{Author's contributions}
-    Text for this section \ldots
-
-\section*{Acknowledgements}
-  Text for this section \ldots
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   \begin{figure}[h!]
+  \includegraphics[width=\maxwidth]{figure/sim_results-1}
   \caption{\csentence{Sample figure title.}

### 2016-08-03 bmc_template/figures.Rnw
everything before supplementary is done

@@
 
-<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Simulation test for BIC-optimal approach",fig.width=6,fig.height=4>>=
+<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.",fig.width=6,fig.height=4>>=
 ICdat <- readRDS("../sim_results/ICdat.rds")
@@
                          levels=c("HMM","HMMsin"),
-                         labels=c("homogeneous","heterogeneous"))) %>% filter(IC=="BIC")
-simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,
-                            linetype=IC,shape=IC))+
+                         labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
+simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model))+
                         geom_point(size=3) +
@@
                         scale_colour_brewer(palette="Set1")+
-                        labs(y="counts",x="# latent states")
+                        labs(y="Counts",x="BIC-optimal States")
 print(simplot)
@@
 
-<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
+<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4>>=
 allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
@@
             turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
-            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
+            homog=ifelse(grepl("TH",type),"Time-inhomogeneous HMM","Homogeneous HMM"))
 (ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
@@
     + facet_wrap(~ homog)
-    + scale_colour_brewer(name="step length\ndistribution",palette="Set1")
+    + scale_colour_brewer(name="Step-length\ndistribution",palette="Set1")
     ## scale_size_continuous( name="# parameters") +
@@
 
-<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
+<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types.
+The left panel shows FMM, FMM with a sin prior and HMM.
+The right panel shows HMMs with different temporal transitions.",fig.width=8>>=
 #L <- load("cat1sumdat.RData")
@@
 
-<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=
+<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.",fig.width=8>>=
 
@@
 
-<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE>>=
+<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ">>=
 acffun <- function(sim){
@@
 
-<<r msdlist,message=FALSE, echo=FALSE, warning=FALSE>>=
+<<r msdlist,message=FALSE, echo=FALSE, fig.cap="State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.", warning=FALSE>>=
 get_cat_pars2 <- function(catid){
@@
 
-\section{sup}
+\section{Additional files}
 

### 2016-08-03 bmc_template/figures.tex
everything before supplementary is done

@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
-\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[Simulation test for BIC-optimal approach]{Simulation test for BIC-optimal approach}\label{fig:sim_results}
+\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations]{BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}\label{fig:sim_results}
 \end{figure}
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/BICred_plot-1} \caption[Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles]{Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.}\label{fig:BICred_plot}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1} \caption[Adjusted BIC by number of free parameters for HMM model types]{Adjusted BIC by number of free parameters for HMM model types.
+The left panel shows FMM, FMM with a sin prior and HMM.
+The right panel shows HMMs with different temporal transitions.}\label{fig:adj_BIC_comparisons}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1} \caption[Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states]{Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}\label{fig:avg_step_length_by_time}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/acf_plot-1} \caption[Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ]{Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states }\label{fig:acf_plot}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/r_msdlist-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/r_msdlist-1} \caption[State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition]{State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.}\label{fig:r msdlist}
+\end{figure}
 

### 2016-08-03 bmc_template/manuscript.tex
everything before supplementary is done

@@
 
-\
-
 Given a sequence of animal movements, movement models aim to find a parsimonious description
@@
 
-\
-
 Ever-increasing capabilities of remote
@@
 not captured in the data.
-
-\
 
@@
 
-\
-
 For movement behaviours that change on a fast time scale, such
@@
 
-\
-
 Our primary goal in this paper is to introduce the use of
@@
 behavioural states of Florida panthers (\emph{Puma concolor coryi}).
-% 
-% %%\todo{we will clarify below that they actually used HSMMs} 
-% %%to analyze hourly movement data
-% %%from 18 Florida panthers (\emph{Puma concolor coryi}).
-% 
-\
 
@@
 models with three states. 
-% (their criterion for choosing ...\todo{check \cite{zhao_knee_2008}!}).
 As van de Kerk et al. \cite{kerk2015hidden} comment, and as
@@
 identifying the number of latent states.
-
-\
 
@@
 
-\
-
 \section*{Methods}
 
-\
-
 \subsection*{Data and previous analyses}
-
-\
 
@@
 
-\
-
 van de Kerk et al.\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
@@
 to making precisely this assumption.
-
-\
 
@@
 
-\
-
 van de Kerk et al.\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
@@
 of underlying movement states.
-% \todo{it occurs to me that we haven't really included any
-% information about simulation testing in this version!
-% We should probably add a short section; if don't, then we
-% should delete following sentence }
+
 Using BIC also simplifies 
@@
 
-\
-
 \subsection*{Model description}
-
-\
 
@@
 
-\
-
 \begin{description}
@@
 
-\
-
 \item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\cov_{k}$ at time $t=k$.
-
-\
 
@@
 \end{description}
-
-\
 
@@
 
-\
-
 For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
@@
 
-\
-
 In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
@@
 we have only $n \times (n-1)$ distinct parameters; we index $j$ from 2 to $n$ for notational clarity):
-
-\
 
@@
 
-\
+\begin{description}
 
-\begin{description}
 \item[Multiple block transition] Here we assume piecewise-constant transition probabilities.
@@
 $$
+
 where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).
 
-\
+\item[Quadratic transition model] 
+We assume the elements of the linear predictor
+are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there 
+is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity 
+constraint would collapse the model to a constant.
 
-\item[Quadratic transition model] We assume the elements of the linear predictor
-are quadratic functions of hour.
-The quadratic model is not diurnally continuous, i.e.
-there is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity constraint would collapse the model to a constant.
 $$
@@
 \item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
+
 $$
@@
 
-\
-
 Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could be useful directions for future exploration.
 
-\
-
-\subsection{Model evaluation}
-
-\
+\subsection*{Model evaluation}
 
@@
 algorithm.
-
-\
 
@@
 of models captures serial correlation at both short and long scales.
-
-\
 
@@
 
-\
-
 We used simulations to predict hourly step length and ACF 
@@
 
-\
+\section*{Results}
 
-\section{Results}
-
-\
+%% fig 1 goes somewhere here
 
 We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.
-Fig.
-~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach.
+Heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach (Figure 1).
 
-\
+%% fig 2 goes somewhere here
 
-Fig.
-~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles)
-
-\
+The BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2)
 
@@
 
-\
+%% fig 3 goes somewhere here
 
-Figure~\ref{fig:adj_BIC_comparisons} shows that models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity.
+Models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3).
 Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
@@
 
-\
-
-Fig.
-~\ref{fig:avg_step_length_by_time} shows a clear diurnal pattern from the average hourly step lengths from the observed data.
+The panthers exhibits a clear diurnal pattern from the average hourly step lengths from the observed data (Figure 4).
 As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle.
@@
 
-\
-
-Like the diurnal pattern shown in Fig. ~\ref{fig:avg_step_length_by_time},
-the strong autocorrelation of the observed step lengths at a 24-hour lag (Fig.
-~\ref{fig:acf_plot}) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
+Like the diurnal pattern (Figure 4),
+the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 5) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
 Because there are a huge number of potential complexities that can be added to movement models (e.g.
@@
 FMM without temporal heterogeneity, although it captures the diurnal pattern well, underpredicts the degree of short-term autocorrelation.
-
-\
 
@@
 step length a traveling state.
-% \todo{do we need this?}
-Figure~\ref{fig:state_plots} compares the estimated parameter values
+
+The estimated parameter values
 for several cats
 (mean and standard deviation of the step length in each state) between
-the time-homogeneous and time-heterogeneous models. 
+the time-homogeneous and time-heterogeneous models are similar across all cats (Figure 7). 
 In general, the states with longer mean step lengths are relatively
@@
 
-\
-
-\section{Discussion}
-
-\
+\section*{Discussion}
 
@@
 number of states that are difficult to interpret biologically.
-
-\
 
@@
 but it can be useful as an approximate upper limit estimate.
-
-\
 
@@
 
-\
-
-\section{Conclusion}
-
-\
+\section*{Conclusion}
 
@@
 
-\section{Declarations}
+\section*{Declarations}
 
@@
 
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
-
-\section*{Figures}
-  \begin{figure}[h!]
-  \includegraphics[width=\maxwidth]{figure/sim_results-1}
-  \caption{\csentence{Sample figure title.}
-      A short description of the figure content
-      should go here.}
-      \end{figure}
-
-\begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      Figure legend text.}
-      \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use of \listoftables is discouraged.
-%%
-\section*{Tables}
-\begin{table}[h!]
-\caption{Sample table title. This is where the description of the table should go.}
-      \begin{tabular}{cccc}
-        \hline
-           & B1  &B2   & B3\\ \hline
-        A1 & 0.1 & 0.2 & 0.3\\
-        A2 & ... & ..  & .\\
-        A3 & ..  & .   & .\\ \hline
-      \end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
-
 \end{backmatter}

### 2016-08-03 bmc_template/manuscript.tex
tex problems

@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
-
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%          <8 June 2012>              %%
-%%                                     %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
-%% http://www.biomedcentral.com/info/authors/                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% http://www.miktex.org                                           %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
-%%% loading packages, author definitions
-
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
 \documentclass{bmcart}
@@
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+\usepackage{graphicx}
+% \def\includegraphic{}
+% \def\includegraphics{}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                             %%
-%%  If you wish to display your graphics for   %%
-%%  your own use using includegraphic or       %%
-%%  includegraphics, then comment out the      %%
-%%  following two lines of code.               %%
-%%  NB: These line *must* be included when     %%
-%%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
-%%  separate graphics through the BMC          %%
-%%  submission process, not included in the    %%
-%%  submitted article.                         %%
-%%                                             %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\def\includegraphic{}
-\def\includegraphics{}
-
-%%% Put your definitions there:
 \startlocaldefs
@@
 \begin{document}
+
+\titlepage
+\
 
@@
 step-length distributions, with and without 
-a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).
+a von Mises-distributed turning angle included in the model (Figure 2).
 (Note that most movement analyses, including van de Kerk et al.
@@
 \section{Data accessibility}
-Hourly step lengths and turning angles of male and female Florida pan- thers available at: http://ufdc.ufl.edu//IR00004241/00001.
+Hourly step lengths and turning angles of male and female Florida panthers available at: http://ufdc.ufl.edu//IR00004241/00001.
 
@@
   The authors declare that they have no competing interests.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
+\bibliography{panthers}      % Bibliography file (usually '*.bib' )
 % for author-year bibliography (bmc-mathphys or spbasic)

### 2016-08-03 bmc_template/manuscript.tex
I'll fill in additional files and sup
Minor touch ups on additional files later

@@
 Here we focus on the four cats with the most data (all with approximately 10,000-15,000
-observations: see Table~1 in Supplementary Material), 
+observations: see Table~1), 
 in part because our goal is to understand
@@
 
+\section*{Funding}
+ NSERC?
+
 \section{Data accessibility}
@@
 
+\section*{Figures}
+  \begin{figure}[h!]
+  % \includegraphics[width=\maxwidth]{figure/acf_plot15-1}
+  \caption{\csentence{Simulation Results} BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}
+      \end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{Sample figure title.}
+      Figure legend text.}
+      \end{figure}
+
+\section*{Tables}
+\begin{table}[h!]
+\caption{Cat ID}
+      \begin{tabular}{cccc}
+        \hline
+           & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
+         & 130 & 1 & 10286\\
+         & 131 & 2  & 9458\\
+         & 48 & 14  & 14645\\
+         & 94  & 15   & 10250\\ \hline
+      \end{tabular}
+\end{table}
+
 \end{backmatter}

### 2016-08-03 bmc_template/manuscript.tex
figures fixed

@@
   \begin{figure}[h!]
-  % \includegraphics[width=\maxwidth]{figure/acf_plot15-1}
+   \includegraphics[width=5in]{figure/sim_results-1}
   \caption{\csentence{Simulation Results} BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}
@@
   \begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      Figure legend text.}
+   \includegraphics[width=5in]{figure/BICred_plot-1}
+  \caption{\csentence{Adjusted BIC Emission Distribution Comparison} Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.}
+      \end{figure}
+
+  \begin{figure}[h!]
+   \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
+  \caption{\csentence{Adjusted BIC Transition Comparison} Adjusted BIC by number of free parameters for HMM model types.
+The left panel shows FMM, FMM with a sin prior and HMM.
+The right panel shows HMMs with different temporal transitions.}
+      \end{figure}
+
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/avg_step_length_by_time-1}
+  \caption{\csentence{Diurnal Step Lengths Plot} Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
+      \end{figure}
+
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/acf_plot-1}
+  \caption{\csentence{Autocorrelation Plot} Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
+      \end{figure}
+
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/r_msdlist-1}
+  \caption{\csentence{State Identification Plot} Log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.}
       \end{figure}

### 2016-08-08 bmc_template/manuscript.tex
acknowledgements

@@
 \usepackage{setspace}
-\doublespacing
 \usepackage{lineno}
+\usepackage{url}
 \linenumbers
-%\RequirePackage{natbib}
+\RequirePackage[sort]{natbib}
 %\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
@@
 
-\titlepage
-\
+%\setcounter{page}{0}  %% ugh, article template assumes starting on odd page!
+% \newpage
 
@@
 \dochead{Research}
+
+%% TO DO
+% author references (maybe fix programmatically?)
+% acknowledgments
 
@@
 \begin{artnotes}
- \note{Sample of title note}     % note to the article
- \note[id=n1]{Equal contributor} % note, connected to author
+% \note{Sample of title note}     % note to the article
+% \note[id=n1]{Equal contributor} % note, connected to author
 \end{artnotes}
@@
 \parttitle{Conclusion} %if any
-Incorportating temporal heterogeneity reduce the selected number of behavioural states closer 
+Incorporating temporal heterogeneity reduce the selected number of behavioural states closer 
 to a biologically interpretable level, improved goodness of fit and predictability. Our 
@@
 % Text and results for this section, as per the individual journal's instructions for authors. %\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}
+
+\doublespacing
 
@@
 we also re-analyze data from 
-van de Kerk et al.\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
+van de Kerk et al. \cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
 models (HSMMs: an extension of HMMs that allow
@@
 
-van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
+van de Kerk et al. \cite{kerk2015hidden} found that the best-fitting HSMMs
 incorporated a surprisingly large number of hidden
@@
 trained hounds and houndsmen. Of these animals, 13 had 
-sufficient data to be used by van de Kerk et al.\cite{kerk2015hidden}.
+sufficient data to be used by van de Kerk et al. \cite{kerk2015hidden}.
 Here we focus on the four cats with the most data (all with approximately 10,000-15,000
@@
 
-van de Kerk et al.\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
+van de Kerk et al. \cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
 an extension of HMM that permits explicit modelling of dwell times 
@@
 negative binomial distributions for dwell times.  As shown 
-by van de Kerk et al.\cite{kerk2015hidden} (Figure S3b, top row, middle panel),
+by van de Kerk et al. \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
 the estimated shape parameter of the negative binomial dwell time
@@
 
-van de Kerk et al.\cite{kerk2015hidden} considered time-homogeneous models with 
+van de Kerk et al. \cite{kerk2015hidden} considered time-homogeneous models with 
 a variety of candidate distributions %
@@
 
-van de Kerk et al.\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
+van de Kerk et al. \cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
 to test the relative penalized goodness of fit for
@@
 of their analyses for ease of biological interpretation.
-We follow van de Kerk et al.\cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
+We follow van de Kerk et al. \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
 BIC across a family of models) as the criterion for identifying
@@
 and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model
-parameters \boldmath{$\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
+parameters {\boldmath $\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
 
@@
 
-We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of Morales et al.\cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
+We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of Morales et al.
+\cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
 
@@
 
-As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas HMMs with temporal heterogenity will increase quadratically (Eq.~\ref{eq:numparam}).
+As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogeneity will increase linearly, whereas HMMs with temporal heterogeneity will increase quadratically (Eq.~\ref{eq:numparam}).
 When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
@@
 
-The BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2)
+The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2)
 
@@
 
-Models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3).
+Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3).
 Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
@@
 movement states seem biologically interpretable for Florida panthers
-according to van de Kerk et al.\cite{kerk2015hidden}: Short step length suggests resting
+according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
 states, intermediate step length a foraging state, and long
@@
 better for out of sample prediction because Viterbi can only estimate state 
-occupancy if observed movements are available (within sample preditions). 
+occupancy if observed movements are available (within sample predictions). 
 Finally, if we want to estimate the number of states, BIC is not necessarily 
@@
 
-Incorporating temporal heterogenity in animal movement is one step in
+Incorporating temporal heterogeneity in animal movement is one step in
 the right direction, but much remains to be done. 
@@
 
-\section*{Declarations}
-
 \begin{backmatter}
@@
 \section*{Acknowledgements}
-  Florida Wildlife and Fisheries?
+
+We would like to thank Madelon van de Kerk, Madan Oli, and David Onorato for their previous work on Florida panthers.
+We also would like to thank McMaster University, Florida Fish and Wildlife Conservation Commission and many individuals for data collection and fieldwork.
+Lastly, to Madelon van de Kerk for making the data available at the Institutional Repository at the University of Florida (IR@UF).
+
+\section*{Ethics approval}
+
+All data used are secondary, drawn from an existing institutional
+data repository.
+
+\section*{Consent for publication}
+
+Not applicable.
 
 \section*{Funding}
- NSERC?
+
+This study was funded by NSERC Discovery Grant 386590-2010 to BMB.
 
 \section{Data accessibility}
-Hourly step lengths and turning angles of male and female Florida panthers available at: http://ufdc.ufl.edu//IR00004241/00001.
+Hourly step lengths and turning angles of male and female Florida panthers available at: \url{http://ufdc.ufl.edu//IR00004241/00001}.
 
-\section*{Author's contributions}
-    Equally contributed.
+\section*{Authors' contributions}
+ML designed analyses and simulations; ran analyses and simulations; and co-wrote the text of the paper.  BMB designed analyses and simulations and co-wrote the text of the paper.
 
@@
 %\nocite{label}
+
+\section*{Tables}
+\begin{table}[h!]
+\caption{Cat ID}
+      \begin{tabular}{cccc}
+        \hline
+           & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
+         & 130 & 1 & 10286\\
+         & 131 & 2  & 9458\\
+         & 48 & 14  & 14645\\
+         & 94  & 15   & 10250\\ \hline
+      \end{tabular}
+\end{table}
 
@@
 
-\section*{Tables}
-\begin{table}[h!]
-\caption{Cat ID}
-      \begin{tabular}{cccc}
-        \hline
-           & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
-         & 130 & 1 & 10286\\
-         & 131 & 2  & 9458\\
-         & 48 & 14  & 14645\\
-         & 94  & 15   & 10250\\ \hline
-      \end{tabular}
-\end{table}
-
 \end{backmatter}

### 2016-08-10 bmc_template/manuscript.tex
working on ms

@@
 \startlocaldefs
+\newcommand{\obs}{\mathbf{y}}
+\newcommand{\cov}{\mathbf{x}}
+\newcommand{\state}{z}
+\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}
+\newcommand{\bmb}[1]{\textbf{[BMB: #1]}}
 \endlocaldefs
@@
 
-\parttitle{Background} %if any
+\parttitle{Background} 
 Clustering time-series data into discrete groups can improve 
 prediction and provide insight into the nature of underlying,
-unobservable states of the system. However, temporal variation in the
-rates at which individuals move between groups can obscure
+unobservable states of the system. However, temporal variation in 
+probabilities of group occupancy, or the rates at which individuals move between groups, can obscure
 such signals. We use finite mixture and hidden Markov models (HMMs), two
-standard clustering techniques, to model high-resolution hourly
+standard clustering techniques, to model long-term hourly
 movement data from Florida panthers (\emph{Puma concolor coryi}).
@@
 
-\parttitle{Results} %if any
-Simulations and Florida panthers data showed model misspecification (omitting important 
-sources of variation) can lead to overfitting and over-estimating number of behavioural states.
-Models incorporating temporal heterogeneity have lower number of states with slightly higher variation in short movement states, and able to make out of sample predictions that captures observed diurnal and autocorrelation patterns exhibited by Florida panthers.
+\parttitle{Results}
 
-\parttitle{Conclusion} %if any
-Incorporating temporal heterogeneity reduce the selected number of behavioural states closer 
-to a biologically interpretable level, improved goodness of fit and predictability. Our 
-suggest that incorporating previously neglected structure in statistical models can
+Simulations and analyses of panther data showed that model
+misspecification (omitting important sources of variation) can lead to
+overfitting/overestimating the underlying number of behavioural
+states.  Models incorporating temporal heterogeneity identify fewer
+underlying states, and can make out-of-sample predictions that
+capture observed diurnal and autocorrelated movement patterns exhibited by
+Florida panthers.
+
+\parttitle{Conclusion}
+
+Incorporating temporal heterogeneity improved goodness of fit and predictive
+capability as well as
+reducing the selected number of behavioural states 
+to a more biologically interpretable level. Our results
+suggest that incorporating additional structure in statistical models 
+of movement behaviour can
 allow more accurate assessment of appropriate model complexity.
@@
 Conditional on the state occupied by an individual, HMMs typically
-assume that animals follow a standard correlated walk model
+assume that animals follow a standard correlated random walk model
 \cite{okubo_diffusion_1980,turchin1998quantitative}.
@@
 scales, where the probability of transitioning between movement states is
-approximately constant. Changes in latent/hidden behavioural state/mode transition 
-% BMB: pick one (perhaps include a parenthetical statement/definition
-% saying that we treat these as a synonyms.  Just stringing them together
-% with slashes is awkward)
+approximately constant. Changes in transition
 probabilities based on
-the local environment can be accounted for incorporating environmental
-covariates in the HMM \cite{patterson_classifying_2009}, or by more
-direct comparisons between inferred states and environmental
+the local environment can be accounted for by incorporating environmental
+covariates in the HMM \cite{patterson_classifying_2009}, or 
+inferred from direct comparisons between inferred states and environmental
 conditions \cite{fryxell_multiple_2008}. Schliehe-Diecks et al. 
 \cite{schliehe-diecks_application_2012}
-consider temporal trends in behavioural transitions over the
+considered temporal trends in behavioural transitions over the
 time scales of a six-hour
@@
 \cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
-In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
-states, but like \cite{kerk2015hidden} they chose to consider only 
+In contrast, Dean et al.~\cite{dean_behavioural_2013} evaluated models with up to 10
+states, but like van de Kerk et al. they chose to consider only 
 models with three states. 
@@
 repertoires with more than three distinct states are difficult to
-interpret --- one possible reason that other authors have not adopted
+interpret --- one reason that other authors have not adopted
 van de Kerk et al.'s model-based approach to
@@
 small models to sparse data into such traps.  We examine whether
-allowing for diurnal variation in the Florida panther data leads to
-selection of models with smaller numbers of latent states; we also
-fit models to simulated data with varying numbers of latent states
-and degrees of temporal heterogeneity to test our conjecture that
+allowing for diurnal variation in the Florida panther data allows us to
+select models with fewer latent states; we also
+fit models to simulated data with varying numbers of latent states,
+and with and without temporal heterogeneity, to test our conjecture that
 heterogeneity can be misidentified as behavioural complexity.
@@
 
-\newcommand{\obs}{\mathbf{y}}
-\newcommand{\cov}{\mathbf{x}}
-\newcommand{\state}{z}
-
 In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $ \mathbf Y = \obs_{1},..., \obs_{T}$ 
@@
 \begin{description}
-\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$, times the vector of observations $\obs_{1}$ conditioned on covariates $\cov_{1}$ and state $\state_{1}$.
+\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ given that the covariates are $\cov_{1}$, times the vector of observations $\obs_{1}$ conditioned on state $\state_{1}$ and covariates $\cov_{1}$.
 
-\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\cov_{k}$ at time $t=k$.
+\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ at time $t=k$, given covariates $\cov_{k}$ .
 
-\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observations $\obs_{k}$ conditioned on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
+\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observations $\obs_{k}$ given state $\state_{k}$ at time $t=k$ and covariates $\cov_{k}$ .
 \end{description}
@@
 
-For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
+For an $n$-state HMM, we need to define an $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
 The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy.
@@
 
-where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).
+where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block corresponding to time $t$, and 0 otherwise).
 
@@
 We assume the elements of the linear predictor
-are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there 
+are quadratic functions of hour:
+
+$$
+\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2} .
+$$
+
+The quadratic model is not diurnally continuous, i.e. there 
 is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity 
@@
 
-$$
-\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
-$$
-
-\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
+\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint:
 
 $$
-\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
+\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right) .
 $$
@@
 \item[Hourly model] Lastly, we extended the multi-block approach and assign  a different transition matrix for every hour of the day.
-This model is included for comparative purposes due to the large number of parameters in the model which makes it not really practical.
+This model is included for comparative purposes; due to the large number of parameters in the model (more than $24n(n-1)$ for a HMM with $n$ states), it is not really practical.
 We only fitted up to four states using the hourly model.
@@
 
-Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could be useful directions for future exploration.
+Other periodic functions, such as Fourier series (i.e., the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could also be considered.
 
@@
 
-\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}
-
 We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
@@
 with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model 
-within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall 
+within each type of transition complexity \bmb{clarify?} ($\dbic$ = BIC - min(BIC)) assesses the overall 
 goodness of fit of each model type. (2) Comparing average step-length by hour of day for the 
 observed data and for data simulated from the models shows how well a particular class of 
-models can capture the diurnal variation in behaviour. (3) Comparing temporal autocorrelations
+models can capture diurnal variation in behaviour. (3) Comparing temporal autocorrelations
 for the observed data and for data simulated from the models shows how well a particular class
-of models captures serial correlation at both short and long scales.
+of models can capture serial correlation at both short and long time scales.
 
-Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity.
-The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components.
+Model complexity and the number of parameters increase as the number of latent states increase.
+For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity.
+In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components.
 Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$.
@@
 \begin{equation}
-\textrm{Number of Free Parameter}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
+\textrm{Number of Free Parameters}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
 {+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
-{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
+{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}} .
 \label{eq:numparam}
@@
 
-As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogeneity will increase linearly, whereas HMMs with temporal heterogeneity will increase quadratically (Eq.~\ref{eq:numparam}).
+As the number of states increases, the number of free parameters in (homogeneous or heterogenous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}).
 When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
+\bmb{what does last sentence mean??
+delete?}
 
-We used simulations to predict hourly step length and ACF 
-because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
-We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions, which use the most probable sequence of movement states based on the observations \cite{zucchini_hidden_2009,langrock_flexible_2012}, double-count the observed data.
-It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not reliably evaluate goodness of fit for the different structural complexities of HMM models.
+We used simulations to predict expected hourly step lengths and
+autocorrelation functions (ACF).  While the computation of expected
+step length and ACF is straightforward for FMMs, and feasible for
+homogeneous HMMs, the interaction between the geometric dwell time
+within each state and the temporally varying interaction probabilities
+makes it infeasible for more complex models.  We used this approach to
+validate our models, comparing our simulated predictions with the
+observed movements.  The more usual approach, generating predictions
+from the expected step lengths conditional on the most likely state
+sequence predicted by the Viterbi algorithm \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
+because the states predicted by the Viterbi algorithm already rely on the observed data. This approach is useful to predict
+missing data in the observation sequence, but because it is
+conditional on the observed values, it can not reliably evaluate
+goodness of fit for the different structural complexities of HMM
+models.
+\bmb{we might need to be more careful here. Apparently Zucchini/Langrock et al also define ``pseudo-residuals'';
+can you look these up and see if that approach suffers from the same issues as Viterbi ... ?}
 
@@
 Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
-However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
+However, this similarity is overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
 normally be interpreted as an enormous difference in goodness of fit,
@@
 
-The panthers exhibits a clear diurnal pattern from the average hourly step lengths from the observed data (Figure 4).
+The average hourly step lengths from the observed panther data exhibit a clear diurnal pattern (Figure 4).
 As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle.
 All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns.
-However, the block model does markedly worse than the other temporal models (changing the block definitions might help), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
-We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points).
+However, the block model does markedly worse than the other temporal models (changing the block definitions might help \bmb{clarify?}), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
+We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points \bmb{clarify?}).
 
@@
 We also would like to thank McMaster University, Florida Fish and Wildlife Conservation Commission and many individuals for data collection and fieldwork.
-Lastly, to Madelon van de Kerk for making the data available at the Institutional Repository at the University of Florida (IR@UF).
+Lastly, we thank Madelon van de Kerk for making the data available at the Institutional Repository at the University of Florida (IR@UF).
 
@@
 \begin{table}[h!]
-\caption{Cat ID}
+\caption{Cat ID and number of observations; ID numbers are given matching
+those shown by van de Kerk et al. 2014 and those 
+in the data located at the UF Institutional repository (IR@UF).}
       \begin{tabular}{cccc}

### 2016-08-10 bmc_template/manuscript.tex
a little more ms stuff

@@
 
-We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.
-Heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach (Figure 1).
+\bmb{first sentence belongs in ``Methods'', with a few more words of motivation/description. I know that makes the Results for this part very short, but that's OK}
+
+We simulated 100 realizations of a two-state HMM with sinusoidal temporal transitions and fitted it with 2- to 5-state HMMs with and without temporal heterogeneity in the transition probabilities.
+Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).
 
@@
 
-The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2)
+The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2).
+\bmb{Clarify parenthetical clause here. Also, more discussion: in particular, discuss/clarify the points that the homogeneous-HMM models do vary a bit among models
+(log-Normal vs Weibull, with/without turning angles) in how many states they identify, but the heterogeneous-HMM ones don't ---
+which is more important for our conclusions here}
 
+\bmb{doesn't this para. belong in methods??}
 As a complement, we also fitted FMM and FMM with sinusoidal variation in state occupancy probabilities to compare the temporal effects in goodness of fit (dashed lines).
@@
 Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3).
-Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
+Turning to the temporally heterogeneous HMMs (Figure 3, right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
 The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.
@@
 spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.
-In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects.
+\bmb{move prev sentence to Discussion??} In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects.
 HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.

### 2016-08-10 bmc_template/figures.Rnw
BMB done for now

@@
                          labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
-simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model))+
-                        geom_point(size=3) +
+simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
     geom_line() +
+    geom_point(size=3,bg="white") +
+    scale_shape_manual(values=c(21,16))+
     scale_colour_brewer(palette="Set1")+
-                        labs(y="Counts",x="BIC-optimal States")
+    labs(y="Counts",x="BIC-optimal number of states")
 print(simplot)
@@
             turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
-            homog=ifelse(grepl("TH",type),"Time-inhomogeneous HMM","Homogeneous HMM"))
-(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
+            homog=ifelse(grepl("TH",type),"Time-heterogeneous","Homogeneous"))
+slname <- "Step-length\ndistribution"
+(ggplot(df2, aes(x=nstates,y=deltaBIC,size=distrib,
+                 colour=distrib,lty=turn_angle))
     + geom_line()
     + facet_wrap(~ homog)
-    + scale_colour_brewer(name="Step-length\ndistribution",palette="Set1")
+    + scale_colour_brewer(name=slname,palette="Set1")
+    + scale_size_manual(name=slname,values=c(1.75,0.75))
     ## scale_size_continuous( name="# parameters") +
@@
 <<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types.
-The left panel shows FMM, FMM with a sin prior and HMM.
+The left panel shows FMM, temporally heterogeneous FMM with a sinusoidal prior and HMM.
 The right panel shows HMMs with different temporal transitions.",fig.width=8>>=
@@
             "hmmquad5","hmmsin5","obs","viterbi")
-mnames2 <- c("fmm-homogeneous  (n=5)"
-             , "fmm-sin (n=4)"
-             , "homogeneous (n=6)"
-             , "block (n=4)"
-             , "hourly (n=3)"
-             , "quadratic (n=5)"
-             , "sin (n=5)"
+mnames2 <- c("FMM-homogeneous  (n=5)"
+             , "FMM-sin (n=4)"
+             , "HMM-homogeneous (n=6)"
+             , "HMM-block (n=4)"
+             , "HMM-hourly (n=3)"
+             , "HMM-quadratic (n=5)"
+             , "HMM-sin (n=5)"
              , "observed"
@@
 all_cat_pars2 <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars2)
-
-g0 <- ggplot(all_cat_pars2,aes(log_mean,log_sd,color=Model))+
-     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
+levels(all_cat_pars2$Model) <- c("homogeneous","heterogeneous (sin)")
+g0 <- ggplot(all_cat_pars2,
+             aes(10^log_mean,log_sd*10^log_mean,color=Model,shape=Model))+
+    scale_x_log10()+scale_y_log10()+
+    labs(x="mean step length (m)",y="sd of step length (m)")+
+    scale_shape_manual(values=c(21,16)) +
+    geom_line() +
+    geom_point(bg="white")+
+    facet_wrap(~catID,labeller=label_both)+zmargin
  (g0c <- g0 +
@@
                                (catID!="15" & log_mean<1.8)),
-                   colour="black",linetype=2,expand=0.05))
-
+                   colour="black",linetype=2,expand=0.05,
+                   group=1))
 @

### 2016-08-10 bmc_template/manuscript.tex
BMB done for now

@@
 \bmb{doesn't this para. belong in methods??}
-As a complement, we also fitted FMM and FMM with sinusoidal variation in state occupancy probabilities to compare the temporal effects in goodness of fit (dashed lines).
+As a complement, we also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to compare the temporal effects in goodness of fit (dashed lines).
 % \todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
@@
 Turning to the temporally heterogeneous HMMs (Figure 3, right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
-The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.
-Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
+The multiple-block model gives approximately the same BIC as the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.
+Finally, the quadratic and sinusoidal models are the best models tested by far; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
 However, this similarity is overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
@@
 HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
-Temporally homogeneous FMM, by definition, produces neither short- nor long-term autocorrelation.
-FMM without temporal heterogeneity, although it captures the diurnal pattern well, underpredicts the degree of short-term autocorrelation.
+Temporally homogeneous FMMs, by definition, produce no autocorrelation (neither short- nor long-term autocorrelation).
+FMMs without temporal heterogeneity, although they capture the diurnal pattern well, underpredict the degree of short-term autocorrelation.
 
+\bmb{next para might belong in Discussion?}
 The hardest problem with multiple latent states is interpreting them
@@
 
-The estimated parameter values
-for several cats
-(mean and standard deviation of the step length in each state) between
-the time-homogeneous and time-heterogeneous models are similar across all cats (Figure 7). 
-In general, the states with longer mean step lengths are relatively
-similar between model classes. For cats 14 and 15, the states with
+The estimated emission parameter values
+(mean and standard deviation of the step length in each state)
+are similar, for both homogeneous and heterogeneous
+models, across all cats (Figure 7 shows a subset of cats). 
+In general, the states with longer mean step lengths are
+similar between homogeneous and heterogeneous models. For cats 14 and 15, the states with
 the longest or next-longest mean step lengths have similar means
@@
 
+\bmb{where did the BIC stuff (Celeux, finding the ``knee point''
+à la Dean et al., etc.) go? I think we really have to mention this
+(maybe not until discussion)}
 \section*{Conclusion}
@@
 \section{Data accessibility}
-Hourly step lengths and turning angles of male and female Florida panthers available at: \url{http://ufdc.ufl.edu//IR00004241/00001}.
+Hourly step lengths and turning angles of male and female Florida panthers available at \url{http://ufdc.ufl.edu//IR00004241/00001}.
 
@@
 \caption{Cat ID and number of observations; ID numbers are given matching
-those shown by van de Kerk et al. 2014 and those 
+those shown by van de Kerk et al. 2015 and those 
 in the data located at the UF Institutional repository (IR@UF).}
@@
    \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
-  \caption{\csentence{Adjusted BIC Transition Comparison} Adjusted BIC by number of free parameters for HMM model types.
-The left panel shows FMM, FMM with a sin prior and HMM.
+  \caption{\csentence{Adjusted BICs Across All Models} Adjusted BIC by number of free parameters for HMM model types.
+The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM.
 The right panel shows HMMs with different temporal transitions.}
@@
    \includegraphics[width=5in]{figure/r_msdlist-1}
-  \caption{\csentence{State Identification Plot} Log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.}
+  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
+\bmb{do I need to explain transformation of LN SD ?}}
       \end{figure}

### 2016-08-29 bmc_template/manuscript.tex
not done, continue on line 253 in the am

@@
 
+A simple simulation was conducted to to see whether fitting homogeneous transition probability HMMs to heterogeneous transitions simulated data will over estimate the number of states and estimate the correct number of states using the heterogeneous transitioning model.
+We used 100 realizations of a two-state HMM with sinusoidal temporal transitions and fitted it with 2- to 5-state HMMs with and without temporal heterogeneity in the transition probabilities.
+
 We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
 HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
-with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model 
-within each type of transition complexity \bmb{clarify?} ($\dbic$ = BIC - min(BIC)) assesses the overall 
-goodness of fit of each model type. (2) Comparing average step-length by hour of day for the 
+with the most data ($> 9000 $ observations).
+(1) BIC was used to compare the goodness of fit of each model type.
+The model with the lowest BIC was selected to be the optimal-BIC model and all BICs were adjusted to $\dbic$ based on the optimal-BIC model ($\dbic$ = BIC - min(BIC)) .
+(2) Comparing average step-length by hour of day for the
 observed data and for data simulated from the models shows how well a particular class of 
@@
 of models can capture serial correlation at both short and long time scales.
+As a complement, we also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to compare the temporal effects in goodness of fit As a reminder, FMMs assume that the latent state in each time step is \emph{independent} of the latent state at the previous time step; time-varying FMMs can accurately describe movement when behaviour can
+change on a short time scale, but the average propensity for different
+behaviours changes over time.
 
@@
 When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
-\bmb{what does last sentence mean??
-delete?}
 
@@
 from the expected step lengths conditional on the most likely state
-sequence predicted by the Viterbi algorithm \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
-because the states predicted by the Viterbi algorithm already rely on the observed data. This approach is useful to predict
+sequence predicted by the Viterbi algorithm or pseduo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
+because the predictions by these methods rely on the observed data. These approach is useful to predict
 missing data in the observation sequence, but because it is
@@
 models.
-\bmb{we might need to be more careful here. Apparently Zucchini/Langrock et al also define ``pseudo-residuals'';
-can you look these up and see if that approach suffers from the same issues as Viterbi ... ?}
 
@@
 %% fig 1 goes somewhere here
-
-\bmb{first sentence belongs in ``Methods'', with a few more words of motivation/description. I know that makes the Results for this part very short, but that's OK}
-
-We simulated 100 realizations of a two-state HMM with sinusoidal temporal transitions and fitted it with 2- to 5-state HMMs with and without temporal heterogeneity in the transition probabilities.
+The simulation agrees with our hypothesis that homogeneous transition HMMs can over estimate the number of hidden states.
 Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).
@@
 which is more important for our conclusions here}
-
-\bmb{doesn't this para. belong in methods??}
-As a complement, we also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to compare the temporal effects in goodness of fit (dashed lines).
-% \todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
-As a reminder, FMMs assume that the latent state in each time step is 
-\emph{independent} of the latent state at the previous time step; 
-time-varying FMMs can accurately describe movement when behaviour can
-change on a short time scale, but the average propensity for different
-behaviours changes over time.
 

### 2016-08-29 bmc_template/manuscript.tex
need to finish up sup plots

@@
 \cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
-In contrast, Dean et al.~\cite{dean_behavioural_2013} evaluated models with up to 10
+In contrast, Dean et al.~\cite{dean2012behavioural} evaluated models with up to 10
 states, but like van de Kerk et al. they chose to consider only 
@@
 
-As the number of states increases, the number of free parameters in (homogeneous or heterogenous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}).
+As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}).
 When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
@@
 
-The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2).
-\bmb{Clarify parenthetical clause here. Also, more discussion: in particular, discuss/clarify the points that the homogeneous-HMM models do vary a bit among models
-(log-Normal vs Weibull, with/without turning angles) in how many states they identify, but the heterogeneous-HMM ones don't ---
-which is more important for our conclusions here}
+The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results.
+For time homogeneous models, the Weibull-wrapped-Cauchy \cite{kerk2015hidden}, Weibull von Mises, and log Normal without turning angles all identify the same BIC-optimal number of states.
+Furthermore, the homogeneous-HMM models do vary among models (different step-length distributions, with/without turning angles) in how many states they identify, whereas, the heterogeneous-HMM models are consistent among models (Figure 2).
 
@@
 All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns.
-However, the block model does markedly worse than the other temporal models (changing the block definitions might help \bmb{clarify?}), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
-We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points \bmb{clarify?}).
+However, the block model does markedly worse than the other temporal models (changing the block definitions might help by re-clustering/grouping different hours or increasing the number of blocks), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
+We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction to illustrate within sample predictions can capture the diurnal patterns, but fail to capture out of sample predictions.
 
@@
 the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 5) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
-Because there are a huge number of potential complexities that can be added to movement models (e.g.
-spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.
-\bmb{move prev sentence to Discussion??} In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects.
+In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects.
 HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
@@
 FMMs without temporal heterogeneity, although they capture the diurnal pattern well, underpredict the degree of short-term autocorrelation.
-
-\bmb{next para might belong in Discussion?}
-The hardest problem with multiple latent states is interpreting them
-biologically. We have no way of knowing what panthers are actually
-thinking (it is certainly more complex than being in one of a small
-number of discrete latent states); 
-we don't know the ``true'' number of latent states,
-nor are we able to observe them directly, although
-incorporating additional direct observations of 
-behaviour (if available) can at least partially
-address this problem \cite{fryxell_multiple_2008}.
-Three distinct
-movement states seem biologically interpretable for Florida panthers
-according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
-states, intermediate step length a foraging state, and long
-step length a traveling state.
 
@@
 match Viterbi-based states estimates with environmental variation 
-\cite{fryxell_multiple_2008}. On the other side of the 
+\cite{fryxell_multiple_2008}. In our panther study, it is really hard to interpret multiple latent states biologically.
+We have no way of knowing what panthers are actually
+thinking (it is certainly more complex than being in one of a small
+number of discrete latent states); 
+we don't know the ``true'' number of latent states,
+nor are we able to observe them directly, although
+incorporating additional direct observations of 
+behaviour (if available) can at least partially
+address this problem \cite{fryxell_multiple_2008}.
+Three distinct
+movement states seem biologically interpretable for Florida panthers
+according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
+states, intermediate step length a foraging state, and long
+step length a traveling state.
+
+On the other side of the 
 spectrum, if the goal of interest is to make predictions (out of sample), 
@@
 occupancy if observed movements are available (within sample predictions).
+
 Finally, if we want to estimate the number of states, BIC is not necessarily 
@@
 but it can be useful as an approximate upper limit estimate.
+This is a common problem when using BICs, thus, the "knee point" (jump point) was proposed for BIC in partitioning based clustering via angle-based method \cite{zhao_knee_2008}.
+A similar method was used with likelihood \cite{dean2012behavioural} where both are applicable in our case, and suggest fewer number of states, but we do not know the reliability when fitting more complex models (ie.
+the shape of the BIC curve can be affected by the non-linear increase of parameters with every additional state).
 
@@
 include all possible features of interest.
+Because there are a huge number of potential complexities that can be added to movement models (e.g.
+spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.
 
-\bmb{where did the BIC stuff (Celeux, finding the ``knee point''
-à la Dean et al., etc.) go? I think we really have to mention this
-(maybe not until discussion)}
 \section*{Conclusion}

### 2016-08-29 bmc_template/figures.Rnw
finish additional file edits

--- 2016-08-02 bmc_template/figures.Rnw
@@
 \begin{document}
+
+\section*{Additional files}
+
+  In the main text of this manuscript, we showed the fitting and simulating results for Cat 1.
+Here, we have the adjusted BIC emission distribution comparisons, adjusted BIC across all FMM and HMM models, diurnal step lengths and autocorrelation plots for Cat 2, 14, and 15.
+Given we have not account for sex, we can see in that our fitting and predictions are better for male cats (Cat 1 and 2) than female cats (Cat 14, 15).
+
+\newpage
 
@@
 
-<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Simulation test for BIC-optimal approach",fig.width=6,fig.height=4>>=
+<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.",fig.width=6,fig.height=4,eval=FALSE>>=
 ICdat <- readRDS("../sim_results/ICdat.rds")
@@
                          levels=c("HMM","HMMsin"),
-                         labels=c("homogeneous","heterogeneous"))) %>% filter(IC=="BIC")
-simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,
-                            linetype=IC,shape=IC))+
-                        geom_point(size=3) +
+                         labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
+simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
     geom_line() +
+    geom_point(size=3,bg="white") +
+    scale_shape_manual(values=c(21,16))+
     scale_colour_brewer(palette="Set1")+
-                        labs(y="counts",x="# latent states")
-print(simplot)
+    labs(y="Counts",x="BIC-optimal number of states")
+# print(simplot)
 
@@
 
-\clearpage
-
-<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
+<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4,eval=FALSE>>=
 allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
@@
             turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
-            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
-(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
+            homog=ifelse(grepl("TH",type),"Time-heterogeneous","Homogeneous"))
+slname <- "Step-length\ndistribution"
+(ggplot(df2, aes(x=nstates,y=deltaBIC,size=distrib,
+                 colour=distrib,lty=turn_angle))
     + geom_line()
     + facet_wrap(~ homog)
-    + scale_colour_brewer(name="step length\ndistribution",palette="Set1")
+    + scale_colour_brewer(name=slname,palette="Set1")
+    + scale_size_manual(name=slname,values=c(1.75,0.75))
     ## scale_size_continuous( name="# parameters") +
@@
 
-\clearpage
-
-<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
+<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types.
+The left panel shows FMM, temporally heterogeneous FMM with a sinusoidal prior and HMM.
+The right panel shows HMMs with different temporal transitions.",fig.width=8,eval=FALSE>>=
 #L <- load("cat1sumdat.RData")
@@
    + zmargin)
+
+# print(adj_BIC_plot)
 @
 
-\clearpage
-
-<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=
+<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.",fig.width=8,eval=FALSE>>=
 
@@
 
-(ggplot(combdf,aes(x=time,y=steplen,colour=model,
+(avg_plot <- ggplot(combdf,aes(x=time,y=steplen,colour=model,
                    lwd=model,
@@
  )
+
+# print(avg_plot)
 @
 
-\clearpage
-
-<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE>>=
+<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ",eval=FALSE>>=
 acffun <- function(sim){
@@
 
-(ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
+(acf_plot <- ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
                   lwd=model,alpha=model,linetype=model,shape=model))
@@
 ) 
-
+# print(acf_plot)
 ## vitacf messes up legend ...
@@
 
-\clearpage
-
-<<r msdlist,message=FALSE, echo=FALSE, warning=FALSE>>=
+<<r msdlist,message=FALSE, echo=FALSE, fig.cap="State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.", warning=FALSE,eval=FALSE>>=
 get_cat_pars2 <- function(catid){
@@
 all_cat_pars2 <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars2)
-
-g0 <- ggplot(all_cat_pars2,aes(log_mean,log_sd,color=Model))+
-     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
- (g0c <- g0 +
+levels(all_cat_pars2$Model) <- c("homogeneous","heterogeneous (sin)")
+g0 <- ggplot(all_cat_pars2,
+             aes(10^log_mean,log_sd*10^log_mean,color=Model,shape=Model))+
+    scale_x_log10()+scale_y_log10()+
+    labs(x="mean step length (m)",y="sd of step length (m)")+
+    scale_shape_manual(values=c(21,16)) +
+    geom_line() +
+    geom_point(bg="white")+
+    facet_wrap(~catID,labeller=label_both)+zmargin
+(all_msd_plot <- g0 +
      geom_encircle(data=subset(all_cat_pars2,
@@
                                (catID!="15" & log_mean<1.8)),
-                   colour="black",linetype=2,expand=0.05))
+                   colour="black",linetype=2,expand=0.05,
+                   group=1))
 
+# print(all_msd_plot)
 @
 
-\clearpage
-
-\section{sup}
-
-<<BICred_plot2,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
+<<BICred_plot2,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4,fig.cap="Adjusted BIC Emission Distribution Comparison for Cat 2">>=
 
@@
 \clearpage
-
-<<adj_BIC_comparisons2, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
+<<adj_BIC_comparisons2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BICs Across All Models for Cat 2",fig.width=8>>=
 #L <- load("cat1sumdat.RData")
@@
 
-<<avg_step_length_by_time2, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=
+<<avg_step_length_by_time2, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8,fig.cap="Diurnal Step Lengths Plot for Cat 2">>=
 
@@
 
-<<acf_plot2, message=FALSE, echo=FALSE, warning=FALSE>>=
+<<acf_plot2, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation Plot for Cat 2">>=
 acffun <- function(sim){
@@
 
-<<BICred_plot14,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
+<<BICred_plot14,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4,fig.cap="Adjusted BIC Emission Distribution Comparison for Cat 14">>=
 
@@
 
-<<adj_BIC_comparisons14, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
+<<adj_BIC_comparisons14, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8,fig.cap="Adjusted BICs Across All Models for Cat 14">>=
 #L <- load("cat1sumdat.RData")
@@
 
-<<avg_step_length_by_time14, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=
+<<avg_step_length_by_time14, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8,fig.cap="Diurnal Step Lengths Plot for Cat 14">>=
 
@@
 
-<<acf_plot14, message=FALSE, echo=FALSE, warning=FALSE>>=
+<<acf_plot14, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation Plot for Cat 14">>=
 acffun <- function(sim){
@@
 
-<<BICred_plot15,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
+<<BICred_plot15,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4,fig.cap="Adjusted BIC Emission Distribution Comparison for Cat 15">>=
 
@@
 
-<<adj_BIC_comparisons15, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
+<<adj_BIC_comparisons15, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8, fig.cap="Adjusted BICs Across All Models for Cat 15">>=
 #L <- load("cat1sumdat.RData")
@@
 
-<<avg_step_length_by_time15, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=
+<<avg_step_length_by_time15, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8,fig.cap="Diurnal Step Lengths Plot for Cat 15">>=
 
@@
 
-<<acf_plot15, message=FALSE, echo=FALSE, warning=FALSE>>=
+<<acf_plot15, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation Plot for Cat 15">>=
 acffun <- function(sim){

### 2016-08-29 bmc_template/figures.tex
BMB done for now
finish additional file edits

--- 2016-08-02 bmc_template/figures.tex
@@
 
+\section*{Additional files}
+
+  In the main text of this manuscript, we showed the fitting and simulating results for Cat 1.
+Here, we have the adjusted BIC emission distribution comparisons, adjusted BIC across all FMM and HMM models, diurnal step lengths and autocorrelation plots for Cat 2, 14, and 15.
+Given we have not account for sex, we can see in that our fitting and predictions are better for male cats (Cat 1 and 2) than female cats (Cat 14, 15).
+
+\newpage
+
 \begin{knitrout}
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
-\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[Simulation test for BIC-optimal approach]{Simulation test for BIC-optimal approach}\label{fig:sim_results}
+\includegraphics[width=\maxwidth]{figure/BICred_plot2-1} \caption[Adjusted BIC Emission Distribution Comparison for Cat 2]{Adjusted BIC Emission Distribution Comparison for Cat 2}\label{fig:BICred_plot2}
+\end{figure}
+
+\end{knitrout}
+
+\clearpage
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons2-1} \caption[Adjusted BICs Across All Models for Cat 2]{Adjusted BICs Across All Models for Cat 2}\label{fig:adj_BIC_comparisons2}
 \end{figure}
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time2-1} \caption[Diurnal Step Lengths Plot for Cat 2]{Diurnal Step Lengths Plot for Cat 2}\label{fig:avg_step_length_by_time2}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/r_msdlist-1}
-
-\end{knitrout}
-
-\clearpage
-
-\section{sup}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot2-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons2-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time2-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot2-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/acf_plot2-1} \caption[Autocorrelation Plot for Cat 2]{Autocorrelation Plot for Cat 2}\label{fig:acf_plot2}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/BICred_plot14-1} \caption[Adjusted BIC Emission Distribution Comparison for Cat 14]{Adjusted BIC Emission Distribution Comparison for Cat 14}\label{fig:BICred_plot14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons14-1} \caption[Adjusted BICs Across All Models for Cat 14]{Adjusted BICs Across All Models for Cat 14}\label{fig:adj_BIC_comparisons14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time14-1} \caption[Diurnal Step Lengths Plot for Cat 14]{Diurnal Step Lengths Plot for Cat 14}\label{fig:avg_step_length_by_time14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/acf_plot14-1} \caption[Autocorrelation Plot for Cat 14]{Autocorrelation Plot for Cat 14}\label{fig:acf_plot14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/BICred_plot15-1} \caption[Adjusted BIC Emission Distribution Comparison for Cat 15]{Adjusted BIC Emission Distribution Comparison for Cat 15}\label{fig:BICred_plot15}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons15-1} \caption[Adjusted BICs Across All Models for Cat 15]{Adjusted BICs Across All Models for Cat 15}\label{fig:adj_BIC_comparisons15}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time15-1} \caption[Diurnal Step Lengths Plot for Cat 15]{Diurnal Step Lengths Plot for Cat 15}\label{fig:avg_step_length_by_time15}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/acf_plot15-1} \caption[Autocorrelation Plot for Cat 15]{Autocorrelation Plot for Cat 15}\label{fig:acf_plot15}
+\end{figure}
 

### 2016-09-08 bmc_template/manuscript.tex
working on ms, deleted bmc_article

@@
 Conditional on the state occupied by an individual, HMMs typically
-assume that animals follow a standard correlated random walk model
+assume that animals follow a correlated random walk model
 \cite{okubo_diffusion_1980,turchin1998quantitative}.
@@
 from GPS collars vs. daily or weekly fixes for radio or VHF
-collars) and longer extent (e.g. from a few days to significant
-fractions of a year, or longer).  When analyzing such
+collars) and longer extent (e.g. from a few days to months or years).  When analyzing such
 long-term data, ecologists will more
@@
 information-theoretic model selection methods will
-typically choose complex, highly parameterized models; when there is
+typically choose complex, highly parameterized models. When there is
 only one way in which models can become more complex (e.g. by
@@
 of considering multivariate (i.e., step length plus turning angle)
-observations.  However, we do briefly compare log-Normal and Weibull
+observations.  To check that this simplification
+does not distort our conclusions
+we do briefly compare log-Normal and Weibull
 step-length distributions, with and without 
@@
 usual in ecological systems the true model will be far more
-complicated than any candidate model \cite{BurnhamAnderson1998};
-we believe that the qualitative conclusions stated here
-for BIC-optimality will carry over to analyses using AIC instead.
+complicated than any candidate model \cite{BurnhamAnderson1998}.
+We have repeated some of our analyses using AIC rather than BIC
+(not shown); for our examples, the qualitative conclusions stated here
+for BIC-optimality carry over to analyses using AIC.
 
@@
 
-A simple simulation was conducted to to see whether fitting homogeneous transition probability HMMs to heterogeneous transitions simulated data will over estimate the number of states and estimate the correct number of states using the heterogeneous transitioning model.
-We used 100 realizations of a two-state HMM with sinusoidal temporal transitions and fitted it with 2- to 5-state HMMs with and without temporal heterogeneity in the transition probabilities.
+We ran a simulation experiment in which we fitted HMMs with
+both homogeneous and heterogeneous 
+transition probabilities to simulated data with
+heterogeneous transition probabilities, to see whether
+the correct (heterogeneous-transition) model correctly
+identified the number of states while the misspecified
+(homogeneous-transition) model overestimated the number of states.
+We used 100 realizations of a two-state HMM with sinusoidally varying
+transition probabilities and fitted it with HMMs ranging from
+2 to 5 hidden states, with and
+without temporal heterogeneity in the transition probabilities.
 
@@
 Then the number of free parameters of an HMM is:
-
-\begin{equation}
-\textrm{Number of Free Parameters}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
-{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
-{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}} .
-\label{eq:numparam}
-\end{equation}
-
-As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}).
+$k_{i}\cdot (n-1)$ (\emph{Initial states})
+$+ k_{t}\cdot n\cdot (n-1)$ (\emph{Transition probabilities})
+$+ k_{e}\cdot n$ (\emph{Emission parameters}).
+As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically.
 When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
@@
 from the expected step lengths conditional on the most likely state
-sequence predicted by the Viterbi algorithm or pseduo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
-because the predictions by these methods rely on the observed data. These approach is useful to predict
+sequence predicted by the Viterbi algorithm or pseudo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
+because the predictions by these methods rely on the observed data. 
+This approach is useful to predict
 missing data in the observation sequence, but because it is
 conditional on the observed values, it can not reliably evaluate
-goodness of fit for the different structural complexities of HMM
-models.
+goodness of fit for HMM models with
+different degrees of structural complexity.
 
@@
 %% fig 1 goes somewhere here
-The simulation agrees with our hypothesis that homogeneous transition HMMs can over estimate the number of hidden states.
+The simulation supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states.
 Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).
+
+\bmb{need to describe in words what the figure says}
 
@@
 This is a common problem when using BICs, thus, the "knee point" (jump point) was proposed for BIC in partitioning based clustering via angle-based method \cite{zhao_knee_2008}.
-A similar method was used with likelihood \cite{dean2012behavioural} where both are applicable in our case, and suggest fewer number of states, but we do not know the reliability when fitting more complex models (ie.
+A similar method was used with likelihood \cite{dean2012behavioural} where both are applicable in our case, and suggest fewer number of states, but we do not know the reliability when fitting more complex models (IA.
 the shape of the BIC curve can be affected by the non-linear increase of parameters with every additional state).
@@
 the right direction, but much remains to be done. 
-Our model neglects other predictors, such as habitat type or
+Our model neglects other predictors, such as sex, habitat type or
 location with respect to environmental features such as roads,
@@
 
+\clearpage
+
   \begin{figure}[h!]
@@
       \end{figure}
+
+\clearpage
 
@@
 
+\clearpage
+
 \begin{figure}[h!]

### 2016-09-09 bmc_template/manuscript.tex
almost-last pass on MS?

@@
 
+Model complexity and the number of parameters increase as the number of latent states increase.
+For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity.
+In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components (initial states, transition probabilities, and emissions).
+Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$.
+Then the number of free parameters of an HMM is:
+[\emph{Initial states}] $k_{i}\cdot (n-1)$ 
++ [\emph{Transition probabilities}] $k_{t}\cdot n\cdot (n-1)$ 
++ [\emph{Emission parameters}] $k_{e}\cdot n$.
+As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically.
+
 \subsection*{Model evaluation}
 
-We used the {\tt depmixS4} package
+We used the {\tt depmixS4} package \cite{visser2010depmixs4}
 to fit covariate-dependent transition HMMs, 
@@
 
-Model complexity and the number of parameters increase as the number of latent states increase.
-For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity.
-In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components.
-Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$.
-Then the number of free parameters of an HMM is:
-$k_{i}\cdot (n-1)$ (\emph{Initial states})
-$+ k_{t}\cdot n\cdot (n-1)$ (\emph{Transition probabilities})
-$+ k_{e}\cdot n$ (\emph{Emission parameters}).
-As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically.
-When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.
-
 We used simulations to predict expected hourly step lengths and
@@
 from the expected step lengths conditional on the most likely state
-sequence predicted by the Viterbi algorithm or pseudo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
+sequence predicted by the Viterbi algorithm or pseudo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is problematic
 because the predictions by these methods rely on the observed data. 
@@
 %% fig 1 goes somewhere here
-The simulation supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states.
-Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).
+The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states (Figure~1).
+Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states) whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).
 
-\bmb{need to describe in words what the figure says}
+\bmb{caption needs a little work}
 
@@
 
-The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results.
-For time homogeneous models, the Weibull-wrapped-Cauchy \cite{kerk2015hidden}, Weibull von Mises, and log Normal without turning angles all identify the same BIC-optimal number of states.
-Furthermore, the homogeneous-HMM models do vary among models (different step-length distributions, with/without turning angles) in how many states they identify, whereas, the heterogeneous-HMM models are consistent among models (Figure 2).
+The BIC-optimal number of states for time-homogeneous models is
+consistent with van de Kerk et al.'s \cite{kerk2015hidden}
+results. For time-homogeneous models, the Weibull-wrapped-Cauchy
+\cite{kerk2015hidden}, Weibull-von Mises, and log Normal without
+turning angles all identify the same BIC-optimal number of
+states. While the number of states identified
+by homogeneous-HMM models varies according to the emissions
+distributions chosen, ranging from $n=5$ for Weibull
+steps alone to $n=7$ for the log Normal-von Mises emissions
+model, the number of states identified
+by heterogeneous-HMM models
+is consistent among emissions models ($n=5$: Figure 2).
 
@@
 
-Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3).
-Turning to the temporally heterogeneous HMMs (Figure 3, right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states.
-The multiple-block model gives approximately the same BIC as the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.
-Finally, the quadratic and sinusoidal models are the best models tested by far; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
-However, this similarity is overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
-normally be interpreted as an enormous difference in goodness of fit,
-between the sinusoidal and quadratic models (both of which have 90 
-parameters).
+Models with temporal heterogeneity provide better fits to the data
+(lower BIC) than homogeneous models in both FMM and HMM frameworks,
+but time-homogeneous HMMs are better than FMMs with sinusoidal
+temporal heterogeneity (Figure 3). Turning to the temporally
+heterogeneous HMMs (Figure 3, right panel), we see that the model with
+different transition probabilities for each hour of the day (HMM +
+THhourly) is overparameterized; it underperforms homogeneous HMM with
+even 3 states, and gets much worse with 4 states. The multiple-block
+model gives approximately the same BIC as the homogeneous HMM,
+although it gives the BIC-optimal number of states as 4, in contrast
+to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal
+models are the best models tested by far; they both give the
+BIC-optimal number of states as 5, and they have similar goodness of
+fit.  However, the similarity between the quadratic 
+and sinusoidal models may be overstated in Figure~3 due to the very large
+variation in BIC (over thousands of units) across the full range of
+models; the best-fit sinusoidal ($n=5$) model is approximately 80 BIC units
+better than the best quadratic model (also $n=5$), which
+would normally be interpreted as an enormous improvement in goodness of
+fit (both models have 90 parameters).
 
-The average hourly step lengths from the observed panther data exhibit a clear diurnal pattern (Figure 4).
-As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle.
-All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns.
-However, the block model does markedly worse than the other temporal models (changing the block definitions might help by re-clustering/grouping different hours or increasing the number of blocks), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).
-We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction to illustrate within sample predictions can capture the diurnal patterns, but fail to capture out of sample predictions.
+The average hourly step lengths from the observed panther data exhibit
+a clear diurnal pattern (Figure 4). As expected, temporally
+homogeneous models (whether FMM or HMM) predict the same mean step
+length regardless of time of day, failing to capture the diurnal
+activity cycle. All of the models incorporating temporal
+heterogeneity, including the temporally heterogeneous FMM, can capture
+the observed patterns. However, the block model does markedly worse
+than the other temporal models (changing the block definitions might
+help by re-clustering/grouping different hours or increasing the
+number of blocks), and the (overparameterized) hourly model does
+better than any other model at capturing the early-evening peak (but
+worse at capturing the mid-day trough). We also included average
+hourly step lengths from three-state temporally homogeneous HMM
+Viterbi prediction to illustrate within sample predictions can capture
+the diurnal patterns, but fail to capture out of sample predictions.
 
@@
 to estimate the number of behavioural states will generally
-overfit their data, in the sense of ``discovering'' large
+overfit their data, i.e. they will ``discover'' large
 number of states that are difficult to interpret biologically.
 
-On a broad spectrum, it really depends on what kind of question 
-that is being answered. On one side of the spectrum, if the goal is 
-to identify states, it might be sufficient to use a 
-simple/traditional HMM model and pre-specify the number of states and, post hoc, 
-match Viterbi-based states estimates with environmental variation 
-\cite{fryxell_multiple_2008}. In our panther study, it is really hard to interpret multiple latent states biologically.
-We have no way of knowing what panthers are actually
-thinking (it is certainly more complex than being in one of a small
-number of discrete latent states); 
-we don't know the ``true'' number of latent states,
-nor are we able to observe them directly, although
-incorporating additional direct observations of 
-behaviour (if available) can at least partially
-address this problem \cite{fryxell_multiple_2008}.
-Three distinct
-movement states seem biologically interpretable for Florida panthers
-according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
-states, intermediate step length a foraging state, and long
-step length a traveling state.
+As usual, the appropriate approach depends on the goal of the
+analysis.  If ecologists simply want to identify states
+and associate them with environmental characteristics, it might be
+sufficient to use a simple (homogeneous) HMM model, pre-specifying the
+number of states to a biologically sensible value, and then match
+\emph{post hoc} Viterbi estimates of state occupancy with
+environmental variation in space and time \cite{fryxell_multiple_2008}.
 
-On the other side of the 
-spectrum, if the goal of interest is to make predictions (out of sample), 
-it might be better to fit a covariate-dependent model so that we can explicitly
-model the switching process. In that case, fitting a covariate-dependent model is
-better for out of sample prediction because Viterbi can only estimate state 
-occupancy if observed movements are available (within sample predictions).
+%% In our panther
+%% study, it is really hard to interpret multiple latent states
+%% biologically.  We have no way of knowing what panthers are actually
+%% thinking (it is certainly more complex than being in one of a small
+%% number of discrete latent states); we don't know the ``true'' number
+%% of latent states, nor are we able to observe them directly, although
+%% incorporating additional direct observations of behaviour (if
+%% available) can at least partially address this problem
+%% \cite{fryxell_multiple_2008}.  Three distinct movement states seem
+%% biologically interpretable for Florida panthers according to van de
+%% Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
+%% states, intermediate step length a foraging state, and long step
+%% length a traveling state.
 
-Finally, if we want to estimate the number of states, BIC is not necessarily 
-good for estimation of number of states \cite{biernacki2000assessing}, 
-but it can be useful as an approximate upper limit estimate.
-This is a common problem when using BICs, thus, the "knee point" (jump point) was proposed for BIC in partitioning based clustering via angle-based method \cite{zhao_knee_2008}.
-A similar method was used with likelihood \cite{dean2012behavioural} where both are applicable in our case, and suggest fewer number of states, but we do not know the reliability when fitting more complex models (IA.
-the shape of the BIC curve can be affected by the non-linear increase of parameters with every additional state).
+On the other hand, if the goal of analysis is to make out-of-sample
+predictions about animal behaviour, such as in a management context,
+it is necessary to fit a covariate-dependent model that
+explicitly incorporates the switching process.
+While the Viterbi algorithm can be applied to
+work backward from observed behaviour
+to variations in state occupancy with environmental conditions
+even when using a homogeneous model,
+a homogeneous model 
+can never \emph{predict} behaviour that varies 
+with environmental conditions.
 
-Incorporating temporal heterogeneity in animal movement is one step in
-the right direction, but much remains to be done. 
-Our model neglects other predictors, such as sex, habitat type or
-location with respect to environmental features such as roads,
-that can potentially improve goodness of fit and predictions and 
+Finally, if our goal is actually to estimate the number of discrete
+behavioural modes underlying observed behaviour --- keeping in mind
+that these discrete behavioural states are certainly an oversimplified
+representation of animals' real internal states --- then, as we have
+shown above, covariate-dependent models will generally be required to
+avoid overestimating the number of states.  We note that researchers
+in cluster analysis (of which HMMs are a special case) have shown that
+the technical conditions required for BIC to apply may be violated
+\cite{biernacki2000assessing}; however, BIC can be useful as an
+approximate \emph{upper} limit on the number of states. Various
+solutions to this problem have been proposed, including the
+``integrated classification likelihood''
+\cite{biernacki2000assessing}, as well as a simpler ``knee point''
+method \cite{zhao_knee_2008} that looks for the cluster size that
+corresponds to the largest change in BIC rather than to the smallest
+overall BIC.  \cite{dean2012behavioural} took a similar approach, but
+based on the log-likelihood curve rather than the BIC.
+However, at least in our simulations the BIC does correctly identify
+the number of states when appropriate heterogeneity is included
+in the model; the best practical criterion for identifying 
+numbers of states underlying animal movement data is an open question.
+
+\bmb{mike, can we take a quick try at computing ICL for our data/sims?}
+
+\begin{equation*}
+\begin{split}
+ICL(m,K) & = \underbrace{\max_{a} \log f(x|\bar z, m, K, a)}_{\textrm{max log-likelihood}} - \frac{\lambda_m,K}{2} \log n \\
+ & + \log \Gamma(K/2) + \sum_{k=1}^{K} \log \Gamma(\bar n_k + 1/2) \\
+ & - K \log \Gamma(1/2) - \log \Gamma(n + K/2)
+\end{split}
+\end{equation*}
+
+($K$=number of clusters; $n$=number of obs.; $\lambda_{m,K}$ is ``the
+number of free components in $a$'' (i.e., number of free parameters:
+p. 8); $\bar n_k$ is, I think, the number of data points which have
+their max. probability of being in cluster $k$ [i.e., we should be
+  able to get this at least approximately from Viterbi, although I'm
+  not sure whether it's more appropriate to use Viterbi or an
+  ``instantaneous'' max probability]). The {\tt rebmix} package has an
+ICL criterion for its particular objects ({\tt
+  sos::findFn("\{integrated classification likelihood\}"}), but I
+don't know if it'll be useful.  I think Raftery and Fraley also have a
+modified IC for clusters?
+
+We have shown the value of incorporating temporal heterogeneity in 
+animal movement models, but much remains to be done. 
+Our model neglects other predictors --- such as sex, habitat type or
+location with respect to environmental features such as roads --- %
+that can potentially improve predictive accuracy and 
 further reduce the estimated number of states. While adding
-more covariates is in principle straightforward using existing
-frameworks, including all possible biological complexities in 
-a HMM with state-dependent transitions may rapidly become intractable
-in terms of both computational time and complexity of choosing
-among possible reduced models and numbers of states.
+more covariates is in principle straightforward, 
+including even a reasonable range of biological complexity in 
+a HMM with state-dependent transitions rapidly becomes intractable
+in both computational time and the complexity of model selection.
 Better diagnostic procedures and tests are needed:

### 2016-09-20 bmc_template/manuscript.tex
pushing text

@@
 solutions to this problem have been proposed, including the
-``integrated classification likelihood''
+``integrated classification likelihood'' (ICL)
 \cite{biernacki2000assessing}, as well as a simpler ``knee point''
@@
 overall BIC.  \cite{dean2012behavioural} took a similar approach, but
-based on the log-likelihood curve rather than the BIC.
-However, at least in our simulations the BIC does correctly identify
+based on the log-likelihood curve rather than the BIC. However, HMMs are 
+a special case even for the various solutions. ICL uses the likelihood of the 
+most probable state sequence via the Viterbi algorithm to account for relative 
+state occupancies between hidden states as a penalizing factor \cite{celeux2008selecting}, but it does
+not have the same properties as the overall unconditional likelihood, such as, 
+monotonic increase in likelihood as the number of hidden states increase. Furthermore, 
+both log-likelihood and BIC based ``knee point'' methods look at the changes correponding to
+a ``linear'' increase in number of states rather than the non-linear increase in parameters
+corresponding to the linear increase in states. 
+Nevertheless, at least in our simulations the BIC does correctly identify
 the number of states when appropriate heterogeneity is included
@@
 numbers of states underlying animal movement data is an open question.
-
-\bmb{mike, can we take a quick try at computing ICL for our data/sims?}
-
-\begin{equation*}
-\begin{split}
-ICL(m,K) & = \underbrace{\max_{a} \log f(x|\bar z, m, K, a)}_{\textrm{max log-likelihood}} - \frac{\lambda_m,K}{2} \log n \\
- & + \log \Gamma(K/2) + \sum_{k=1}^{K} \log \Gamma(\bar n_k + 1/2) \\
- & - K \log \Gamma(1/2) - \log \Gamma(n + K/2)
-\end{split}
-\end{equation*}
-
-($K$=number of clusters; $n$=number of obs.; $\lambda_{m,K}$ is ``the
-number of free components in $a$'' (i.e., number of free parameters:
-p. 8); $\bar n_k$ is, I think, the number of data points which have
-their max. probability of being in cluster $k$ [i.e., we should be
-  able to get this at least approximately from Viterbi, although I'm
-  not sure whether it's more appropriate to use Viterbi or an
-  ``instantaneous'' max probability]). The {\tt rebmix} package has an
-ICL criterion for its particular objects ({\tt
-  sos::findFn("\{integrated classification likelihood\}"}), but I
-don't know if it'll be useful.  I think Raftery and Fraley also have a
-modified IC for clusters?
 

### 2016-09-20 ICLpanther.R

@@ -4,7 +4,7 @@ library(dplyr)
 library(reshape2)
 library(ggplot2)
 
-load('.cat.15.hmm.RData')
+load('.cat.1.hmm.RData')
 
 source("stateseqLL.R")
 
@@ -17,7 +17,7 @@ sumdf <- function(lst){
   nstates <-ldply(lst,nstates)
   para <- ldply(lst,freepars)
   n <- ldply(lst,ntimes)
-  model <- c('HMM','HMM','HMM','HMM','HMM'
+  model <- c('HMM','HMM','HMM','HMM','HMM','HMM'
              # ,'HMMsin','HMMsin','HMMsin','HMMsin'
   )
   temp <- data.frame(LL=LL$V1
@@ -34,7 +34,7 @@ sumdf <- function(lst){
 
 }
 
-fitlist <- list(fithmm3s,fithmm4s,fithmm5s,fithmm6s,fithmm7s
+fitlist <- list(fithmm2s,fithmm3s,fithmm4s,fithmm5s,fithmm6s,fithmm7s
                 # ,hmmsin2s,hmmsin3s,hmmsin4s,hmmsin5s
 )
 

### 2016-09-21 bmc_template/manuscript.tex
tweak

@@
 %% fig 1 goes somewhere here
-The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states (Figure~1).
-Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states) whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).
+The simulation experiment supports our hypothesis that homogeneous transition HMMs overestimate the number of hidden states when the model is misspecified (Figure~1).
+Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).
 
@@
 ``integrated classification likelihood'' (ICL)
-\cite{biernacki2000assessing}, as well as a simpler ``knee point''
+\cite{biernacki2000assessing,celeux2008selecting}, as well as a simpler ``knee point''
 method \cite{zhao_knee_2008} that looks for the cluster size that
 corresponds to the largest change in BIC rather than to the smallest
-overall BIC.  \cite{dean2012behavioural} took a similar approach, but
+overall BIC.  Dean \emph{et al.} \cite{dean2012behavioural} took a similar approach, but
 based on the log-likelihood curve rather than the BIC. However, HMMs are 
-a special case even for the various solutions. ICL uses the likelihood of the 
-most probable state sequence via the Viterbi algorithm to account for relative 
-state occupancies between hidden states as a penalizing factor \cite{celeux2008selecting}, but it does
-not have the same properties as the overall unconditional likelihood, such as, 
-monotonic increase in likelihood as the number of hidden states increase. Furthermore, 
-both log-likelihood and BIC based ``knee point'' methods look at the changes correponding to
-a ``linear'' increase in number of states rather than the non-linear increase in parameters
-corresponding to the linear increase in states. 
-Nevertheless, at least in our simulations the BIC does correctly identify
+a special case even for the various solutions. 
+% ICL uses the likelihood of the  most probable state sequence via the Viterbi algorithm to account for relative  state occupancies between hidden states as a penalizing factor \cite{celeux2008selecting}, but it does not have the same properties as the overall unconditional likelihood, such as,  monotonic increase in likelihood as the number of hidden states increase.
+Furthermore,  both log-likelihood and BIC based ``knee point'' methods look at the changes corresponding to a ``linear'' increase in number of states rather than the non-linear increase in parameters corresponding to the linear increase in states.
+
+Nevertheless, in our simulations the BIC does correctly identify
 the number of states when appropriate heterogeneity is included
 in the model; the best practical criterion for identifying 
-numbers of states underlying animal movement data is an open question.
+numbers of states underlying animal movement data remains an open question.
 
@@
   \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
-\bmb{do I need to explain transformation of LN SD ?}}
       \end{figure}

### 2016-09-21 bmc_template/manuscript.tex
putting figs at the appropriate place, going to replace sim results with a table

@@
 
-%% fig 1 goes somewhere here
-The simulation experiment supports our hypothesis that homogeneous transition HMMs overestimate the number of hidden states when the model is misspecified (Figure~1).
+  \begin{figure}[h!]
+   \includegraphics[width=5in]{figure/sim_results-1}
+  \caption{\csentence{Frequency of choosing BIC-optimal state on simulated data.} 45 out of 100 sets of simulated data from a 2-state HMM with BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}
+      \end{figure}
+
+The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states when the model is misspecified (Figure~1).
 Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).
 
-\bmb{caption needs a little work}
-
-%% fig 2 goes somewhere here
+  \begin{figure}[h!]
+   \includegraphics[width=5in]{figure/BICred_plot-1}
+  \caption{\csentence{Adjusted BIC Emission Distribution Comparison} Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.}
+      \end{figure}
 
@@
 
-%% fig 3 goes somewhere here
+  \begin{figure}[h!]
+   \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
+  \caption{\csentence{Adjusted BICs Across All Models} Adjusted BIC by number of free parameters for HMM model types.
+The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM.
+The right panel shows HMMs with different temporal transitions.}
+      \end{figure}
 
@@
 
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/avg_step_length_by_time-1}
+  \caption{\csentence{Diurnal Step Lengths Plot} Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
+      \end{figure}
+
 The average hourly step lengths from the observed panther data exhibit
@@
 
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/acf_plot-1}
+  \caption{\csentence{Autocorrelation Plot} Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
+      \end{figure}
+
 Like the diurnal pattern (Figure 4),
@@
 
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/r_msdlist-1}
+  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
+  }
+      \end{figure}
+
 The estimated emission parameter values
@@
 are similar, for both homogeneous and heterogeneous
-models, across all cats (Figure 7 shows a subset of cats). 
+models, across all cats (Figure 6 shows a subset of cats). 
 In general, the states with longer mean step lengths are
@@
 
-%% In our panther
-%% study, it is really hard to interpret multiple latent states
-%% biologically.  We have no way of knowing what panthers are actually
-%% thinking (it is certainly more complex than being in one of a small
-%% number of discrete latent states); we don't know the ``true'' number
-%% of latent states, nor are we able to observe them directly, although
-%% incorporating additional direct observations of behaviour (if
-%% available) can at least partially address this problem
-%% \cite{fryxell_multiple_2008}.  Three distinct movement states seem
-%% biologically interpretable for Florida panthers according to van de
-%% Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
-%% states, intermediate step length a foraging state, and long step
-%% length a traveling state.
-
 On the other hand, if the goal of analysis is to make out-of-sample
@@
 a special case even for the various solutions.
-% ICL uses the likelihood of the  most probable state sequence via the Viterbi algorithm to account for relative  state occupancies between hidden states as a penalizing factor \cite{celeux2008selecting}, but it does not have the same properties as the overall unconditional likelihood, such as,  monotonic increase in likelihood as the number of hidden states increase.
-Furthermore,  both log-likelihood and BIC based ``knee point'' methods look at the changes corresponding to a ``linear'' increase in number of states rather than the non-linear increase in parameters corresponding to the linear increase in states.
 
@@
 
-\section*{Figures}
-  \begin{figure}[h!]
-   \includegraphics[width=5in]{figure/sim_results-1}
-  \caption{\csentence{Simulation Results} BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}
-      \end{figure}
-
-\clearpage
-
-  \begin{figure}[h!]
-   \includegraphics[width=5in]{figure/BICred_plot-1}
-  \caption{\csentence{Adjusted BIC Emission Distribution Comparison} Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.}
-      \end{figure}
-
-\clearpage
-
-  \begin{figure}[h!]
-   \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
-  \caption{\csentence{Adjusted BICs Across All Models} Adjusted BIC by number of free parameters for HMM model types.
-The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM.
-The right panel shows HMMs with different temporal transitions.}
-      \end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-   \includegraphics[width=5in]{figure/avg_step_length_by_time-1}
-  \caption{\csentence{Diurnal Step Lengths Plot} Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
-      \end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-   \includegraphics[width=5in]{figure/acf_plot-1}
-  \caption{\csentence{Autocorrelation Plot} Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
-      \end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-   \includegraphics[width=5in]{figure/r_msdlist-1}
-  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
-      \end{figure}
-
 \end{backmatter}

### 2016-09-22 bmc_template/manuscript.tex
manuscript done, working on additional files and cover letter

@@
 
-  \begin{figure}[h!]
-   \includegraphics[width=5in]{figure/sim_results-1}
-  \caption{\csentence{Frequency of choosing BIC-optimal state on simulated data.} 45 out of 100 sets of simulated data from a 2-state HMM with BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}
-      \end{figure}
+\begin{table}[h!]
+\caption{Frequency of choosing BIC-optimal state on simulated data.}
+\begin{tabular}{ccccccc}
+  & BIC-Optimal States       & \bf{2}  & \bf{3}  & \bf{4} & \bf{5} & \bf{6} \\ \hline
+  & Homogeneous Transition   & 45 & 21 & 20 & 14 & 0 \\ \hline
+  & Heterogeneous Transition & 100 & 0 & 0 & 0 & 0 \\ \hline
+\end{tabular}
+\end{table}
 
-The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states when the model is misspecified (Figure~1).
+The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states when the model is misspecified (Table~1).
 Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).
@@
    \includegraphics[width=5in]{figure/BICred_plot-1}
-  \caption{\csentence{Adjusted BIC Emission Distribution Comparison} Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.}
+  \caption{Comparison of BIC-optimal state predictions for homogeneous transition HMMs (left panel) and and heterogeneous transition HMMs (right panel) with different emission complexities on panther data.
+Solid line represents univariate response/emission HMMs (without turning angles) and dotted line represents multivariate response HMMs (including turning angles with von-Mises distribution).
+Red lines represents log-normal step-length distribution and blue lines represents Weibull step-length distribution.
+}
       \end{figure}
@@
    \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
-  \caption{\csentence{Adjusted BICs Across All Models} Adjusted BIC by number of free parameters for HMM model types.
+  \caption{Comparison of adjusted BIC for all HMM transition complexities.
 The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM.
-The right panel shows HMMs with different temporal transitions.}
+The right panel shows HMMs with different temporal transitions.
+Dashed lines represents FMMs and solid lines represents HMMs.}
       \end{figure}
@@
    \includegraphics[width=5in]{figure/avg_step_length_by_time-1}
-  \caption{\csentence{Diurnal Step Lengths Plot} Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
+  \caption{Comparison of average step-length by time of day out of sample predictions for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
+Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
       \end{figure}
@@
    \includegraphics[width=5in]{figure/acf_plot-1}
-  \caption{\csentence{Autocorrelation Plot} Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}
+  \caption{Comparison of out of sample predictions autocorrelations for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
+Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
       \end{figure}
@@
    \includegraphics[width=5in]{figure/r_msdlist-1}
-  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
+  \caption{State identification of BIC-optimal HMMs for all panthers.
+Mean (x-axis) and standard deviation (y-axis) of step length by state $\textrm{log}_{10} m$ for BIC-optimal homogeneous HMMs (red line) and heterogeneous HMMs with sinusoidal transition (blue line).
+Short movement states (step-length \textless $10^{1.5}$m) are circled together.
   }

### 2016-09-22 bmc_template/cover.tex
crappy cover

\documentclass[10pt]{letter}

\usepackage{graphicx}
\usepackage{times}
%\usepackage[sort&compress]{natbib}

\bibliographystyle{abbrv}

\makeatletter
% http://tex.stackexchange.com/questions/18033/using-bibtex-with-letter-class
\newenvironment{thebibliography}[1]
     {\list{\@biblabel{\@arabic\c@enumiv}}%
           {\settowidth\labelwidth{\@biblabel{#1}}%
            \leftmargin\labelwidth
            \advance\leftmargin\labelsep
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\newcommand\newblock{\hskip .11em\@plus.33em\@minus.07em}
\makeatother

\begin{document}

\date{\today}

\begin{letter}{
}

\opening{To the editor,}

We (Michael Li and Ben Bolker) are submitting our manuscript, ``Incorporating Periodic Variability in Hidden Markov Models for Animal Movement'', for consideration for publication in \emph{Movement Ecology}. Using high resolution movement data from Florida panthers, our study explore the uncertainty of identifying unobservable behavioural states and forcasting using different complexities of finite mixture and hidden Markov models (HMMs). Allowing for temporal heterogeneity in transition probabilities, a straightforward 
but little-used extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2016-09-22 bmc_template/cover.tex
cover.tex

@@
 
-We (Michael Li and Ben Bolker) are submitting our manuscript, ``Incorporating Periodic Variability in Hidden Markov Models for Animal Movement'', for consideration for publication in \emph{Movement Ecology}.
-Using high resolution movement data from Florida panthers, our study explore the uncertainty of identifying unobservable behavioural states and forcasting using different complexities of finite mixture and hidden Markov models (HMMs).
-Allowing for temporal heterogeneity in transition probabilities, a straightforward
-but little-used extension of the standard HMM framework, 
-resolves some shortcomings of current
-models and clarifies the behavioural patterns of
-panthers.
+We (Michael Li and Ben Bolker) would like to submit our manuscript,
+``Incorporating Periodic Variability in Hidden Markov Models for
+Animal Movement'', for consideration for publication in \emph{Movement
+  Ecology}. Using high resolution movement data from Florida panthers,
+and inspired by previous analyses using these data, we explore the
+ability of hidden Markov models (HMMs) to correctly estimate the
+number of latent behavioural states.  We use simulation studies and
+analyses of the panther data to show that allowing for diurnal
+heterogeneity in transition probabilities among behavioural states, a
+straightforward but little-used extension of the HMM framework,
+resolves some shortcomings of current models and clarifies the
+behavioural patterns of panthers.
+
+We confirm that we have no competing interests and that both authors
+have approved the manuscript, which has not been published or
+submitted for publication elsewhere, for publication.
+
+We suggest the following peer reviewers: Roland Langrock (),
+Luca Borger (), and Juan Morales ().
 

### 2016-09-23 bmc_template/figures.tex
ready to submit

--- 2016-08-02 bmc_template/figures.tex
@@
 \documentclass{article}\usepackage[]{graphicx}\usepackage{xcolor}
+\usepackage{lineno}
+\linenumbers
+
 %% maxwidth is the original width if it is less than linewidth
@@
 \makeatletter
+\renewcommand{\thetable}{A\arabic{table}}
+
+\renewcommand{\thefigure}{A\arabic{figure}}
 \def\maxwidth{ %
@@
 
+\section*{Additional files}
+
+  In the main text of this manuscript, we showed the fitting and simulating results for Cat 1.
+Here, we have the adjusted BIC emission distribution comparisons, adjusted BIC across all FMM and HMM models, diurnal step lengths and autocorrelation plots for Cat 2, 14, and 15.
+Table A1 gives the name of panther ID (Cat ID) in Institutional Repository at the University of Florida (IR@UF), \cite{kerk2015hidden}, and number of observations for each panther.
+
+\begin{table}[h!]
+\caption{Cat ID and number of observations; ID numbers are given matching
+those shown by van de Kerk et al. 2015 and those 
+in the data located at the UF Institutional repository (IR@UF).}
+\begin{tabular}{cccc}
+  \hline
+  & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
+  & 130 & 1 & 10286\\
+  & 131 & 2  & 9458\\
+  & 48 & 14  & 14645\\
+  & 94  & 15   & 10250\\ \hline
+\end{tabular}
+\end{table}
+
+\newpage
+
 \begin{knitrout}
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
-\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[Simulation test for BIC-optimal approach]{Simulation test for BIC-optimal approach}\label{fig:sim_results}
+\includegraphics[width=\maxwidth]{figure/BICred_plot2-1} \caption{Cat 2: Comparison of BIC-optimal state predictions for homogeneous transition HMMs (left panel) and heterogeneous transition HMMs (right panel) with different emission complexities on panther data.
+Solid line represents univariate response/emission HMMs (without turning angles) and dotted line represents multivariate response HMMs (including turning angles with von-Mises distribution).
+Red lines represents log-normal step-length distribution and blue lines represents Weibull step-length distribution.}
+\label{fig:BICred_plot2}
+\end{figure}
+
+\end{knitrout}
+
+\clearpage
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons2-1} \caption{Cat 2: Comparison of adjusted BIC for all HMM transition complexities.
+The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM.
+The right panel shows HMMs with different temporal transitions.
+Dashed lines represents FMMs and solid lines represents HMMs.}\label{fig:adj_BIC_comparisons2}
 \end{figure}
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time2-1} \caption{Cat 2: Comparison of average step-length by time of day out of sample predictions for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
+Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}\label{fig:avg_step_length_by_time2}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/r_msdlist-1}
-
-\end{knitrout}
-
-\clearpage
-
-\section{sup}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot2-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons2-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time2-1}
-
-\end{knitrout}
-
-\clearpage
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot2-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/acf_plot2-1} \caption{Cat 2: Comparison of out of sample predictions autocorrelations for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
+Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/BICred_plot14-1} \caption{This figure mataches Figure A1, but for Cat 14.}\label{fig:BICred_plot14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons14-1} \caption{This figure mataches Figure A2, but for Cat 14.}\label{fig:adj_BIC_comparisons14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time14-1} \caption{This figure mataches Figure A3, but for Cat 14.}\label{fig:avg_step_length_by_time14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot14-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/acf_plot14-1} \caption{This figure mataches Figure A4, but for Cat 14.}\label{fig:acf_plot14}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/BICred_plot15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/BICred_plot15-1} \caption{This figure mataches Figure A1, but for Cat 15.}\label{fig:BICred_plot15}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons15-1} \caption{This figure mataches Figure A2, but for Cat 15.}\label{fig:adj_BIC_comparisons15}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time15-1} \caption{This figure mataches Figure A3, but for Cat 15.}\label{fig:avg_step_length_by_time15}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/acf_plot15-1}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\maxwidth]{figure/acf_plot15-1} \caption{This figure mataches Figure A4, but for Cat 15.}\label{fig:acf_plot15}
+\end{figure}
+
+\clearpage
 
 \end{knitrout}
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{panthers}
 

### 2016-09-23 bmc_template/manuscript.tex
ready to submit

@@
 \dochead{Research}
-
-%% TO DO
-% author references (maybe fix programmatically?)
-% acknowledgments
 
@@
 step-length distributions, with and without 
-a von Mises-distributed turning angle included in the model (Figure 2).
+a von Mises-distributed turning angle included in the model (Figure 1).
 (Note that most movement analyses, including van de Kerk et al.
@@
 
-For an $n$-state HMM, we need to define an $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.
-The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy.
-It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.
+For an $n$-state HMM, we need to define an $n \times n$ matrix that specifies the probabilities
+$\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$. The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are 
+identical across all states --- i.e., the probability of occupying a state at the next time step 
+is independent of the current state occupancy. It can be modelled in the HMM framework by setting 
+the transition probabilities  $\pi_{ij} = \pi_{i*}$.
 
-In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
-As is standard for HMMs with covariates \cite{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without loss of generality (i.e.
-we have only $n \times (n-1)$ distinct parameters; we index $j$ from 2 to $n$ for notational clarity):
+In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all 
+probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
+As is standard for HMMs with covariates \cite{visser2010depmixs4}, we define this multinomial 
+logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without 
+loss of generality (i.e. we have only $n \times (n-1)$ distinct parameters; we index $j$ from 2 to 
+$n$ for notational clarity):
 
@@
 
-We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of Morales et al.
+We considered four different transition models for diurnal variation in behaviour, incorporating 
+hour-of-day as a covariate following the general approach of Morales et al. 
 \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.
@@
 
-\item[Hourly model] Lastly, we extended the multi-block approach and assign  a different transition matrix for every hour of the day.
-This model is included for comparative purposes; due to the large number of parameters in the model (more than $24n(n-1)$ for a HMM with $n$ states), it is not really practical.
-We only fitted up to four states using the hourly model.
+\item[Hourly model] Lastly, we extended the multi-block approach and assign  a different 
+transition matrix for every hour of the day. This model is included for comparative purposes; due 
+to the large number of parameters in the model (more than $24n(n-1)$ for a HMM with $n$ states), 
+it is not really practical. We only fitted up to four states using the hourly model.  
 \end{description}
 
-Other periodic functions, such as Fourier series (i.e., the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could also be considered.
+Other periodic functions, such as Fourier series (i.e., the sinusoidal transition model augmented 
+by additional sinusoidal components at higher frequencies) or periodic splines, could also be 
+considered.
 
 Model complexity and the number of parameters increase as the number of latent states increase.
-For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity.
-In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components (initial states, transition probabilities, and emissions).
-Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$.
-Then the number of free parameters of an HMM is:
+For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and 
+finally by FMMs and HMMs incorporating temporal heterogeneity. In general, the number of free 
+parameters in an HMM is the sum of the number of free parameters for each of the three model 
+components (initial states, transition probabilities, and emissions). Let $n$ be the number of 
+hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the 
+covariate-dependence of the prior distribution, transition function and emission distributions; 
+that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical 
+predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:
 [\emph{Initial states}] $k_{i}\cdot (n-1)$ 
@@
 + [\emph{Emission parameters}] $k_{e}\cdot n$.
-As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically.
+As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) 
+FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity 
+(or covariate-dependent transitions more generally) the number increases quadratically.
 
@@
 HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
-with the most data ($> 9000 $ observations).
-(1) BIC was used to compare the goodness of fit of each model type.
-The model with the lowest BIC was selected to be the optimal-BIC model and all BICs were adjusted to $\dbic$ based on the optimal-BIC model ($\dbic$ = BIC - min(BIC)) .
-(2) Comparing average step-length by hour of day for the
+with the most data ($> 9000 $ observations). (1) BIC was used to compare the goodness of fit of 
+each model type. The model with the lowest BIC was selected to be the optimal-BIC model and all 
+BICs were adjusted to $\dbic$ based on the optimal-BIC model ($\dbic$ = BIC - min(BIC)) . (2) 
+Comparing average step-length by hour of day for the 
 observed data and for data simulated from the models shows how well a particular class of 
@@
 for the observed data and for data simulated from the models shows how well a particular class
-of models can capture serial correlation at both short and long time scales.
-As a complement, we also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to compare the temporal effects in goodness of fit As a reminder, FMMs assume that the latent state in each time step is \emph{independent} of the latent state at the previous time step; time-varying FMMs can accurately describe movement when behaviour can
+of models can capture serial correlation at both short and long time scales. As a complement, we 
+also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to 
+compare the temporal effects in goodness of fit As a reminder, FMMs assume that the latent state 
+in each time step is \emph{independent} of the latent state at the previous time step; 
+time-varying FMMs can accurately describe movement when behaviour can
 change on a short time scale, but the average propensity for different
@@
 
-\begin{table}[h!]
-\caption{Frequency of choosing BIC-optimal state on simulated data.}
-\begin{tabular}{ccccccc}
-  & BIC-Optimal States       & \bf{2}  & \bf{3}  & \bf{4} & \bf{5} & \bf{6} \\ \hline
-  & Homogeneous Transition   & 45 & 21 & 20 & 14 & 0 \\ \hline
-  & Heterogeneous Transition & 100 & 0 & 0 & 0 & 0 \\ \hline
-\end{tabular}
-\end{table}
-
-The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states when the model is misspecified (Table~1).
-Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).
-
-  \begin{figure}[h!]
-   \includegraphics[width=5in]{figure/BICred_plot-1}
-  \caption{Comparison of BIC-optimal state predictions for homogeneous transition HMMs (left panel) and and heterogeneous transition HMMs (right panel) with different emission complexities on panther data.
-Solid line represents univariate response/emission HMMs (without turning angles) and dotted line represents multivariate response HMMs (including turning angles with von-Mises distribution).
-Red lines represents log-normal step-length distribution and blue lines represents Weibull step-length distribution.
-}
-      \end{figure}
+The simulation experiment supports our hypothesis that homogeneous transition HMMs can 
+overestimate the number of hidden states when the model is misspecified (Table~1). Heterogeneous 
+transition models can always predict the correct number of states (in 100/100 simulations, BIC 
+correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models 
+overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer 
+than half of the simulations; values up to $n=5$ are frequently chosen).
 
@@
 by heterogeneous-HMM models
-is consistent among emissions models ($n=5$: Figure 2).
-
-  \begin{figure}[h!]
-   \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
-  \caption{Comparison of adjusted BIC for all HMM transition complexities.
-The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM.
-The right panel shows HMMs with different temporal transitions.
-Dashed lines represents FMMs and solid lines represents HMMs.}
-      \end{figure}
+is consistent among emissions models ($n=5$: Figure 1).
 
@@
 but time-homogeneous HMMs are better than FMMs with sinusoidal
-temporal heterogeneity (Figure 3). Turning to the temporally
-heterogeneous HMMs (Figure 3, right panel), we see that the model with
+temporal heterogeneity (Figure 2). Turning to the temporally
+heterogeneous HMMs (Figure 2, right panel), we see that the model with
 different transition probabilities for each hour of the day (HMM +
@@
 fit.  However, the similarity between the quadratic 
-and sinusoidal models may be overstated in Figure~3 due to the very large
+and sinusoidal models may be overstated in Figure~2 due to the very large
 variation in BIC (over thousands of units) across the full range of
@@
 
-\begin{figure}[h!]
-   \includegraphics[width=5in]{figure/avg_step_length_by_time-1}
-  \caption{Comparison of average step-length by time of day out of sample predictions for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
-Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
-      \end{figure}
-
 The average hourly step lengths from the observed panther data exhibit
-a clear diurnal pattern (Figure 4). As expected, temporally
+a clear diurnal pattern (Figure 3). As expected, temporally
 homogeneous models (whether FMM or HMM) predict the same mean step
@@
 
-\begin{figure}[h!]
-   \includegraphics[width=5in]{figure/acf_plot-1}
-  \caption{Comparison of out of sample predictions autocorrelations for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
-Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
-      \end{figure}
-
-Like the diurnal pattern (Figure 4),
-the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 5) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
+Like the diurnal pattern (Figure 3),
+the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 4) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
 In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects.
@@
 
-\begin{figure}[h!]
-   \includegraphics[width=5in]{figure/r_msdlist-1}
-  \caption{State identification of BIC-optimal HMMs for all panthers.
-Mean (x-axis) and standard deviation (y-axis) of step length by state $\textrm{log}_{10} m$ for BIC-optimal homogeneous HMMs (red line) and heterogeneous HMMs with sinusoidal transition (blue line).
-Short movement states (step-length \textless $10^{1.5}$m) are circled together.
-  }
-      \end{figure}
-
 The estimated emission parameter values
@@
 are similar, for both homogeneous and heterogeneous
-models, across all cats (Figure 6 shows a subset of cats). 
+models, across all cats (Figure 5 shows a subset of cats). 
 In general, the states with longer mean step lengths are
@@
 
-\section{Data accessibility}
+\section*{Data accessibility}
 Hourly step lengths and turning angles of male and female Florida panthers available at \url{http://ufdc.ufl.edu//IR00004241/00001}.
@@
 The authors declare that they have no competing interests.
+
+\end{backmatter}
 
@@
 \section*{Tables}
+
 \begin{table}[h!]
-\caption{Cat ID and number of observations; ID numbers are given matching
-those shown by van de Kerk et al. 2015 and those 
-in the data located at the UF Institutional repository (IR@UF).}
-\begin{tabular}{cccc}
-  \hline
-  & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
-  & 130 & 1 & 10286\\
-  & 131 & 2  & 9458\\
-  & 48 & 14  & 14645\\
-  & 94  & 15   & 10250\\ \hline
+\caption{Frequency of choosing BIC-optimal state on simulated data.}
+\begin{tabular}{ccccccc}
+  & BIC-Optimal States       & \bf{2}  & \bf{3}  & \bf{4} & \bf{5} & \bf{6} \\ \hline
+  & Homogeneous Transition   & 45 & 21 & 20 & 14 & 0 \\ \hline
+  & Heterogeneous Transition & 100 & 0 & 0 & 0 & 0 \\ \hline
 \end{tabular}
@@
 
-\end{backmatter}
+\clearpage
+
+\section*{Figures}
+  \begin{figure}[h!]
+   \includegraphics[width=5in]{figure/Figure-1}
+  \caption{Comparison of BIC-optimal state predictions for homogeneous transition HMMs (left panel) and heterogeneous transition HMMs (right panel) with different emission complexities on panther data.
+Solid line represents univariate response/emission HMMs (without turning angles) and dotted line represents multivariate response HMMs (including turning angles with von-Mises distribution).
+Red lines represents log-normal step-length distribution and blue lines represents Weibull step-length distribution.
+}
+      \end{figure}
+
+\clearpage
+
+  \begin{figure}[h!]
+   \includegraphics[width=5in]{figure/Figure-2}
+  \caption{Comparison of adjusted BIC for all HMM transition complexities.
+The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM.
+The right panel shows HMMs with different temporal transitions.
+Dashed lines represents FMMs and solid lines represents HMMs.}
+      \end{figure}
+
+\clearpage
+
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/Figure-3}
+  \caption{Comparison of average step-length by time of day out of sample predictions for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
+Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
+      \end{figure}
+
+\clearpage
+
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/Figure-4}
+  \caption{Comparison of out of sample predictions autocorrelations for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities.
+Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
+      \end{figure}
+
+\clearpage
+
+\begin{figure}[h!]
+   \includegraphics[width=5in]{figure/Figure-5}
+  \caption{State identification of BIC-optimal HMMs for all panthers.
+Mean (x-axis) and standard deviation (y-axis) of step length by state $\textrm{log}_{10} m$ for BIC-optimal homogeneous HMMs (red line) and heterogeneous HMMs with sinusoidal transition (blue line).
+Short movement states (step-length \textless $10^{1.5}$m) are circled together.
+  }
+      \end{figure}
+
 \end{document}

### 2016-12-22 bmc_template/figures.Rnw
need turning angle picture

@@
 
-<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.",fig.width=6,fig.height=4,eval=FALSE>>=
-ICdat <- readRDS("../sim_results/ICdat.rds")
-aa <- readRDS("../sim_results/sim.101.RDS")
+<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.",eval=TRUE>>=
+ICdat <- readRDS("../simsin_results/ICdat.rds")
 
-ICdat <- mutate(ICdat,
-                model=factor(model,
+ICdat <- (ICdat 
+          %>% ungroup() 
+          %>% mutate(model=factor(model,
                          levels=c("HMM","HMMsin"),
-                         labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
-simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
+                         labels=c("Homogeneous","Heterogeneous")))
+          %>% filter(IC == "BIC")
+)
+simsin <- (ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
     geom_line() +
@@
     scale_colour_brewer(palette="Set1")+
-    labs(y="Counts",x="BIC-optimal number of states")
-# print(simplot)
+    ggtitle("2-state HMM with Time-dependent Transition")+
+    labs(y="Counts",x="BIC-optimal number of states")+
+    theme(legend.position="none"))
+
+ICdat <- readRDS("../sim_results/ICdat.rds")
+
+ICdat <- (ICdat 
+          %>% ungroup() 
+          %>% mutate(model=factor(model,
+                         levels=c("HMM","HMMsin"),
+                         labels=c("Homogeneous","Heterogeneous")))
+          %>% filter(IC == "BIC")
+)
+simgps <- (ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
+    geom_line() +
+    geom_point(size=3,bg="white") +
+    scale_shape_manual(values=c(21,16))+
+    scale_colour_brewer(palette="Set1")+
+    ggtitle("2-state HMM with GPS Error")+
+    labs(y="Counts",x="BIC-optimal number of states")+
+      theme(legend.position="none"))
+
+ICdat <- readRDS("../simtime_results/ICdat.rds")
+
+ICdat <- (ICdat 
+          %>% ungroup() 
+          %>% mutate(model=factor(model,
+                         levels=c("HMM","HMMsin"),
+                         labels=c("Homogeneous","Heterogeneous")))
+          %>% filter(IC == "BIC")
+)
+simtimegps <- (ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
+    geom_line() +
+    geom_point(size=3,bg="white") +
+    scale_shape_manual(values=c(21,16))+
+    scale_colour_brewer(palette="Set1")+
+    ggtitle("2-state HMM with Time-dependent Transition and GPS Error")+
+    labs(y="Counts",x="BIC-optimal number of states")+
+      theme(legend.position="none"))
+
+print(grid.arrange(simsin,simgps,simtimegps,ncol=2))
 
@@
 
-<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4,eval=FALSE>>=
+\clearpage
+<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4,eval=TRUE>>=
 allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
@@
 
+\clearpage
+
 <<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types.
 The left panel shows FMM, temporally heterogeneous FMM with a sinusoidal prior and HMM.
-The right panel shows HMMs with different temporal transitions.",fig.width=8,eval=FALSE>>=
+The right panel shows HMMs with different temporal transitions.",fig.width=8,eval=TRUE>>=
 #L <- load("cat1sumdat.RData")
@@
 linetypev <- c(rep(2,2),rep(1,5))
-(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
+(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                   , y=deltaBIC
@@
    + scale_size_continuous(name="# of states")
-   + scale_x_continuous(breaks=seq(0,300,by=30))
-   + geom_point(aes(shape=factor(nstates),alpha=nstates))
+   # + scale_x_continuous(breaks=seq(0,300,by=30))
+   + geom_point()
    + scale_linetype_manual(values=linetypev, name="Model")
    + scale_colour_brewer(palette="Dark2", name="Model")
-   + scale_shape_discrete(name="Number of states")
+   # + scale_shape_discrete(name="Number of states")
    + scale_alpha_continuous(name="Number of states")
@@
 
-# print(adj_BIC_plot)
+#print(adj_BIC_plot)
 @
 
-<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.",fig.width=8,eval=FALSE>>=
+\clearpage
+
+<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.",fig.width=8,eval=TRUE>>=
 
@@
   %>% group_by(Time) 
-  %>% summarise_each(funs(mean))
-  %>% transmute(time = Time, variable = "obs", value=LogDist)
+  %>% summarise_each(funs(mean,sd))
+  %>% transmute(time=Time,model="Observed",steplen=mean,smin=mean-2*sd,smax=mean+2*sd)
 )
@@
              , "HMM-sin (n=5)"
-             , "observed"
-             , "viterbi homogenous (n=3)")
+             , "Observed"
+             , "viterbi homogenous (n=3)"
+             )
 
-combdf <- (rbind(avgdf,obsdf,vitdf2)
+combdf <- (rbind(avgdf,vitdf2)
            %>% transmute(time=time,model=variable,steplen=value)
            %>% filter(model %in% mnames)
-           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2)
+                      , smin=steplen,smax=steplen)
 )
-linetypev <- c(rep(c(2,1),c(2,6))
-                ,0
+
+combdf2 <- rbind(combdf,obsdf)
+linetypev <- c(rep(c(2,1),c(2,5))
+                ,1,0
                )
-colvec <- c(c(brewer.pal(7,"Dark2"),"black")
-             ,"black"
+colvec <- c(c(brewer.pal(7,"Dark2"))
+             ,"black","black"
             )
-lwdvec <- c(rep(c(1,4),c(7,1))
-             ,4
+
+lwdvec <- c(rep(c(1),c(7))
+             ,1,4
             )
-alphavec <- c(rep(c(1,0.2),c(7,1))
-               ,1
+alphavec <- c(rep(0,7)
+               ,0.1,0
               )
@@
 
-(avg_plot <- ggplot(combdf,aes(x=time,y=steplen,colour=model,
-                   lwd=model,
-                   alpha=model,linetype=model, shape=model))
-  + geom_point()
-  + geom_line()
-  + xlab("Time of Day") 
-  + scale_shape_manual(values=shapevec, name="Model")
-  + ylab(expression('Mean step length'~(log[10]*m)))
+avg_plot <- ggplot(combdf2,aes(x=time))
+print(avg_plot 
+  + geom_ribbon(aes(ymin=smin,ymax=smax,alpha=model))
+  + geom_line(aes(y=steplen,linetype=model,colour=model))
   + scale_colour_manual(values=colvec, name="Model")
-  + scale_alpha_manual(values=alphavec, name="Model")
   + scale_size_manual(values=lwdvec, name="Model")
   + scale_linetype_manual(values=linetypev, name="Model")
+  + scale_alpha_manual(values=alphavec, name="Model")
+  + geom_point(aes(y=steplen,colour=model,shape=model,lwd=model))
+  + scale_shape_manual(values=shapevec,name="Model")
+  + ylab(expression('Mean step length'~(log[10]*m)))
+  + xlab("Time of Day") 
 )
@@
 
-<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ",eval=FALSE>>=
+\clearpage
+
+<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ",eval=TRUE>>=
 acffun <- function(sim){
@@
 
+linetypev <- c(rep(c(2,1),c(2,6))
+                ,0
+               )
+colvec <- c(c(brewer.pal(7,"Dark2"),"black")
+             ,"black"
+            )
+lwdvec <- c(rep(c(1,4),c(7,1))
+             ,4
+            )
+alphavec <- c(rep(c(1,0.2),c(7,1))
+               ,1
+              )
+
+shapevec <- c(rep(32,8),118)
+
 (acf_plot <- ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
@@
 ) 
-# print(acf_plot)
+print(acf_plot)
 ## vitacf messes up legend ...
@@
 
-<<r msdlist,message=FALSE, echo=FALSE, fig.cap="State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.", warning=FALSE,eval=FALSE>>=
+\clearpage
+<<r msdlist,message=FALSE, echo=FALSE, fig.cap="State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.", warning=FALSE,eval=TRUE>>=
 get_cat_pars2 <- function(catid){
@@
 @
+
+\clearpage
+
+Everything Below goes into the appendix
 
@@
    + scale_size_continuous(name="# of states")
-   + scale_x_continuous(breaks=seq(0,300,by=30))
+   # + scale_x_continuous(breaks=seq(0,300,by=30))
    + geom_point(aes(shape=factor(nstates),alpha=nstates))
@@
   %>% group_by(Time) 
-  %>% summarise_each(funs(mean))
-  %>% transmute(time = Time, variable = "obs", value=LogDist)
+  %>% summarise_each(funs(mean,sd))
+  %>% transmute(time=Time,model="Observed",steplen=mean,smin=mean-2*sd,smax=mean+2*sd)
 )
+
 vit <- readRDS("../summary_stats/cat.2.vit.RDS")
@@
             "hmmquad4","hmmsin4","obs","viterbi")
-mnames2 <- c("fmm-homogeneous  (n=4)"
-             , "fmm-sin (n=4)"
-             , "homogeneous (n=5)"
-             , "block (n=4)"
-             , "hourly (n=3)"
-             , "quadratic (n=4)"
-             , "sin (n=4)"
-             , "observed"
-             , "viterbi homogenous (n=3)")
-
-combdf <- (rbind(avgdf,obsdf,vitdf2)
+mnames2 <- c("Fmm-homogeneous  (n=4)"
+             , "Fmm-sin (n=4)"
+             , "HMM-homogeneous (n=5)"
+             , "HMM-block (n=4)"
+             , "HMM-hourly (n=3)"
+             , "HMM-quadratic (n=4)"
+             , "HMM-sin (n=4)"
+             , "Observed"
+             , "Viterbi homogenous (n=3)")
+combdf <- (rbind(avgdf,vitdf2)
            %>% transmute(time=time,model=variable,steplen=value)
            %>% filter(model %in% mnames)
-           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2)
+                      , smin=steplen,smax=steplen)
 )
-linetypev <- c(rep(c(2,1),c(2,6))
-                ,0
+
+combdf2 <- rbind(combdf,obsdf)
+linetypev <- c(rep(c(2,1),c(2,5))
+                ,1,0
                )
-colvec <- c(c(brewer.pal(7,"Dark2"),"black")
-             ,"black"
+colvec <- c(c(brewer.pal(7,"Dark2"))
+             ,"black","black"
             )
-lwdvec <- c(rep(c(1,4),c(7,1))
-             ,4
+
+lwdvec <- c(rep(c(1),c(7))
+             ,1,4
             )
-alphavec <- c(rep(c(1,0.2),c(7,1))
-               ,1
+alphavec <- c(rep(0,7)
+               ,0.1,0
               )
@@
 
-(ggplot(combdf,aes(x=time,y=steplen,colour=model,
-                   lwd=model,
-                   alpha=model,linetype=model, shape=model))
-  + geom_point()
-  + geom_line()
-  + xlab("Time of Day") 
-  + scale_shape_manual(values=shapevec, name="Model")
-  + ylab(expression('Mean step length'~(log[10]*m)))
+avg_plot <- ggplot(combdf2,aes(x=time))
+(avg_plot 
+  + geom_ribbon(aes(ymin=smin,ymax=smax,alpha=model))
+  + geom_line(aes(y=steplen,linetype=model,colour=model))
   + scale_colour_manual(values=colvec, name="Model")
-  + scale_alpha_manual(values=alphavec, name="Model")
   + scale_size_manual(values=lwdvec, name="Model")
   + scale_linetype_manual(values=linetypev, name="Model")
+  + scale_alpha_manual(values=alphavec, name="Model")
+  + geom_point(aes(y=steplen,colour=model,shape=model,lwd=model))
+  + scale_shape_manual(values=shapevec,name="Model")
+  + ylab(expression('Mean step length'~(log[10]*m)))
+  + xlab("Time of Day") 
 )
+
 @
@@
 )
+
+linetypev <- c(rep(c(2,1),c(2,6))
+                ,0
+               )
+colvec <- c(c(brewer.pal(7,"Dark2"),"black")
+             ,"black"
+            )
+lwdvec <- c(rep(c(1,4),c(7,1))
+             ,4
+            )
+alphavec <- c(rep(c(1,0.2),c(7,1))
+               ,1
+              )
+
+shapevec <- c(rep(32,8),118)
 
@@
             "hmmquad5","hmmsin4","obs","viterbi")
-mnames2 <- c("fmm-homogeneous  (n=4)"
-             , "fmm-sin (n=3)"
-             , "homogeneous (n=5)"
-             , "block (n=4)"
-             , "hourly (n=3)"
-             , "quadratic (n=5)"
-             , "sin (n=4)"
-             , "observed"
-             , "viterbi homogenous (n=3)")
-
-combdf <- (rbind(avgdf,obsdf,vitdf2)
+mnames2 <- c("Fmm-homogeneous  (n=4)"
+             , "Fmm-sin (n=3)"
+             , "HMM-homogeneous (n=5)"
+             , "HMM-block (n=4)"
+             , "HMM-hourly (n=3)"
+             , "HMM-quadratic (n=5)"
+             , "HMM-sin (n=4)"
+             , "Observed"
+             , "Viterbi homogenous (n=3)")
+combdf <- (rbind(avgdf,vitdf2)
            %>% transmute(time=time,model=variable,steplen=value)
            %>% filter(model %in% mnames)
-           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2)
+                      , smin=steplen,smax=steplen)
 )
-linetypev <- c(rep(c(2,1),c(2,6))
-                ,0
+
+combdf2 <- rbind(combdf,obsdf)
+linetypev <- c(rep(c(2,1),c(2,5))
+                ,1,0
                )
-colvec <- c(c(brewer.pal(7,"Dark2"),"black")
-             ,"black"
+colvec <- c(c(brewer.pal(7,"Dark2"))
+             ,"black","black"
             )
-lwdvec <- c(rep(c(1,4),c(7,1))
-             ,4
+
+lwdvec <- c(rep(c(1),c(7))
+             ,1,4
             )
-alphavec <- c(rep(c(1,0.2),c(7,1))
-               ,1
+alphavec <- c(rep(0,7)
+               ,0.1,0
               )
@@
   %>% group_by(Time) 
-  %>% summarise_each(funs(mean))
-  %>% transmute(time = Time, variable = "obs", value=LogDist)
+  %>% summarise_each(funs(mean,sd))
+  %>% transmute(time=Time,model="Observed",steplen=mean,smin=mean-2*sd,smax=mean+2*sd)
 )
+
 vit <- readRDS("../summary_stats/cat.15.vit.RDS")
@@
             "hmmquad5","hmmsin5","obs","viterbi")
-mnames2 <- c("fmm-homogeneous  (n=3)"
-             , "fmm-sin (n=4)"
-             , "homogeneous (n=5)"
-             , "block (n=4)"
-             , "hourly (n=3)"
-             , "quadratic (n=5)"
-             , "sin (n=5)"
-             , "observed"
-             , "viterbi homogenous (n=3)")
+mnames2 <- c("Fmm-homogeneous  (n=3)"
+             , "Fmm-sin (n=4)"
+             , "HMM-homogeneous (n=5)"
+             , "HMM-block (n=4)"
+             , "HMM-hourly (n=3)"
+             , "HMM-quadratic (n=5)"
+             , "HMM-sin (n=5)"
+             , "Observed"
+             , "Viterbi homogenous (n=3)")
 
-combdf <- (rbind(avgdf,obsdf,vitdf2)
+combdf <- (rbind(avgdf,vitdf2)
            %>% transmute(time=time,model=variable,steplen=value)
            %>% filter(model %in% mnames)
-           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2)
+                      , smin=steplen,smax=steplen)
 )
-linetypev <- c(rep(c(2,1),c(2,6))
-                ,0
+
+combdf2 <- rbind(combdf,obsdf)
+linetypev <- c(rep(c(2,1),c(2,5))
+                ,1,0
                )
-colvec <- c(c(brewer.pal(7,"Dark2"),"black")
-             ,"black"
+colvec <- c(c(brewer.pal(7,"Dark2"))
+             ,"black","black"
             )
-lwdvec <- c(rep(c(1,4),c(7,1))
-             ,4
+
+lwdvec <- c(rep(c(1),c(7))
+             ,1,4
             )
-alphavec <- c(rep(c(1,0.2),c(7,1))
-               ,1
+alphavec <- c(rep(0,7)
+               ,0.1,0
               )
@@
 )
+combacfdf <- (rbind(acfdf,vitacf)
+           %>% transmute(lag=lag,model=model,ACF=ACF)
+           %>% filter(model %in% mnames)
+           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
+)
+
+linetypev <- c(rep(c(2,1),c(2,6))
+                ,0
+               )
+colvec <- c(c(brewer.pal(7,"Dark2"),"black")
+             ,"black"
+            )
+lwdvec <- c(rep(c(1,4),c(7,1))
+             ,4
+            )
+alphavec <- c(rep(c(1,0.2),c(7,1))
+               ,1
+              )
+
+shapevec <- c(rep(32,8),118)
 

### 2016-12-22 simtestnoerror.R
sim test no error

@@ -1,0 +1,132 @@
+library(depmixS4)
+library(plyr)
+
+t <- rep(0:23,500)
+y <- rep(1,12000)
+
+tempdat <- data.frame(y=y,t=t)
+
+system.time(mod <- depmix(y~1
+                          , data=tempdat
+                          , transition=~1 #cos((2*pi*t)/24)+ sin((2*pi*t)/24)
+                          , nstate=2
+                          , family=gaussian())
+)
+getpars(mod)
+
+seed = unlist(strsplit(rtargetname,"[.]"))[2]
+set.seed(seed)
+
+#randpars <- sample(-3:3,length(getpars(mod))-6,replace=TRUE)
+transprobs <- runif(2,0.001,0.999)
+ranpars <- c(1-transprobs[1],transprobs[1],1-transprobs[2],transprobs[2])
+
+newmod <- setpars(mod,c(0.5,0.5,randpars,0,1,2,1))
+newmod
+sim <- simhmm(newmod)
+error <- 0 ## real scale
+oldsl <- sim@response[[1]][[1]]@y
+newsl <- log10(abs(10^oldsl + error)) 
+hist(oldsl)
+hist(newsl)
+df <- data.frame(obs= newsl,states=sim@states,time=t)
+
+system.time(hmm2 <- depmix(obs~1
+                           , data=df
+                           , transition=~1
+                           , nstate=2
+                           , family=gaussian())
+)
+
+system.time(hmm2s <- fit(hmm2,verbose=FALSE))
+
+system.time(hmm3 <- depmix(obs~1
+                           , data=df
+                           , transition=~1
+                           , nstate=3
+                           , family=gaussian())
+)
+
+system.time(hmm3s <- fit(hmm3,verbose=FALSE))
+
+system.time(hmm4 <- depmix(obs~1
+                           , data=df
+                           , transition=~1
+                           , nstate=4
+                           , family=gaussian())
+)
+
+system.time(hmm4s <- fit(hmm4,verbose=FALSE))
+
+
+
+# system.time(hmm5 <- depmix(obs~1
+#                            , data=df
+#                            , transition=~1
+#                            , nstate=5
+#                            , family=gaussian())
+# )
+# 
+# system.time(hmm5s <- fit(hmm5,verbose=FALSE))
+
+system.time(hmmsin2 <- depmix(obs~1
+                              , data=df
[…]

### 2016-12-22 simtestnoerror.R
don't want to wait for the results, need 6 more

@@ -19,7 +19,7 @@ set.seed(seed)
 
 #randpars <- sample(-3:3,length(getpars(mod))-6,replace=TRUE)
 transprobs <- runif(2,0.001,0.999)
-ranpars <- c(1-transprobs[1],transprobs[1],1-transprobs[2],transprobs[2])
+randpars <- c(1-transprobs[1],transprobs[1],1-transprobs[2],transprobs[2])
 
 newmod <- setpars(mod,c(0.5,0.5,randpars,0,1,2,1))
 newmod
